---
title: "What tf$Variable-backed data nodes cost"
format: gfm
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(here)
library(fs)
library(dplyr)
library(tidyr)
library(knitr)

run_dir <- here("2026-09-21-data-variables-cost")
files <- dir_ls(run_dir, glob = "*results-*.rds")
raw <- lapply(files, readRDS)
results <- bind_rows(lapply(raw, \(x) x$results))

design <- raw[[1]]
shas <- vapply(raw, \(x) substr(x$sha, 1, 10), character(1))
names(shas) <- vapply(raw, \(x) x$label, character(1))
```

greta-dev/greta#739 gives every data node a persistent `tf$Variable` instead of
folding its value into each trace. That buys swapping data without a retrace.
This is what it costs everyone who never swaps anything.

`r design$n_reps` reps per model, `r design$warmup` warmup and
`r design$n_samples` samples on `r design$chains` chains.

```{r}
#| label: shas
tibble::tibble(branch = names(shas), sha = unname(shas)) |>
  kable()
```

## Time

```{r}
#| label: time
results |>
  group_by(model, label) |>
  summarise(
    build = median(build_secs),
    mcmc = median(mcmc_secs),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = label, values_from = c(build, mcmc)) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

Ratios, branch relative to main. Above 1 is slower.

```{r}
#| label: time-ratio
results |>
  group_by(model, label) |>
  summarise(
    build = median(build_secs),
    mcmc = median(mcmc_secs),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = label, values_from = c(build, mcmc)) |>
  transmute(
    model,
    build_ratio = round(`build_data-variables` / build_main, 2),
    mcmc_ratio = round(`mcmc_data-variables` / mcmc_main, 2)
  ) |>
  kable()
```

## Memory

RSS deltas in MB: at dag construction, where the variables are created, and
across sampling. Total is what the process actually grew by.

```{r}
#| label: memory
results |>
  group_by(model, label) |>
  summarise(
    build_mb = median(build_delta_mb),
    mcmc_mb = median(mcmc_delta_mb),
    total_mb = median(build_delta_mb + mcmc_delta_mb),
    .groups = "drop"
  ) |>
  pivot_wider(
    names_from = label,
    values_from = c(build_mb, mcmc_mb, total_mb)
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  kable()
```

## Is the spread small enough to read these?

Min and max across reps, so a ratio computed from medians can be judged against
the noise it sits in.

```{r}
#| label: spread
results |>
  group_by(model, label) |>
  summarise(
    mcmc_min = round(min(mcmc_secs), 2),
    mcmc_max = round(max(mcmc_secs), 2),
    total_mb_min = round(min(build_delta_mb + mcmc_delta_mb), 1),
    total_mb_max = round(max(build_delta_mb + mcmc_delta_mb), 1),
    .groups = "drop"
  ) |>
  kable()
```
