---
title: "mcmc() on main and thinning-i318 (greta#318)"
format:
  gfm: default
  html: default
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(here)
library(fs)
library(dplyr)
library(tidyr)
library(knitr)

run_dir <- here("2026-09-29-thinning-i318")
results_rds <- path(run_dir, "results.rds")
if (!file_exists(results_rds)) {
  stop(
    "no results.rds - run 2026-09-29-thinning-i318/01-measure.R first",
    call. = FALSE
  )
}
results <- readRDS(results_rds)

sha_link <- function(sha) {
  sprintf(
    "[%s](https://github.com/greta-dev/greta/commit/%s)",
    substr(sha, 1, 8),
    sha
  )
}
shas <- results$standard$shas

mcmc_ratios <- bind_rows(
  lapply(names(results), function(tier) {
    results[[tier]]$timings |>
      filter(task == "mcmc") |>
      select(example, branch, ms) |>
      pivot_wider(names_from = branch, values_from = ms) |>
      mutate(tier = tier, speedup = main / `thinning-i318`)
  })
)
speedup_range <- range(mcmc_ratios$speedup)

agreement <- bind_rows(
  lapply(names(results), function(tier) {
    mutate(results[[tier]]$agreement, tier = tier)
  })
)
n_disagree <- sum(agreement$n_disagree)
```

On main (`r sha_link(shas$sha[shas$branch == "main"])`), greta passed `thin` to
TensorFlow Probability as `num_steps_between_results`, which keeps one draw in
`thin + 1`: every draw and every warmup step cost two HMC iterations at the
default `thin = 1`. thinning-i318
(`r sha_link(shas$sha[shas$branch == "thinning-i318"])`) passes `thin - 1`.

## Fixed work: `mcmc()` with the same arguments

The deterministic task runs `mcmc()` with fixed settings on each branch. The
branch is `r sprintf("%.2f", speedup_range[1])` to
`r sprintf("%.2f", speedup_range[2])` times faster across both tiers. Less than
two is expected, since each run also pays costs outside the HMC iterations -
tracing, and the R-to-TensorFlow round trip per burst - but this run does not
measure how much of each call they take.

```{r}
mcmc_ratios |>
  transmute(
    tier,
    example,
    main_s = main / 1000,
    branch_s = `thinning-i318` / 1000,
    speedup
  ) |>
  kable(digits = 2)
```

## Do the branches sample the same posterior?

Across `r nrow(agreement)` example-tier comparisons, `r n_disagree` variables
disagree beyond Monte Carlo error.

```{r}
agreement |>
  select(tier, example, variables, max_abs_z, n_disagree, verdict) |>
  kable(digits = 2)
```

## Sampling to a target ESS

One replicate per branch in both tiers, which `R/tiers.R` notes cannot resolve
a sampling-speed difference. These swing both ways by more than the fix could
cause, and the reason is the tuned step size: on the linear example alone,
`atelier/evidence/check-thinning-ess-i318.R` found minimum bulk-ESS varying
more than ten-fold across three seeds on the same branch. Per HMC iteration
actually run, the branches mix alike there; per `n_samples`, the branch gives
about half, since main ran two iterations per draw.

```{r}
bind_rows(
  lapply(names(results), function(tier) {
    mutate(results[[tier]]$sampling, tier = tier)
  })
) |>
  select(
    tier,
    example,
    branch,
    elapsed,
    mcmc_samples,
    seconds_per_1000_ess,
    ess_bulk_min,
    rhat_max
  ) |>
  arrange(tier, example, branch) |>
  kable(digits = 2)
```
