---
title: "Rounding error in chol2symm(chol(x))"
format:
  gfm: default
  html: default
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(here)
library(fs)

run_dir <- here("2026-09-28-chol2symm-tolerance-i842")
results_rds <- path(run_dir, "results.rds")
if (!file_exists(results_rds)) {
  stop(
    "no results.rds - run 2026-09-28-chol2symm-tolerance-i842/01-measure.R first",
    call. = FALSE
  )
}
raw <- readRDS(results_rds)
results <- raw$results
r <- results[results$side == "r", ]
g <- results[results$side == "greta", ]

pct <- function(x) paste0(formatC(100 * mean(x), format = "f", digits = 1), "%")
sci <- function(x) formatC(x, format = "e", digits = 1)
```

## The question

greta's "chol2symm inverts chol" test rebuilt a 9 x 9 Wishart matrix from its
Cholesky factor and compared at `tolerance = .Machine$double.eps`
(`r sci(raw$tolerances[["eps"]])`). It failed on Windows CI for greta #842 with
differences in the last digit. Is that tolerance tighter than the rounding error
of the rebuild itself, and what tolerance is safe?

## Results

`all.equal()` compares the mean relative difference over the elements that
differ against its tolerance, leaving out elements that match exactly. That is
the error below, taken from `all.equal(giveErr = TRUE)`.

```{r}
#| label: consistency
# the recorded error has to agree with the recorded failures, or it is the
# wrong statistic
stopifnot(
  identical(results$fails_eps, results$error > raw$tolerances[["eps"]]),
  identical(results$fails_fixed, results$error > raw$tolerances[["fixed"]])
)
```

| | draws | largest error | fails at `double.eps` | fails at `1e-12` |
|---|---|---|---|---|
| `chol2symm()` in R | `r nrow(r)` | `r sci(max(r$error))` | `r pct(r$fails_eps)` | `r pct(r$fails_fixed)` |
| greta `calculate()` | `r nrow(g)` | `r sci(max(g$error))` | `r pct(g$fails_eps)` | `r pct(g$fails_fixed)` |

The largest rounding error across all draws is
`r sci(max(results$error))`, so `.Machine$double.eps` fails on ordinary
rounding alone. A tolerance of `1e-12` leaves a margin of
`r formatC(1e-12 / max(results$error), format = "d", big.mark = ",")` times
over the largest error seen, while a real bug in `chol2symm()` - a transpose, a
wrong triangle - gives errors of order 1.

## Environment

greta at `r raw$greta_sha`.

```{r}
#| label: environment
#| results: asis
source(here("R", "provenance.R"))
cat(format_provenance(raw$provenance), sep = "\n")
```
