---
title: "greta benchmark suite"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    fig-width: 8
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
source("packages.R")
tar_source()

tar_load(c(
  timings,
  timings_relative,
  sampling,
  pooled_posterior,
  posterior_comparison,
  agreement,
  rss,
  shas,
  reference,
  under_test,
  target_ess
))

ratio <- timings_relative |>
  select(example, task, branch, ms) |>
  pivot_wider(names_from = branch, values_from = ms) |>
  mutate(ratio = .data[[under_test]] / .data[[reference]])
```

**`{r} under_test`** against **`{r} reference`**.

```{r}
#| label: shas
shas |>
  mutate(role = c("reference", "under test"), sha = substr(sha, 1, 10)) |>
  select(role, branch, sha) |>
  kable()
```

# Summary

## 1. Is it faster?

`model()` and `opt()` are deterministic, so every iteration is the same work
and the distribution of iteration times is the measurement. The minimum is the
robust estimator for timing, because timing noise is one-sided
[@chen2016].

```{r}
#| label: fig-speed
#| fig-cap: "Every timed iteration. Each row is a task, each column a branch. Clouds sitting further right are slower."
#| fig-height: 5
autoplot(timings) +
  facet_grid(example ~ branch, scales = "free_x") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")
```

```{r}
#| label: speed-ratio
ratio |>
  transmute(
    example,
    task,
    reference = round(.data[[reference]], 1),
    under_test = round(.data[[under_test]], 1),
    ratio = round(ratio, 2),
    direction = if_else(ratio > 1, "slower", "faster")
  ) |>
  kable(
    col.names = c(
      "example",
      "task",
      paste(reference, "(ms)"),
      paste(under_test, "(ms)"),
      "ratio",
      ""
    )
  )
```

## 2. Does it sample more efficiently?

Seconds to reach a minimum bulk-ESS of `{r} target_ess`. Quality is held
fixed and cost is measured, so there is no convergence gate discarding runs -
a branch that cannot reach the target appears as `hit_cap` rather than
vanishing.

```{r}
#| label: fig-sampling
#| fig-cap: "Seconds of sampling per 1000 effective draws. Lower is better."
#| fig-height: 3.5
sampling |>
  ggplot(aes(x = seconds_per_1000_ess, y = example, fill = branch)) +
  geom_col(position = position_dodge(preserve = "single"), width = 0.7) +
  labs(x = "seconds per 1000 effective draws", y = NULL, fill = NULL) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")
```

### Memory

End-of-run resident set size - the physical RAM the measuring process held,
not residual sum of squares. `bench`'s `mem_alloc` is deliberately not used:
it comes from `Rprofmem` and sees the R heap only, while greta's memory lives
in the TensorFlow graph on the Python side.

```{r}
#| label: rss
rss |>
  mutate(rss_mb = round(rss_mb)) |>
  kable(col.names = c("branch", "RSS (MB)"))
```

## 3. Does it give the same answer?

The question a timing cannot answer: two branches can both converge, both be
fast, and disagree about the estimate. Posterior means always differ a little
between runs, so the comparison is in **Monte Carlo standard error units**
rather than raw differences [@vehtari2021; @magnusson2025]:

$$z = \frac{\bar\theta_{\text{test}} - \bar\theta_{\text{ref}}}{\sqrt{\text{mcse}^2_{\text{test}} + \text{mcse}^2_{\text{ref}}}}$$

```{r}
#| label: fig-agreement
#| fig-cap: "Every variable's difference in means, in MCSE units. Inside the shaded band the branches agree to within the precision either of them has."
#| fig-height: 3
ggplot(posterior_comparison, aes(x = z_mean, y = example)) +
  annotate(
    "rect",
    xmin = -4,
    xmax = 4,
    ymin = -Inf,
    ymax = Inf,
    fill = "grey88"
  ) +
  geom_vline(xintercept = 0, colour = "grey40") +
  stat_dots(side = "both", scale = 0.6) +
  labs(x = "difference in means (MCSE units)", y = NULL) +
  theme_minimal(base_size = 10)
```

```{r}
#| label: agreement-table
agreement |>
  mutate(max_abs_z = round(max_abs_z, 2)) |>
  kable(
    col.names = c(
      "example",
      "variables",
      "largest |z|",
      "disagreeing",
      "worst variable",
      "verdict"
    )
  )
```

# Details

::: panel-tabset

## Timings

```{r}
#| label: timings-detail
timings_relative |>
  mutate(ms = round(ms, 1), relative = round(relative, 3)) |>
  pivot_wider(names_from = branch, values_from = c(ms, relative)) |>
  kable()
```

## Sampling

```{r}
#| label: sampling-detail
sampling |>
  select(
    branch,
    example,
    rep,
    elapsed,
    mcmc_samples,
    ess_bulk_min,
    ess_bulk_median,
    rhat_max,
    n_variables,
    hit_cap
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  kable()
```

A single run gives a zero-width interval, which cannot overlap anything, so
the replicate count decides the verdict before the intervals get a say. `quick`
and `standard` run one replicate and will always say so; only `thorough` can
resolve a sampling speed difference.

```{r}
#| label: sampling-resolution
sampling |>
  filter(!hit_cap) |>
  group_by(example, branch) |>
  summarise(
    mid = median(seconds_per_1000_ess),
    lo = min(seconds_per_1000_ess),
    hi = max(seconds_per_1000_ess),
    n = n(),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = branch, values_from = c(mid, lo, hi, n)) |>
  rename_with(\(nm) sub(paste0("_", under_test, "$"), "_test", nm)) |>
  rename_with(\(nm) sub(paste0("_", reference, "$"), "_ref", nm)) |>
  mutate(
    ratio = round(mid_test / mid_ref, 2),
    runs = pmin(n_test, n_ref),
    verdict = case_when(
      runs < 2 ~ "too few runs to say",
      lo_test > hi_ref | hi_test < lo_ref ~ "resolved",
      .default = "unresolved"
    )
  ) |>
  select(example, ratio, runs, verdict) |>
  kable()
```

## Posteriors

The marginal for every variable, both branches overlaid. Intervals sitting on
top of each other are the expected result.

```{r}
#| label: fig-intervals
#| fig-height: 7
pooled_posterior |>
  group_by(example) |>
  slice_max(q95 - q5, n = 6) |>
  ungroup() |>
  ggplot(aes(y = variable, colour = branch)) +
  geom_pointinterval(
    aes(x = mean, xmin = q5, xmax = q95),
    position = position_dodge(width = 0.5)
  ) +
  facet_wrap(~example, scales = "free", ncol = 2) +
  labs(x = NULL, y = NULL, colour = NULL) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")
```

## Provenance

```{r}
#| label: provenance
str(tar_read(provenance))
```

:::

# Methods

**Diagnostics.** Rank-normalised split-R̂ and bulk/tail effective sample size,
with R̂ < 1.01 and ESS > 400 as reference thresholds [@vehtari2021]. ESS > 400
is where the diagnostics become trustworthy, not where accuracy becomes
sufficient. `coda::effectiveSize()` is deliberately not used: it fits an AR
model per chain with no between-chain information, so it reports a large ESS
for chains stuck in different modes.

**Comparing posteriors.** Differences in posterior means are divided by the
combined Monte Carlo standard error, so bias is separable from Monte Carlo
noise [@magnusson2025]. A two-sample Kolmogorov-Smirnov test on the draws
would be wrong here: it assumes both samples are iid, and MCMC draws are
autocorrelated, so a correct sampler is rejected too often [@talts2018].

**Timing.** Timing noise is one-sided, so the minimum is the robust estimator
[@chen2016]. Effect-size intervals are reported rather than significance tests
[@kalibera2013]; a ratio whose spread is wider than the effect is labelled
unresolved rather than quoted.

**What is not measured.** Gradient evaluations are not counted, so a change can
be attributed to wall time but not to "needs fewer gradients" versus "each
gradient got cheaper" [@hoffman2021]. There is no external reference posterior,
so a bug present in both branches would pass unnoticed; posteriordb
[@magnusson2025] and Inference Gym are the obvious sources if that becomes
worth closing. `rhat_nested()` [@margossian2025] would be the right diagnostic
if greta's chains were ever run many-and-short.

# References

::: {#refs}
:::
