---
title: "greta benchmark suite"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    fig-width: 8
bibliography: references.bib
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
source("packages.R")
tar_source()

tar_load(c(
  timings,
  timings_relative,
  sampling,
  pooled_posterior,
  posterior_comparison,
  agreement,
  rss,
  shas,
  reference,
  under_test,
  target_ess,
  fit_draws,
  fitted_values
))

example_names <- unique(fit_draws$example)

ratio <- timings_relative |>
  select(example, task, branch, ms) |>
  pivot_wider(names_from = branch, values_from = ms) |>
  mutate(ratio = .data[[under_test]] / .data[[reference]])
```

Two **methods** are compared throughout: a reference and the one under test.
Here they are git branches of greta, but nothing below depends on that - they
are simply two versions of the same code measured on the same machine, in the
same session, against the same examples.

```{r}
#| label: shas
shas |>
  mutate(role = c("reference", "under test"), sha = substr(sha, 1, 10)) |>
  select(role, branch, sha) |>
  kable(col.names = c("role", "git branch", "commit"))
```

# Summary

## 1. Is it faster?

`model()`, `opt()` and a fixed-length `mcmc()` are deterministic - every
iteration is the same work - so the distribution of iteration times is the
measurement. The minimum is the robust estimator for timing, because timing
noise is one-sided [@chen2016].

`mcmc` here is 1000 warmup and 1000 samples across 4 chains on 4 cores, which
is greta's default shape. A fixed iteration count
is the only way to ask whether one iteration got more expensive: the sampling
tier below lets the number of draws float until it reaches a quality target, so
its wall time confounds per-iteration cost with how well the chain happened to
mix.

```{r}
#| label: fig-speed
#| fig-cap: "Every timed iteration, one row per method. The two rows in a panel do identical work, so the one further right is slower. Panels are different tasks and examples, which are not comparable with each other - only the two rows within a panel are."
#| fig-height: 5
autoplot(timings) +
  aes(y = branch) +
  facet_grid(example ~ task, scales = "free_x") +
  labs(y = NULL) +
  theme(legend.position = "bottom")
```

```{r}
#| label: speed-ratio
ratio |>
  transmute(
    example,
    task,
    reference = round(.data[[reference]], 1),
    under_test = round(.data[[under_test]], 1),
    ratio = round(ratio, 2),
    direction = if_else(ratio > 1, "slower", "faster")
  ) |>
  kable(
    col.names = c(
      "example",
      "task",
      paste(reference, "(ms)"),
      paste(under_test, "(ms)"),
      "ratio",
      ""
    )
  )
```

## 2. Does it sample more efficiently?

Seconds to reach a minimum bulk-ESS of `{r} target_ess` - a different question
from the fixed-length `mcmc` above, which measures per-iteration cost. Quality
is held fixed and cost is measured, so there is no convergence gate discarding runs -
a method that cannot reach the target appears as `hit_cap` rather than
vanishing.

```{r}
#| label: fig-sampling
#| fig-cap: "Seconds of sampling per 1000 effective draws, one bar per method. Lower is better. Each panel is one example; the comparison is between the two bars within a panel."
#| fig-height: 3.5
sampling |>
  ggplot(aes(x = seconds_per_1000_ess, y = branch, fill = branch)) +
  geom_point(aes(colour = branch)) +
  facet_wrap(~example, ncol = 2, scales = "free_x") 
  labs(x = "seconds per 1000 effective draws", y = NULL) +
  theme_minimal(base_size = 10)
```

### Memory

End-of-run resident set size - the physical RAM the measuring process held,
not residual sum of squares. `bench`'s `mem_alloc` is deliberately not used:
it comes from `Rprofmem` and sees the R heap only, while greta's memory lives
in the TensorFlow graph on the Python side.

```{r}
#| label: rss
rss |>
  mutate(rss_mb = round(rss_mb)) |>
  kable(col.names = c("method", "RSS (MB)"))
```

## 3. Does it give the same answer?

The question a timing cannot answer: two methods can both converge, both be
fast, and disagree about the estimate. Posterior means always differ a little
between runs, so the comparison is in **Monte Carlo standard error units**
rather than raw differences [@vehtari2021; @magnusson2025]:

$$z = \frac{\bar\theta_{\text{test}} - \bar\theta_{\text{ref}}}{\sqrt{\text{mcse}^2_{\text{test}} + \text{mcse}^2_{\text{ref}}}}$$

```{r}
#| label: fig-agreement
#| fig-cap: "Every variable's difference in means, in MCSE units. Inside the shaded band the two methods agree to within the precision either of them has."
#| fig-height: 3
ggplot(posterior_comparison, aes(x = z_mean, y = example)) +
  annotate(
    "rect",
    xmin = -4,
    xmax = 4,
    ymin = -Inf,
    ymax = Inf,
    fill = "grey88"
  ) +
  geom_vline(xintercept = 0, colour = "grey40") +
  stat_dots(side = "both", scale = 0.6) +
  labs(x = "difference in means (MCSE units)", y = NULL) +
  theme_minimal(base_size = 10)
```

```{r}
#| label: agreement-table
agreement |>
  mutate(max_abs_z = round(max_abs_z, 2)) |>
  kable(
    col.names = c(
      "example",
      "variables",
      "largest |z|",
      "disagreeing",
      "worst variable",
      "verdict"
    )
  )
```

## 4. What do the fits look like?

One seeded run of each example on each branch: 4 chains of 1000 warmup and
1000 samples. The plots put the branches side by side, so a difference in how
a model samples or what it estimates shows as a difference between panels.

### Fitted values against the data

The fitted values are each model's expected value for the data, such as the
regression line, not new data simulated from the model. The ribbons are 50%,
80% and 95% intervals; the points and crosses are the data.

```{r}
#| label: fig-fits
#| fig-height: 4
for (example_name in example_names) {
  print(gg_fit(fitted_values, example_name))
}
```

### Posterior densities

The first eight variables of each example, as densities with their median and
66% and 95% intervals, one row per branch.

```{r}
#| label: fig-densities
#| fig-height: 4.5
for (example_name in example_names) {
  print(gg_densities(fit_draws, example_name))
}
```

### Trace plots

The first six variables of each example, each chain in its own colour. Chains
that overlap and wander around a common level are what mixing looks like.

```{r}
#| label: fig-traces
#| fig-height: 6
for (example_name in example_names) {
  print(gg_traces(fit_draws, example_name))
}
```

# Details

::: panel-tabset

## Timings

```{r}
#| label: timings-detail
timings_relative |>
  mutate(ms = round(ms, 1), relative = round(relative, 3)) |>
  pivot_wider(names_from = branch, values_from = c(ms, relative)) |>
  rename_with(\(nm) sub("^ms_", "ms: ", sub("^relative_", "relative: ", nm))) |>
  kable()
```

## Sampling

```{r}
#| label: sampling-detail
sampling |>
  select(
    branch,
    example,
    rep,
    elapsed,
    mcmc_samples,
    ess_bulk_min,
    ess_bulk_median,
    rhat_max,
    n_variables,
    hit_cap
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  rename(method = branch) |>
  kable()
```

A single run gives a zero-width interval, which cannot overlap anything, so
the replicate count decides the verdict before the intervals get a say. `quick`
and `standard` run one replicate and will always say so; only `thorough` can
resolve a sampling speed difference.

```{r}
#| label: sampling-resolution
sampling |>
  filter(!hit_cap) |>
  group_by(example, branch) |>
  summarise(
    mid = median(seconds_per_1000_ess),
    lo = min(seconds_per_1000_ess),
    hi = max(seconds_per_1000_ess),
    n = n(),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = branch, values_from = c(mid, lo, hi, n)) |>
  rename_with(\(nm) sub(paste0("_", under_test, "$"), "_test", nm)) |>
  rename_with(\(nm) sub(paste0("_", reference, "$"), "_ref", nm)) |>
  mutate(
    ratio = round(mid_test / mid_ref, 2),
    runs = pmin(n_test, n_ref),
    verdict = case_when(
      runs < 2 ~ "too few runs to say",
      lo_test > hi_ref | hi_test < lo_ref ~ "resolved",
      .default = "unresolved"
    )
  ) |>
  select(example, ratio, runs, verdict) |>
  kable()
```

## Posteriors

The marginal for every variable, both methods overlaid. Intervals sitting on
top of each other are the expected result.

```{r}
#| label: fig-intervals
#| fig-height: 7
pooled_posterior |>
  group_by(example) |>
  slice_max(q95 - q5, n = 6) |>
  ungroup() |>
  ggplot(aes(y = variable, colour = branch)) +
  geom_pointinterval(
    aes(x = mean, xmin = q5, xmax = q95),
    position = position_dodge(width = 0.5)
  ) +
  facet_wrap(~example, scales = "free", ncol = 2) +
  labs(x = NULL, y = NULL, colour = NULL) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")
```

## Provenance

```{r}
#| label: provenance
str(tar_read(provenance))
```

:::

# Methods

**Diagnostics.** Rank-normalised split-R̂ and bulk/tail effective sample size,
with R̂ < 1.01 and ESS > 400 as reference thresholds [@vehtari2021]. ESS > 400
is where the diagnostics become trustworthy, not where accuracy becomes
sufficient. `coda::effectiveSize()` is deliberately not used: it fits an AR
model per chain with no between-chain information, so it reports a large ESS
for chains stuck in different modes.

**Comparing posteriors.** Differences in posterior means are divided by the
combined Monte Carlo standard error, so bias is separable from Monte Carlo
noise [@magnusson2025]. A two-sample Kolmogorov-Smirnov test on the draws
would be wrong here: it assumes both samples are iid, and MCMC draws are
autocorrelated, so a correct sampler is rejected too often [@talts2018].

**Timing.** Timing noise is one-sided, so the minimum is the robust estimator
[@chen2016]. Effect-size intervals are reported rather than significance tests
[@kalibera2013]; a ratio whose spread is wider than the effect is labelled
unresolved rather than quoted.

**What is not measured.** Gradient evaluations are not counted, so a change can
be attributed to wall time but not to "needs fewer gradients" versus "each
gradient got cheaper" [@hoffman2021]. There is no external reference posterior,
so a bug present in both methods would pass unnoticed; posteriordb
[@magnusson2025] and Inference Gym are the obvious sources if that becomes
worth closing. `rhat_nested()` [@margossian2025] would be the right diagnostic
if greta's chains were ever run many-and-short.

# References

::: {#refs}
:::
