---
title: "Greta benchmark suite"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    fig-width: 8
bibliography: references.bib
params:
  store: "_targets"
  further: ""
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
source("packages.R")
tar_source()

tar_load(
  c(
    shas,
    comparisons,
    tier,
    example_names,
    target_ess,
    reps,
    bench_repeats,
    mcmc_repeats,
    warmup_iterations,
    sample_iterations,
    mcmc_chains,
    mcmc_cores,
    opt_iterations,
    iterations,
    every_run,
    mcmc_runs,
    speed,
    sampling,
    sampling_speed,
    rss,
    posterior_comparison,
    agreement,
    pooled_posterior,
    fit_draws,
    fitted_values,
    fit_diagnostics,
    provenance
  ),
  store = params$store
)

versions <- shas$label

# stores measured before the run order was recorded have no run_order target
run_order <- tryCatch(
  tar_read(run_order, store = params$store),
  error = function(e) NULL
)

check_shown_code(c(
  linear = "code-linear",
  multiple_linear = "code-multiple-linear",
  hierarchical_linear = "code-hierarchical-linear",
  eight_schools = "code-eight-schools",
  cjs = "code-cjs"
))
```

This report compares versions of greta, measured on one machine, at the `{r} tier` tier. Every number in it is read from the targets store `{r} params$store`. The first version listed is the reference.

```{r}
#| label: versions
versions_table <- shas |>
  mutate(
    sha = substr(sha, 1, 10),
    iterations_per_draw = iterations$iterations_per_draw[
      match(label, iterations$branch)
    ]
  )
if (!is.null(run_order)) {
  versions_table$measured_order <- run_order$measured_order[
    match(versions_table$label, run_order$branch)
  ]
}
kable(
  versions_table,
  col.names = c(
    "version",
    "git ref",
    "commit",
    "iterations per draw",
    if (!is.null(run_order)) "measured"
  )
)
```

Each version is measured in a process of its own, one after another. `{r} if (is.null(run_order)) "This run did not record the order." else "The last column is the order: it is drawn afresh each time the versions are measured, so that whatever changes on the machine over a run does not always fall on the same version."`

`mcmc()` takes a number of draws, and versions of greta can run different numbers of iterations per draw: greta 0.6.0 runs two at `thin = 1`. Each version's iterations per draw is measured (`R/iterations.R`), and every `mcmc()` call is given the number of draws that makes the same number of iterations on every version: `{r} warmup_iterations` warmup and `{r} sample_iterations` sampling iterations, on `{r} mcmc_chains` chains and `{r} mcmc_cores` cores.

For each model, each version is measured on:

1. **Speed:** the time it takes to build the model, `opt()`, and `mcmc()` of a fixed number of iterations.
2. **Efficiency:** effective samples per second from those `mcmc()` runs, and the time `mcmc()` takes to reach a target effective sample size, `{r} reps` times.
3. **Agreement:** the difference between versions' posterior means, in Monte Carlo standard errors.
4. **Convergence:** R-hat and effective sample size of `{r} reps` seeded fits.

Section 5 plots each model's fitted values and posterior densities, and section 6 holds any measurements particular to this run. Every `mcmc()` call uses `hmc()`, greta's default sampler.

This report compares versions with each other. Whether a version's answers are correct is tested in greta itself, in `tests/testthat/test_posteriors_*.R`: each sampler's draws are compared with draws taken from the distribution directly, and Geweke tests check that alternating between sampling a model's parameters and simulating its data keeps their joint distribution.

# The models

Each model is defined in maths and in greta code, and comes from greta's `inst/examples/`. Normal and log-normal distributions are written with their mean and standard deviation, as greta's `normal()` and `lognormal()` take them; $\text{Cauchy}^{+}$ is a Cauchy truncated to positive values. A model's parameter count is the number of values `mcmc()` returns for it.

## linear

A regression of each department's rating in `attitude` on its number of complaints. `{r} parameter_text(fit_draws, "linear", tier)`

$$
\begin{aligned}
\text{rating}_i &\sim \text{Normal}(\mu_i, \sigma), \quad i = 1, \dots, 30 \\
\mu_i &= \alpha + \beta \, \text{complaints}_i \\
\alpha, \beta &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Cauchy}^{+}(0, 3)
\end{aligned}
$$

```{r}
#| label: code-linear
#| echo: true
#| eval: false
int <- normal(0, 10)
coef <- normal(0, 10)
sd <- cauchy(0, 3, truncation = c(0, Inf))
mu <- int + coef * attitude$complaints
distribution(attitude$rating) <- normal(mu, sd)
m <- model(int, coef, sd)
```

## multiple_linear

The same rating on all six of `attitude`'s other columns, $X$, through a matrix multiply. `{r} parameter_text(fit_draws, "multiple_linear", tier)`

$$
\begin{aligned}
\text{rating}_i &\sim \text{Normal}(\mu_i, \sigma), \quad i = 1, \dots, 30 \\
\mu_i &= \alpha + \sum_{j = 1}^{6} X_{ij} \, \beta_j \\
\alpha, \beta_j &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Cauchy}^{+}(0, 3)
\end{aligned}
$$

```{r}
#| label: code-multiple-linear
#| echo: true
#| eval: false
design <- as.matrix(attitude[, 2:7])
int <- normal(0, 10)
coefs <- normal(0, 10, dim = ncol(design))
sd <- cauchy(0, 3, truncation = c(0, Inf))
mu <- int + design %*% coefs
distribution(attitude$rating) <- normal(mu, sd)
m <- model(int, coefs, sd)
```

## hierarchical_linear

Sepal length on sepal width in `iris`, with an offset for each species, $s[i]$ being flower $i$'s. Setosa's offset is zero and the other two share a scale, $\tau$. The offsets are indexed with `rbind()` and integer indexing. `{r} parameter_text(fit_draws, "hierarchical_linear", tier)`

$$
\begin{aligned}
\text{Sepal.Length}_i &\sim \text{Normal}(\mu_i, \sigma), \quad i = 1, \dots, 150 \\
\mu_i &= \alpha + \beta \, \text{Sepal.Width}_i + \gamma_{s[i]} \\
\gamma_1 &= 0, \qquad \gamma_2, \gamma_3 \sim \text{Normal}(0, \tau) \\
\tau &\sim \text{LogNormal}(0, 1) \\
\alpha, \beta &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Cauchy}^{+}(0, 3)
\end{aligned}
$$

```{r}
#| label: code-hierarchical-linear
#| echo: true
#| eval: false
int <- normal(0, 10)
coef <- normal(0, 10)
sd <- cauchy(0, 3, truncation = c(0, Inf))
species_sd <- lognormal(0, 1)
species_offset <- normal(0, species_sd, dim = 2)
species_effect <- rbind(0, species_offset)
species_id <- as.numeric(iris$Species)
mu <- int + coef * iris$Sepal.Width + species_effect[species_id]
distribution(iris$Sepal.Length) <- normal(mu, sd)
m <- model(int, coef, sd, species_sd, species_offset)
```

## eight_schools

The effect of coaching in eight schools, $y_j$, each measured with a known standard error, $\sigma_j$. The scale of the school effects, $\sigma_\eta$, is itself a parameter, which makes the posterior a funnel. `{r} parameter_text(fit_draws, "eight_schools", tier)`

$$
\begin{aligned}
y_j &\sim \text{Normal}(\theta_j, \sigma_j), \quad j = 1, \dots, 8 \\
\theta_j &= \mu + \xi \, \eta_j \\
\eta_j &\sim \text{Normal}(0, \sigma_\eta) \\
\sigma_\eta &\sim \text{InverseGamma}(1, 1) \\
\mu &\sim \text{Normal}(0, 100) \\
\xi &\sim \text{Normal}(0, 5)
\end{aligned}
$$

```{r}
#| label: code-eight-schools
#| echo: true
#| eval: false
y <- c(28, 8, -3, 7, -1, 1, 18, 12)
sigma_y <- c(15, 10, 16, 11, 9, 11, 10, 18)
N <- length(y)
sigma_eta <- inverse_gamma(1, 1)
eta <- normal(0, sigma_eta, dim = N)
mu_theta <- normal(0, 100)
xi <- normal(0, 5)
theta <- mu_theta + xi * eta
distribution(y) <- normal(theta, sigma_y)
m <- model(sigma_eta, eta, mu_theta, xi)
```

## cjs

A Cormack-Jolly-Seber capture-recapture model: survival, $\phi_t$, and recapture, $p_t$, on each of 20 occasions, from the simulated capture histories $y_{it}$ of 100 animals. Animal $i$ is first caught on occasion $f_i$ and last on $l_i$, so it is known to be alive on every occasion between. $\chi_t$ is the probability that an animal alive on occasion $t$ is never seen again, built by a recursion over the 19 earlier occasions. `{r} parameter_text(fit_draws, "cjs", tier)`

$$
\begin{aligned}
\phi_t, \, p_t &\sim \text{Beta}(1, 1), \quad t = 1, \dots, 20 \\
\chi_{20} &= 1, \qquad \chi_t = (1 - \phi_t) + \phi_t \, (1 - p_{t + 1}) \, \chi_{t + 1} \\
1 &\sim \text{Bernoulli}(\phi_{t - 1}), \quad f_i < t \le l_i \\
y_{it} &\sim \text{Bernoulli}(p_t), \quad f_i < t \le l_i \\
1 &\sim \text{Bernoulli}(\chi_{l_i}), \quad \text{if } l_i < 20
\end{aligned}
$$

```{r}
#| label: code-cjs
#| echo: true
#| eval: false
set.seed(2026)
n_obs <- 100
n_time <- 20
y <- matrix(
  sample(c(0, 1), size = n_obs * n_time, replace = TRUE),
  ncol = n_time
)

first_obs <- apply(y, 1, function(x) min(which(x > 0)))
final_obs <- apply(y, 1, function(x) max(which(x > 0)))
obs_id <- apply(
  y,
  1,
  function(x) {
    seq(min(which(x > 0)), max(which(x > 0)), by = 1)[-1]
  }
)
obs_id <- unlist(obs_id)
capture_vec <- apply(
  y,
  1,
  function(x) x[min(which(x > 0)):max(which(x > 0))][-1]
)
capture_vec <- unlist(capture_vec)

phi <- beta(1, 1, dim = n_time)
p <- beta(1, 1, dim = n_time)

chi <- ones(n_time)
for (i in seq_len(n_time - 1)) {
  tn <- n_time - i
  chi[tn] <- (1 - phi[tn]) + phi[tn] * (1 - p[tn + 1]) * chi[tn + 1]
}

alive_data <- ones(length(obs_id))
not_seen_last <- final_obs != n_time
final_observation <- ones(sum(not_seen_last))

distribution(alive_data) <- bernoulli(phi[obs_id - 1])
distribution(capture_vec) <- bernoulli(p[obs_id])
distribution(final_observation) <- bernoulli(
  chi[final_obs[not_seen_last]]
)

m <- model(phi, p)
```

# 1. Speed

Three tasks are timed on each model, all in one process per version, after each model has been built, optimised and sampled once so that nothing timed includes tracing.

**`model`** runs the model's whole code from "The models" above, from its first line to its call to `model()`, such as `m <- model(int, coef, sd)` for linear. It builds a new model each time, and `bench::mark()` repeats it between `{r} bench_repeats` and `{r} 2 * bench_repeats` times.

**`opt`**, repeated the same way, on the model built once beforehand:

```{r}
#| label: timed-opt
#| output: asis
code_block(sprintf(
  "opt(m, optimiser = adam(), max_iterations = %d)",
  opt_iterations
))
```

**`mcmc`**, run `{r} mcmc_repeats` times and timed by hand, so that each run's draws are kept for section 2. `k` is the version's iterations per draw:

```{r}
#| label: timed-mcmc
#| output: asis
code_block(sprintf(
  "mcmc(m, warmup = %d / k, n_samples = %d / k, chains = %d, n_cores = %d, verbose = FALSE)",
  warmup_iterations,
  sample_iterations,
  mcmc_chains,
  mcmc_cores
))
```

```{r}
#| label: fig-speed
#| fig-cap: "Every timed run, one row per version: the small dots are single runs and the large dot is their mean. The rows in a panel do the same calls, so a row further right took longer. Panels are different tasks and models, on their own scales."
#| fig-height: 6
every_run |>
  ggplot(aes(x = seconds, y = branch, colour = branch)) +
  geom_point(size = 1.5, alpha = 0.4) +
  stat_summary(fun = mean, geom = "point", size = 4) +
  facet_grid(example ~ task, scales = "free_x") +
  labs(x = "seconds", y = NULL) +
  theme(legend.position = "none")
```

<details>
<summary>Table of the median, minimum and maximum, in milliseconds</summary>

```{r}
#| label: speed-range-table
every_run |>
  group_by(task, example, branch) |>
  summarise(
    ms = sprintf(
      "%.1f (%.1f-%.1f)",
      1000 * median(seconds),
      1000 * min(seconds),
      1000 * max(seconds)
    ),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = branch, values_from = ms) |>
  arrange(task, example) |>
  kable()
```

</details>

<details>
<summary>Table of the ratio of medians for each pair of versions: the first version named over the second</summary>

```{r}
#| label: speed-ratio-table
speed |>
  select(example, task, all_of(comparisons$comparison)) |>
  mutate(across(all_of(comparisons$comparison), \(ratio) round(ratio, 2))) |>
  arrange(task, example) |>
  kable()
```

</details>

# 2. Efficiency

## At a fixed number of iterations

The `{r} mcmc_repeats` timed `mcmc()` runs from section 1, each of `{r} warmup_iterations` warmup and `{r} sample_iterations` sampling iterations: the smallest bulk ESS over the model's variables, per second of the run, warmup included.

```{r}
#| label: fig-fixed-efficiency
#| fig-cap: "Effective samples per second for every fixed-length mcmc() run: the small dots are single runs and the large dot is their mean."
#| fig-height: 5
mcmc_runs |>
  mutate(ess_per_second = ess_bulk_min / seconds) |>
  ggplot(aes(x = ess_per_second, y = branch, colour = branch)) +
  geom_point(size = 1.5, alpha = 0.4) +
  stat_summary(fun = mean, geom = "point", size = 4) +
  facet_wrap(~example, ncol = 1, scales = "free_x") +
  expand_limits(x = 0) +
  labs(x = "smallest bulk ESS per second", y = NULL) +
  theme(legend.position = "none")
```

<details>
<summary>Table of the minimum, median and maximum ESS and ESS per second over the runs</summary>

```{r}
#| label: fixed-efficiency-table
mcmc_runs |>
  mutate(ess_per_second = ess_bulk_min / seconds) |>
  group_by(example, branch) |>
  summarise(
    runs = n(),
    ess = sprintf(
      "%.0f (%.0f-%.0f)",
      median(ess_bulk_min),
      min(ess_bulk_min),
      max(ess_bulk_min)
    ),
    ess_per_second = sprintf(
      "%.1f (%.1f-%.1f)",
      median(ess_per_second),
      min(ess_per_second),
      max(ess_per_second)
    ),
    runs_rhat_above_1.01 = sum(rhat_max > 1.01),
    .groups = "drop"
  ) |>
  kable(
    col.names = c(
      "model",
      "version",
      "runs",
      "smallest bulk ESS, median (min-max)",
      "ESS per second, median (min-max)",
      "runs with R-hat > 1.01"
    )
  )
```

</details>

## Time to a target ESS

Seconds of sampling per 1000 effective draws: `mcmc()` with `{r} warmup_iterations` warmup and `{r} sample_iterations` sampling iterations, then `extra_samples()` until the smallest bulk ESS over the model's variables reaches `{r} target_ess` and every R-hat is below 1.01, or until the tier's time limit. Each version runs this `{r} reps` times per model. A run that reaches the time limit is marked `hit_cap` in the table of sampling runs under Details.

```{r}
#| label: fig-sampling
#| fig-cap: "Seconds of sampling per 1000 effective draws. With one run per version and model, a bar is that run; with several, each dot is a run."
#| fig-height: 3.5
one_run_each <- all(count(sampling, example, branch)$n == 1)
if (one_run_each) {
  sampling_plot <- sampling |>
    ggplot(aes(x = seconds_per_1000_ess, y = example, fill = branch)) +
    geom_col(position = "dodge")
} else {
  sampling_plot <- sampling |>
    ggplot(aes(x = seconds_per_1000_ess, y = example, colour = branch)) +
    geom_point(size = 2.5, alpha = 0.7, position = position_dodge(width = 0.6))
}
sampling_plot +
  labs(
    x = "seconds per 1000 effective draws",
    y = NULL,
    fill = NULL,
    colour = NULL
  ) +
  theme(legend.position = "bottom")
```

<details>
<summary>Table of the median for each version, the ratio of medians for each pair, and the number of runs</summary>

```{r}
#| label: sampling-table
sampling_speed |>
  mutate(
    across(all_of(versions), \(seconds) round(seconds, 2)),
    across(all_of(comparisons$comparison), \(ratio) round(ratio, 2))
  ) |>
  select(example, all_of(versions), all_of(comparisons$comparison), runs) |>
  kable(
    col.names = c(
      "model",
      paste(versions, "(s)"),
      comparisons$comparison,
      "runs"
    )
  )
```

</details>

## Memory

```{r}
#| label: rss
rss |>
  mutate(rss_mb = round(rss_mb)) |>
  kable(col.names = c("version", "RSS (MB)"))
```

- RSS is resident set size: the physical memory the measuring process held, in MB.
- It is read once, at the end of each version's run, so it covers every task together, and it is not a peak.
- It is used rather than `bench`'s `mem_alloc`, which sees only R's memory. Most of greta's memory is held by TensorFlow, on the Python side.

# 3. Agreement

For each variable, the difference between two versions' posterior means, divided by the Monte Carlo standard error of that difference [@vehtari2021; @magnusson2025], from the runs in section 2:

$$z = \frac{\bar\theta_{\text{test}} - \bar\theta_{\text{ref}}}{\sqrt{\text{mcse}^2_{\text{test}} + \text{mcse}^2_{\text{ref}}}}$$

The table counts the variables with $|z| > 4$.

```{r}
#| label: fig-agreement
#| fig-cap: "Every variable's difference in means, in MCSE units, for each pair of versions. The shaded band is |z| up to 4."
#| fig-height: 6
ggplot(posterior_comparison, aes(x = z_mean, y = example)) +
  annotate(
    "rect",
    xmin = -4,
    xmax = 4,
    ymin = -Inf,
    ymax = Inf,
    fill = "grey88"
  ) +
  geom_vline(xintercept = 0, colour = "grey40") +
  stat_dots(side = "both", scale = 0.6) +
  facet_wrap(~comparison, ncol = 1) +
  expand_limits(x = c(-6, 6)) +
  labs(x = "difference in means (MCSE units)", y = NULL)
```

<details>
<summary>Table of the largest |z| for each model and pair of versions</summary>

```{r}
#| label: agreement-table
agreement |>
  select(
    comparison,
    example,
    variables,
    max_abs_z,
    n_disagree,
    worst_variable
  ) |>
  mutate(max_abs_z = round(max_abs_z, 2)) |>
  kable(
    col.names = c(
      "comparison",
      "model",
      "variables",
      "largest |z|",
      "variables with |z| > 4",
      "variable with the largest |z|"
    )
  )
```

</details>

# 4. Convergence

R-hat and effective sample size over each model's variables, for `{r} reps` seeded fits of each model on each version, with seeds 1 to `{r} reps`: `{r} mcmc_chains` chains of `{r} warmup_iterations` warmup and `{r} sample_iterations` sampling iterations each. Versions whose `mcmc()` follows `set.seed()` start each seed from the same random numbers; greta 0.6.0's does not.

```{r}
#| label: fig-diagnostics
#| fig-cap: "Bulk ESS of each fit over the model's variables, one line per seed: the line runs from the smallest to the largest, the point is the median. The dashed line is 400. Log scale."
#| fig-height: 6
fit_diagnostics |>
  ggplot(aes(y = branch, colour = branch, group = seed)) +
  geom_vline(xintercept = 400, linetype = "dashed", colour = "grey50") +
  geom_pointrange(
    aes(x = ess_bulk_median, xmin = ess_bulk_min, xmax = ess_bulk_max),
    position = position_dodge(width = 0.6),
    size = 0.3
  ) +
  scale_x_log10() +
  facet_wrap(~example, ncol = 1) +
  labs(x = "bulk ESS", y = NULL) +
  theme(legend.position = "none")
```

<details>
<summary>Table of R-hat and ESS for each fit</summary>

```{r}
#| label: diagnostics-table
fit_diagnostics |>
  mutate(
    rhat_max = round(rhat_max, 3),
    across(starts_with("ess"), round)
  ) |>
  arrange(example, branch, seed) |>
  kable(
    col.names = c(
      "model",
      "version",
      "seed",
      "largest R-hat",
      "bulk ESS, min",
      "bulk ESS, median",
      "bulk ESS, max",
      "tail ESS, min"
    )
  )
```

</details>

# 5. Fits

The seed 1 fit of each model from section 4. Each version is a colour. The fitted values are each model's expected value for the data, such as its regression line, not new data simulated from the model. The data are the black points.

## linear

`{r} unmeasured_note("linear", example_names, tier)`

```{r}
#| label: fig-fit-linear
#| fig-cap: "The fitted line's median and 95% interval."
#| fig-height: 4
if ("linear" %in% example_names) {
  gg_fit(fitted_values, "linear")
}
```

```{r}
#| label: fig-densities-linear
#| fig-cap: "Posterior densities."
#| fig-height: 3
if ("linear" %in% example_names) {
  gg_densities(fit_draws, "linear")
}
```

## multiple_linear

`{r} unmeasured_note("multiple_linear", example_names, tier)`

```{r}
#| label: fig-fit-multiple-linear
#| fig-cap: "Each fitted value's median, 50% and 95% intervals, against the value it fits. The dashed line is where the two are equal."
#| fig-height: 4
if ("multiple_linear" %in% example_names) {
  gg_fit(fitted_values, "multiple_linear")
}
```

```{r}
#| label: fig-densities-multiple-linear
#| fig-cap: "Posterior densities."
#| fig-height: 4.5
if ("multiple_linear" %in% example_names) {
  gg_densities(fit_draws, "multiple_linear")
}
```

## hierarchical_linear

`{r} unmeasured_note("hierarchical_linear", example_names, tier)`

```{r}
#| label: fig-fit-hierarchical-linear
#| fig-cap: "The fitted line's median and 95% interval, by species."
#| fig-height: 4
if ("hierarchical_linear" %in% example_names) {
  gg_fit(fitted_values, "hierarchical_linear")
}
```

```{r}
#| label: fig-densities-hierarchical-linear
#| fig-cap: "Posterior densities."
#| fig-height: 4.5
if ("hierarchical_linear" %in% example_names) {
  gg_densities(fit_draws, "hierarchical_linear")
}
```

## eight_schools

`{r} unmeasured_note("eight_schools", example_names, tier)`

```{r}
#| label: fig-fit-eight-schools
#| fig-cap: "Each school's estimated effect: median, 50% and 95% intervals. The black points are the observed effects."
#| fig-height: 4
if ("eight_schools" %in% example_names) {
  gg_fit(fitted_values, "eight_schools")
}
```

```{r}
#| label: fig-densities-eight-schools
#| fig-cap: "Posterior densities of the first eight variables."
#| fig-height: 4.5
if ("eight_schools" %in% example_names) {
  gg_densities(fit_draws, "eight_schools")
}
```

## cjs

`{r} unmeasured_note("cjs", example_names, tier)`

```{r}
#| label: fig-fit-cjs
#| fig-cap: "Survival on each occasion: median, 50% and 95% intervals. There is no data to compare it with."
#| fig-height: 4
if ("cjs" %in% example_names) {
  gg_fit(fitted_values, "cjs")
}
```

```{r}
#| label: fig-densities-cjs
#| fig-cap: "Posterior densities of the first eight variables."
#| fig-height: 4.5
if ("cjs" %in% example_names) {
  gg_densities(fit_draws, "cjs")
}
```

# 6. Further measurements

```{r}
#| label: further
#| output: asis
if (nzchar(params$further)) {
  cat(knitr::knit_child(params$further, quiet = TRUE, envir = environment()))
} else {
  cat("None for this run.\n")
}
```

# Details

::: panel-tabset

## Sampling runs

```{r}
#| label: sampling-detail
sampling |>
  select(
    branch,
    example,
    rep,
    elapsed,
    mcmc_iterations,
    ess_bulk_min,
    ess_bulk_median,
    rhat_max,
    n_variables,
    hit_cap
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  rename(version = branch) |>
  kable()
```

## Posteriors

The mean and 90% interval, the 5% to 95% quantiles, of each model's six widest variables, from the runs in section 2.

```{r}
#| label: widest-variables
widest_variables <- pooled_posterior |>
  group_by(example) |>
  slice_max(q95 - q5, n = 6) |>
  ungroup()
```

```{r}
#| label: fig-intervals
#| fig-height: 7
ggplot(widest_variables, aes(y = variable, colour = branch)) +
  geom_pointinterval(
    aes(x = mean, xmin = q5, xmax = q95),
    position = position_dodge(width = 0.5)
  ) +
  facet_wrap(~example, scales = "free", ncol = 2) +
  labs(x = NULL, y = NULL, colour = NULL) +
  theme(legend.position = "bottom")
```

## Provenance

```{r}
#| label: provenance
str(provenance)
```

:::

# Methods

**Diagnostics.** Rank-normalised split-R̂ and bulk and tail effective sample size [@vehtari2021]. ESS above 400 is where the diagnostics become reliable, not where accuracy becomes sufficient. `coda::effectiveSize()` is not used: it fits an AR model per chain with no between-chain information, so it reports a large ESS for chains stuck in different modes.

**Comparing posteriors.** Differences in posterior means are divided by the combined Monte Carlo standard error, so bias is separable from Monte Carlo noise [@magnusson2025]. A two-sample Kolmogorov-Smirnov test on the draws is not used: it assumes both samples are independent draws, and MCMC draws are autocorrelated, so a correct sampler is rejected too often [@talts2018]. With one $z$ per variable, the largest of many exceeds 2 routinely, so the table counts variables beyond 4.

**Timing.** Timing noise is one-sided [@chen2016]: a run can be slowed by something else on the machine, never sped up. The plots show every run so the spread can be read beside the mean. Sampling measurements vary between runs of the same version, so each is repeated: three times at the quick tier and five at the standard and thorough. Effect sizes are reported rather than significance tests [@kalibera2013].

**What is not measured.** Gradient evaluations are not counted, so a difference in wall time cannot be split into fewer gradients and cheaper gradients [@hoffman2021]. There is no external reference posterior, so a bug present in every version would not show; posteriordb [@magnusson2025] and Inference Gym hold reference posteriors.

# References

::: {#refs}
:::
