---
title: "What driving the optimiser loop from R costs"
format: gfm
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(here)
library(fs)
library(bench)
library(dplyr)
library(tibble)
library(rlang)
library(knitr)

run_dir <- here("2026-08-22-optimiser-r-loop")
results_rds <- path(run_dir, "results.rds")
if (!file_exists(results_rds)) {
  stop("no results.rds - run 2026-08-22-optimiser-r-loop/01-measure.R first", call. = FALSE)
}
results <- readRDS(results_rds)
knitr::read_chunk(path(run_dir, "01-measure.R"))
summary(results$loop_cost, relative = TRUE)
med <- as.numeric(results$loop_cost$median)
speedup <- med[1] / med[2]
per_step_ms <- 1000 * (med[1] - med[2]) / results$n_steps
opt_per_iter <- results$opt_scaling$per_iter_ms
```

## The question

`tf_optimiser` and `tf_compat_optimiser` run a `while` loop in R, stepping into
TensorFlow once per iteration. `tfp_optimiser` does not. What does the R-driven
loop cost, and what would moving it into TF actually buy?

## Part 1: `opt()` cost per iteration does not move with model size

```{r}
#| label: scaling
results$opt_scaling |>
  transmute(n_free, iterations, `per iteration (ms)` = round(per_iter_ms, 2)) |>
  kable()
```

`r sprintf("%.2f", min(opt_per_iter))`–`r sprintf("%.2f", max(opt_per_iter))` ms
per iteration across a 500-fold change in the number of free parameters. The
per-iteration cost is therefore fixed overhead, not gradient work.

## Part 2: the ceiling, on identical arithmetic

greta removed. The same `r results$n_steps` gradient-descent steps, driven from
R and driven inside TF, so the difference is the cost of crossing back into R:

```{r loops}
#| eval: false
#| echo: true
```

```{r}
#| label: loop-cost
results$loop_cost |>
  select(expression, min, median, `itr/sec`) |>
  mutate(across(c(min, median), format)) |>
  kable(align = "r")
```

Driving the loop inside TF is **`r sprintf("%.0f", speedup)`x** faster here,
which puts the bare round trip at about **`r sprintf("%.3f", per_step_ms)` ms
per step**.

## What that means for the fix

The round trip is real but it is *not* most of what `opt()` spends per
iteration: `r sprintf("%.3f", per_step_ms)` ms of round trip against
`r sprintf("%.2f", median(opt_per_iter))` ms per `opt()` iteration, so roughly
`r sprintf("%.0f", 100 * per_step_ms / median(opt_per_iter))`% of it.

The rest is greta's own loop body, which runs in R: the convergence check, the
`$numpy()` conversion of the objective, and `check_numerical_overflow()`. So
moving the `while` loop into TF only pays if the body moves with it. Wrapping
the loop alone and leaving the body in R would recover the smaller share.

## Environment

```{r}
#| label: provenance
link_source <- function(source) {
  sub(
    "([A-Za-z0-9_.-]+/[A-Za-z0-9_.-]+)@([0-9a-f]{7,40})",
    "[\\1@\\2](https://github.com/\\1/commit/\\2)",
    source
  )
}

platform <- results$provenance$platform
packages <- as_tibble(results$provenance$packages)

bind_rows(
  tibble(
    item = c("run at", "OS", "system", "CPU", "cores detected", "R"),
    value = c(
      results$provenance$run_at, platform$os, platform$system,
      results$provenance$cpu, as.character(results$provenance$cores),
      platform$version
    )
  ),
  packages |>
    transmute(item = paste("R package:", package),
              value = paste0(ondiskversion, " (", link_source(source), ")"))
) |>
  set_names(c("", "")) |>
  kable()
```
