---
title: "Wiring `model(compile=)` through to `jit_compile`"
format: gfm
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(here)
library(fs)
library(bench)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(rlang)
library(knitr)

run_dir <- here("2026-08-19-jit-compile-vs-main")
results_rds <- path(run_dir, "results.rds")
if (!file_exists(results_rds)) {
  stop("no results.rds - run 2026-08-19-jit-compile-vs-main/01-measure.R first", call. = FALSE)
}
results <- readRDS(results_rds)

branch_current <- results$raw$branch[[1]]
branch_reference <- results$raw$branch[[2]]

# one bench_mark carrying both branches, so bench's own summary() and
# autoplot() can be used rather than hand-rolled ratios
combined <- map2(results$raw$branch, results$raw$result, \(b, res) {
  tb <- as_tibble(res$timings)
  tb$branch <- b
  tb
}) |>
  list_rbind() |>
  as_bench_mark()

# relative within a task, not across tasks: build and mcmc_short are different
# work and normalising them together would say nothing
relative_for <- function(task) {
  sub <- combined[as.character(combined$expression) == task, ]
  sub$expression <- sub$branch
  summary(as_bench_mark(sub), relative = TRUE) |>
    transmute(branch = as.character(expression), task = task,
              min = round(min, 3), median = round(median, 3))
}
```

## The question

`model(compile = TRUE)` is documented as applying XLA JIT compilation but has
been inert since January 2023. The `wire-jit-compile` branch passes it through at
the two `tf_function()` sites in `dag_class.R`. Does that make greta faster?

Both branches resolve their own Python stack, recorded below, so a difference
here is the effect of the wiring rather than of a version bump.

The model is a plain regression on purpose: XLA cannot compile the gradients of
`FillScaleTriL` or `CorrelationCholesky`, so `wishart()`, `lkj_correlation()` and
`cholesky_variable()` models error on the wired branch instead of producing a
timing.

## Results

```{r}
#| label: absolute
combined |>
  as_tibble() |>
  transmute(task = as.character(expression), branch,
            median = format(median), `itr/sec` = round(`itr/sec`, 1)) |>
  pivot_wider(names_from = branch, values_from = c(median, `itr/sec`)) |>
  kable(align = "r")
```

Relative within each task, taking the faster branch as 1:

```{r}
#| label: relative
map(unique(as.character(combined$expression)), relative_for) |>
  list_rbind() |>
  select(task, branch, min, median) |>
  kable(align = "r")
```

```{r}
#| label: fig-dist
#| fig-cap: "Every iteration of the sampling task, both branches, log scale. The distributions overlap heavily - the difference is a shift in a wide, noisy spread, not a clean separation."
#| fig-height: 2.5
#| fig-width: 7
mcmc_times <- map2(results$raw$branch, results$raw$result, \(b, res) {
  i <- which(as.character(res$timings$expression) == "mcmc_short")
  tibble(branch = b, seconds = as.numeric(res$timings$time[[i]]))
}) |>
  list_rbind()

ggplot(mcmc_times, aes(x = seconds, y = branch)) +
  geom_jitter(height = 0.15, alpha = 0.5, colour = "#F8766D") +
  scale_x_continuous(labels = \(x) paste0(round(1000 * x), "ms")) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 11)
```

## What this says

```{r}
#| label: verdict
w <- mcmc_times$seconds[mcmc_times$branch == branch_current] * 1000
m <- mcmc_times$seconds[mcmc_times$branch == branch_reference] * 1000
tt <- t.test(w, m)
diff_ms <- median(m) - median(w)
pct <- 100 * (1 - median(w) / median(m))
ci <- sort(-tt$conf.int)
```

Sampling is **`r sprintf("%.1f", pct)`% faster** on the wired branch:
`r sprintf("%.0f", median(w))` ms against `r sprintf("%.0f", median(m))` ms,
a difference of `r sprintf("%.0f", diff_ms)` ms.

The spread is wide — a within-branch standard deviation of about
`r sprintf("%.0f", sd(m))` ms, comparable to the effect itself — so this needs
`r length(w)` iterations per branch to see. A Welch test gives
p = `r signif(tt$p.value, 2)`, with a 95% interval on the difference of
`r sprintf("%.0f", ci[1])` to `r sprintf("%.0f", ci[2])` ms. **An earlier run at
10 iterations per branch could not resolve it**, and gave estimates ranging from
27 ms to 62 ms; treat any single small run of this comparison as uninformative.

Model definition is unchanged, which is what you would expect: XLA compiles at
first call, not at definition.

So the upside is real but modest, and it applies only to the models XLA can
handle.

## Environment

```{r}
#| label: provenance
link_source <- function(source) {
  sub("([A-Za-z0-9_.-]+/[A-Za-z0-9_.-]+)@([0-9a-f]{7,40})",
      "[\\1@\\2](https://github.com/\\1/commit/\\2)", source)
}

platform <- results$provenance$platform
packages <- as_tibble(results$provenance$packages)

stacks <- map2(results$raw$branch, results$raw$result, \(b, res) {
  tibble(item = paste("stack,", b),
         value = paste("python", res$stack[["python"]],
                       "| tensorflow", res$stack[["tensorflow"]],
                       "| tfp", res$stack[["tfp"]]))
}) |>
  list_rbind()

bind_rows(
  tibble(
    item = c("run at", "OS", "system", "CPU", "cores detected", "R"),
    value = c(results$provenance$run_at, platform$os, platform$system,
              results$provenance$cpu, as.character(results$provenance$cores),
              platform$version)
  ),
  packages |>
    transmute(item = paste("R package:", package),
              value = paste0(ondiskversion, " (", link_source(source), ")")),
  stacks,
  tibble(item = paste("greta,", c("current", "reference")),
         value = sprintf("[`%s`](https://github.com/greta-dev/greta/commit/%s)",
                         unname(results$shas), unname(results$shas)))
) |>
  set_names(c("", "")) |>
  kable()
```
