---
title: "Wiring `model(compile=)` through to `jit_compile`"
format: gfm
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(here)
library(fs)
library(bench)
library(dplyr)
library(tidyr)
library(purrr)
library(rlang)
library(tibble)
library(knitr)

run_dir <- here("2026-08-19-jit-compile-vs-main")
results_rds <- path(run_dir, "results.rds")
if (!file_exists(results_rds)) {
  stop("no results.rds - run 2026-08-19-jit-compile-vs-main/01-measure.R first", call. = FALSE)
}
results <- readRDS(results_rds)
```

## The question

`model(compile = TRUE)` is documented as applying XLA JIT compilation but has
been inert since January 2023. The `wire-jit-compile` branch passes it through
at the two `tf_function()` sites in `dag_class.R`. Does that make greta faster?

A ratio below 1 means the branch is faster. Both branches resolve their own
Python stack, recorded below, so a ratio here is the effect of the wiring rather
than of a version bump.

The model is a plain regression on purpose: XLA cannot compile the gradients of
`FillScaleTriL` or `CorrelationCholesky`, so anything using `wishart()`,
`lkj_correlation()` or `cholesky_variable()` errors on the wired branch instead
of producing a timing.

## Results

```{r}
#| label: timings
timings <- results$raw |>
  mutate(timings = map(result, "timings")) |>
  select(branch, timings) |>
  unnest(timings) |>
  mutate(task = as.character(expression))

timings |>
  select(task, branch, median) |>
  pivot_wider(names_from = branch, values_from = median) |>
  kable(digits = 4)
```

```{r}
#| label: ratios
timings |>
  select(task, branch, median, min) |>
  pivot_longer(c(median, min), names_to = "statistic") |>
  pivot_wider(names_from = branch, values_from = value) |>
  mutate(ratio = round(as.numeric(.data[[names(.)[3]]]) / as.numeric(.data[[names(.)[4]]]), 3)) |>
  select(task, statistic, ratio) |>
  pivot_wider(names_from = statistic, values_from = ratio) |>
  kable()
```

`bench::mark()`. `min` is the statistic least contaminated by garbage collection
and scheduling; where the two branches share byte-identical code, it is the one
to read.

## Environment

```{r}
#| label: provenance
# sessioninfo reports a GitHub source as "Github (owner/repo@sha)". Turn the
# sha into a link so it can be clicked rather than copied.
link_source <- function(source) {
  sub(
    "([A-Za-z0-9_.-]+/[A-Za-z0-9_.-]+)@([0-9a-f]{7,40})",
    "[\\1@\\2](https://github.com/\\1/commit/\\2)",
    source
  )
}

platform <- results$provenance$platform
packages <- as_tibble(results$provenance$packages)

stacks <- results$raw |>
  mutate(stack = map_chr(result, \(x) paste(
    "python", x$stack[["python"]],
    "| tensorflow", x$stack[["tensorflow"]],
    "| tfp", x$stack[["tfp"]]
  ))) |>
  transmute(item = paste("stack,", branch), value = stack)

bind_rows(
  tibble(
    item = c("run at", "OS", "system", "CPU", "cores detected", "R"),
    value = c(
      results$provenance$run_at,
      platform$os,
      platform$system,
      results$provenance$cpu,
      as.character(results$provenance$cores),
      platform$version
    )
  ),
  packages |>
    transmute(
      item = paste("R package:", package),
      value = paste0(ondiskversion, " (", link_source(source), ")")
    ),
  stacks,
  tibble(
    item = paste("greta,", c("current", "reference")),
    # linked so a reader can click through to the commit that was measured
    value = sprintf(
      "[`%s`](https://github.com/greta-dev/greta/commit/%s)",
      unname(results$shas), unname(results$shas)
    )
  )
) |>
  set_names(c("", "")) |>
  kable()
```
