---
title: "Warmup schemes on the tf-warmup-i547 branch"
description: "greta's warmup tuning, the same with greta#841's count fixed, Stan's windowed scheme from greta#853, and TFP's own windowed HMC and NUTS, run inside TensorFlow on the same models: the step sizes they tune, how long they take, and the effective samples they give."
date: 2026-10-07
bibliography: ../../references.bib
---

The tf-warmup-i547 branch of greta tunes warmup inside TensorFlow ([greta#547](https://github.com/greta-dev/greta/issues/547)), which makes it cheap to try other ways of tuning. This post compares five, on the same five models:

- **greta**: greta's own tuning, as on main. It adapts the step size, `epsilon`, by dual averaging over the first 10% and last 60% of warmup [@hoffman2014], and `diag_sd` from the variance of every warmup draw so far between 10% and 40%, shrunk by a count of the *rejected* proposals.
- **greta_841**: the same, shrunk by a count of the accepted proposals, as [greta#841](https://github.com/greta-dev/greta/issues/841) says it was meant to be.
- **windowed**: Stan's windowed adaptation, as proposed in [greta#853](https://github.com/greta-dev/greta/issues/853). After a buffer that adapts `epsilon` only, `diag_sd` is estimated afresh in windows ending at 25%, 45% and 90% of warmup, each from its own draws, and dual averaging restarts after each.
- **tfp_hmc** and **tfp_nuts**: TFP's own windowed adaptation of HMC and NUTS, as its `windowed_adaptive_hmc()` and `windowed_adaptive_nuts()` use it, driven from the same log density. TFP's HMC takes a fixed number of leapfrog steps, where greta's draws one between `Lmin` and `Lmax` every iteration.

The first three are an internal choice on the branch, not something `mcmc()` offers yet.

```{r}
#| label: setup
#| message: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(bench)
library(here)

post_path <- "posts/2026-10-07-warmup-adaptation-i853"

# each `# ---- label ----` section of the scripts, shown below by its label
knitr::read_chunk(here(post_path, "benchmark.R"))
knitr::read_chunk(here(post_path, "run.R"))

# links to the scripts on GitHub, at the commit being rendered when GitHub
# Actions renders the site
github_ref <- Sys.getenv("GITHUB_SHA", "main")
github_file_url <- function(file) {
  paste0(
    "https://github.com/greta-dev/greta.benchmarks/blob/",
    github_ref,
    "/",
    post_path,
    "/",
    file
  )
}
# Quarto escapes the markdown in an inline `{r}` result unless it is wrapped
# in I(), so both link helpers return I()
file_link <- function(file) {
  I(paste0("[`", file, "`](", github_file_url(file), ")"))
}

# a link to the lines of a script's `# ---- label ----` section
section_link <- function(file, label) {
  lines <- readLines(here(post_path, file))
  headings <- grep("^# ---- .+ ----$", lines)
  start <- headings[lines[headings] == paste("# ----", label, "----")]
  end <- c(headings, length(lines) + 1)[match(start, headings) + 1] - 1
  while (end > start && !nzchar(trimws(lines[end]))) {
    end <- end - 1
  }
  I(paste0(
    "[`", label, "`](", github_file_url(file), "#L", start, "-L", end, ")"
  ))
}

theme_set(theme_bw(base_size = 13))

sessions <- list.files(
  here(post_path, "results"),
  pattern = "^session-[0-9]+[.]rds$",
  full.names = TRUE
) |>
  lapply(readRDS)
n_sessions <- length(sessions)
results <- lapply(sessions, \(session_file) session_file$results$branch)
bench_settings <- results[[1]]$bench_settings
model_levels <- names(results[[1]]$schemes)
method_levels <- c(bench_settings$schemes, paste0("tfp_", bench_settings$tfp_kinds))
method_colours <- setNames(
  c("#999999", "#56B4E9", "#009E73", "#E69F00", "#CC79A7"),
  method_levels
)
method_scale <- scale_colour_manual(values = method_colours, guide = "none")
method_fill <- scale_fill_manual(values = method_colours, guide = "none")
jitter <- position_jitter(width = 0, height = 0.2, seed = 2026)

# one row per timed call, from every session, model and method
every_call <- lapply(seq_along(results), function(i) {
  lapply(model_levels, function(model_name) {
    schemes <- results[[i]]$schemes[[model_name]]$marks
    tfp <- results[[i]]$tfp[[model_name]]$marks
    marks <- list(
      tibble(method = schemes$scheme, time = schemes$time),
      tibble(method = paste0("tfp_", tfp$kind), time = tfp$time)
    )
    bind_rows(marks) |>
      mutate(call = lapply(time, seq_along)) |>
      unnest(c(time, call)) |>
      mutate(
        session = sessions[[i]]$session,
        model = model_name,
        seconds = as.numeric(time),
        .keep = "unused"
      )
  }) |>
    bind_rows()
}) |>
  bind_rows() |>
  mutate(
    method = factor(method, levels = method_levels),
    model = factor(model, levels = model_levels),
    calls = factor(
      ifelse(call == 1, "first call on a new model", "second call on the same model"),
      levels = c("first call on a new model", "second call on the same model")
    )
  )

# a posterior summary of every method's second-call draws, in every session
posterior_summaries <- lapply(seq_along(results), function(i) {
  lapply(model_levels, function(model_name) {
    draws <- c(
      results[[i]]$schemes[[model_name]]$draws,
      setNames(
        results[[i]]$tfp[[model_name]]$draws,
        paste0("tfp_", names(results[[i]]$tfp[[model_name]]$draws))
      )
    )
    lapply(names(draws), function(method) {
      tibble(
        session = sessions[[i]]$session,
        model = model_name,
        method = method,
        posterior::summarise_draws(
          draws[[method]],
          "mean",
          "sd",
          "rhat",
          "ess_bulk"
        )
      )
    }) |>
      bind_rows()
  }) |>
    bind_rows()
}) |>
  bind_rows() |>
  mutate(
    method = factor(method, levels = method_levels),
    model = factor(model, levels = model_levels)
  )

# three significant figures each, where format() would pad a vector to the
# decimal places of its smallest value
format_number <- function(x) {
  formatC(signif(x, 3), format = "fg", digits = 3, big.mark = ",")
}

# a searchable table, its numbers rounded for reading
dt_table <- function(data, digits = 3) {
  table <- DT::datatable(
    data,
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 10, autoWidth = TRUE)
  )
  decimal_columns <- names(data)[vapply(data, is.double, logical(1))]
  if (length(decimal_columns) == 0) {
    return(table)
  }
  DT::formatRound(table, decimal_columns, digits)
}

session_caption <- "Each small dot is one session's second call. The large faint dot behind them is their mean."
```

Two scripts made every result here. `{r} file_link("benchmark.R")` is what the branch runs, in a fresh R session, and `{r} file_link("run.R")` runs it in each session. Each section below shows the part of a script it is about, read out of the script by its `# ---- label ----` heading and linked to those lines on GitHub. All the code on this page is folded: click a fold to open it.

```{r}
#| label: run-script
#| file: run.R
#| eval: false
#| code-summary: "run.R"
```

```{r}
#| label: benchmark-script
#| file: benchmark.R
#| eval: false
#| code-summary: "benchmark.R"
```

## What was compared

The branch alone, installed as `{r} section_link("run.R", "run-versions")` says, in `{r} n_sessions` sessions by `{r} section_link("run.R", "run-sessions")`. It ran on TensorFlow `{r} results[[1]]$provenance$tensorflow_version` and TensorFlow Probability `{r} results[[1]]$provenance$tfp_version`, on `{r} results[[1]]$provenance$machine` with `{r} results[[1]]$provenance$cores` cores.

```{r}
#| label: run-versions
#| eval: false
#| code-summary: "run.R: run-versions"
```

```{r}
#| label: run-sessions
#| eval: false
#| code-summary: "run.R: run-sessions"
```

Each session ran `benchmark.R` from top to bottom, with the seed and settings in its `{r} section_link("benchmark.R", "benchmark-setup")` and `{r} section_link("benchmark.R", "benchmark-settings")` sections:

```{r}
#| label: benchmark-setup
#| eval: false
#| code-summary: "benchmark.R: benchmark-setup"
```

```{r}
#| label: benchmark-settings
#| eval: false
#| code-summary: "benchmark.R: benchmark-settings"
```

greta's three schemes run through `mcmc()`, in `{r} section_link("benchmark.R", "benchmark-schemes")`; TFP's two through its windowed sampling, in `{r} section_link("benchmark.R", "benchmark-tfp")`:

```{r}
#| label: benchmark-schemes
#| eval: false
#| code-summary: "benchmark.R: benchmark-schemes"
```

```{r}
#| label: benchmark-tfp
#| eval: false
#| code-summary: "benchmark.R: benchmark-tfp"
```

## The models

::: panel-tabset

### normals

Twenty independent normals with standard deviations from 0.01 to 100, the test in [greta#853](https://github.com/greta-dev/greta/issues/853): a posterior whose scales span four orders of magnitude, so a `diag_sd` that is off on one scale shows. Its section is `{r} section_link("benchmark.R", "model-normals")`.

```{r}
#| label: model-normals
#| code-summary: "benchmark.R: model-normals"
```

### linear

```{r}
#| label: model-linear
#| code-summary: "benchmark.R: model-linear"
```

### multiple_linear

```{r}
#| label: model-multiple-linear
#| code-summary: "benchmark.R: model-multiple-linear"
```

### hierarchical_linear

```{r}
#| label: model-hierarchical-linear
#| code-summary: "benchmark.R: model-hierarchical-linear"
```

### eight_schools

```{r}
#| label: model-eight-schools
#| code-summary: "benchmark.R: model-eight-schools"
```

:::

The other four are greta's examples, as in the [thinning post](../2026-10-06-thinning-i318/#the-models), where they are also written out in maths. Each model's runs are its `bench-` section: `{r} section_link("benchmark.R", "bench-normals")`, `{r} section_link("benchmark.R", "bench-linear")`, `{r} section_link("benchmark.R", "bench-multiple-linear")`, `{r} section_link("benchmark.R", "bench-hierarchical-linear")` and `{r} section_link("benchmark.R", "bench-eight-schools")`.

## The scales each scheme found, on the normals

For greta's three schemes, the tuned `diag_sd` of each of the twenty normals, divided by its true standard deviation: 1 is a perfect estimate. TFP's schemes keep their estimate inside their kernel results, which the benchmark does not return, so they are left out here; their posterior standard deviations are in the next plot.

```{r}
#| label: tuned-scales
true_sd <- results[[1]]$normals_sd
tuned_scales <- lapply(seq_along(results), function(i) {
  tuned <- results[[i]]$schemes$normals$tuned
  lapply(names(tuned), function(scheme) {
    ratio <- tuned[[scheme]]$diag_sd / true_sd
    tibble(
      session = sessions[[i]]$session,
      method = scheme,
      true_sd = true_sd,
      ratio = ratio
    )
  }) |>
    bind_rows()
}) |>
  bind_rows() |>
  mutate(method = factor(method, levels = method_levels))

posterior_scales <- posterior_summaries |>
  filter(model == "normals") |>
  mutate(
    true_sd = true_sd[as.integer(sub("x\\[([0-9]+),1\\]", "\\1", variable))],
    ratio = sd / true_sd
  )
```

::: panel-tabset

### Tuned diag_sd

```{r}
#| label: plot-tuned-scales
#| fig-cap: "Each dot is one parameter in one session. The grey line is a perfect estimate."
#| fig-height: 5
ggplot(tuned_scales, aes(x = true_sd, y = ratio, colour = method)) +
  geom_hline(yintercept = 1, colour = "grey60") +
  geom_point(alpha = 0.7) +
  scale_x_log10() +
  scale_y_log10() +
  scale_colour_manual(values = method_colours) +
  facet_wrap(~method) +
  labs(x = "True standard deviation", y = "Tuned diag_sd / true sd") +
  theme(legend.position = "none")
```

### Posterior sd

```{r}
#| label: plot-posterior-scales
#| fig-cap: "Each dot is one parameter in one session: the standard deviation of its draws over the true one. The grey line is a perfect estimate."
#| fig-height: 5
ggplot(posterior_scales, aes(x = true_sd, y = ratio, colour = method)) +
  geom_hline(yintercept = 1, colour = "grey60") +
  geom_point(alpha = 0.7) +
  scale_x_log10() +
  scale_y_log10() +
  scale_colour_manual(values = method_colours) +
  facet_wrap(~method) +
  labs(x = "True standard deviation", y = "Posterior sd / true sd") +
  theme(legend.position = "none")
```

:::

## Speed

Seconds for each call. The first call on a new model traces its TensorFlow functions; the second reuses them.

```{r}
#| label: speed
speed_medians <- every_call |>
  filter(call == 2) |>
  group_by(model, method) |>
  summarise(seconds = median(seconds), .groups = "drop")
```

```{r}
#| label: plot-speed
#| fig-cap: "Each dot is one call, from every session. The box marks their median and quartiles."
#| fig-height: 14
ggplot(every_call, aes(x = seconds, y = method, colour = method)) +
  geom_boxplot(aes(fill = method), width = 0.4, outlier.shape = NA, alpha = 0.2) +
  geom_point(position = jitter, alpha = 0.7) +
  method_scale +
  method_fill +
  facet_wrap(vars(model, calls), ncol = 2, scales = "free_x") +
  labs(x = "Seconds", y = NULL)
```

<details><summary>Table: median seconds of the second call</summary>

```{r}
#| label: table-speed
speed_medians |>
  pivot_wider(names_from = method, values_from = seconds) |>
  dt_table(digits = 2)
```

</details>

## Effective samples

The bulk effective sample size of the parameter with the smallest, as @hoffman2014 report, and that per second of the second call, warmup included [@girolami2011]. Every method ran `{r} bench_settings$chains` chains of `{r} bench_settings$warmup` warmup iterations and `{r} bench_settings$n_samples` draws.

```{r}
#| label: efficiency
second_calls <- every_call |>
  filter(call == 2) |>
  select(session, model, method, seconds)

efficiency <- posterior_summaries |>
  group_by(session, model, method) |>
  summarise(
    smallest_ess_bulk = min(ess_bulk),
    largest_rhat = max(rhat),
    .groups = "drop"
  ) |>
  left_join(second_calls, by = c("session", "model", "method")) |>
  mutate(smallest_ess_per_second = smallest_ess_bulk / seconds)

efficiency_means <- efficiency |>
  group_by(model, method) |>
  summarise(
    across(c(smallest_ess_bulk, smallest_ess_per_second, largest_rhat), mean),
    .groups = "drop"
  )

gg_sessions <- function(x, x_label) {
  ggplot(efficiency, aes(x = {{ x }}, y = method, colour = method)) +
    geom_point(data = efficiency_means, size = 5, alpha = 0.25) +
    geom_point(position = jitter, size = 1.2) +
    method_scale +
    facet_wrap(~model, ncol = 1, scales = "free_x") +
    labs(x = x_label, y = NULL)
}
```

::: panel-tabset

### Smallest bulk ESS

```{r}
#| label: plot-ess
#| fig-cap: !expr session_caption
#| fig-height: 12
gg_sessions(smallest_ess_bulk, "Smallest bulk ESS")
```

### Smallest bulk ESS per second

```{r}
#| label: plot-ess-per-second
#| fig-cap: !expr session_caption
#| fig-height: 12
gg_sessions(smallest_ess_per_second, "Smallest bulk ESS per second")
```

### Largest R-hat

```{r}
#| label: plot-rhat
#| fig-cap: !expr session_caption
#| fig-height: 12
gg_sessions(largest_rhat, "Largest R-hat")
```

:::

<details><summary>Table: mean over sessions</summary>

```{r}
#| label: table-efficiency-means
dt_table(efficiency_means, digits = 2)
```

</details>

<details><summary>Table: every session</summary>

```{r}
#| label: table-efficiency
dt_table(efficiency, digits = 2)
```

</details>

## Session details

What the branch reported about itself in session 1, from `benchmark.R`'s `{r} section_link("benchmark.R", "benchmark-provenance")` section:

```{r}
#| label: benchmark-provenance
#| eval: false
#| code-summary: "benchmark.R: benchmark-provenance"
```

<details><summary>`greta_sitrep()`</summary>

```{r}
#| label: sitrep
writeLines(results[[1]]$greta_sitrep_report)
```

</details>

<details><summary>`sessioninfo::session_info()`</summary>

```{r}
#| label: session-info
results[[1]]$session_info
```

</details>

## References
