---
title: "greta#853 (windowed-adaptation-i853): warmup schemes compared"
description: "Five warmup schemes run inside TensorFlow on five models: greta's own, greta's with greta#841's count fixed, Stan's windowed scheme from greta#853, and TFP's windowed HMC and NUTS. Stan's scheme finds the scales of the normals that greta's own misses, and gives more effective samples per second than greta's on every model but eight_schools."
date: 2026-10-07
bibliography: ../../references.bib
---

```{r}
#| label: setup
#| code-summary: "Setup"
# the packages' start-up messages, such as where here() finds the project, are
# not part of the results
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(bench)
  library(here)
})

post_path <- "posts/2026-10-07-warmup-adaptation-i853"

# each `# ---- label ----` section of the scripts, shown below by its label
knitr::read_chunk(here(post_path, "benchmark.R"))
knitr::read_chunk(here(post_path, "run.R"))

# links to the scripts on GitHub, at the commit being rendered when GitHub
# Actions renders the site
github_ref <- Sys.getenv("GITHUB_SHA", "main")
github_file_url <- function(file) {
  paste0(
    "https://github.com/greta-dev/greta.benchmarks/blob/",
    github_ref,
    "/",
    post_path,
    "/",
    file
  )
}

# a link to the lines of a script's `# ---- label ----` section. Quarto escapes
# the markdown in an inline `{r}` result unless it is wrapped in I()
section_link <- function(file, label) {
  lines <- readLines(here(post_path, file))
  headings <- grep("^# ---- .+ ----$", lines)
  start <- headings[lines[headings] == paste("# ----", label, "----")]
  end <- c(headings, length(lines) + 1)[match(start, headings) + 1] - 1
  while (end > start && !nzchar(trimws(lines[end]))) {
    end <- end - 1
  }
  I(paste0(
    "[`", label, "`](", github_file_url(file), "#L", start, "-L", end, ")"
  ))
}

# "a, b and c"
paste_and <- function(x) {
  if (length(x) < 2) {
    return(x)
  }
  paste(paste(head(x, -1), collapse = ", "), "and", tail(x, 1))
}

# links to several sections of a script, as "a, b and c"
section_links <- function(file, labels) {
  links <- vapply(labels, \(label) section_link(file, label), character(1))
  I(paste_and(links))
}

# three significant figures each, where format() would pad a vector to the
# decimal places of its smallest value. formatC() pads a number with fewer
# than three significant figures with spaces, which trimws() removes
format_number <- function(x) {
  trimws(formatC(signif(x, 3), format = "fg", digits = 3, big.mark = ","))
}

# "3.25 to 4.81": the smallest and largest of x
format_range <- function(x) {
  paste(format_number(min(x)), "to", format_number(max(x)))
}

# a searchable table, its numbers rounded for reading
dt_table <- function(data, digits = 3) {
  table <- DT::datatable(
    data,
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 10, autoWidth = TRUE)
  )
  decimal_columns <- names(data)[vapply(data, is.double, logical(1))]
  if (length(decimal_columns) == 0) {
    return(table)
  }
  DT::formatRound(table, decimal_columns, digits)
}

theme_set(theme_bw(base_size = 13))

sessions <- list.files(
  here(post_path, "results"),
  pattern = "^session-[0-9]+[.]rds$",
  full.names = TRUE
) |>
  lapply(readRDS)
n_sessions <- length(sessions)
results <- lapply(sessions, \(session_file) session_file$results$branch)
bench_settings <- results[[1]]$bench_settings
model_levels <- names(results[[1]]$schemes)
method_levels <- c(
  bench_settings$schemes,
  paste0("tfp_", bench_settings$tfp_kinds)
)
method_colours <- setNames(
  c("#999999", "#56B4E9", "#009E73", "#E69F00", "#CC79A7"),
  method_levels
)
method_scale <- scale_colour_manual(values = method_colours, guide = "none")
method_fill <- scale_fill_manual(values = method_colours, guide = "none")
jitter <- position_jitter(width = 0, height = 0.2, seed = 2026)

# each model's runs are its `bench-` section of benchmark.R
bench_links <- section_links(
  "benchmark.R",
  paste0("bench-", gsub("_", "-", model_levels))
)

# A raincloud for each method: the density of its sessions above, their box
# plot in the middle, and every session as a dot below. The summaries are
# drawn first and faint, so the dots sit on top.
raincloud <- list(
  ggdist::stat_halfeye(
    aes(fill = method),
    width = 0.4,
    justification = -0.3,
    .width = 0,
    point_colour = NA,
    alpha = 0.3
  ),
  geom_boxplot(
    aes(fill = method),
    width = 0.25,
    outlier.shape = NA,
    alpha = 0.3
  ),
  # a fixed dot size, so a session looks the same size in every panel
  ggdist::stat_dots(
    aes(fill = method),
    side = "bottom",
    layout = "swarm",
    justification = 1.3,
    scale = 0.75,
    binwidth = unit(2.6, "mm"),
    overflow = "compress",
    colour = NA,
    alpha = 0.6
  ),
  method_scale,
  method_fill,
  theme_bw(base_size = 13)
)

# one row per timed call, from every session, model and method
every_call <- lapply(seq_along(results), function(i) {
  lapply(model_levels, function(model_name) {
    schemes <- results[[i]]$schemes[[model_name]]$marks
    tfp <- results[[i]]$tfp[[model_name]]$marks
    marks <- list(
      tibble(method = schemes$scheme, time = schemes$time),
      tibble(method = paste0("tfp_", tfp$kind), time = tfp$time)
    )
    bind_rows(marks) |>
      mutate(call = lapply(time, seq_along)) |>
      unnest(c(time, call)) |>
      mutate(
        session = sessions[[i]]$session,
        model = model_name,
        seconds = as.numeric(time),
        .keep = "unused"
      )
  }) |>
    bind_rows()
}) |>
  bind_rows() |>
  mutate(
    method = factor(method, levels = method_levels),
    model = factor(model, levels = model_levels),
    calls = factor(
      ifelse(call == 1, "first call", "second call"),
      levels = c("first call", "second call")
    )
  )

# posterior caps a bulk ESS at S × log10(S) for S draws, and warns each time it
# does. Count those warnings here, keep the cap beside each estimate, and say
# what they mean under the effective samples plot.
ess_cap_warnings <- 0
summarise_counting_caps <- function(draws) {
  withCallingHandlers(
    posterior::summarise_draws(draws, "mean", "sd", "rhat", "ess_bulk"),
    warning = function(w) {
      if (grepl("The ESS has been capped", conditionMessage(w), fixed = TRUE)) {
        ess_cap_warnings <<- ess_cap_warnings + 1
        invokeRestart("muffleWarning")
      }
    }
  )
}

# a posterior summary of every method's second-call draws, in every session
posterior_summaries <- lapply(seq_along(results), function(i) {
  lapply(model_levels, function(model_name) {
    draws <- c(
      results[[i]]$schemes[[model_name]]$draws,
      setNames(
        results[[i]]$tfp[[model_name]]$draws,
        paste0("tfp_", names(results[[i]]$tfp[[model_name]]$draws))
      )
    )
    lapply(names(draws), function(method) {
      n_draws <- posterior::ndraws(draws[[method]])
      tibble(
        session = sessions[[i]]$session,
        model = model_name,
        method = method,
        summarise_counting_caps(draws[[method]]),
        n_draws = n_draws,
        ess_cap = n_draws * log10(n_draws)
      )
    }) |>
      bind_rows()
  }) |>
    bind_rows()
}) |>
  bind_rows() |>
  mutate(
    method = factor(method, levels = method_levels),
    model = factor(model, levels = model_levels),
    ess_capped = ess_bulk >= ess_cap * (1 - 1e-9)
  )

second_calls <- every_call |>
  filter(call == 2) |>
  select(session, model, method, seconds)

# each run's smallest bulk ESS and largest R-hat, and that ESS per second of
# the run's second call, warmup included
efficiency <- posterior_summaries |>
  group_by(session, model, method) |>
  summarise(
    smallest_ess_bulk = min(ess_bulk),
    smallest_ess_capped = ess_capped[which.min(ess_bulk)],
    largest_rhat = max(rhat),
    .groups = "drop"
  ) |>
  left_join(second_calls, by = c("session", "model", "method")) |>
  mutate(smallest_ess_per_second = smallest_ess_bulk / seconds)

# each method's median over sessions, as a multiple of greta's own scheme's
# median on the same model
relative_to_greta <- function(data, value) {
  data |>
    group_by(model, method) |>
    summarise(median = median({{ value }}), .groups = "drop") |>
    group_by(model) |>
    mutate(times_greta = median / median[method == "greta"]) |>
    ungroup()
}

# one method's multiples of greta's, named by model
times_greta <- function(ratios, scheme) {
  scheme_ratios <- filter(ratios, method == scheme)
  setNames(scheme_ratios$times_greta, scheme_ratios$model)
}

ess_per_second_ratios <- relative_to_greta(efficiency, smallest_ess_per_second)
windowed_ess_per_second <- times_greta(ess_per_second_ratios, "windowed")
```

[greta#853](https://github.com/greta-dev/greta/issues/853) is an issue, not a PR: it proposes that `hmc()` adapt `diag_sd` in windows during warmup, as Stan does, and the windowed-adaptation-i853 branch adds that scheme to greta's sampler as an internal choice. This post runs it beside greta's own scheme, greta's with [greta#841](https://github.com/greta-dev/greta/issues/841)'s count fixed, and TFP's windowed HMC and NUTS, all inside TensorFlow on one greta commit and five models; it is not part of PR [greta#855](https://github.com/greta-dev/greta/pull/855). The windowed scheme gives `{r} format_number(windowed_ess_per_second[["normals"]])` times the smallest bulk ESS per second of greta's own scheme on twenty normals of very different scales, `{r} format_range(windowed_ess_per_second[c("linear", "multiple_linear", "hierarchical_linear")])` times on linear, multiple_linear and hierarchical_linear, and `{r} format_number(windowed_ess_per_second[["eight_schools"]])` times on eight_schools.

## The five schemes

- **greta**: greta's own tuning. It adapts the step size, `epsilon`, by dual averaging over the first 10% and last 60% of warmup [@hoffman2014], and estimates `diag_sd` between 10% and 40% of warmup from every warmup draw so far, shrunk by a count of the *rejected* proposals, then fixes it.
- **greta_841**: the same, shrunk by a count of the *accepted* proposals, as [greta#841](https://github.com/greta-dev/greta/issues/841) says it was meant to be.
- **windowed**: Stan's windowed adaptation, from greta#853. After a buffer that adapts `epsilon` only, it estimates `diag_sd` afresh in windows ending at 25%, 45% and 90% of warmup, each from its own draws, and restarts dual averaging after each.
- **tfp_hmc** and **tfp_nuts**: TFP's own windowed adaptation of HMC and NUTS, as its `windowed_adaptive_hmc()` and `windowed_adaptive_nuts()` use it, on the same log density. TFP's HMC takes `{r} bench_settings$tfp_leapfrog_steps` leapfrog steps every iteration, where greta's draws a number between `Lmin` and `Lmax`.

## Does each scheme find the normals' scales?

For greta's three schemes, each normal's tuned `diag_sd` over its true standard deviation, so 1 is a perfect estimate. TFP keeps its estimate inside its kernel results, which the benchmark does not return.

```{r}
#| label: tuned-scales
true_sd <- results[[1]]$normals_sd
tuned_scales <- lapply(seq_along(results), function(i) {
  tuned <- results[[i]]$schemes$normals$tuned
  lapply(names(tuned), function(scheme) {
    tibble(
      session = sessions[[i]]$session,
      method = scheme,
      true_sd = true_sd,
      ratio = tuned[[scheme]]$diag_sd / true_sd
    )
  }) |>
    bind_rows()
}) |>
  bind_rows() |>
  mutate(method = factor(method, levels = method_levels))

# each scheme's tuned diag_sd over the true sd, named by scheme
tuned_ratios <- split(tuned_scales$ratio, tuned_scales$method, drop = TRUE)
```

```{r}
#| label: plot-tuned-scales
#| fig-cap: "Each dot is one normal in one session, on log scales. The grey line is a perfect estimate."
#| fig-height: 4
ggplot(tuned_scales, aes(x = true_sd, y = ratio, colour = method)) +
  geom_hline(yintercept = 1, colour = "grey60") +
  geom_point(alpha = 0.7) +
  scale_x_log10() +
  scale_y_log10() +
  method_scale +
  facet_wrap(~method) +
  labs(x = "True standard deviation", y = "Tuned diag_sd / true sd")
```

Tuned in `benchmark.R`'s `{r} section_link("benchmark.R", "bench-normals")` section, by `run_schemes()` from its `{r} section_link("benchmark.R", "benchmark-schemes")` section, which keeps each scheme's `diag_sd`.

The windowed scheme's `diag_sd` is `{r} format_range(tuned_ratios$windowed)` times the true standard deviation. greta's own is `{r} format_range(tuned_ratios$greta)` times, furthest off on the widest normals, and greta#841's count leaves it much the same, at `{r} format_range(tuned_ratios$greta_841)` times.

## How long does each scheme take?

Seconds for the second call on each model, which reuses the first call's trace. The effective samples per second below divide by these.

```{r}
#| label: speed
speed_ratios <- relative_to_greta(second_calls, seconds)
```

```{r}
#| label: plot-speed
#| fig-cap: "Each dot is one session's second call, on a log scale. Above the dots, the curve is the density of the sessions and the box marks their median and quartiles."
#| fig-height: 12
ggplot(second_calls, aes(x = seconds, y = method, colour = method)) +
  raincloud +
  scale_x_log10() +
  facet_wrap(~model, ncol = 1, scales = "free_x") +
  labs(x = "Seconds", y = NULL)
```

Timed by `benchmark.R`'s `{r} bench_links` sections, which call `run_schemes()` from its `{r} section_link("benchmark.R", "benchmark-schemes")` section and `run_tfp()` from its `{r} section_link("benchmark.R", "benchmark-tfp")` section.

On the five models, the windowed scheme takes `{r} format_range(times_greta(speed_ratios, "windowed"))` times as long as greta's own, and greta_841 `{r} format_range(times_greta(speed_ratios, "greta_841"))` times, so the three cost about the same. TFP's HMC takes `{r} format_range(times_greta(speed_ratios, "tfp_hmc"))` times as long as greta's, and TFP's NUTS `{r} format_range(times_greta(speed_ratios, "tfp_nuts"))` times.

<details><summary>Table: median seconds of the first and second calls</summary>

```{r}
#| label: table-speed
every_call |>
  group_by(model, calls, method) |>
  summarise(seconds = median(seconds), .groups = "drop") |>
  pivot_wider(names_from = method, values_from = seconds) |>
  dt_table(digits = 2)
```

</details>

## How many effective samples per second does each give?

The bulk ESS of each run's worst-mixing parameter, as @hoffman2014 report it, per second of the second call, warmup included [@girolami2011].

```{r}
#| label: efficiency
capped_estimates <- filter(posterior_summaries, ess_capped)
capped_runs <- distinct(capped_estimates, session, model, method)

# every run as a small dot over a large faint mean, for the smallest ESS here
# and the largest R-hat in the checks
efficiency_means <- efficiency |>
  group_by(model, method) |>
  summarise(
    across(c(smallest_ess_bulk, largest_rhat), mean),
    .groups = "drop"
  )

gg_sessions <- function(x, x_label) {
  ggplot(efficiency, aes(x = {{ x }}, y = method, colour = method)) +
    geom_point(data = efficiency_means, size = 5, alpha = 0.25) +
    geom_point(position = jitter, size = 1.2) +
    method_scale +
    facet_wrap(~model, ncol = 1, scales = "free_x") +
    labs(x = x_label, y = NULL)
}
```

```{r}
#| label: plot-ess-per-second
#| fig-cap: "Each dot is one session's run, on a log scale. Above the dots, the curve is the density of the sessions and the box marks their median and quartiles."
#| fig-height: 12
ggplot(
  efficiency,
  aes(x = smallest_ess_per_second, y = method, colour = method)
) +
  raincloud +
  scale_x_log10() +
  facet_wrap(~model, ncol = 1, scales = "free_x") +
  labs(x = "Smallest bulk ESS per second", y = NULL)
```

Draws and times from `benchmark.R`'s `{r} bench_links` sections, which call `run_schemes()` from its `{r} section_link("benchmark.R", "benchmark-schemes")` section and `run_tfp()` from its `{r} section_link("benchmark.R", "benchmark-tfp")` section.

The windowed scheme gives `{r} format_number(windowed_ess_per_second[["normals"]])` times the smallest bulk ESS per second of greta's own scheme on the normals, `{r} format_range(windowed_ess_per_second[c("linear", "multiple_linear", "hierarchical_linear")])` times on linear, multiple_linear and hierarchical_linear, and `{r} format_number(windowed_ess_per_second[["eight_schools"]])` times on eight_schools. On the five models, greta_841 gives `{r} format_range(times_greta(ess_per_second_ratios, "greta_841"))` times greta's, TFP's HMC `{r} format_range(times_greta(ess_per_second_ratios, "tfp_hmc"))` times, and TFP's NUTS `{r} format_range(times_greta(ess_per_second_ratios, "tfp_nuts"))` times.

Anticorrelated draws, which `hmc()` can give on independent normals, have an ESS above their number, and posterior caps it at S × log10(S), or `{r} format_number(unique(posterior_summaries$ess_cap))` for these `{r} format_number(unique(posterior_summaries$n_draws))` draws, with a warning each time. It capped `{r} nrow(capped_estimates)` estimates in `{r} nrow(capped_runs)` runs, all on `{r} paste_and(unique(as.character(capped_runs$model)))` with `{r} paste_and(unique(as.character(capped_runs$method)))` (its `{r} ess_cap_warnings` warnings are counted and silenced in Setup), and `{r} sum(efficiency$smallest_ess_capped)` of the `{r} nrow(efficiency)` runs had a capped estimate as their smallest ESS, the number compared here.

<details><summary>Figure: smallest bulk ESS, not per second</summary>

```{r}
#| label: plot-ess
#| fig-cap: "Each small dot is one session's run, on a log scale. The large faint dot behind them is their mean."
#| fig-height: 12
gg_sessions(smallest_ess_bulk, "Smallest bulk ESS") +
  scale_x_log10()
```

</details>

<details><summary>Table: median smallest bulk ESS per second, and as a multiple of greta's</summary>

```{r}
#| label: table-ess-per-second
ess_per_second_ratios |>
  rename(smallest_ess_per_second = median) |>
  dt_table(digits = 2)
```

</details>

<details><summary>Table: every run</summary>

```{r}
#| label: table-efficiency
efficiency |>
  select(
    session,
    model,
    method,
    smallest_ess_bulk,
    seconds,
    smallest_ess_per_second,
    largest_rhat
  ) |>
  dt_table(digits = 2)
```

</details>

## Checks

### Do the chains converge?

The largest R-hat of each run, against the 1.01 that @vehtari2021 recommend.

```{r}
#| label: rhat
rhat_threshold <- 1.01
runs_over_threshold <- efficiency |>
  group_by(method) |>
  summarise(over = sum(largest_rhat > rhat_threshold), runs = n())
runs_over_text <- I(paste_and(paste0(
  "`",
  runs_over_threshold$method,
  "` ",
  runs_over_threshold$over
)))
```

```{r}
#| label: plot-rhat
#| fig-cap: "Each small dot is one session's run. The large faint dot behind them is their mean, and the grey line is 1.01."
#| fig-height: 12
gg_sessions(largest_rhat, "Largest R-hat") +
  geom_vline(xintercept = rhat_threshold, colour = "grey60")
```

From the same draws, made in `benchmark.R`'s `{r} bench_links` sections.

Of each scheme's `{r} unique(runs_over_threshold$runs)` runs, this many have a largest R-hat above `{r} rhat_threshold`: `{r} runs_over_text`. At `{r} bench_settings$warmup` warmup iterations and `{r} bench_settings$n_samples` draws per chain, `{r} sum(runs_over_threshold$over)` of the `{r} nrow(efficiency)` runs miss that threshold, so most of the effective samples above come from chains that have not converged by that rule.

### Do the draws have the right spread?

On the normals, where the truth is known: each normal's posterior standard deviation over its true one.

```{r}
#| label: posterior-scales
posterior_scales <- posterior_summaries |>
  filter(model == "normals") |>
  mutate(
    true_sd = true_sd[as.integer(sub("x\\[([0-9]+),1\\]", "\\1", variable))],
    ratio = sd / true_sd
  )

spread_ranges <- posterior_scales |>
  group_by(method) |>
  summarise(range = format_range(ratio))
spread_text <- I(paste_and(paste0(
  "`",
  spread_ranges$method,
  "` ",
  spread_ranges$range
)))
```

```{r}
#| label: plot-posterior-scales
#| fig-cap: "Each dot is one normal in one session, on log scales. The grey line is a perfect estimate."
#| fig-height: 6
ggplot(posterior_scales, aes(x = true_sd, y = ratio, colour = method)) +
  geom_hline(yintercept = 1, colour = "grey60") +
  geom_point(alpha = 0.7) +
  scale_x_log10() +
  scale_y_log10() +
  method_scale +
  facet_wrap(~method) +
  labs(x = "True standard deviation", y = "Posterior sd / true sd")
```

From the draws made in `benchmark.R`'s `{r} section_link("benchmark.R", "bench-normals")` section.

Each normal's posterior standard deviation is this many times its true one: `{r} spread_text`.

## How it was measured

```{r}
#| label: provenance
provenance <- bind_rows(lapply(results, \(result) result$provenance))
session_numbers <- vapply(sessions, \(session) session$session, numeric(1))

# a session that pak installed from GitHub records the commit; one installed
# from a local worktree records its path, and run.R's run-versions section
# names the commit that worktree was clean at
installed_from <- vapply(
  sessions,
  \(session) session$versions[["branch"]],
  character(1)
)
installed_locally <- startsWith(installed_from, "local::")
measured_commit <- unique(sub(
  "^greta-dev/greta@",
  "",
  installed_from[!installed_locally]
))
commit_link <- I(paste0(
  "[",
  substr(measured_commit, 1, 7),
  "](https://github.com/greta-dev/greta/commit/",
  measured_commit,
  ")"
))
```

Every result here is from one greta commit, `{r} commit_link`, on tf-warmup-i547, the branch of PR greta#855, which then carried the greta_841 and windowed schemes as an internal choice for comparison. Those two have since moved to windowed-adaptation-i853, a local branch built on the PR's branch, and are not part of the PR; TFP's two schemes are driven from `benchmark.R` itself. `run.R`'s `{r} section_link("run.R", "run-versions")` section pins the commit: sessions `{r} paste_and(session_numbers[installed_locally])` installed it from a local worktree clean at the commit, and sessions `{r} paste_and(session_numbers[!installed_locally])` from GitHub.

`run.R`'s `{r} section_link("run.R", "run-sessions")` section ran `benchmark.R` in `{r} n_sessions` fresh R sessions, one after another, on `{r} unique(provenance$machine)` with `{r} unique(provenance$cores)` cores, TensorFlow `{r} unique(provenance$tensorflow_version)` and TensorFlow Probability `{r} unique(provenance$tfp_version)`. Each session ran every scheme on every model, with a seed of its own from `benchmark.R`'s `{r} section_link("benchmark.R", "benchmark-setup")` section: `{r} bench_settings$chains` chains of `{r} bench_settings$warmup` warmup iterations and `{r} bench_settings$n_samples` draws, called `{r} bench_settings$mcmc_calls` times, as its `{r} section_link("benchmark.R", "benchmark-settings")` section sets. greta's three schemes run through `mcmc()`, with the scheme set on greta's internal sampler class, in its `{r} section_link("benchmark.R", "benchmark-schemes")` section; TFP's two run through TFP's windowed sampling on the same log density, in its `{r} section_link("benchmark.R", "benchmark-tfp")` section.

```{r}
#| label: benchmark-schemes
#| eval: false
#| code-summary: "benchmark.R: benchmark-schemes"
```

```{r}
#| label: benchmark-tfp
#| eval: false
#| code-summary: "benchmark.R: benchmark-tfp"
```

Both scripts, in full:

```{r}
#| label: run-script
#| file: run.R
#| eval: false
#| code-summary: "run.R"
```

```{r}
#| label: benchmark-script
#| file: benchmark.R
#| eval: false
#| code-summary: "benchmark.R"
```

### The models

Twenty independent normals with standard deviations from `{r} min(true_sd)` to `{r} max(true_sd)`, greta#853's test: their scales span four orders of magnitude, so a `diag_sd` that is off on one scale shows. The other four are greta's examples, written out in maths in [the PR's main post](../2026-10-07-tf-warmup-i547/#the-models).

::: panel-tabset

#### normals

```{r}
#| label: model-normals
#| eval: false
#| code-summary: "benchmark.R: model-normals"
```

#### linear

```{r}
#| label: model-linear
#| eval: false
#| code-summary: "benchmark.R: model-linear"
```

#### multiple_linear

```{r}
#| label: model-multiple-linear
#| eval: false
#| code-summary: "benchmark.R: model-multiple-linear"
```

#### hierarchical_linear

```{r}
#| label: model-hierarchical-linear
#| eval: false
#| code-summary: "benchmark.R: model-hierarchical-linear"
```

#### eight_schools

```{r}
#| label: model-eight-schools
#| eval: false
#| code-summary: "benchmark.R: model-eight-schools"
```

:::

## Session details

What the branch reported about itself in session `{r} sessions[[1]]$session`, from `benchmark.R`'s `{r} section_link("benchmark.R", "benchmark-provenance")` section.

<details><summary>`greta_sitrep()`</summary>

```{r}
#| label: sitrep
writeLines(results[[1]]$greta_sitrep_report)
```

</details>

<details><summary>`sessioninfo::session_info()`</summary>

```{r}
#| label: session-info
# a saved session_info() prints as a plain list unless sessioninfo's print
# method is registered, which loading its namespace does
invisible(loadNamespace("sessioninfo"))
results[[1]]$session_info
```

</details>

## References
