---
title: "greta#855 (tf-warmup-i547): warmup and sampling in TensorFlow"
description: "How long mcmc() and opt() with adam() take on greta 0.6.0 from CRAN, main, greta#850 and greta#855, on four of greta's example models. greta#855 is the fastest of the four at both, and its mcmc() samples the same posteriors."
date: 2026-10-07
bibliography: ../../references.bib
---

```{r}
#| label: setup
#| message: false
#| code-summary: "Setup"
library(dplyr)
library(tidyr)
library(ggplot2)
library(bench)
library(here)

post_path <- "posts/2026-10-07-tf-warmup-i547"

# each `# ---- label ----` section of the scripts, shown below by its label
knitr::read_chunk(here(post_path, "benchmark.R"))
knitr::read_chunk(here(post_path, "run.R"))

# links to the scripts on GitHub, at the commit being rendered when GitHub
# Actions renders the site
github_ref <- Sys.getenv("GITHUB_SHA", "main")
github_file_url <- function(file) {
  paste0(
    "https://github.com/greta-dev/greta.benchmarks/blob/",
    github_ref,
    "/",
    post_path,
    "/",
    file
  )
}
# Quarto escapes the markdown in an inline `{r}` result unless it is wrapped
# in I(), so the link helpers return I()
file_link <- function(file) {
  I(paste0("[`", file, "`](", github_file_url(file), ")"))
}

# a link to the lines of a script's `# ---- label ----` section
section_link <- function(file, label) {
  lines <- readLines(here(post_path, file))
  headings <- grep("^# ---- .+ ----$", lines)
  start <- headings[lines[headings] == paste("# ----", label, "----")]
  end <- c(headings, length(lines) + 1)[match(start, headings) + 1] - 1
  while (end > start && !nzchar(trimws(lines[end]))) {
    end <- end - 1
  }
  I(paste0(
    "[`", label, "`](", github_file_url(file), "#L", start, "-L", end, ")"
  ))
}

# links to several sections of a script: "`a`, `b` and `c`"
section_links <- function(file, labels) {
  links <- vapply(labels, \(label) section_link(file, label), character(1))
  I(paste_and(unname(links)))
}

# white panels, so the grey band in Agreement shows
theme_set(theme_bw())

sessions <- list.files(
  here(post_path, "results"),
  pattern = "^session-[0-9]+[.]rds$",
  full.names = TRUE
) |>
  lapply(readRDS)

# run.R names the versions as they were when this ran. The PR, greta#855, has
# since taken the tf-warmup-i547 branch, so the page relabels them: tf-warmup
# is an earlier commit of the PR, and branch is the PR at 5bdab31e
version_labels <- c(
  CRAN = "CRAN",
  main = "main",
  `#850` = "#850",
  `tf-warmup` = "#855 early",
  branch = "#855"
)
pr_label <- version_labels[["branch"]]

versions <- sessions[[1]]$versions
# pak installed branch from a worktree, which the results record as a local
# path; run.R names the same commit on GitHub
on_github <- c(
  branch = "greta-dev/greta@5bdab31e9ec9295c845a4fac9d1201e1aabb9c93"
)
installed_locally <- startsWith(versions, "local::")
versions[installed_locally] <- on_github[names(versions)[installed_locally]]
names(versions) <- version_labels[names(versions)]
version_levels <- names(versions)
version_colours <- setNames(
  c("#E69F00", "#56B4E9", "#009E73", "#D55E00", "#CC79A7")[seq_along(
    version_levels
  )],
  version_levels
)
bench_settings <- sessions[[1]]$results[[1]]$bench_settings
n_sessions <- length(sessions)

# one row per session and version, holding that version's results
runs <- lapply(sessions, function(session_file) {
  result_names <- names(session_file$results)
  tibble(
    session = session_file$session,
    version = factor(
      unname(version_labels[result_names]),
      levels = version_levels
    ),
    position = match(result_names, session_file$order),
    results = unname(session_file$results)
  )
}) |>
  bind_rows()

model_levels <- names(runs$results[[1]]$mcmc_same)
# benchmark.R's section labels spell each model with hyphens
model_sections <- gsub("_", "-", model_levels)
setting_levels <- c("the same arguments", "each version's defaults")

# a part of the results that is a table or a value per run, stacked, with the
# run it came from
stack_tables <- function(part) {
  pieces <- lapply(seq_len(nrow(runs)), function(i) {
    values <- runs$results[[i]][[part]]
    if (!is.data.frame(values)) {
      values <- tibble(value = values)
    }
    tibble(session = runs$session[i], version = runs$version[i], values)
  })
  bind_rows(pieces)
}

# a part of the results that is a bench_mark per model, or one per run,
# stacked into one bench_mark, with the run and model each came from
stack_marks <- function(part) {
  pieces <- list()
  for (i in seq_len(nrow(runs))) {
    marks <- runs$results[[i]][[part]]
    if (inherits(marks, "bench_mark")) {
      marks <- list(marks)
    }
    for (j in seq_along(marks)) {
      mark <- marks[[j]]
      mark$session <- runs$session[i]
      mark$version <- runs$version[i]
      mark$model <- if (is.null(names(marks))) NA_character_ else names(marks)[j]
      pieces[[length(pieces) + 1]] <- mark
    }
  }
  vctrs::vec_rbind(!!!pieces)
}

provenance <- stack_tables("provenance")
sampler_iterations_per_draw <- stack_tables("sampler_iterations_per_draw")
opt_marks <- stack_marks("opt_adam")
opt_marks$setting <- NA_character_

# the two sets of timed mcmc() calls, stacked, with which arguments each used
same_marks <- stack_marks("mcmc_same")
same_marks$setting <- setting_levels[1]
defaults_marks <- stack_marks("mcmc_defaults")
defaults_marks$setting <- setting_levels[2]
mcmc_marks <- vctrs::vec_rbind(same_marks, defaults_marks)

# one row per timed call: bench keeps every call's time in the `time` column
every_call <- function(marks) {
  marks |>
    as_tibble() |>
    select(session, version, model, setting, time) |>
    mutate(call = lapply(time, seq_along)) |>
    unnest(c(time, call)) |>
    mutate(
      seconds = as.numeric(time),
      model = factor(model, levels = model_levels),
      setting = factor(setting, levels = setting_levels),
      .keep = "unused"
    )
}

# bench's own summary() of each mark, with the mean and longest of its calls
# added, all in seconds
bench_summary <- function(marks) {
  summaries <- summary(marks, filter_gc = FALSE)
  tibble(
    session = summaries$session,
    version = summaries$version,
    model = summaries$model,
    setting = summaries$setting,
    min = as.numeric(summaries$min),
    median = as.numeric(summaries$median),
    mean = vapply(marks$time, \(time) mean(as.numeric(time)), numeric(1)),
    max = vapply(marks$time, \(time) max(as.numeric(time)), numeric(1)),
    `itr/sec` = summaries$`itr/sec`,
    n_itr = summaries$n_itr,
    total_time = as.numeric(summaries$total_time)
  )
}

# The first call on a model traces its TensorFlow functions; later calls on the
# same model reuse them.
label_calls <- function(call) {
  last <- max(call)
  later <- if (last == 2) {
    "second call"
  } else if (last == 3) {
    "calls 2 and 3"
  } else {
    paste0("calls 2 to ", last)
  }
  factor(
    ifelse(call == 1, "first call", later),
    levels = c("first call", later)
  )
}

# three significant figures each, where format() would pad a vector to the
# decimal places of its smallest value; formatC() pads to a common width, so
# trim that
format_number <- function(x) {
  trimws(formatC(signif(x, 3), format = "fg", digits = 3, big.mark = ","))
}

# "CRAN 10.6, main 11.2, #843 0.643"
versions_text <- function(version, value) {
  paste(paste(version, format_number(value)), collapse = ", ")
}

# "CRAN, main and #843"
paste_and <- function(x) {
  if (length(x) <= 1) {
    return(x)
  }
  paste(paste(x[-length(x)], collapse = ", "), "and", x[length(x)])
}

# "2.49 to 3.21": the range over models of a ratio with #855 as its unit, from
# a table of each version's `lowest` and `highest`, filtered by `...`
ratio_range <- function(ratios, version_label, ...) {
  row <- filter(ratios, version == version_label, ...)
  paste(format_number(row$lowest), "to", format_number(row$highest))
}

version_scale <- scale_colour_manual(values = version_colours, guide = "none")
version_fill <- scale_fill_manual(values = version_colours, guide = "none")
jitter <- position_jitter(width = 0, height = 0.2, seed = 2026)

# A raincloud for each version: the density of its calls above, their box plot
# in the middle, and every call as a dot below. The summaries are drawn first
# and faint, so the dots sit on top.
raincloud <- list(
  ggdist::stat_halfeye(
    aes(fill = version),
    width = 0.4,
    justification = -0.3,
    .width = 0,
    point_colour = NA,
    alpha = 0.3
  ),
  geom_boxplot(
    aes(fill = version),
    width = 0.25,
    outlier.shape = NA,
    alpha = 0.3
  ),
  # a fixed dot size, so a call looks the same size in every panel
  ggdist::stat_dots(
    aes(fill = version),
    side = "bottom",
    layout = "swarm",
    justification = 1.3,
    scale = 0.75,
    binwidth = unit(2.6, "mm"),
    overflow = "compress",
    colour = NA,
    alpha = 0.6
  ),
  version_scale,
  version_fill,
  theme_bw(base_size = 13)
)
raincloud_caption <- "Each dot is one call, from one session. Above the dots, the curve is the density of the calls and the box marks their median and quartiles."
session_caption <- "Each small dot is one session's second call. The large faint dot behind them is their mean."

# a searchable table, its numbers rounded for reading
dt_table <- function(data, digits = 3) {
  table <- DT::datatable(
    data,
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 10, autoWidth = TRUE)
  )
  decimal_columns <- names(data)[vapply(data, is.double, logical(1))]
  if (length(decimal_columns) == 0) {
    return(table)
  }
  DT::formatRound(table, decimal_columns, digits)
}

# The PR at a glance, which the opening quotes: this post's mcmc() calls and
# the call overhead post's, on CRAN and the PR
glance_rows <- c(
  "`mcmc()`, each version's defaults, first call",
  "`mcmc()`, each version's defaults, calls after the first",
  "`mcmc()`, 1000 warmup and 1000 draws, calls after the first",
  "`mcmc()` with the progress bar, 1000 and 1000, calls after the first",
  "`mcmc()` with `one_by_one = TRUE`, 200 and 200, calls after the first",
  "`opt()` with `adam()`, 2000 iterations, first call",
  "`opt()` with `adam()`, 2000 iterations, calls after the first"
)

this_post_calls <- every_call(mcmc_marks) |>
  filter(version %in% c("CRAN", pr_label)) |>
  mutate(
    row = case_when(
      setting == setting_levels[2] & call == 1 ~ glance_rows[1],
      setting == setting_levels[2] & call > 1 ~ glance_rows[2],
      setting == setting_levels[1] & call > 1 ~ glance_rows[3]
    ),
    post = "[this post](#speed)"
  ) |>
  filter(!is.na(row)) |>
  select(row, post, version, model, seconds)

# the call overhead post names the PR `branch`. Each bench_mark keeps every
# call's time, and opt()'s has one row per iteration limit, of which this
# table takes 2000
overhead_path <- "posts/2026-10-07-call-overhead-i547"
overhead_sessions <- list.files(
  here(overhead_path, "results"),
  pattern = "^session-[0-9]+[.]rds$",
  full.names = TRUE
) |>
  lapply(readRDS)
calls_of <- function(time) {
  tibble(call = seq_along(time), seconds = as.numeric(time))
}
overhead_calls <- lapply(overhead_sessions, function(session_file) {
  lapply(c("CRAN", "branch"), function(version) {
    result <- session_file$results[[version]]
    lapply(names(result$mcmc), function(model) {
      mcmc_runs <- result$mcmc[[model]]
      opt_runs <- result$opt[[model]]
      bind_rows(
        calls_of(mcmc_runs$progress$time[[1]]) |>
          mutate(row = if_else(call > 1, glance_rows[4], NA)),
        calls_of(mcmc_runs$one_by_one$time[[1]]) |>
          mutate(row = if_else(call > 1, glance_rows[5], NA)),
        calls_of(opt_runs$time[[which(opt_runs$max_iterations == 2000)]]) |>
          mutate(row = if_else(call == 1, glance_rows[6], glance_rows[7]))
      ) |>
        mutate(version = version, model = model)
    }) |>
      bind_rows()
  }) |>
    bind_rows()
}) |>
  bind_rows() |>
  filter(!is.na(row)) |>
  mutate(
    version = if_else(version == "branch", pr_label, version),
    post = "[Call overhead](../2026-10-07-call-overhead-i547/#summary)"
  ) |>
  select(row, post, version, model, seconds)

glance_calls <- bind_rows(this_post_calls, overhead_calls) |>
  mutate(
    version = if_else(version == "CRAN", "CRAN", "PR"),
    model = as.character(model)
  )

glance_seconds <- glance_calls |>
  group_by(row, post, version) |>
  summarise(seconds = median(seconds), .groups = "drop") |>
  pivot_wider(names_from = version, values_from = seconds)

glance_ratios <- glance_calls |>
  group_by(row, version, model) |>
  summarise(seconds = median(seconds), .groups = "drop") |>
  pivot_wider(names_from = version, values_from = seconds) |>
  group_by(row) |>
  summarise(
    models = n(),
    lowest = min(CRAN / PR),
    highest = max(CRAN / PR),
    .groups = "drop"
  ) |>
  mutate(version = "CRAN")
```

greta#855 runs `mcmc()`'s warmup and sampling, and `opt()`'s loop with an optimiser such as `adam()`, inside TensorFlow, rather than returning to R every few iterations. This post compares it with greta 0.6.0 on CRAN, with main, and with [greta#850](https://github.com/greta-dev/greta/pull/850), which it supersedes, on four of greta's example models. Once the first call has traced the model:

- `mcmc()` with each version's defaults:
    - CRAN takes `{r} ratio_range(glance_ratios, "CRAN", row == glance_rows[2])` times as long as greta#855, which returns twice the draws
- `opt()` with `adam()`:
    - CRAN takes `{r} ratio_range(glance_ratios, "CRAN", row == glance_rows[7])` times as long as greta#855

Every version samples the same posteriors.

## The PR at a glance

[]{#summary}Three posts benchmark greta#855: this one, on the PR at 5bdab31e; [Call overhead](../2026-10-07-call-overhead-i547/), on its last two changes, which keep `opt()`'s traced loop on the model and the chain's state as tensors between calls, at f30ab146; and [TFP's loops](../2026-10-07-tfp-alternatives-i547/), on two TensorFlow Probability loops tried and not adopted. [Warmup schemes](../2026-10-07-warmup-adaptation-i853/) is about [greta#853](https://github.com/greta-dev/greta/issues/853), a branch built on this one and not part of the PR.

The PR's timings against CRAN, from the first two posts. Seconds are the median over every call on every model; the last numeric column is how many times as long CRAN takes as the PR, as the range over models of each model's median. Each method was called several times on each model: the first call traces the model, and the calls after it reuse what it traced.

```{r}
#| label: pr-at-a-glance
glance_seconds |>
  left_join(glance_ratios, by = "row") |>
  mutate(row = factor(row, levels = glance_rows)) |>
  arrange(row) |>
  transmute(
    call = as.character(row),
    models,
    `CRAN 0.6.0, s` = format_number(CRAN),
    `this PR, s` = format_number(PR),
    `CRAN ÷ this PR` = paste(format_number(lowest), "to", format_number(highest)),
    `measured in` = post
  ) |>
  knitr::kable(align = "lrrrrl")
```

With each version's defaults, the PR runs 2000 warmup iterations and 2000 draws where CRAN runs 1000 of each: about the same number of sampler iterations, since each of CRAN's draws costs two, so CRAN takes those times for half the draws. In the rows with the same warmup and draws, CRAN runs twice the sampler iterations, so those rows include [greta#850](https://github.com/greta-dev/greta/pull/850)'s thinning fix as well as the PR's own changes. With `verbose = FALSE`, as here, the PR's last two changes leave `mcmc()`'s time as it was, which [Call overhead](../2026-10-07-call-overhead-i547/#verbose-false-measured-again) measured on its own.

## How long does `mcmc()` take? {#speed}

Every `mcmc()` call on each model, with two sets of arguments: the same `warmup` and `n_samples` (`{r} bench_settings$warmup` and `{r} bench_settings$n_samples`) on every version, which is twice the sampler iterations on CRAN and main as on the others, and each version's defaults, which is about the same number of iterations on all of them. The first call on a new model traces its TensorFlow functions, and the second call reuses them.

```{r}
#| label: mcmc-speed
mcmc_timings <- every_call(mcmc_marks) |>
  mutate(calls = label_calls(call))
mcmc_summary <- bench_summary(mcmc_marks)

# for each model and call, each version's median seconds over #855's
mcmc_ratios <- mcmc_timings |>
  group_by(setting, calls, model, version) |>
  summarise(seconds = median(seconds), .groups = "drop_last") |>
  mutate(times_as_long = seconds / seconds[version == pr_label]) |>
  group_by(setting, calls, version) |>
  summarise(
    lowest = min(times_as_long),
    highest = max(times_as_long),
    .groups = "drop"
  )
```

```{r}
#| label: plot-mcmc-speed
#| fig-cap: !expr raincloud_caption
#| fig-width: 11
#| fig-height: 10
ggplot(mcmc_timings, aes(x = seconds, y = version, colour = version)) +
  raincloud +
  scale_x_log10() +
  facet_grid(vars(model), vars(setting, calls)) +
  labs(x = "Seconds, on a log scale", y = NULL)
```

Timed by `benchmark.R`'s `{r} section_links("benchmark.R", paste0("bench-mcmc-same-", model_sections))` sections (the same arguments), and `{r} section_links("benchmark.R", paste0("bench-mcmc-defaults-", model_sections))` (each version's defaults).

With each version's defaults, a second call takes `{r} ratio_range(mcmc_ratios, "CRAN", setting == setting_levels[2], calls == "second call")` times as long on CRAN as on #855, `{r} ratio_range(mcmc_ratios, "main", setting == setting_levels[2], calls == "second call")` on main and `{r} ratio_range(mcmc_ratios, "#850", setting == setting_levels[2], calls == "second call")` on #850; a first call takes `{r} ratio_range(mcmc_ratios, "CRAN", setting == setting_levels[2], calls == "first call")` times as long on CRAN. With the same arguments, a second call takes `{r} ratio_range(mcmc_ratios, "CRAN", setting == setting_levels[1], calls == "second call")` times as long on CRAN, which runs twice the iterations.

<details><summary>Table: bench's summary() of every mark</summary>

```{r}
#| label: table-mcmc-summary
dt_table(mcmc_summary, digits = 4)
```

</details>

## How long does `opt()` with `adam()` take? {#opt-with-adam}

Every call of `opt()` with `adam()` and up to `{r} bench_settings$opt_max_iterations` iterations, `{r} bench_settings$opt_calls` on each model, from the same seed each time.

```{r}
#| label: opt-speed
opt_timings <- every_call(opt_marks) |>
  mutate(calls = label_calls(call))
opt_summary <- bench_summary(opt_marks)

later_opt_medians <- opt_timings |>
  filter(call > 1) |>
  group_by(version) |>
  summarise(seconds = median(seconds))

# for each model and call, each version's median seconds over #855's
opt_ratios <- opt_timings |>
  group_by(calls, model, version) |>
  summarise(seconds = median(seconds), .groups = "drop_last") |>
  mutate(times_as_long = seconds / seconds[version == pr_label]) |>
  group_by(calls, version) |>
  summarise(
    lowest = min(times_as_long),
    highest = max(times_as_long),
    .groups = "drop"
  )
```

```{r}
#| label: plot-opt-speed
#| fig-cap: !expr raincloud_caption
#| fig-height: 9
ggplot(opt_timings, aes(x = seconds, y = version, colour = version)) +
  raincloud +
  scale_x_log10() +
  facet_grid(vars(model), vars(calls)) +
  labs(x = "Seconds, on a log scale", y = NULL)
```

Timed by `benchmark.R`'s `{r} section_links("benchmark.R", paste0("bench-opt-adam-", model_sections))` sections.

For a later call, CRAN takes `{r} ratio_range(opt_ratios, "CRAN", calls != "first call")` times as long as #855, and main `{r} ratio_range(opt_ratios, "main", calls != "first call")`. The median later call takes `{r} versions_text(later_opt_medians$version, later_opt_medians$seconds)` seconds. The PR's last commit also keeps `opt()`'s traced loop on the model, and on it CRAN takes `{r} ratio_range(glance_ratios, "CRAN", row == glance_rows[7])` times as long for a later call ([the PR at a glance](#the-pr-at-a-glance)).

<details><summary>Table: bench's summary() of every mark</summary>

```{r}
#| label: table-opt-summary
dt_table(opt_summary, digits = 4)
```

</details>

## How many effective samples per second? {#effective-samples}

The bulk effective sample size (ESS) [@vehtari2021] of the parameter with the smallest, as @hoffman2014 report, per second of the second `mcmc()` call on each model, warmup included: the time-normalised ESS of @girolami2011.

```{r}
#| label: efficiency
# every model's posterior summary, from each run's draws; the checks below use
# it too
setting_names <- c(same = setting_levels[1], defaults = setting_levels[2])
posterior <- lapply(seq_len(nrow(runs)), function(i) {
  draws <- runs$results[[i]]$mcmc_draws
  lapply(names(setting_names), function(setting_name) {
    lapply(names(draws[[setting_name]]), function(model_name) {
      tibble(
        session = runs$session[i],
        version = runs$version[i],
        setting = setting_names[[setting_name]],
        model = model_name,
        posterior::summarise_draws(
          draws[[setting_name]][[model_name]],
          "mean",
          "mcse_mean",
          "rhat",
          "ess_bulk"
        )
      )
    }) |>
      bind_rows()
  }) |>
    bind_rows()
}) |>
  bind_rows() |>
  mutate(
    model = factor(model, levels = model_levels),
    setting = factor(setting, levels = setting_levels)
  )

second_calls <- mcmc_timings |>
  filter(call == 2) |>
  select(session, version, model, setting, seconds)

efficiency <- posterior |>
  group_by(session, version, model, setting) |>
  summarise(smallest_ess_bulk = min(ess_bulk), .groups = "drop") |>
  left_join(second_calls, by = c("session", "version", "model", "setting")) |>
  mutate(smallest_ess_per_second = smallest_ess_bulk / seconds)

efficiency_means <- efficiency |>
  group_by(model, setting, version) |>
  summarise(
    smallest_ess_per_second = mean(smallest_ess_per_second),
    .groups = "drop"
  )

# for each model, #855's mean over each version's: how many times that
# version's ESS per second #855 gives
ess_ratios <- efficiency_means |>
  group_by(setting, model) |>
  mutate(
    pr_gives = smallest_ess_per_second[version == pr_label] /
      smallest_ess_per_second
  ) |>
  group_by(setting, version) |>
  summarise(lowest = min(pr_gives), highest = max(pr_gives), .groups = "drop")

gg_sessions <- function(data, means, x, x_label) {
  ggplot(data, aes(x = {{ x }}, y = version, colour = version)) +
    geom_point(data = means, size = 5, alpha = 0.25) +
    geom_point(position = jitter, size = 1.2) +
    version_scale +
    facet_wrap(vars(model, setting), ncol = 2, scales = "free_x") +
    labs(x = x_label, y = NULL)
}
```

```{r}
#| label: plot-efficiency
#| fig-cap: !expr session_caption
#| fig-height: 8
gg_sessions(
  efficiency,
  efficiency_means,
  smallest_ess_per_second,
  "Smallest bulk ESS per second"
)
```

The draws are kept by `benchmark.R`'s `{r} section_link("benchmark.R", "mcmc-draws")` section, and the seconds are the second calls timed in [How long does `mcmc()` take?](#speed).

With each version's defaults, #855 gives `{r} ratio_range(ess_ratios, "CRAN", setting == setting_levels[2])` times CRAN's smallest bulk ESS per second, and `{r} ratio_range(ess_ratios, "#850", setting == setting_levels[2])` times #850's. With the same arguments, it gives `{r} ratio_range(ess_ratios, "CRAN", setting == setting_levels[1])` times CRAN's.

<details><summary>Table: every session</summary>

```{r}
#| label: table-efficiency
dt_table(efficiency, digits = 1)
```

</details>

## Are more chains worth it? {#the-number-of-chains}

`mcmc()` with each version's defaults on `{r} paste_and(bench_settings$scaling_chains)` chains, the second call on each model: [greta#294](https://github.com/greta-dev/greta/issues/294) found that TensorFlow runs many chains at nearly the cost of a few.

```{r}
#| label: chains
chain_seconds <- lapply(seq_len(nrow(runs)), function(i) {
  marks <- runs$results[[i]]$chains
  lapply(names(marks), function(model_name) {
    mark <- marks[[model_name]]
    tibble(
      session = runs$session[i],
      version = runs$version[i],
      model = model_name,
      chains = mark$chains,
      # each mark's calls are the first and the second on its model
      seconds = vapply(mark$time, \(time) as.numeric(time)[2], numeric(1))
    )
  }) |>
    bind_rows()
}) |>
  bind_rows()

chain_ess <- lapply(seq_len(nrow(runs)), function(i) {
  ess <- runs$results[[i]]$chains_ess
  lapply(names(ess), function(model_name) {
    tibble(
      session = runs$session[i],
      version = runs$version[i],
      model = model_name,
      ess[[model_name]]
    )
  }) |>
    bind_rows()
}) |>
  bind_rows()

chain_measures <- chain_seconds |>
  left_join(chain_ess, by = c("session", "version", "model", "chains")) |>
  mutate(smallest_ess_per_second = smallest_ess_bulk / seconds) |>
  select(session, version, model, chains, seconds, smallest_ess_per_second) |>
  pivot_longer(
    c(seconds, smallest_ess_per_second),
    names_to = "measure",
    values_to = "value"
  ) |>
  mutate(
    model = factor(model, levels = intersect(model_levels, model)),
    chains = factor(chains, levels = bench_settings$scaling_chains),
    measure = factor(
      measure,
      levels = c("seconds", "smallest_ess_per_second"),
      labels = c("seconds", "smallest bulk ESS per second")
    )
  )

chain_means <- chain_measures |>
  group_by(model, measure, chains, version) |>
  summarise(value = mean(value), .groups = "drop")

fewest_chains <- as.character(min(bench_settings$scaling_chains))
most_chains <- as.character(max(bench_settings$scaling_chains))

# for each model and measure, #855's mean at the most chains over its mean at
# the fewest
pr_growth <- chain_means |>
  filter(version == pr_label) |>
  group_by(model, measure) |>
  summarise(
    growth = value[chains == most_chains] / value[chains == fewest_chains],
    .groups = "drop"
  )

pr_time_growth <- with(
  filter(pr_growth, measure == "seconds"),
  setNames(format_number(growth), model)
)
pr_ess_growth <- with(
  filter(pr_growth, measure != "seconds"),
  setNames(format_number(growth), model)
)

# at the most chains, #855's mean ESS per second over CRAN's
pr_over_cran <- chain_means |>
  filter(chains == most_chains, measure == "smallest bulk ESS per second") |>
  group_by(model) |>
  summarise(
    ratio = value[version == pr_label] / value[version == "CRAN"],
    .groups = "drop"
  ) |>
  with(setNames(format_number(ratio), model))
```

```{r}
#| label: plot-chains
#| fig-cap: "Each small dot is one session's second call. The large faint dot behind them is their mean."
#| fig-width: 10
#| fig-height: 7
ggplot(chain_measures, aes(x = value, y = version, colour = version)) +
  geom_point(data = chain_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_grid(
    vars(chains),
    vars(model, measure),
    scales = "free_x",
    labeller = labeller(chains = label_both)
  ) +
  labs(x = NULL, y = NULL)
```

Timed by `benchmark.R`'s `{r} section_links("benchmark.R", c("bench-chains-linear", "bench-chains-eight-schools"))` sections, which also keep each call's smallest bulk ESS.

From `{r} fewest_chains` chains to `{r} most_chains`, a call on #855 takes `{r} pr_time_growth[["linear"]]` times as long on linear and `{r} pr_time_growth[["eight_schools"]]` on eight_schools, and gives `{r} pr_ess_growth[["linear"]]` and `{r} pr_ess_growth[["eight_schools"]]` times the effective samples per second. At `{r} most_chains` chains, #855 gives `{r} pr_over_cran[["linear"]]` times CRAN's effective samples per second on linear, and `{r} pr_over_cran[["eight_schools"]]` times on eight_schools.

<details><summary>Table: every session</summary>

```{r}
#| label: table-chains
chain_measures |>
  pivot_wider(names_from = measure, values_from = value) |>
  dt_table(digits = 2)
```

</details>

## Do the versions sample the same posterior? {#checks-on-the-draws}

Faster sampling is only worth having if it samples the same posterior. Three checks, on the draws of each model's second `mcmc()` call, which `benchmark.R`'s `{r} section_link("benchmark.R", "mcmc-draws")` section keeps.

### Agreement

Whether two versions estimate the same posterior means. Each version's means and Monte Carlo standard errors (MCSE) are pooled over its sessions, and the difference between two versions' means is divided by the MCSE of that difference [@vehtari2021; @magnusson2025],

$$
z = \frac{\bar\theta_{\text{b}} - \bar\theta_{\text{a}}}{\sqrt{\text{se}^2_{\text{a}} + \text{se}^2_{\text{b}}}},
$$

where a version's pooled standard error is $\sqrt{\sum \text{mcse}^2} / n$ over its $n$ sessions, each with a seed of its own. If both versions sample the same posterior, each $z$ is close to a standard normal draw.

```{r}
#| label: agreement
pooled <- posterior |>
  group_by(setting, version, model, variable) |>
  summarise(
    sessions = n(),
    mean = mean(mean),
    se = sqrt(sum(mcse_mean^2)) / n(),
    .groups = "drop"
  )

pairs <- utils::combn(version_levels, 2)
agreement <- lapply(seq_len(ncol(pairs)), function(i) {
  reference <- filter(pooled, version == pairs[1, i])
  compared <- filter(pooled, version == pairs[2, i])
  inner_join(
    reference,
    compared,
    by = c("setting", "model", "variable"),
    suffix = c("_a", "_b")
  ) |>
    mutate(
      comparison = paste(pairs[2, i], "vs", pairs[1, i]),
      z = (mean_b - mean_a) / sqrt(se_a^2 + se_b^2)
    )
}) |>
  bind_rows() |>
  mutate(comparison = factor(comparison, levels = rev(unique(comparison))))

# the band all of a comparison's z fall inside with probability 0.95 when both
# versions sample the same posterior: a Bonferroni bound over its parameters
parameters_compared <- n_distinct(paste(agreement$model, agreement$variable))
band <- qnorm(1 - 0.05 / (2 * parameters_compared))

comparisons <- n_distinct(agreement$comparison, agreement$setting)
outside_band <- filter(agreement, abs(z) > band)
outside_text <- paste0(
  outside_band$comparison,
  "'s `",
  outside_band$variable,
  "` on ",
  outside_band$model,
  " with ",
  outside_band$setting,
  " (z = ",
  format_number(outside_band$z),
  ")"
)
```

```{r}
#| label: plot-agreement
#| fig-cap: "Each dot is one parameter. The grey band is where every parameter's z falls, with probability at least 0.95, when both versions sample the same posterior."
#| fig-height: 6
ggplot(agreement, aes(x = z, y = comparison)) +
  annotate(
    "rect",
    xmin = -band,
    xmax = band,
    ymin = -Inf,
    ymax = Inf,
    fill = "grey80",
    alpha = 0.5
  ) +
  geom_point(position = jitter, alpha = 0.5) +
  facet_wrap(vars(setting)) +
  labs(x = "z", y = NULL)
```

Computed from the draws kept by `benchmark.R`'s `{r} section_link("benchmark.R", "mcmc-draws")` section.

The grey band is $\pm$ `{r} format_number(band)`: if two versions sample the same posterior, all `{r} parameters_compared` parameters' $z$ fall inside it with probability at least 0.95. Outside it, `{r} nrow(outside_band)` of the `{r} nrow(agreement)` $z$: `{r} paste_and(outside_text)`. Each of the `{r} comparisons` comparisons has at most a 0.05 chance of a $z$ outside the band, so up to `{r} format_number(0.05 * comparisons)` such comparison is expected by chance.

<details><summary>Table: every parameter</summary>

```{r}
#| label: table-agreement
agreement |>
  select(setting, comparison, model, variable, mean_a, se_a, mean_b, se_b, z) |>
  dt_table()
```

</details>

### Convergence

For each session's second `mcmc()` call, the largest rank-normalised split-$\hat{R}$ over the model's parameters [@vehtari2021].

```{r}
#| label: convergence
convergence <- posterior |>
  group_by(session, version, model, setting) |>
  summarise(largest_rhat = max(rhat), .groups = "drop")

convergence_means <- convergence |>
  group_by(model, setting, version) |>
  summarise(largest_rhat = mean(largest_rhat), .groups = "drop")

rhat_threshold <- 1.01
unconverged <- convergence |>
  group_by(version) |>
  summarise(runs = n(), above = sum(largest_rhat > rhat_threshold))
worst_run <- slice_max(convergence, largest_rhat, n = 1)
```

```{r}
#| label: plot-rhat
#| fig-cap: !expr session_caption
#| fig-height: 8
gg_sessions(convergence, convergence_means, largest_rhat, "Largest R-hat")
```

Computed from the draws kept by `benchmark.R`'s `{r} section_link("benchmark.R", "mcmc-draws")` section.

The largest $\hat{R}$ is above `{r} rhat_threshold` in about as many runs on every version, of `{r} unique(unconverged$runs)` each: `{r} versions_text(unconverged$version, unconverged$above)`. No version converges in every run, as four chains of these lengths are short for these models. The highest, `{r} format_number(worst_run$largest_rhat)`, is `{r} worst_run$version`'s on `{r} worst_run$model` with `{r} worst_run$setting`.

<details><summary>Table: every session</summary>

```{r}
#| label: table-convergence
dt_table(convergence)
```

</details>

### Sampler iterations per kept draw

`?mcmc` says each kept draw is `thin` iterations of the sampler after the one before. `benchmark.R` counts the iterations between draws at `thin = 1` and `thin = 3`, with a random walk that accepts every proposal.

```{r}
#| label: sampler-iterations
iterations_by_version <- sampler_iterations_per_draw |>
  group_by(version, thin) |>
  summarise(
    iterations = paste(sort(unique(iterations_per_draw)), collapse = " or "),
    .groups = "drop"
  )

as_documented <- sampler_iterations_per_draw |>
  group_by(version) |>
  summarise(matches = all(iterations_per_draw == thin)) |>
  filter(matches) |>
  pull(version) |>
  as.character()

iterations_by_version |>
  pivot_wider(
    names_from = thin,
    values_from = iterations,
    names_prefix = "thin = "
  ) |>
  knitr::kable()
```

Counted by `benchmark.R`'s `{r} section_link("benchmark.R", "benchmark-sampler-iterations")` section, in every session.

Each kept draw costs `thin` iterations, as `?mcmc` says, on `{r} paste_and(as_documented)`.

<details><summary>Traceplots and posterior densities</summary>

Every draw of each chain from each model's second `mcmc()` call with each version's defaults: traceplots from session 1, numbered from the end of warmup, with a row of panels for each version; and densities pooled over sessions, every version's overlaid in one panel per parameter. CRAN and main kept every second iteration, so their traces have half the draws.

```{r}
#| label: draws
# every draw of every run with each version's defaults, one row per draw and
# parameter
draws <- lapply(seq_len(nrow(runs)), function(i) {
  model_draws <- runs$results[[i]]$mcmc_draws$defaults
  lapply(names(model_draws), function(model_name) {
    posterior::as_draws_df(model_draws[[model_name]]) |>
      as.data.frame() |>
      pivot_longer(
        -c(.chain, .iteration, .draw),
        names_to = "variable",
        values_to = "value"
      ) |>
      mutate(
        session = runs$session[i],
        version = runs$version[i],
        model = model_name
      )
  }) |>
    bind_rows()
}) |>
  bind_rows()

# a model's draws array is draws by chains by parameters
parameter_count <- function(name) {
  dim(runs$results[[1]]$mcmc_draws$same[[name]])[3]
}

# wide enough for a column of panels per parameter
draws_width <- function(name) {
  max(8, 1.6 * parameter_count(name))
}

densities_height <- function(name) {
  1.5 + 2 * ceiling(parameter_count(name) / 4)
}

gg_trace <- function(name) {
  model_draws <- draws |>
    filter(session == 1, model == name) |>
    mutate(variable = factor(variable, levels = unique(variable)))
  # each parameter's panels share its range across the versions, so they can
  # be compared down a column
  parameter_ranges <- model_draws |>
    group_by(variable) |>
    reframe(value = range(value)) |>
    crossing(version = unique(model_draws$version)) |>
    mutate(.iteration = 1L)
  ggplot(model_draws, aes(x = .iteration, y = value, colour = factor(.chain))) +
    geom_line(linewidth = 0.2) +
    geom_blank(
      data = parameter_ranges,
      aes(x = .iteration, y = value),
      inherit.aes = FALSE
    ) +
    facet_wrap(
      vars(version, variable),
      ncol = n_distinct(model_draws$variable),
      scales = "free_y"
    ) +
    labs(x = "Draw", y = NULL, colour = "Chain")
}

# every version's density in one panel per parameter, each with its own line
# type as well as colour, so a line drawn on top of an identical one still shows
gg_densities <- function(name) {
  draws |>
    filter(model == name) |>
    mutate(variable = factor(variable, levels = unique(variable))) |>
    ggplot(aes(x = value, colour = version, linetype = version)) +
    geom_density(linewidth = 0.7) +
    scale_colour_manual(values = version_colours) +
    facet_wrap(~variable, scales = "free", ncol = 4) +
    labs(x = NULL, y = NULL, colour = NULL, linetype = NULL) +
    theme(legend.position = "top")
}
```

::: panel-tabset

#### linear

```{r}
#| label: plot-trace-linear
#| fig-height: 6
#| fig-width: !expr draws_width("linear")
#| column: page
gg_trace("linear")
```

```{r}
#| label: plot-densities-linear
#| fig-height: !expr densities_height("linear")
gg_densities("linear")
```

#### multiple_linear

```{r}
#| label: plot-trace-multiple-linear
#| fig-height: 6
#| fig-width: !expr draws_width("multiple_linear")
#| column: page
gg_trace("multiple_linear")
```

```{r}
#| label: plot-densities-multiple-linear
#| fig-height: !expr densities_height("multiple_linear")
gg_densities("multiple_linear")
```

#### hierarchical_linear

```{r}
#| label: plot-trace-hierarchical-linear
#| fig-height: 6
#| fig-width: !expr draws_width("hierarchical_linear")
#| column: page
gg_trace("hierarchical_linear")
```

```{r}
#| label: plot-densities-hierarchical-linear
#| fig-height: !expr densities_height("hierarchical_linear")
gg_densities("hierarchical_linear")
```

#### eight_schools

```{r}
#| label: plot-trace-eight-schools
#| fig-height: 6
#| fig-width: !expr draws_width("eight_schools")
#| column: page
gg_trace("eight_schools")
```

```{r}
#| label: plot-densities-eight-schools
#| fig-height: !expr densities_height("eight_schools")
gg_densities("eight_schools")
```

:::

</details>

## How it was measured

Two scripts made every result here: `{r} file_link("benchmark.R")` is what each version runs, in a fresh R session of its own, and `{r} file_link("run.R")` runs it for each version, in `{r} n_sessions` sessions, with the versions in a new random order each session. Each was on `{r} unique(provenance$machine)` with `{r} unique(provenance$cores)` cores. `mcmc()` ran with `{r} bench_settings$chains` chains on `{r} bench_settings$n_cores` cores and `verbose = FALSE`, as set in `benchmark.R`'s `{r} section_link("benchmark.R", "benchmark-settings")` section.

```{r}
#| label: versions
version_notes <- c(
  CRAN = "greta 0.6.0, the release users upgrade from",
  main = "greta's main branch",
  `#850` = "makes each kept draw cost `thin` iterations, and doubles the default `warmup` and `n_samples` to 2000",
  `#855 early` = "an earlier commit of the PR: tunes warmup, draws `hmc()`'s leapfrog steps and runs `opt()`'s loop inside TensorFlow",
  `#855` = "the PR at 5bdab31e: also traces warmup and sampling as one function, for the number of chains, and keeps it on the model"
)

provenance |>
  distinct(version, greta_version, greta_sha, tensorflow_version, tfp_version, r_version) |>
  arrange(version) |>
  mutate(
    what_it_is = version_notes[as.character(version)],
    installed_from = versions[as.character(version)],
    # pak records a CRAN install's version in place of a commit, and a local
    # install nothing, so where it recorded no commit, take the one run.R names
    commit = ifelse(
      grepl("^[0-9a-f]{40}$", greta_sha),
      greta_sha,
      sub("^.*@", "", installed_from)
    ),
    installed_from = ifelse(
      grepl("^[0-9a-f]{40}$", commit),
      paste0(
        "[",
        sub("@.*", "", installed_from),
        "@",
        substr(commit, 1, 8),
        "](https://github.com/greta-dev/greta/commit/",
        commit,
        ")"
      ),
      installed_from
    )
  ) |>
  select(
    version,
    `what it is` = what_it_is,
    `installed from` = installed_from,
    greta = greta_version,
    TensorFlow = tensorflow_version,
    TFP = tfp_version,
    R = r_version
  ) |>
  knitr::kable()
```

The versions are pinned to these commits in `run.R`'s `{r} section_link("run.R", "run-versions")` section. Every version's `greta_sitrep()` report and `sessioninfo::session_info()` are in [Session details](#session-details).

<details><summary>Table: the order of each session</summary>

```{r}
#| label: table-order
runs |>
  select(session, position, version) |>
  arrange(session, position) |>
  dt_table()
```

</details>

```{r}
#| label: run-script
#| file: run.R
#| eval: false
#| code-summary: "run.R"
```

```{r}
#| label: benchmark-script
#| file: benchmark.R
#| eval: false
#| code-summary: "benchmark.R"
```

### The models

Four of greta's `inst/examples/`, each defined in maths and by its section of `benchmark.R`. Normal and log-normal distributions are written with their mean and standard deviation, as greta's `normal()` and `lognormal()` take them; $\text{Cauchy}^{+}$ is a Cauchy truncated to positive values.

::: panel-tabset

#### linear

A regression of each department's rating in `attitude` on its number of complaints. `{r} parameter_count("linear")` parameters. Its section is `{r} section_link("benchmark.R", "model-linear")`.

$$
\begin{aligned}
\text{rating}_i &\sim \text{Normal}(\mu_i, \sigma), \quad i = 1, \dots, 30 \\
\mu_i &= \alpha + \beta \, \text{complaints}_i \\
\alpha, \beta &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Cauchy}^{+}(0, 3)
\end{aligned}
$$

```{r}
#| label: model-linear
#| code-summary: "benchmark.R: model-linear"
```

#### multiple_linear

The same rating on all six of `attitude`'s other columns, $X$, through a matrix multiply. `{r} parameter_count("multiple_linear")` parameters. Its section is `{r} section_link("benchmark.R", "model-multiple-linear")`.

$$
\begin{aligned}
\text{rating}_i &\sim \text{Normal}(\mu_i, \sigma), \quad i = 1, \dots, 30 \\
\mu_i &= \alpha + \sum_{j = 1}^{6} X_{ij} \, \beta_j \\
\alpha, \beta_j &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Cauchy}^{+}(0, 3)
\end{aligned}
$$

```{r}
#| label: model-multiple-linear
#| code-summary: "benchmark.R: model-multiple-linear"
```

#### hierarchical_linear

Sepal length on sepal width in `iris`, with an offset for each species, $s[i]$ being flower $i$'s. Setosa's offset is zero and the other two share a scale, $\tau$. `{r} parameter_count("hierarchical_linear")` parameters. Its section is `{r} section_link("benchmark.R", "model-hierarchical-linear")`.

$$
\begin{aligned}
\text{Sepal.Length}_i &\sim \text{Normal}(\mu_i, \sigma), \quad i = 1, \dots, 150 \\
\mu_i &= \alpha + \beta \, \text{Sepal.Width}_i + \gamma_{s[i]} \\
\gamma_1 &= 0, \qquad \gamma_2, \gamma_3 \sim \text{Normal}(0, \tau) \\
\tau &\sim \text{LogNormal}(0, 1) \\
\alpha, \beta &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Cauchy}^{+}(0, 3)
\end{aligned}
$$

```{r}
#| label: model-hierarchical-linear
#| code-summary: "benchmark.R: model-hierarchical-linear"
```

#### eight_schools

The effect of coaching in eight schools, $y_j$, each measured with a known standard error, $\sigma_j$. The scale of the school effects, $\sigma_\eta$, is itself a parameter, which makes the posterior a funnel. `{r} parameter_count("eight_schools")` parameters. Its section is `{r} section_link("benchmark.R", "model-eight-schools")`.

$$
\begin{aligned}
y_j &\sim \text{Normal}(\theta_j, \sigma_j), \quad j = 1, \dots, 8 \\
\theta_j &= \mu + \xi \, \eta_j \\
\eta_j &\sim \text{Normal}(0, \sigma_\eta) \\
\sigma_\eta &\sim \text{InverseGamma}(1, 1) \\
\mu &\sim \text{Normal}(0, 100) \\
\xi &\sim \text{Normal}(0, 5)
\end{aligned}
$$

```{r}
#| label: model-eight-schools
#| code-summary: "benchmark.R: model-eight-schools"
```

:::

## Session details

What each version reported about itself in session 1, from `benchmark.R`'s `{r} section_link("benchmark.R", "benchmark-provenance")` section.

```{r}
#| label: session-one
# a saved session_info() prints as a plain list unless sessioninfo's print
# method is registered, which loading its namespace does
invisible(loadNamespace("sessioninfo"))
session_one <- runs |>
  filter(session == 1) |>
  arrange(version)
session_result <- function(version_label) {
  session_one$results[[which(session_one$version == version_label)]]
}
```

::: panel-tabset

### CRAN

<details><summary>`greta_sitrep()`</summary>

```{r}
#| label: sitrep-cran
writeLines(session_result("CRAN")$greta_sitrep_report)
```

</details>

<details><summary>`sessioninfo::session_info()`</summary>

```{r}
#| label: session-info-cran
session_result("CRAN")$session_info
```

</details>

### main

<details><summary>`greta_sitrep()`</summary>

```{r}
#| label: sitrep-main
writeLines(session_result("main")$greta_sitrep_report)
```

</details>

<details><summary>`sessioninfo::session_info()`</summary>

```{r}
#| label: session-info-main
session_result("main")$session_info
```

</details>

### #850

<details><summary>`greta_sitrep()`</summary>

```{r}
#| label: sitrep-850
writeLines(session_result("#850")$greta_sitrep_report)
```

</details>

<details><summary>`sessioninfo::session_info()`</summary>

```{r}
#| label: session-info-850
session_result("#850")$session_info
```

</details>

### #855 early

<details><summary>`greta_sitrep()`</summary>

```{r}
#| label: sitrep-855-early
writeLines(session_result("#855 early")$greta_sitrep_report)
```

</details>

<details><summary>`sessioninfo::session_info()`</summary>

```{r}
#| label: session-info-855-early
session_result("#855 early")$session_info
```

</details>

### #855

<details><summary>`greta_sitrep()`</summary>

```{r}
#| label: sitrep-855
writeLines(session_result("#855")$greta_sitrep_report)
```

</details>

<details><summary>`sessioninfo::session_info()`</summary>

```{r}
#| label: session-info-855
session_result("#855")$session_info
```

</details>

:::

## References

::: {#refs}
:::
