---
title: "greta#833 (wire-jit-compile): XLA compilation"
description: "Building a model, opt() and mcmc() on greta's CRAN release, main and greta#833, all with model(compile = TRUE), which only #833 acts on. With XLA, opt() is slower, building a model and mcmc() take about as long, and the posteriors are the same."
date: 2026-10-06
bibliography: ../../references.bib
---

```{r}
#| label: setup
#| code-summary: "Setup"
# the packages' start-up messages say nothing about the results
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(bench)
  library(here)
})

post_path <- "posts/2026-10-06-jit-compile-i833"

# each `# ---- label ----` section of the scripts, shown below by its label
knitr::read_chunk(here(post_path, "benchmark.R"))
knitr::read_chunk(here(post_path, "run.R"))

# links to the scripts on GitHub, at the commit being rendered when GitHub
# Actions renders the site
github_ref <- Sys.getenv("GITHUB_SHA", "main")
github_file_url <- function(file) {
  paste0(
    "https://github.com/greta-dev/greta.benchmarks/blob/",
    github_ref,
    "/",
    post_path,
    "/",
    file
  )
}

# a link to the lines of a script's `# ---- label ----` section. Quarto escapes
# the markdown in an inline `{r}` result unless it is wrapped in I()
section_link <- function(file, label) {
  lines <- readLines(here(post_path, file))
  headings <- grep("^# ---- .+ ----$", lines)
  start <- headings[lines[headings] == paste("# ----", label, "----")]
  end <- c(headings, length(lines) + 1)[match(start, headings) + 1] - 1
  while (end > start && !nzchar(trimws(lines[end]))) {
    end <- end - 1
  }
  I(paste0(
    "[`", label, "`](", github_file_url(file), "#L", start, "-L", end, ")"
  ))
}

# "CRAN, main and #833"
paste_and <- function(x) {
  if (length(x) <= 1) {
    return(x)
  }
  paste(paste(x[-length(x)], collapse = ", "), "and", x[length(x)])
}

# links to several sections of a script, as "a, b and c"
section_links <- function(file, labels) {
  links <- vapply(
    labels,
    \(label) unclass(section_link(file, label)),
    character(1)
  )
  I(paste_and(unname(links)))
}

# white panels, so the grey band in the agreement figure shows
theme_set(theme_bw())

sessions <- list.files(
  here(post_path, "results"),
  pattern = "^session-[0-9]+[.]rds$",
  full.names = TRUE
) |>
  lapply(readRDS)

versions <- sessions[[1]]$versions
version_levels <- names(versions)
version_colours <- setNames(
  c("#E69F00", "#56B4E9", "#009E73")[seq_along(version_levels)],
  version_levels
)
bench_settings <- sessions[[1]]$results[[1]]$bench_settings
n_sessions <- length(sessions)

# one row per session and version, holding that version's results
runs <- lapply(sessions, function(session_file) {
  tibble(
    session = session_file$session,
    version = factor(names(session_file$results), levels = version_levels),
    position = match(names(session_file$results), session_file$order),
    results = unname(session_file$results)
  )
}) |>
  bind_rows()

model_levels <- names(runs$results[[1]]$build)
# each model's part of a section label in benchmark.R, such as
# bench-opt-multiple-linear
model_sections <- gsub("_", "-", model_levels)

# a part of the results that is a table or a value per run, stacked, with the
# run it came from
stack_tables <- function(part) {
  pieces <- lapply(seq_len(nrow(runs)), function(i) {
    values <- runs$results[[i]][[part]]
    if (!is.data.frame(values)) {
      values <- tibble(value = values)
    }
    tibble(session = runs$session[i], version = runs$version[i], values)
  })
  bind_rows(pieces)
}

# a part of the results that is a bench_mark per model, stacked into one
# bench_mark, with the run and model each came from
stack_marks <- function(part) {
  pieces <- list()
  for (i in seq_len(nrow(runs))) {
    marks <- runs$results[[i]][[part]]
    for (j in seq_along(marks)) {
      mark <- marks[[j]]
      mark$session <- runs$session[i]
      mark$version <- runs$version[i]
      mark$model <- names(marks)[j]
      pieces[[length(pieces) + 1]] <- mark
    }
  }
  vctrs::vec_rbind(!!!pieces)
}

provenance <- stack_tables("provenance")
sampler_iterations_per_draw <- stack_tables("sampler_iterations_per_draw")
build_marks <- stack_marks("build")
opt_marks <- stack_marks("opt")
mcmc_marks <- stack_marks("mcmc")

cran_version <- unique(provenance$greta_version[provenance$version == "CRAN"])

# one row per timed call: bench keeps every call's time in the `time` column
every_call <- function(marks) {
  marks |>
    as_tibble() |>
    select(session, version, model, time) |>
    mutate(call = lapply(time, seq_along)) |>
    unnest(c(time, call)) |>
    mutate(
      seconds = as.numeric(time),
      model = factor(model, levels = model_levels),
      .keep = "unused"
    )
}

# The first call on a model traces its TensorFlow functions, and on #833
# compiles them with XLA; later calls on the same model reuse them.
label_calls <- function(call) {
  last <- max(call)
  later <- if (last == 2) {
    "second call on the same model"
  } else {
    paste0("calls 2 to ", last, " on the same model")
  }
  factor(
    ifelse(call == 1, "first call on a new model", later),
    levels = c("first call on a new model", later)
  )
}

build <- every_call(build_marks)
opt_timings <- every_call(opt_marks) |>
  mutate(calls = label_calls(call))
mcmc_timings <- every_call(mcmc_marks) |>
  mutate(calls = label_calls(call))

# bench's own summary() of each mark, with the mean and longest of its calls
# added, all in seconds
bench_summary <- function(marks) {
  summaries <- summary(marks, filter_gc = FALSE)
  tibble(
    session = summaries$session,
    version = summaries$version,
    model = summaries$model,
    min = as.numeric(summaries$min),
    median = as.numeric(summaries$median),
    mean = vapply(marks$time, \(time) mean(as.numeric(time)), numeric(1)),
    max = vapply(marks$time, \(time) max(as.numeric(time)), numeric(1)),
    `itr/sec` = summaries$`itr/sec`,
    n_itr = summaries$n_itr,
    total_time = as.numeric(summaries$total_time)
  )
}

# each version's median call in each group, and how many times as long as
# #833's median it takes, so #833 is the unit
relative_medians <- function(timings, ...) {
  timings |>
    group_by(..., version) |>
    summarise(median_seconds = median(seconds), .groups = "drop_last") |>
    mutate(
      times_as_long_as_833 = median_seconds /
        median_seconds[version == "#833"]
    ) |>
    ungroup()
}

# three significant figures each, trailing zeros kept, where format() would pad
# a vector to the decimal places of its smallest value
format_number <- function(x) {
  formatC(signif(x, 3), format = "fg", digits = 3, flag = "#", big.mark = ",")
}

# how many times as long a version's median call takes as #833's
times_as_long <- function(timings, version = "main") {
  ratios <- relative_medians(timings)
  format_number(ratios$times_as_long_as_833[ratios$version == version])
}

version_scale <- scale_colour_manual(values = version_colours, guide = "none")
version_fill <- scale_fill_manual(values = version_colours, guide = "none")
jitter <- position_jitter(width = 0, height = 0.2, seed = 2026)

# A raincloud for each version: the density of its calls above, their box plot
# in the middle, and every call as a dot below. The summaries are drawn first
# and faint, so the dots sit on top.
raincloud <- list(
  ggdist::stat_halfeye(
    aes(fill = version),
    width = 0.4,
    justification = -0.3,
    .width = 0,
    point_colour = NA,
    alpha = 0.3
  ),
  geom_boxplot(
    aes(fill = version),
    width = 0.25,
    outlier.shape = NA,
    alpha = 0.3
  ),
  # a fixed dot size, so a call looks the same size in every panel
  ggdist::stat_dots(
    aes(fill = version),
    side = "bottom",
    layout = "swarm",
    justification = 1.3,
    scale = 0.75,
    binwidth = unit(2.6, "mm"),
    overflow = "compress",
    colour = NA,
    alpha = 0.6
  ),
  version_scale,
  version_fill,
  theme_bw(base_size = 13)
)
raincloud_caption <- "Each dot is one call, from every session. Above the dots, the curve is the density of the calls and the box marks their median and quartiles."
session_caption <- "Each small dot is one session's second mcmc() call. The large faint dot behind them is their mean."

# a searchable table, its numbers rounded for reading
dt_table <- function(data, digits = 3) {
  table <- DT::datatable(
    data,
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 10, autoWidth = TRUE)
  )
  decimal_columns <- names(data)[vapply(data, is.double, logical(1))]
  if (length(decimal_columns) == 0) {
    return(table)
  }
  DT::formatRound(table, decimal_columns, digits)
}

# a model's draws array is draws by chains by parameters
parameter_count <- function(name) {
  dim(runs$results[[1]]$mcmc_draws[[name]])[3]
}

# wide enough for a column of panels per parameter
draws_width <- function(name) {
  max(8, 1.6 * parameter_count(name))
}
```

[greta#833](https://github.com/greta-dev/greta/pull/833) makes `model(compile = TRUE)` compile the model's TensorFlow functions with [XLA](https://openxla.org/xla), and makes `compile = FALSE` the default. Every model here is built with `compile = TRUE`, on CRAN's greta `{r} cran_version`, on main and on #833; CRAN and main ignore the argument, so the difference between main and #833 is XLA. Building a model takes as long on all three, but `opt()` is slower with XLA: main takes `{r} times_as_long(filter(opt_timings, call == 1))` times as long as #833 on a model's first call, and `{r} times_as_long(filter(opt_timings, call > 1))` times as long on later calls. `mcmc()` takes about as long, main taking `{r} times_as_long(filter(mcmc_timings, call == 1))` times as long as #833 on the first call and `{r} times_as_long(filter(mcmc_timings, call == 2))` times as long on the second, and all three sample the same posterior.

## How long does building a model take? {#build}

Calling each model's function, which creates its greta arrays and calls `model()`, `{r} bench_settings$build_calls` times in each session.

```{r}
#| label: plot-build
#| fig-cap: !expr raincloud_caption
#| fig-height: 13
build_summary <- bench_summary(build_marks)

ggplot(build, aes(x = seconds, y = version, colour = version)) +
  raincloud +
  facet_wrap(~model, ncol = 1, scales = "free_x") +
  labs(x = "Seconds", y = NULL)
```

Timed by `benchmark.R`'s `{r} section_links("benchmark.R", paste0("bench-build-", model_sections))` sections.

Building a model compiles nothing, since XLA compiles on a model's first call: over all models, main takes `{r} times_as_long(build)` times as long as #833, and CRAN `{r} times_as_long(build, "CRAN")` times as long.

<details><summary>Table: bench's summary() of every mark</summary>

```{r}
#| label: table-build-summary
dt_table(build_summary, digits = 4)
```

</details>

## How long does opt() take? {#opt}

`opt()`, `{r} bench_settings$opt_calls` times on the same model: the first call traces the model's TensorFlow functions, which #833 also compiles with XLA, and the later calls reuse them.

```{r}
#| label: plot-opt
#| fig-cap: !expr raincloud_caption
#| fig-height: 16
opt_summary <- bench_summary(opt_marks)
opt_by_model <- relative_medians(opt_timings, model, calls)

# the models whose later calls are faster on #833 than on main
opt_later_faster <- opt_by_model |>
  filter(
    version == "main",
    calls != "first call on a new model",
    times_as_long_as_833 > 1
  ) |>
  pull(model) |>
  as.character()

ggplot(opt_timings, aes(x = seconds, y = version, colour = version)) +
  raincloud +
  facet_wrap(vars(model, calls), ncol = 2, scales = "free_x") +
  labs(x = "Seconds", y = NULL)
```

Timed by `benchmark.R`'s `{r} section_links("benchmark.R", paste0("bench-opt-", model_sections))` sections.

Over all models, main takes `{r} times_as_long(filter(opt_timings, call == 1))` times as long as #833 on a model's first call, and `{r} times_as_long(filter(opt_timings, call > 1))` times as long on its later calls. #833's later calls are faster than main's only on `{r} paste_and(opt_later_faster)`.

<details><summary>Table: median seconds per model and call, and how many times as long as #833's</summary>

```{r}
#| label: table-opt-ratios
dt_table(opt_by_model)
```

</details>

<details><summary>Table: bench's summary() of every mark</summary>

```{r}
#| label: table-opt-summary
dt_table(opt_summary, digits = 4)
```

</details>

## How long does mcmc() take? {#mcmc}

`mcmc()`, with `{r} bench_settings$chains` chains of `{r} bench_settings$warmup` warmup and `{r} bench_settings$n_samples` kept draws, twice on the same model: the first call traces the model's TensorFlow functions, which #833 also compiles with XLA, and the second reuses them.

```{r}
#| label: plot-mcmc
#| fig-cap: !expr raincloud_caption
#| fig-height: 16
mcmc_summary <- bench_summary(mcmc_marks)
mcmc_by_model <- relative_medians(mcmc_timings, model, calls)

main_second_call <- mcmc_by_model |>
  filter(version == "main", calls == "second call on the same model")
mcmc_least <- main_second_call |>
  slice_min(times_as_long_as_833, with_ties = FALSE)
mcmc_most <- main_second_call |>
  slice_max(times_as_long_as_833, with_ties = FALSE)

ggplot(mcmc_timings, aes(x = seconds, y = version, colour = version)) +
  raincloud +
  facet_wrap(vars(model, calls), ncol = 2, scales = "free_x") +
  labs(x = "Seconds", y = NULL)
```

Timed by `benchmark.R`'s `{r} section_links("benchmark.R", paste0("bench-mcmc-", model_sections))` sections.

Over all models, main takes `{r} times_as_long(filter(mcmc_timings, call == 1))` times as long as #833 on a model's first call, and `{r} times_as_long(filter(mcmc_timings, call == 2))` times as long on its second. Model by model, on the second call, main takes from `{r} format_number(mcmc_least$times_as_long_as_833)` times as long as #833, on `{r} as.character(mcmc_least$model)`, to `{r} format_number(mcmc_most$times_as_long_as_833)` times as long, on `{r} as.character(mcmc_most$model)`.

<details><summary>Table: median seconds per model and call, and how many times as long as #833's</summary>

```{r}
#| label: table-mcmc-ratios
dt_table(mcmc_by_model)
```

</details>

<details><summary>Table: bench's summary() of every mark</summary>

```{r}
#| label: table-mcmc-summary
dt_table(mcmc_summary, digits = 4)
```

</details>

## Does #833 give more effective samples per second? {#efficiency}

The bulk effective sample size (ESS) per second [@girolami2011; @vehtari2021] of each model's parameter with the smallest bulk ESS, as @hoffman2014 report it, from each session's second `mcmc()` call, warmup included.

```{r}
#| label: plot-efficiency
#| fig-cap: !expr session_caption
#| fig-height: 6
# each model's posterior summary in each run, from every draw of every chain
# of its second mcmc() call
posterior <- lapply(seq_len(nrow(runs)), function(i) {
  draws <- runs$results[[i]]$mcmc_draws
  lapply(names(draws), function(model_name) {
    tibble(
      session = runs$session[i],
      version = runs$version[i],
      model = model_name,
      posterior::summarise_draws(
        draws[[model_name]],
        "mean",
        "sd",
        "mcse_mean",
        "rhat",
        "ess_bulk",
        "ess_tail"
      )
    )
  }) |>
    bind_rows()
}) |>
  bind_rows() |>
  mutate(model = factor(model, levels = model_levels))

second_calls <- mcmc_timings |>
  filter(call == 2) |>
  select(session, version, model, seconds)

efficiency <- posterior |>
  group_by(session, version, model) |>
  summarise(
    smallest_ess_bulk = min(ess_bulk),
    median_ess_bulk = median(ess_bulk),
    .groups = "drop"
  ) |>
  left_join(second_calls, by = c("session", "version", "model")) |>
  mutate(
    smallest_ess_per_second = smallest_ess_bulk / seconds,
    median_ess_per_second = median_ess_bulk / seconds
  )

efficiency_means <- efficiency |>
  group_by(model, version) |>
  summarise(
    smallest_ess_per_second = mean(smallest_ess_per_second),
    median_ess_per_second = mean(median_ess_per_second),
    .groups = "drop"
  )

# how many times main's mean smallest ESS per second #833 gives, per model
efficiency_vs_main <- efficiency_means |>
  group_by(model) |>
  summarise(
    times_main = smallest_ess_per_second[version == "#833"] /
      smallest_ess_per_second[version == "main"]
  )
efficiency_least <- slice_min(efficiency_vs_main, times_main, with_ties = FALSE)
efficiency_most <- slice_max(efficiency_vs_main, times_main, with_ties = FALSE)

# the widest spread between one version's sessions on one model
efficiency_spread <- efficiency |>
  group_by(model, version) |>
  summarise(
    times = max(smallest_ess_per_second) / min(smallest_ess_per_second),
    .groups = "drop"
  ) |>
  slice_max(times, with_ties = FALSE)

ggplot(efficiency, aes(x = smallest_ess_per_second, y = version, colour = version)) +
  geom_point(data = efficiency_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(~model, ncol = 1, scales = "free_x") +
  labs(x = "Smallest bulk ESS per second", y = NULL)
```

Draws kept by `benchmark.R`'s `{r} section_link("benchmark.R", "mcmc-draws")` section, from the calls timed by its `{r} section_links("benchmark.R", paste0("bench-mcmc-", model_sections))` sections.

Averaged over sessions, #833 gives from `{r} format_number(efficiency_least$times_main)` times main's smallest bulk ESS per second, on `{r} as.character(efficiency_least$model)`, to `{r} format_number(efficiency_most$times_main)` times main's, on `{r} as.character(efficiency_most$model)`. One version's sessions differ by up to `{r} format_number(efficiency_spread$times)` times on the same model (`{r} as.character(efficiency_spread$version)` on `{r} as.character(efficiency_spread$model)`), so these differences between versions are within the noise.

<details><summary>Figure: median bulk ESS per second</summary>

```{r}
#| label: plot-efficiency-median
#| fig-cap: !expr session_caption
#| fig-height: 6
ggplot(efficiency, aes(x = median_ess_per_second, y = version, colour = version)) +
  geom_point(data = efficiency_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(~model, ncol = 1, scales = "free_x") +
  labs(x = "Median bulk ESS per second", y = NULL)
```

</details>

<details><summary>Table: mean ESS per second</summary>

```{r}
#| label: table-efficiency-means
dt_table(efficiency_means, digits = 1)
```

</details>

## Do the versions sample the same posterior? {#agreement}

For each parameter, the difference between two versions' posterior means, pooled over sessions, divided by the Monte Carlo standard error of that difference, $z$ [@vehtari2021; @magnusson2025]. If both versions sample the same posterior, each $z$ is close to a standard normal draw.

```{r}
#| label: plot-agreement
#| fig-cap: "Each dot is one parameter. The grey band is where every parameter's z falls, with probability at least 0.95, when both versions sample the same posterior."
#| fig-height: 6
# Each session has a seed of its own, so the sessions are independent, and a
# version's pooled standard error is sqrt(sum(mcse^2)) / n over its n sessions
pooled <- posterior |>
  group_by(version, model, variable) |>
  summarise(
    sessions = n(),
    mean = mean(mean),
    se = sqrt(sum(mcse_mean^2)) / n(),
    .groups = "drop"
  )

pairs <- utils::combn(version_levels, 2)
agreement <- lapply(seq_len(ncol(pairs)), function(i) {
  reference <- filter(pooled, version == pairs[1, i])
  compared <- filter(pooled, version == pairs[2, i])
  inner_join(reference, compared, by = c("model", "variable"), suffix = c("_a", "_b")) |>
    mutate(
      comparison = paste(pairs[2, i], "vs", pairs[1, i]),
      z = (mean_b - mean_a) / sqrt(se_a^2 + se_b^2)
    )
}) |>
  bind_rows() |>
  mutate(comparison = factor(comparison, levels = unique(comparison)))

# the band all of a comparison's z fall inside with probability 0.95 when both
# versions sample the same posterior: a Bonferroni bound over its parameters
parameters_compared <- n_distinct(paste(agreement$model, agreement$variable))
band <- qnorm(1 - 0.05 / (2 * parameters_compared))

outside_band <- agreement |>
  group_by(comparison) |>
  summarise(outside = sum(abs(z) > band), largest_abs_z = max(abs(z)))
main_vs_833 <- filter(outside_band, comparison == "#833 vs main")

ggplot(agreement, aes(x = z, y = model)) +
  annotate(
    "rect",
    xmin = -band,
    xmax = band,
    ymin = -Inf,
    ymax = Inf,
    fill = "grey80",
    alpha = 0.5
  ) +
  geom_point(position = jitter, alpha = 0.7) +
  facet_wrap(~comparison, ncol = 1) +
  labs(x = "z", y = NULL)
```

Draws kept by `benchmark.R`'s `{r} section_link("benchmark.R", "mcmc-draws")` section.

The grey band is $\pm$ `{r} format_number(band)`, which all `{r} parameters_compared` parameters' $z$ fall inside with probability at least 0.95 when two versions sample the same posterior, by a Bonferroni bound. Parameters outside it: `{r} paste(outside_band$comparison, outside_band$outside, collapse = ", ")`; the largest $|z|$ between #833 and main is `{r} format_number(main_vs_833$largest_abs_z)`.

<details><summary>Figure: posterior densities</summary>

Every draw of every chain, from each session's second `mcmc()` call, pooled over sessions, with each version's density overlaid in one panel per parameter.

```{r}
#| label: densities
# every draw of every run, one row per draw and parameter
draws <- lapply(seq_len(nrow(runs)), function(i) {
  model_draws <- runs$results[[i]]$mcmc_draws
  lapply(names(model_draws), function(model_name) {
    posterior::as_draws_df(model_draws[[model_name]]) |>
      as.data.frame() |>
      pivot_longer(
        -c(.chain, .iteration, .draw),
        names_to = "variable",
        values_to = "value"
      ) |>
      mutate(
        session = runs$session[i],
        version = runs$version[i],
        model = model_name
      )
  }) |>
    bind_rows()
}) |>
  bind_rows()

# each version with its own line type as well as colour, so a line drawn on
# top of an identical one still shows
gg_densities <- function(name) {
  draws |>
    filter(model == name) |>
    mutate(variable = factor(variable, levels = unique(variable))) |>
    ggplot(aes(x = value, colour = version, linetype = version)) +
    geom_density(linewidth = 0.7) +
    scale_colour_manual(values = version_colours) +
    facet_wrap(~variable, scales = "free", ncol = 4) +
    labs(x = NULL, y = NULL, colour = NULL, linetype = NULL) +
    theme(legend.position = "top")
}

densities_height <- function(name) {
  1.5 + 2 * ceiling(parameter_count(name) / 4)
}
```

::: panel-tabset

### linear

```{r}
#| label: plot-densities-linear
#| fig-height: !expr densities_height("linear")
gg_densities("linear")
```

### multiple_linear

```{r}
#| label: plot-densities-multiple-linear
#| fig-height: !expr densities_height("multiple_linear")
gg_densities("multiple_linear")
```

### hierarchical_linear

```{r}
#| label: plot-densities-hierarchical-linear
#| fig-height: !expr densities_height("hierarchical_linear")
gg_densities("hierarchical_linear")
```

### eight_schools

```{r}
#| label: plot-densities-eight-schools
#| fig-height: !expr densities_height("eight_schools")
gg_densities("eight_schools")
```

:::

</details>

<details><summary>Table: largest |z| per comparison and model</summary>

```{r}
#| label: table-agreement
agreement |>
  group_by(comparison, model) |>
  summarise(
    parameters = n(),
    largest_abs_z = max(abs(z)),
    parameter = variable[which.max(abs(z))],
    .groups = "drop"
  ) |>
  dt_table(digits = 2)
```

</details>

## Do the chains converge? {#convergence}

The largest rank-normalised split-$\hat{R}$ over a model's parameters, from each session's second `mcmc()` call [@vehtari2021].

```{r}
#| label: plot-rhat
#| fig-cap: !expr session_caption
#| fig-height: 6
convergence <- posterior |>
  group_by(session, version, model) |>
  summarise(
    largest_rhat = max(rhat),
    smallest_ess_bulk = min(ess_bulk),
    .groups = "drop"
  )

convergence_means <- convergence |>
  group_by(model, version) |>
  summarise(
    largest_rhat = mean(largest_rhat),
    smallest_ess_bulk = mean(smallest_ess_bulk),
    .groups = "drop"
  )

# Vehtari et al. (2021) take chains to have converged when every parameter's
# R-hat is below this
rhat_threshold <- 1.01
runs_over_threshold <- convergence |>
  group_by(version) |>
  summarise(runs = sum(largest_rhat > rhat_threshold))
runs_per_version <- n_sessions * length(model_levels)
largest_rhat <- slice_max(posterior, rhat, with_ties = FALSE)

ggplot(convergence, aes(x = largest_rhat, y = version, colour = version)) +
  geom_point(data = convergence_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(~model, ncol = 1, scales = "free_x") +
  labs(x = "Largest R-hat", y = NULL)
```

Draws kept by `benchmark.R`'s `{r} section_link("benchmark.R", "mcmc-draws")` section.

On every version, most runs' largest $\hat{R}$ is above `{r} rhat_threshold`: `{r} paste(runs_over_threshold$version, runs_over_threshold$runs, collapse = ", ")` of `{r} runs_per_version` runs. The largest is `{r} format_number(largest_rhat$rhat)`, for `{r} as.character(largest_rhat$model)`'s `{r} largest_rhat$variable` on `{r} as.character(largest_rhat$version)`.

<details><summary>Figure: traceplots</summary>

Every draw of each chain from session 1's second `mcmc()` call on each model, numbered from the end of warmup, with a row of panels for each version.

```{r}
#| label: traceplots
gg_trace <- function(name) {
  model_draws <- draws |>
    filter(session == 1, model == name) |>
    mutate(variable = factor(variable, levels = unique(variable)))
  # each parameter's panels share its range across the versions, so they can
  # be compared down a column
  parameter_ranges <- model_draws |>
    group_by(variable) |>
    reframe(value = range(value)) |>
    crossing(version = unique(model_draws$version)) |>
    mutate(.iteration = 1L)
  ggplot(model_draws, aes(x = .iteration, y = value, colour = factor(.chain))) +
    geom_line(linewidth = 0.2) +
    geom_blank(
      data = parameter_ranges,
      aes(x = .iteration, y = value),
      inherit.aes = FALSE
    ) +
    facet_wrap(
      vars(version, variable),
      ncol = n_distinct(model_draws$variable),
      scales = "free_y"
    ) +
    labs(x = "Draw", y = NULL, colour = "Chain")
}
```

::: panel-tabset

### linear

```{r}
#| label: plot-trace-linear
#| fig-height: 6
#| fig-width: !expr draws_width("linear")
#| column: page
gg_trace("linear")
```

### multiple_linear

```{r}
#| label: plot-trace-multiple-linear
#| fig-height: 6
#| fig-width: !expr draws_width("multiple_linear")
#| column: page
gg_trace("multiple_linear")
```

### hierarchical_linear

```{r}
#| label: plot-trace-hierarchical-linear
#| fig-height: 6
#| fig-width: !expr draws_width("hierarchical_linear")
#| column: page
gg_trace("hierarchical_linear")
```

### eight_schools

```{r}
#| label: plot-trace-eight-schools
#| fig-height: 6
#| fig-width: !expr draws_width("eight_schools")
#| column: page
gg_trace("eight_schools")
```

:::

</details>

<details><summary>Table: largest R-hat and smallest bulk ESS of every run</summary>

```{r}
#| label: table-convergence
dt_table(convergence)
```

</details>

## How it was measured {#how-it-was-measured}

The versions, pinned to commits in `run.R`'s `{r} section_link("run.R", "run-versions")` section:

```{r}
#| label: versions
provenance |>
  distinct(version, greta_version, greta_sha, tensorflow_version, tfp_version, r_version) |>
  arrange(version) |>
  mutate(
    installed_from = versions[as.character(version)],
    # pak records a CRAN install's version in place of a commit, so only a
    # GitHub install gets a link to its commit
    installed_from = ifelse(
      grepl("^[0-9a-f]{40}$", greta_sha),
      paste0(
        "[",
        sub("@.*", "", installed_from),
        "@",
        substr(greta_sha, 1, 8),
        "](https://github.com/greta-dev/greta/commit/",
        greta_sha,
        ")"
      ),
      installed_from
    )
  ) |>
  select(
    version,
    `installed from` = installed_from,
    greta = greta_version,
    TensorFlow = tensorflow_version,
    TFP = tfp_version,
    R = r_version
  ) |>
  knitr::kable()

python_stacks <- provenance |>
  distinct(version, tensorflow_version, tfp_version) |>
  group_by(tensorflow_version, tfp_version) |>
  summarise(
    versions = paste_and(as.character(sort(version))),
    .groups = "drop"
  )
python_stack_text <- paste0(
  python_stacks$versions,
  " ran on TensorFlow ",
  python_stacks$tensorflow_version,
  " and TensorFlow Probability ",
  python_stacks$tfp_version,
  collapse = "; "
)

xla_compiled <- lapply(seq_len(nrow(runs)), function(i) {
  compiled <- runs$results[[i]]$xla_compiled
  tibble(
    session = runs$session[i],
    version = runs$version[i],
    model = names(compiled),
    xla_compiled = unname(compiled)
  )
}) |>
  bind_rows()

xla_by_version <- xla_compiled |>
  group_by(version) |>
  summarise(compiled = paste(sort(unique(xla_compiled)), collapse = " or "))

sampler_iterations_values <- sampler_iterations_per_draw |>
  group_by(version) |>
  summarise(values = paste(sort(unique(value)), collapse = " or "))
```

`{r} python_stack_text`. Each version ran in `{r} n_sessions` sessions, on `{r} unique(provenance$machine)` with `{r} unique(provenance$cores)` cores. In each session, `run.R`'s `{r} section_link("run.R", "run-sessions")` section ran `benchmark.R` once per version, each in a fresh R session, with the versions in a new random order.

<details><summary>Table: the order of each session</summary>

```{r}
#| label: table-order
runs |>
  select(session, position, version) |>
  arrange(session, position) |>
  dt_table()
```

</details>

Every model is built with `compile = TRUE`, and the settings are in `benchmark.R`'s `{r} section_link("benchmark.R", "benchmark-settings")` section. Whether XLA compiled the log-density function `mcmc()` used, read in `{r} section_link("benchmark.R", "benchmark-xla")`, over every model and session: `{r} paste(xla_by_version$version, xla_by_version$compiled, collapse = ", ")`. Sampler iterations per kept draw, measured in `{r} section_link("benchmark.R", "benchmark-sampler-iterations")`, in every session: `{r} paste(sampler_iterations_values$version, sampler_iterations_values$values, collapse = ", ")`, so the same `mcmc()` arguments are the same work on each version.

### The models

The models are from greta's `inst/examples/`. Normal and log-normal distributions are written with their mean and standard deviation, as greta's `normal()` and `lognormal()` take them; $\text{Cauchy}^{+}$ is a Cauchy truncated to positive values.

::: panel-tabset

#### linear

A regression of each department's rating in `attitude` on its number of complaints, with `{r} parameter_count("linear")` parameters.

$$
\begin{aligned}
\text{rating}_i &\sim \text{Normal}(\mu_i, \sigma), \quad i = 1, \dots, 30 \\
\mu_i &= \alpha + \beta \, \text{complaints}_i \\
\alpha, \beta &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Cauchy}^{+}(0, 3)
\end{aligned}
$$

```{r}
#| label: model-linear
#| eval: false
#| code-summary: "benchmark.R: model-linear"
```

#### multiple_linear

The same rating on all six of `attitude`'s other columns, $X$, through a matrix multiply, with `{r} parameter_count("multiple_linear")` parameters.

$$
\begin{aligned}
\text{rating}_i &\sim \text{Normal}(\mu_i, \sigma), \quad i = 1, \dots, 30 \\
\mu_i &= \alpha + \sum_{j = 1}^{6} X_{ij} \, \beta_j \\
\alpha, \beta_j &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Cauchy}^{+}(0, 3)
\end{aligned}
$$

```{r}
#| label: model-multiple-linear
#| eval: false
#| code-summary: "benchmark.R: model-multiple-linear"
```

#### hierarchical_linear

Sepal length on sepal width in `iris`, with an offset for each species, $s[i]$ being flower $i$'s. Setosa's offset is zero and the other two share a scale, $\tau$. `{r} parameter_count("hierarchical_linear")` parameters.

$$
\begin{aligned}
\text{Sepal.Length}_i &\sim \text{Normal}(\mu_i, \sigma), \quad i = 1, \dots, 150 \\
\mu_i &= \alpha + \beta \, \text{Sepal.Width}_i + \gamma_{s[i]} \\
\gamma_1 &= 0, \qquad \gamma_2, \gamma_3 \sim \text{Normal}(0, \tau) \\
\tau &\sim \text{LogNormal}(0, 1) \\
\alpha, \beta &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Cauchy}^{+}(0, 3)
\end{aligned}
$$

```{r}
#| label: model-hierarchical-linear
#| eval: false
#| code-summary: "benchmark.R: model-hierarchical-linear"
```

#### eight_schools

The effect of coaching in eight schools, $y_j$, each measured with a known standard error, $\sigma_j$. The scale of the school effects, $\sigma_\eta$, is itself a parameter, which makes the posterior a funnel. `{r} parameter_count("eight_schools")` parameters.

$$
\begin{aligned}
y_j &\sim \text{Normal}(\theta_j, \sigma_j), \quad j = 1, \dots, 8 \\
\theta_j &= \mu + \xi \, \eta_j \\
\eta_j &\sim \text{Normal}(0, \sigma_\eta) \\
\sigma_\eta &\sim \text{InverseGamma}(1, 1) \\
\mu &\sim \text{Normal}(0, 100) \\
\xi &\sim \text{Normal}(0, 5)
\end{aligned}
$$

```{r}
#| label: model-eight-schools
#| eval: false
#| code-summary: "benchmark.R: model-eight-schools"
```

:::

### The scripts

`run.R` installs each version and, in each session, runs `benchmark.R` for each version in a fresh R session of its own; `benchmark.R` is everything each version runs.

```{r}
#| label: run-script
#| file: run.R
#| eval: false
#| code-summary: "run.R"
```

```{r}
#| label: benchmark-script
#| file: benchmark.R
#| eval: false
#| code-summary: "benchmark.R"
```

## Session details

What each version reported about itself in session 1, from `benchmark.R`'s `{r} section_link("benchmark.R", "benchmark-provenance")` section.

```{r}
#| label: session-one
# a saved session_info() prints with sessioninfo's print method, which needs
# sessioninfo's namespace loaded
invisible(loadNamespace("sessioninfo"))

session_one <- runs |>
  filter(session == 1) |>
  arrange(version)
```

::: panel-tabset

### CRAN

<details><summary>`greta_sitrep()`</summary>

```{r}
#| label: sitrep-cran
writeLines(session_one$results[[which(session_one$version == "CRAN")]]$greta_sitrep_report)
```

</details>

<details><summary>`sessioninfo::session_info()`</summary>

```{r}
#| label: session-info-cran
session_one$results[[which(session_one$version == "CRAN")]]$session_info
```

</details>

### main

<details><summary>`greta_sitrep()`</summary>

```{r}
#| label: sitrep-main
writeLines(session_one$results[[which(session_one$version == "main")]]$greta_sitrep_report)
```

</details>

<details><summary>`sessioninfo::session_info()`</summary>

```{r}
#| label: session-info-main
session_one$results[[which(session_one$version == "main")]]$session_info
```

</details>

### #833

<details><summary>`greta_sitrep()`</summary>

```{r}
#| label: sitrep-833
writeLines(session_one$results[[which(session_one$version == "#833")]]$greta_sitrep_report)
```

</details>

<details><summary>`sessioninfo::session_info()`</summary>

```{r}
#| label: session-info-833
session_one$results[[which(session_one$version == "#833")]]$session_info
```

</details>

:::

## References

::: {#refs}
:::
