---
title: "opt(hessian = TRUE) on main and greta#843"
format:
  gfm: default
  html: default
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(here)
library(fs)
library(dplyr)
library(tidyr)
library(knitr)

run_dir <- here("2026-09-29-hessian-timing-i546")
results_rds <- path(run_dir, "results.rds")
if (!file_exists(results_rds)) {
  stop(
    "no results.rds - run 2026-09-29-hessian-timing-i546/01-measure.R first",
    call. = FALSE
  )
}
raw <- readRDS(results_rds)
results <- raw$results
reference <- raw$branches$reference
under_test <- raw$branches$under_test

secs <- function(x) formatC(x, format = "f", digits = 2)
shape_label <- function(kind, n) {
  ifelse(kind == "scalars", paste(n, "scalar targets"), paste("one target of", n))
}

# median and range per cell: the range is the noise floor a difference has to
# clear before it counts (AGENTS.md: read resolved/unresolved before ratios)
summarised <- results |>
  mutate(shape = shape_label(kind, n)) |>
  group_by(branch, shape, kind, n, route) |>
  summarise(
    reps = n(),
    first_median = median(first),
    first_min = min(first),
    first_max = max(first),
    second_median = median(second),
    .groups = "drop"
  )
```

## The question

greta#843 changes how `opt(hessian = TRUE)` takes hessians: one graph build for all targets rather than one per target, and TensorFlow's `pfor` jacobian only for targets of at least 100 elements, with a while loop below that. How long does `opt(hessian = TRUE)` take on `main` and on the branch, and does the threshold of 100 sit where the two jacobian routes cross?

Each replicate is a fresh R process that builds the model and times `opt(hessian = TRUE)` twice: the first call, which includes TensorFlow tracing and is what a user waits for, and a second, once traced. `r raw$design$n_reps` replicates per cell.

## main against the branch

```{r}
#| label: branches
default <- summarised |> filter(route == "default")

comparison <- default |>
  select(branch, shape, kind, n, first_median, first_min, first_max) |>
  pivot_wider(
    names_from = branch,
    values_from = c(first_median, first_min, first_max)
  ) |>
  mutate(
    ref_median = .data[[paste0("first_median_", reference)]],
    test_median = .data[[paste0("first_median_", under_test)]],
    ref_min = .data[[paste0("first_min_", reference)]],
    ref_max = .data[[paste0("first_max_", reference)]],
    test_min = .data[[paste0("first_min_", under_test)]],
    test_max = .data[[paste0("first_max_", under_test)]],
    # resolved only when the two ranges do not overlap
    resolved = test_max < ref_min | test_min > ref_max,
    speedup = ref_median / test_median
  ) |>
  arrange(kind, n)

kable(
  comparison |>
    transmute(
      shape,
      !!reference := paste0(secs(ref_median), " (", secs(ref_min), "-", secs(ref_max), ")"),
      !!under_test := paste0(secs(test_median), " (", secs(test_min), "-", secs(test_max), ")"),
      resolved,
      speedup = ifelse(resolved, paste0(formatC(speedup, format = "f", digits = 1), "x"), "not resolved")
    ),
  caption = "First call to opt(hessian = TRUE), seconds: median (range)"
)
```

```{r}
#| label: headline
twenty <- comparison |> filter(kind == "scalars", n == 20)
```

With 20 scalar targets, the case #546 reported, the first call takes `r secs(twenty$ref_median)` s on `r reference` and `r secs(twenty$test_median)` s on the branch (medians, `r raw$design$n_reps` replicates each; ranges `r secs(twenty$ref_min)`-`r secs(twenty$ref_max)` and `r secs(twenty$test_min)`-`r secs(twenty$test_max)`).

## pfor against the while loop, on the branch

```{r}
#| label: routes
routes <- summarised |>
  filter(branch == under_test, route %in% c("pfor", "while")) |>
  select(shape, kind, n, route, first_median, first_min, first_max) |>
  pivot_wider(names_from = route, values_from = c(first_median, first_min, first_max)) |>
  mutate(
    faster = ifelse(first_median_while < first_median_pfor, "while loop", "pfor"),
    resolved = first_max_while < first_min_pfor | first_min_while > first_max_pfor
  ) |>
  arrange(kind, n)

kable(
  routes |>
    transmute(
      shape,
      pfor = paste0(secs(first_median_pfor), " (", secs(first_min_pfor), "-", secs(first_max_pfor), ")"),
      `while loop` = paste0(secs(first_median_while), " (", secs(first_min_while), "-", secs(first_max_while), ")"),
      faster,
      resolved
    ),
  caption = "First call, seconds: median (range), each route forced"
)
```

greta uses `pfor` for targets of at least 100 elements. The table shows which route is faster at each shape measured; the threshold is right if the while loop wins below 100 and `pfor` at or above it.

## Environment

```{r}
#| label: environment
#| results: asis
cat(
  paste0(
    "- `", names(raw$shas), "` at [", substr(raw$shas, 1, 8),
    "](https://github.com/greta-dev/greta/commit/", raw$shas, ")"
  ),
  sep = "\n"
)
cat("\n")
source(here("R", "provenance.R"))
cat(format_provenance(raw$provenance), sep = "\n")
```
