---
title: "Normal or t cut-off for batch-means scores"
format:
  gfm: default
  html: default
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(here)
library(fs)

run_dir <- here("2026-09-28-batch-means-t-threshold")
results_rds <- path(run_dir, "results.rds")
if (!file_exists(results_rds)) {
  stop(
    "no results.rds - run 2026-09-28-batch-means-t-threshold/01-measure.R first",
    call. = FALSE
  )
}
raw <- readRDS(results_rds)
results <- raw$results

# the per-score rate the bivariate test is set for: 1% across five two-sided
# scores
alpha <- 0.01 / (2 * 5)
target <- 2 * alpha

rates <- do.call(
  rbind,
  lapply(split(results, results$n), function(d) {
    df <- d$n_batches[1] - 1
    data.frame(
      n = d$n[1],
      batches = d$n_batches[1],
      reps = nrow(d),
      normal = mean(d$score > qnorm(1 - alpha)),
      t = mean(d$score > qt(1 - alpha, df))
    )
  })
)
# binomial standard error of a rate near the target, at this many replicates
mc_se <- sqrt(target * (1 - target) / rates$reps[1])

fmt <- function(x) formatC(x, format = "f", digits = 4)
ratio <- function(x) formatC(x / target, format = "f", digits = 2)
```

## The question

The bivariate normal test scores each posterior summary as
`|estimate - truth| / MCSE`, with the MCSE from batch means, and fails a
sampler when any of five scores passes `qnorm(1 - 0.01 / 10)`. A standard
error estimated from `a` batch means makes the score t-distributed with `a - 1`
degrees of freedom. Does the normal cut-off reject more often than the
`r fmt(target)` per score it is set for, and does a t cut-off fix it?

Chains are AR(1) with coefficient `r raw$ar` and mean zero,
`r rates$reps[1]` replicates at each length.

## Results

| n | batches | target rate | normal cut-off | t cut-off |
|---|---|---|---|---|
`r paste0("| ", rates$n, " | ", rates$batches, " | ", fmt(target), " | ", fmt(rates$normal), " (", ratio(rates$normal), "x) | ", fmt(rates$t), " (", ratio(rates$t), "x) |", collapse = "\n")`

Each rate has a Monte Carlo standard error of about `r fmt(mc_se)`. The normal
cut-off rejects `r ratio(min(rates$normal))`-`r ratio(max(rates$normal))` times
as often as intended; the t cut-off is within
`r formatC(max(abs(rates$t - target)) / mc_se, format = "f", digits = 1)`
standard errors of the target.

## Environment

`mcse()` from greta at `r raw$greta_sha`.

```{r}
#| label: environment
#| results: asis
source(here("R", "provenance.R"))
cat(format_provenance(raw$provenance), sep = "\n")
```
