---
title: "Retracing on CRAN, main and greta#843"
description: "Building models, opt(), opt(hessian = TRUE) and mcmc() on greta 0.6.0, main and greta#843: how long each takes, how many times each traces its TensorFlow functions, and the posteriors mcmc() returns."
date: 2026-10-05
bibliography: ../../references.bib
resources:
  - run.R
  - benchmark.R
---

[greta#843](https://github.com/greta-dev/greta/pull/843) traces greta's log-density and trace-values functions once per model, whatever the batch size, and takes all of `opt(hessian = TRUE)`'s hessians in one graph build. This post compares it with the version on CRAN and with main.

Two scripts made every result here. [`benchmark.R`](benchmark.R) is what each version runs, in a fresh R session of its own, and [`run.R`](run.R) runs it for each version in each session. Each section below shows the part of a script it is about, read out of the script by its `# ---- label ----` heading, so the code shown is the code that ran. Where a section prints the model it times, that line is run here, from the model's own section. Both scripts are in full at the end.

```{r}
#| label: setup
#| message: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(bench)

# each `# ---- label ----` section of the scripts, shown below by its label
knitr::read_chunk("benchmark.R")
knitr::read_chunk("run.R")

# white panels, so the grey band in Agreement shows
theme_set(theme_bw())

sessions <- list.files(
  "results",
  pattern = "^session-[0-9]+[.]rds$",
  full.names = TRUE
) |>
  lapply(readRDS)

versions <- sessions[[1]]$versions
version_levels <- names(versions)
version_colours <- setNames(
  c("#E69F00", "#56B4E9", "#009E73")[seq_along(version_levels)],
  version_levels
)
bench_settings <- sessions[[1]]$results[[1]]$bench_settings
n_sessions <- length(sessions)

# one row per session and version, holding that version's results
runs <- lapply(sessions, function(session_file) {
  tibble(
    session = session_file$session,
    version = factor(names(session_file$results), levels = version_levels),
    position = match(names(session_file$results), session_file$order),
    results = unname(session_file$results)
  )
}) |>
  bind_rows()

model_levels <- names(runs$results[[1]]$build)

# a part of the results that is a table or a value per run, stacked, with the
# run it came from
stack_tables <- function(part) {
  pieces <- lapply(seq_len(nrow(runs)), function(i) {
    values <- runs$results[[i]][[part]]
    if (!is.data.frame(values)) {
      values <- tibble(value = values)
    }
    tibble(session = runs$session[i], version = runs$version[i], values)
  })
  bind_rows(pieces)
}

# a part of the results that is a bench_mark per model, or one per run,
# stacked into one bench_mark, with the run and model each came from
stack_marks <- function(part) {
  pieces <- list()
  for (i in seq_len(nrow(runs))) {
    marks <- runs$results[[i]][[part]]
    if (inherits(marks, "bench_mark")) {
      marks <- list(marks)
    }
    for (j in seq_along(marks)) {
      mark <- marks[[j]]
      mark$session <- runs$session[i]
      mark$version <- runs$version[i]
      mark$model <- if (is.null(names(marks))) NA_character_ else names(marks)[j]
      pieces[[length(pieces) + 1]] <- mark
    }
  }
  vctrs::vec_rbind(!!!pieces)
}

provenance <- stack_tables("provenance")
sampler_iterations_per_draw <- stack_tables("sampler_iterations_per_draw")
build_marks <- stack_marks("build")
opt_marks <- stack_marks("opt")
hessian_marks <- stack_marks("hessian")
hessian_retracing_warnings <- stack_tables("hessian_retracing_warnings")
mcmc_marks <- stack_marks("mcmc")
mcmc_tracing <- stack_tables("mcmc_tracing")

# one row per timed call: bench keeps every call's time in the `time` column
every_call <- function(marks) {
  marks |>
    as_tibble() |>
    select(session, version, model, time) |>
    mutate(call = lapply(time, seq_along)) |>
    unnest(c(time, call)) |>
    mutate(
      seconds = as.numeric(time),
      model = factor(model, levels = model_levels),
      .keep = "unused"
    )
}

# The first call on a model traces its TensorFlow functions; later calls on the
# same model reuse them.
label_calls <- function(call) {
  last <- max(call)
  later <- if (last == 2) {
    "second call on the same model"
  } else if (last == 3) {
    "calls 2 and 3 on the same model"
  } else {
    paste0("calls 2 to ", last, " on the same model")
  }
  factor(
    ifelse(call == 1, "first call on a new model", later),
    levels = c("first call on a new model", later)
  )
}

# three significant figures each, where format() would pad a vector to the
# decimal places of its smallest value
format_number <- function(x) {
  formatC(signif(x, 3), format = "fg", digits = 3, big.mark = ",")
}

# "CRAN 10.6, main 11.2, #843 0.643"
versions_text <- function(version, value) {
  paste(paste(version, format_number(value)), collapse = ", ")
}

# "CRAN, main and #843"
paste_and <- function(x) {
  if (length(x) <= 1) {
    return(x)
  }
  paste(paste(x[-length(x)], collapse = ", "), "and", x[length(x)])
}

version_scale <- scale_colour_manual(values = version_colours, guide = "none")
jitter <- position_jitter(width = 0, height = 0.2, seed = 2026)
call_caption <- "Each small dot is one call, from every session. The large faint dot behind them is their mean."

# a searchable table, its numbers rounded for reading
dt_table <- function(data, digits = 3) {
  table <- DT::datatable(
    data,
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 10, autoWidth = TRUE)
  )
  decimal_columns <- names(data)[vapply(data, is.double, logical(1))]
  if (length(decimal_columns) == 0) {
    return(table)
  }
  DT::formatRound(table, decimal_columns, digits)
}

# a model's draws array is draws by chains by parameters
parameter_count <- function(name) {
  dim(runs$results[[1]]$mcmc_draws[[name]])[3]
}
```

## What was compared

```{r}
#| label: versions
provenance |>
  distinct(version, greta_version, greta_sha, tensorflow_version, tfp_version, r_version) |>
  arrange(version) |>
  mutate(
    installed_from = versions[as.character(version)],
    # pak records a CRAN install's version in place of a commit, so only a
    # GitHub install gets a link to its commit
    installed_from = ifelse(
      grepl("^[0-9a-f]{40}$", greta_sha),
      paste0(
        "[",
        sub("@.*", "", installed_from),
        "@",
        substr(greta_sha, 1, 8),
        "](https://github.com/greta-dev/greta/commit/",
        greta_sha,
        ")"
      ),
      installed_from
    )
  ) |>
  select(
    version,
    `installed from` = installed_from,
    greta = greta_version,
    TensorFlow = tensorflow_version,
    TFP = tfp_version,
    R = r_version
  ) |>
  knitr::kable()

python_stacks <- provenance |>
  distinct(version, tensorflow_version, tfp_version) |>
  group_by(tensorflow_version, tfp_version) |>
  summarise(
    versions = paste_and(as.character(sort(version))),
    .groups = "drop"
  )
python_stack_text <- paste0(
  python_stacks$versions,
  " ran on TensorFlow ",
  python_stacks$tensorflow_version,
  " and TensorFlow Probability ",
  python_stacks$tfp_version,
  collapse = "; "
)
```

`{r} python_stack_text`. Every version's `greta_sitrep()` report and `sessioninfo::session_info()` are in [Session details](#session-details).

The versions, pinned to commits, from `run.R`'s `run-versions` section:

```{r}
#| label: run-versions
#| eval: false
#| code-fold: false
```

Each version ran in `{r} n_sessions` sessions, on `{r} unique(provenance$machine)` with `{r} unique(provenance$cores)` cores. In each session, `run.R`'s `run-sessions` section ran `benchmark.R` once per version, each in a fresh R session, with the versions in a new random order:

```{r}
#| label: run-sessions
#| eval: false
#| code-fold: false
```

<details><summary>Table: the order of each session</summary>

```{r}
#| label: table-order
runs |>
  select(session, position, version) |>
  arrange(session, position) |>
  dt_table()
```

</details>

Each version's session ran `benchmark.R` from top to bottom. Its seed and settings, from the `benchmark-setup` and `benchmark-settings` sections:

```{r}
#| label: benchmark-setup
#| eval: false
#| code-fold: false
```

```{r}
#| label: benchmark-settings
#| eval: false
#| code-fold: false
```

## The models

Each model is from greta's `inst/examples/`, and is defined in maths and by its section of `benchmark.R`. Normal and log-normal distributions are written with their mean and standard deviation, as greta's `normal()` and `lognormal()` take them; $\text{Cauchy}^{+}$ is a Cauchy truncated to positive values. Defining a model's function needs no greta, so these sections are run here, which lets the timing sections below print each model.

::: panel-tabset

### linear

A regression of each department's rating in `attitude` on its number of complaints. `{r} parameter_count("linear")` parameters.

$$
\begin{aligned}
\text{rating}_i &\sim \text{Normal}(\mu_i, \sigma), \quad i = 1, \dots, 30 \\
\mu_i &= \alpha + \beta \, \text{complaints}_i \\
\alpha, \beta &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Cauchy}^{+}(0, 3)
\end{aligned}
$$

```{r}
#| label: model-linear
#| code-fold: false
```

### multiple_linear

The same rating on all six of `attitude`'s other columns, $X$, through a matrix multiply. `{r} parameter_count("multiple_linear")` parameters.

$$
\begin{aligned}
\text{rating}_i &\sim \text{Normal}(\mu_i, \sigma), \quad i = 1, \dots, 30 \\
\mu_i &= \alpha + \sum_{j = 1}^{6} X_{ij} \, \beta_j \\
\alpha, \beta_j &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Cauchy}^{+}(0, 3)
\end{aligned}
$$

```{r}
#| label: model-multiple-linear
#| code-fold: false
```

### hierarchical_linear

Sepal length on sepal width in `iris`, with an offset for each species, $s[i]$ being flower $i$'s. Setosa's offset is zero and the other two share a scale, $\tau$. `{r} parameter_count("hierarchical_linear")` parameters.

$$
\begin{aligned}
\text{Sepal.Length}_i &\sim \text{Normal}(\mu_i, \sigma), \quad i = 1, \dots, 150 \\
\mu_i &= \alpha + \beta \, \text{Sepal.Width}_i + \gamma_{s[i]} \\
\gamma_1 &= 0, \qquad \gamma_2, \gamma_3 \sim \text{Normal}(0, \tau) \\
\tau &\sim \text{LogNormal}(0, 1) \\
\alpha, \beta &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Cauchy}^{+}(0, 3)
\end{aligned}
$$

```{r}
#| label: model-hierarchical-linear
#| code-fold: false
```

### eight_schools

The effect of coaching in eight schools, $y_j$, each measured with a known standard error, $\sigma_j$. The scale of the school effects, $\sigma_\eta$, is itself a parameter, which makes the posterior a funnel. `{r} parameter_count("eight_schools")` parameters.

$$
\begin{aligned}
y_j &\sim \text{Normal}(\theta_j, \sigma_j), \quad j = 1, \dots, 8 \\
\theta_j &= \mu + \xi \, \eta_j \\
\eta_j &\sim \text{Normal}(0, \sigma_\eta) \\
\sigma_\eta &\sim \text{InverseGamma}(1, 1) \\
\mu &\sim \text{Normal}(0, 100) \\
\xi &\sim \text{Normal}(0, 5)
\end{aligned}
$$

```{r}
#| label: model-eight-schools
#| code-fold: false
```

### hessian model

The model greta#546 reported, used only for `opt(hessian = TRUE)`: twenty observations $y_k$, simulated, each with its own mean $b_k$, so `opt(hessian = TRUE)` takes twenty hessians. greta's `variable()` gives $b_k$ a flat prior, so it has no distribution line.

$$
y_k \sim \text{Normal}(b_k, 1), \quad k = 1, \dots, 20
$$

```{r}
#| label: model-hessian
#| code-fold: false
```

:::

## Sampler iterations per kept draw

The same `mcmc()` arguments are the same work only if each version's sampler runs the same number of iterations for each draw it keeps. `benchmark.R` measures this in its `benchmark-sampler-iterations` section:

```{r}
#| label: benchmark-sampler-iterations
#| eval: false
#| code-fold: false
```

```{r}
#| label: sampler-iterations
sampler_iterations_values <- sampler_iterations_per_draw |>
  group_by(version) |>
  summarise(values = paste(sort(unique(value)), collapse = " or "))
```

Sampler iterations per kept draw, in every session: `{r} paste(sampler_iterations_values$version, sampler_iterations_values$values, collapse = ", ")`.

<details><summary>Table: every session</summary>

```{r}
#| label: table-sampler-iterations
sampler_iterations_per_draw |>
  select(session, version, sampler_iterations_per_draw = value) |>
  dt_table()
```

</details>

## Speed

Every call below is timed with {bench}'s `mark()`, which keeps each call's time in its result's `time` column. `benchmark.R` saves each result as `mark()` made it, so `summary()` and `autoplot()` work on them. The first `opt()`, `opt(hessian = TRUE)` or `mcmc()` call on a new model traces the model's TensorFlow functions; later calls on the same model reuse them. So the plots split the first call from the later ones.

Each timing section starts by printing the model it times. This page runs only that line, which prints the model's function; knitr shows the lines it does not run here with `##` in front of them.

::: panel-tabset

### Building the model

Calling each model's function: creating its greta arrays and calling `model()`. The timed code for each model, its `bench-build-` section of `benchmark.R`, with the model it builds printed beneath:

::: panel-tabset

#### linear

```{r}
#| label: bench-build-linear
#| eval: !expr 1
#| code-fold: false
```

#### multiple_linear

```{r}
#| label: bench-build-multiple-linear
#| eval: !expr 1
#| code-fold: false
```

#### hierarchical_linear

```{r}
#| label: bench-build-hierarchical-linear
#| eval: !expr 1
#| code-fold: false
```

#### eight_schools

```{r}
#| label: bench-build-eight-schools
#| eval: !expr 1
#| code-fold: false
```

:::

```{r}
#| label: build-summaries
build <- every_call(build_marks)

build_means <- build |>
  group_by(model, version) |>
  summarise(seconds = mean(seconds), .groups = "drop")

build_medians <- build |>
  group_by(version) |>
  summarise(seconds = median(seconds))
```

Median seconds per call, over all models and sessions: `{r} versions_text(build_medians$version, build_medians$seconds)`.

```{r}
#| label: plot-build
#| fig-cap: !expr call_caption
#| fig-height: 6
ggplot(build, aes(x = seconds, y = version, colour = version)) +
  geom_point(data = build_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(~model, ncol = 1, scales = "free_x") +
  labs(x = "Seconds", y = NULL)
```

<details><summary>Table: mean seconds</summary>

```{r}
#| label: table-build-means
dt_table(build_means)
```

</details>

<details><summary>Table: every call</summary>

```{r}
#| label: table-build
dt_table(build)
```

</details>

### opt()

`opt()`, `{r} bench_settings$opt_calls` times on a model built for it. The timed code for each model, its `bench-opt-` section of `benchmark.R`, after the note in `bench-opt`:

```{r}
#| label: bench-opt
#| eval: false
#| code-fold: false
```

::: panel-tabset

#### linear

```{r}
#| label: bench-opt-linear
#| eval: !expr 1
#| code-fold: false
```

#### multiple_linear

```{r}
#| label: bench-opt-multiple-linear
#| eval: !expr 1
#| code-fold: false
```

#### hierarchical_linear

```{r}
#| label: bench-opt-hierarchical-linear
#| eval: !expr 1
#| code-fold: false
```

#### eight_schools

```{r}
#| label: bench-opt-eight-schools
#| eval: !expr 1
#| code-fold: false
```

:::

```{r}
#| label: opt-summaries
opt_timings <- every_call(opt_marks) |>
  mutate(calls = label_calls(call))

opt_means <- opt_timings |>
  group_by(model, calls, version) |>
  summarise(seconds = mean(seconds), .groups = "drop")

opt_medians <- opt_timings |>
  group_by(version) |>
  summarise(seconds = median(seconds))
```

Median seconds per call, over all models, calls and sessions: `{r} versions_text(opt_medians$version, opt_medians$seconds)`.

```{r}
#| label: plot-opt
#| fig-cap: !expr call_caption
#| fig-height: 8
ggplot(opt_timings, aes(x = seconds, y = version, colour = version)) +
  geom_point(data = opt_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(vars(model, calls), ncol = 2, scales = "free_x") +
  labs(x = "Seconds", y = NULL)
```

<details><summary>Table: mean seconds</summary>

```{r}
#| label: table-opt-means
dt_table(opt_means)
```

</details>

<details><summary>Table: every call</summary>

```{r}
#| label: table-opt
dt_table(opt_timings)
```

</details>

### opt(hessian = TRUE)

`opt(hessian = TRUE)` on the hessian model, `{r} bench_settings$hessian_calls` times on the same model. The timed code, `benchmark.R`'s `bench-hessian` section, with the model it builds printed beneath:

```{r}
#| label: bench-hessian
#| eval: !expr 1
#| code-fold: false
```

```{r}
#| label: hessian-summaries
hessian <- every_call(hessian_marks) |>
  select(-model) |>
  mutate(calls = label_calls(call))

hessian_means <- hessian |>
  group_by(calls, version) |>
  summarise(seconds = mean(seconds), .groups = "drop")

hessian_medians <- hessian |>
  group_by(version) |>
  summarise(seconds = median(seconds))
```

Median seconds per call, over all calls and sessions: `{r} versions_text(hessian_medians$version, hessian_medians$seconds)`.

```{r}
#| label: plot-hessian
#| fig-cap: !expr call_caption
#| fig-height: 4
ggplot(hessian, aes(x = seconds, y = version, colour = version)) +
  geom_point(data = hessian_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(~calls, ncol = 1) +
  labs(x = "Seconds", y = NULL)
```

<details><summary>Table: mean seconds</summary>

```{r}
#| label: table-hessian-means
dt_table(hessian_means)
```

</details>

<details><summary>Table: every call</summary>

```{r}
#| label: table-hessian
dt_table(hessian)
```

</details>

### mcmc()

`mcmc()` twice on each model: first on a new model, then on the same model again, which reuses the functions the first call traced. The timed code for each model, its `bench-mcmc-` section of `benchmark.R`, after the note in `bench-mcmc`:

```{r}
#| label: bench-mcmc
#| eval: false
#| code-fold: false
```

::: panel-tabset

#### linear

```{r}
#| label: bench-mcmc-linear
#| eval: !expr 1
#| code-fold: false
```

#### multiple_linear

```{r}
#| label: bench-mcmc-multiple-linear
#| eval: !expr 1
#| code-fold: false
```

#### hierarchical_linear

```{r}
#| label: bench-mcmc-hierarchical-linear
#| eval: !expr 1
#| code-fold: false
```

#### eight_schools

```{r}
#| label: bench-mcmc-eight-schools
#| eval: !expr 1
#| code-fold: false
```

:::

```{r}
#| label: mcmc-summaries
mcmc_timings <- every_call(mcmc_marks) |>
  mutate(calls = label_calls(call))

mcmc_means <- mcmc_timings |>
  group_by(model, calls, version) |>
  summarise(seconds = mean(seconds), .groups = "drop")

mcmc_medians <- mcmc_timings |>
  group_by(version) |>
  summarise(seconds = median(seconds))
```

Median seconds per call, over all models, calls and sessions: `{r} versions_text(mcmc_medians$version, mcmc_medians$seconds)`.

```{r}
#| label: plot-mcmc
#| fig-cap: !expr call_caption
#| fig-height: 8
ggplot(mcmc_timings, aes(x = seconds, y = version, colour = version)) +
  geom_point(data = mcmc_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(vars(model, calls), ncol = 2, scales = "free_x") +
  labs(x = "Seconds", y = NULL)
```

<details><summary>Table: mean seconds</summary>

```{r}
#| label: table-mcmc-means
dt_table(mcmc_means)
```

</details>

<details><summary>Table: every call</summary>

```{r}
#| label: table-mcmc
dt_table(mcmc_timings)
```

</details>

:::

## Tracing

After both `mcmc()` calls on each model, `benchmark.R`'s `mcmc-tracing` section asks each of the model's TensorFlow functions how many times it has been traced, and keeps the retracing warnings TensorFlow logged during the two calls:

```{r}
#| label: mcmc-tracing
#| eval: false
#| code-fold: false
```

with the helpers in its `benchmark-helpers` section:

```{r}
#| label: benchmark-helpers
#| eval: false
#| code-fold: false
```

```{r}
#| label: tracing-counts
tracing <- mcmc_tracing |>
  pivot_longer(
    c(retracing_warnings, ends_with("_traces")),
    names_to = "measure",
    values_to = "count"
  ) |>
  mutate(
    model = factor(model, levels = model_levels),
    measure = factor(
      measure,
      levels = c("log_prob_traces", "trace_values_traces", "sampler_traces", "retracing_warnings"),
      labels = c("log density traces", "trace values traces", "sampler traces", "retracing warnings")
    )
  )

# whether every session gave the same count, so one bar can stand for them all
counts_per_session <- tracing |>
  group_by(version, model, measure) |>
  summarise(distinct_counts = n_distinct(count), .groups = "drop")
same_in_every_session <- all(counts_per_session$distinct_counts == 1)
```

`{r} if (same_in_every_session) paste0("Each count was the same in all ", n_sessions, " sessions, so each bar is that count.") else "The counts differ between sessions, so each dot is one session."`

```{r}
#| label: plot-tracing
#| fig-height: 5
#| warning: false
tracing_plot <- ggplot(tracing, aes(x = count, y = model, fill = version, colour = version)) +
  scale_fill_manual(values = version_colours) +
  scale_colour_manual(values = version_colours) +
  scale_x_continuous(breaks = \(limits) unique(floor(pretty(limits)))) +
  facet_wrap(~measure, ncol = 2, scales = "free_x") +
  labs(x = "Count", y = NULL, fill = NULL, colour = NULL)

if (same_in_every_session) {
  tracing_plot +
    geom_col(
      data = distinct(tracing, version, model, measure, count),
      position = position_dodge(width = 0.8),
      width = 0.75,
      colour = NA
    )
} else {
  tracing_plot +
    geom_point(position = position_dodge(width = 0.8))
}
```

<details><summary>Table: every count</summary>

```{r}
#| label: table-tracing
dt_table(tracing)
```

</details>

```{r}
#| label: hessian-warnings
hessian_warning_counts <- hessian_retracing_warnings |>
  group_by(version) |>
  summarise(counts = paste(sort(unique(value)), collapse = " or "))
```

Retracing warnings TensorFlow logged during each session's `{r} bench_settings$hessian_calls` calls of `opt(hessian = TRUE)`, counted in `bench-hessian`: `{r} paste(hessian_warning_counts$version, hessian_warning_counts$counts, collapse = ", ")`.

## Efficiency

The posterior checks from here on use every draw of every chain from each model's second `mcmc()` call, which `benchmark.R`'s `mcmc-draws` section keeps:

```{r}
#| label: mcmc-draws
#| eval: false
#| code-fold: false
```

Each model's posterior summary, in each session, from those draws [@vehtari2021]:

```{r}
#| label: posterior-summaries
posterior <- lapply(seq_len(nrow(runs)), function(i) {
  draws <- runs$results[[i]]$mcmc_draws
  lapply(names(draws), function(model_name) {
    tibble(
      session = runs$session[i],
      version = runs$version[i],
      model = model_name,
      posterior::summarise_draws(
        draws[[model_name]],
        "mean",
        "sd",
        "mcse_mean",
        "rhat",
        "ess_bulk",
        "ess_tail"
      )
    )
  }) |>
    bind_rows()
}) |>
  bind_rows() |>
  mutate(model = factor(model, levels = model_levels))
```

Bulk effective sample size (ESS) per second, for the parameter with the smallest bulk ESS and for the median parameter. The draws are the second call's, so the seconds are the second call's.

```{r}
#| label: efficiency
second_calls <- mcmc_timings |>
  filter(call == 2) |>
  select(session, version, model, seconds)

efficiency <- posterior |>
  group_by(session, version, model) |>
  summarise(
    smallest_ess_bulk = min(ess_bulk),
    median_ess_bulk = median(ess_bulk),
    .groups = "drop"
  ) |>
  left_join(second_calls, by = c("session", "version", "model")) |>
  mutate(
    smallest_ess_per_second = smallest_ess_bulk / seconds,
    median_ess_per_second = median_ess_bulk / seconds
  )

efficiency_means <- efficiency |>
  group_by(model, version) |>
  summarise(
    smallest_ess_per_second = mean(smallest_ess_per_second),
    median_ess_per_second = mean(median_ess_per_second),
    .groups = "drop"
  )

session_caption <- "Each small dot is one session's second call. The large faint dot behind them is their mean."
```

::: panel-tabset

### Smallest bulk ESS per second

```{r}
#| label: plot-efficiency-smallest
#| fig-cap: !expr session_caption
#| fig-height: 6
ggplot(efficiency, aes(x = smallest_ess_per_second, y = version, colour = version)) +
  geom_point(data = efficiency_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(~model, ncol = 1, scales = "free_x") +
  labs(x = "Smallest bulk ESS per second", y = NULL)
```

### Median bulk ESS per second

```{r}
#| label: plot-efficiency-median
#| fig-cap: !expr session_caption
#| fig-height: 6
ggplot(efficiency, aes(x = median_ess_per_second, y = version, colour = version)) +
  geom_point(data = efficiency_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(~model, ncol = 1, scales = "free_x") +
  labs(x = "Median bulk ESS per second", y = NULL)
```

:::

<details><summary>Table: mean ESS per second</summary>

```{r}
#| label: table-efficiency-means
dt_table(efficiency_means, digits = 1)
```

</details>

<details><summary>Table: every session</summary>

```{r}
#| label: table-efficiency
dt_table(efficiency, digits = 1)
```

</details>

## Convergence

For each session's second `mcmc()` call, the largest rank-normalised split-$\hat{R}$ and the smallest bulk ESS over the model's parameters [@vehtari2021].

```{r}
#| label: convergence
convergence <- posterior |>
  group_by(session, version, model) |>
  summarise(
    largest_rhat = max(rhat),
    smallest_ess_bulk = min(ess_bulk),
    .groups = "drop"
  )

convergence_means <- convergence |>
  group_by(model, version) |>
  summarise(
    largest_rhat = mean(largest_rhat),
    smallest_ess_bulk = mean(smallest_ess_bulk),
    .groups = "drop"
  )
```

::: panel-tabset

### Largest R-hat

```{r}
#| label: plot-rhat
#| fig-cap: !expr session_caption
#| fig-height: 6
ggplot(convergence, aes(x = largest_rhat, y = version, colour = version)) +
  geom_point(data = convergence_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(~model, ncol = 1, scales = "free_x") +
  labs(x = "Largest R-hat", y = NULL)
```

### Smallest bulk ESS

```{r}
#| label: plot-ess
#| fig-cap: !expr session_caption
#| fig-height: 6
ggplot(convergence, aes(x = smallest_ess_bulk, y = version, colour = version)) +
  geom_point(data = convergence_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(~model, ncol = 1, scales = "free_x") +
  labs(x = "Smallest bulk ESS", y = NULL)
```

:::

<details><summary>Table: every session</summary>

```{r}
#| label: table-convergence
dt_table(convergence)
```

</details>

## Agreement

Whether two versions estimate the same posterior means. The difference between two versions' means is divided by the Monte Carlo standard error (MCSE) of that difference, so a difference can be told apart from Monte Carlo noise [@vehtari2021; @magnusson2025]. For each parameter, each version's means and MCSEs are pooled over its sessions, and

$$
z = \frac{\bar\theta_{\text{b}} - \bar\theta_{\text{a}}}{\sqrt{\text{se}^2_{\text{a}} + \text{se}^2_{\text{b}}}}.
$$

Each session has a seed of its own, so the sessions are independent, and a version's pooled standard error is $\sqrt{\sum \text{mcse}^2} / n$ over its $n$ sessions. If both versions sample the same posterior, each $z$ is close to a standard normal draw. Two versions that both respect `set.seed()` return the same draws from the same seed, and so a $z$ of zero.

```{r}
#| label: agreement
pooled <- posterior |>
  group_by(version, model, variable) |>
  summarise(
    sessions = n(),
    mean = mean(mean),
    se = sqrt(sum(mcse_mean^2)) / n(),
    .groups = "drop"
  )

pairs <- utils::combn(version_levels, 2)
agreement <- lapply(seq_len(ncol(pairs)), function(i) {
  reference <- filter(pooled, version == pairs[1, i])
  compared <- filter(pooled, version == pairs[2, i])
  inner_join(reference, compared, by = c("model", "variable"), suffix = c("_a", "_b")) |>
    mutate(
      comparison = paste(pairs[2, i], "vs", pairs[1, i]),
      z = (mean_b - mean_a) / sqrt(se_a^2 + se_b^2)
    )
}) |>
  bind_rows() |>
  mutate(comparison = factor(comparison, levels = unique(comparison)))

# the band all of a comparison's z fall inside with probability 0.95 when both
# versions sample the same posterior: a Bonferroni bound over its parameters
parameters_compared <- n_distinct(paste(agreement$model, agreement$variable))
band <- qnorm(1 - 0.05 / (2 * parameters_compared))

outside_band <- agreement |>
  group_by(comparison) |>
  summarise(outside = sum(abs(z) > band), largest_abs_z = max(abs(z)))
```

The grey band is $\pm$ `{r} format_number(band)`: if two versions sample the same posterior, all `{r} parameters_compared` parameters' $z$ fall inside it with probability at least 0.95, by a Bonferroni bound. Parameters outside it, per comparison: `{r} paste(outside_band$comparison, outside_band$outside, collapse = ", ")`.

```{r}
#| label: plot-agreement
#| fig-cap: "Each dot is one parameter. The grey band is where every parameter's z falls, with probability at least 0.95, when both versions sample the same posterior."
#| fig-height: 6
ggplot(agreement, aes(x = z, y = model)) +
  annotate(
    "rect",
    xmin = -band,
    xmax = band,
    ymin = -Inf,
    ymax = Inf,
    fill = "grey80",
    alpha = 0.5
  ) +
  geom_point(position = jitter, alpha = 0.7) +
  facet_wrap(~comparison, ncol = 1) +
  labs(x = "z", y = NULL)
```

<details><summary>Table: largest |z| per comparison and model</summary>

```{r}
#| label: table-agreement-summary
agreement |>
  group_by(comparison, model) |>
  summarise(
    parameters = n(),
    largest_abs_z = max(abs(z)),
    parameter = variable[which.max(abs(z))],
    .groups = "drop"
  ) |>
  dt_table(digits = 2)
```

</details>

<details><summary>Table: every parameter</summary>

```{r}
#| label: table-agreement
agreement |>
  select(comparison, model, variable, mean_a, se_a, mean_b, se_b, z) |>
  dt_table()
```

</details>

## Traceplots

Every draw of each chain from session 1's second `mcmc()` call on each model, numbered from the end of warmup.

```{r}
#| label: traceplots
# every draw of every run, one row per draw and parameter
draws <- lapply(seq_len(nrow(runs)), function(i) {
  model_draws <- runs$results[[i]]$mcmc_draws
  lapply(names(model_draws), function(model_name) {
    posterior::as_draws_df(model_draws[[model_name]]) |>
      as.data.frame() |>
      pivot_longer(
        -c(.chain, .iteration, .draw),
        names_to = "variable",
        values_to = "value"
      ) |>
      mutate(
        session = runs$session[i],
        version = runs$version[i],
        model = model_name
      )
  }) |>
    bind_rows()
}) |>
  bind_rows()

gg_trace <- function(name, version_label) {
  draws |>
    filter(session == 1, model == name, version == version_label) |>
    ggplot(aes(x = .iteration, y = value, colour = factor(.chain))) +
    geom_line(linewidth = 0.2) +
    facet_wrap(~variable, scales = "free_y", ncol = 4) +
    labs(x = "Draw", y = NULL, colour = "Chain")
}

trace_height <- function(name) {
  1 + 1.6 * ceiling(parameter_count(name) / 4)
}
```

::: panel-tabset

### linear

::: panel-tabset

#### CRAN

```{r}
#| label: plot-trace-linear-cran
#| fig-height: !expr trace_height("linear")
gg_trace("linear", "CRAN")
```

#### main

```{r}
#| label: plot-trace-linear-main
#| fig-height: !expr trace_height("linear")
gg_trace("linear", "main")
```

#### #843

```{r}
#| label: plot-trace-linear-843
#| fig-height: !expr trace_height("linear")
gg_trace("linear", "#843")
```

:::

### multiple_linear

::: panel-tabset

#### CRAN

```{r}
#| label: plot-trace-multiple-linear-cran
#| fig-height: !expr trace_height("multiple_linear")
gg_trace("multiple_linear", "CRAN")
```

#### main

```{r}
#| label: plot-trace-multiple-linear-main
#| fig-height: !expr trace_height("multiple_linear")
gg_trace("multiple_linear", "main")
```

#### #843

```{r}
#| label: plot-trace-multiple-linear-843
#| fig-height: !expr trace_height("multiple_linear")
gg_trace("multiple_linear", "#843")
```

:::

### hierarchical_linear

::: panel-tabset

#### CRAN

```{r}
#| label: plot-trace-hierarchical-linear-cran
#| fig-height: !expr trace_height("hierarchical_linear")
gg_trace("hierarchical_linear", "CRAN")
```

#### main

```{r}
#| label: plot-trace-hierarchical-linear-main
#| fig-height: !expr trace_height("hierarchical_linear")
gg_trace("hierarchical_linear", "main")
```

#### #843

```{r}
#| label: plot-trace-hierarchical-linear-843
#| fig-height: !expr trace_height("hierarchical_linear")
gg_trace("hierarchical_linear", "#843")
```

:::

### eight_schools

::: panel-tabset

#### CRAN

```{r}
#| label: plot-trace-eight-schools-cran
#| fig-height: !expr trace_height("eight_schools")
gg_trace("eight_schools", "CRAN")
```

#### main

```{r}
#| label: plot-trace-eight-schools-main
#| fig-height: !expr trace_height("eight_schools")
gg_trace("eight_schools", "main")
```

#### #843

```{r}
#| label: plot-trace-eight-schools-843
#| fig-height: !expr trace_height("eight_schools")
gg_trace("eight_schools", "#843")
```

:::

:::

## Posterior densities

Every draw of every chain, from each session's second `mcmc()` call, pooled over sessions.

```{r}
#| label: densities
gg_densities <- function(name) {
  draws |>
    filter(model == name) |>
    ggplot(aes(x = value, colour = version)) +
    geom_density() +
    scale_colour_manual(values = version_colours) +
    facet_wrap(~variable, scales = "free", ncol = 4) +
    labs(x = NULL, y = NULL, colour = NULL)
}

# whether two versions returned the same draws in every session
draws_identical <- function(pair) {
  all(vapply(
    unique(runs$session),
    function(session_number) {
      first <- runs$results[[which(runs$session == session_number & runs$version == pair[1])]]
      second <- runs$results[[which(runs$session == session_number & runs$version == pair[2])]]
      identical(first$mcmc_draws, second$mcmc_draws)
    },
    logical(1)
  ))
}
identical_pairs <- Filter(
  draws_identical,
  utils::combn(version_levels, 2, simplify = FALSE)
)
identical_text <- if (length(identical_pairs) == 0) {
  ""
} else {
  paste0(
    vapply(identical_pairs, \(pair) paste(pair, collapse = "'s and "), character(1)),
    "'s draws are identical in every session, so their lines lie on top of each other.",
    collapse = " "
  )
}
```

`{r} identical_text`

::: panel-tabset

### linear

```{r}
#| label: plot-densities-linear
#| fig-height: !expr trace_height("linear")
gg_densities("linear")
```

### multiple_linear

```{r}
#| label: plot-densities-multiple-linear
#| fig-height: !expr trace_height("multiple_linear")
gg_densities("multiple_linear")
```

### hierarchical_linear

```{r}
#| label: plot-densities-hierarchical-linear
#| fig-height: !expr trace_height("hierarchical_linear")
gg_densities("hierarchical_linear")
```

### eight_schools

```{r}
#| label: plot-densities-eight-schools
#| fig-height: !expr trace_height("eight_schools")
gg_densities("eight_schools")
```

:::

## Session details

What each version reported about itself in session 1, from `benchmark.R`'s `benchmark-provenance` section:

```{r}
#| label: benchmark-provenance
#| eval: false
#| code-fold: false
```

```{r}
#| label: session-one
session_one <- runs |>
  filter(session == 1) |>
  arrange(version)
```

::: panel-tabset

### CRAN

<details><summary>`greta_sitrep()`</summary>

```{r}
#| label: sitrep-cran
writeLines(session_one$results[[which(session_one$version == "CRAN")]]$greta_sitrep_report)
```

</details>

<details><summary>`sessioninfo::session_info()`</summary>

```{r}
#| label: session-info-cran
session_one$results[[which(session_one$version == "CRAN")]]$session_info
```

</details>

### main

<details><summary>`greta_sitrep()`</summary>

```{r}
#| label: sitrep-main
writeLines(session_one$results[[which(session_one$version == "main")]]$greta_sitrep_report)
```

</details>

<details><summary>`sessioninfo::session_info()`</summary>

```{r}
#| label: session-info-main
session_one$results[[which(session_one$version == "main")]]$session_info
```

</details>

### #843

<details><summary>`greta_sitrep()`</summary>

```{r}
#| label: sitrep-843
writeLines(session_one$results[[which(session_one$version == "#843")]]$greta_sitrep_report)
```

</details>

<details><summary>`sessioninfo::session_info()`</summary>

```{r}
#| label: session-info-843
session_one$results[[which(session_one$version == "#843")]]$session_info
```

</details>

:::

## The code

### run.R

[Download `run.R`](run.R).

```{r}
#| label: run-script
#| file: run.R
#| eval: false
#| code-fold: false
```

### benchmark.R

[Download `benchmark.R`](benchmark.R).

```{r}
#| label: benchmark-script
#| file: benchmark.R
#| eval: false
#| code-fold: false
```

Both, with every session's results in `results/`, are in [greta-dev/greta.benchmarks](https://github.com/greta-dev/greta.benchmarks).

## References

::: {#refs}
:::
