---
title: "Retracing on CRAN, main and greta#843"
description: "Building models, opt(), opt(hessian = TRUE) and mcmc() on greta 0.6.0, main and greta#843: how long each takes, how many times each traces its TensorFlow functions, and the posteriors mcmc() returns."
date: 2026-10-05
bibliography: ../../references.bib
resources:
  - run.R
---

[greta#843](https://github.com/greta-dev/greta/pull/843) traces greta's log-density and trace-values functions once per model, whatever the batch size, and takes all of `opt(hessian = TRUE)`'s hessians in one graph build. This post compares it with the version on CRAN and with main. Everything measured is in [`run.R`](run.R), shown in full at the end.

```{r}
#| label: setup
#| message: false
library(dplyr)
library(tidyr)
library(ggplot2)

# white panels, so the grey band in Agreement shows
theme_set(theme_bw())

sessions <- list.files(
  "results",
  pattern = "^session-[0-9]+[.]rds$",
  full.names = TRUE
) |>
  lapply(readRDS)

versions <- sessions[[1]]$versions
version_levels <- names(versions)
version_colours <- setNames(
  c("#E69F00", "#56B4E9", "#009E73")[seq_along(version_levels)],
  version_levels
)
settings <- sessions[[1]]$results[[1]]$settings
n_sessions <- length(sessions)

# one data frame of `part` from every version in every session
collect <- function(part) {
  rows <- list()
  for (session in sessions) {
    for (version in names(session$results)) {
      values <- session$results[[version]][[part]]
      if (!is.data.frame(values)) {
        values <- data.frame(value = values)
      }
      rows[[length(rows) + 1]] <- data.frame(
        session = session$session,
        version = version,
        position = match(version, session$order),
        values,
        row.names = NULL
      )
    }
  }
  bind_rows(rows) |>
    mutate(version = factor(version, levels = version_levels))
}

# The first call on a model traces its TensorFlow functions; later calls on the
# same model reuse them.
label_calls <- function(call) {
  last <- max(call)
  later <- if (last == 2) {
    "second call on the same model"
  } else if (last == 3) {
    "calls 2 and 3 on the same model"
  } else {
    paste0("calls 2 to ", last, " on the same model")
  }
  factor(
    ifelse(call == 1, "first call on a new model", later),
    levels = c("first call on a new model", later)
  )
}

provenance <- collect("provenance")
sampler_iterations_per_draw <- collect("sampler_iterations_per_draw")
build <- collect("build")
opt_timings <- collect("opt") |> mutate(calls = label_calls(call))
hessian <- collect("hessian") |> mutate(calls = label_calls(call))
mcmc_calls <- collect("mcmc") |> mutate(calls = label_calls(call))
posterior <- collect("posterior") |> mutate(calls = label_calls(call))
draws <- collect("draws")

model_levels <- unique(build$model)
tidy_models <- function(data) {
  mutate(data, model = factor(model, levels = model_levels))
}
build <- tidy_models(build)
opt_timings <- tidy_models(opt_timings)
mcmc_calls <- tidy_models(mcmc_calls)
posterior <- tidy_models(posterior)
draws <- tidy_models(draws)

# three significant figures each, where format() would pad a vector to the
# decimal places of its smallest value
format_number <- function(x) {
  formatC(signif(x, 3), format = "fg", digits = 3, big.mark = ",")
}

# "CRAN 10.6, main 11.2, #843 0.643"
versions_text <- function(version, value) {
  paste(paste(version, format_number(value)), collapse = ", ")
}

# "CRAN, main and #843"
paste_and <- function(x) {
  if (length(x) <= 1) {
    return(x)
  }
  paste(paste(x[-length(x)], collapse = ", "), "and", x[length(x)])
}

version_scale <- scale_colour_manual(values = version_colours, guide = "none")
jitter <- position_jitter(width = 0, height = 0.2, seed = 2026)
call_caption <- "Each small dot is one call, from every session. The large faint dot behind them is their mean."

# a searchable table, its numbers rounded for reading
dt_table <- function(data, digits = 3) {
  table <- DT::datatable(
    data,
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 10, autoWidth = TRUE)
  )
  decimal_columns <- names(data)[vapply(data, is.double, logical(1))]
  if (length(decimal_columns) == 0) {
    return(table)
  }
  DT::formatRound(table, decimal_columns, digits)
}

# code shown on this page is read out of run.R, so it is the code that ran
script <- parse("run.R", keep.source = TRUE)

# the first call in run.R, depth first, for which `matches()` is TRUE
find_in_script <- function(matches) {
  search <- function(expr) {
    if (!is.call(expr)) {
      return(NULL)
    }
    if (matches(expr)) {
      return(expr)
    }
    for (i in seq_along(expr)) {
      if (is.call(expr[[i]])) {
        found <- search(expr[[i]])
        if (!is.null(found)) {
          return(found)
        }
      }
    }
    NULL
  }
  for (expr in script) {
    found <- search(expr)
    if (!is.null(found)) {
      return(found)
    }
  }
  NULL
}

assigned_to <- function(name) {
  find_in_script(
    \(call) identical(call[[1]], as.name("<-")) && identical(call[[2]], as.name(name))
  )[[3]]
}

# a parsed function() call keeps its source as its fourth element
function_body_source <- function(function_call) {
  lines <- as.character(function_call[[4]])
  body <- lines[-c(1, length(lines))]
  code_lines <- body[nzchar(trimws(body))]
  indent <- min(nchar(sub("\\S.*", "", code_lines)))
  substring(body, indent + 1)
}

mcmc_call <- find_in_script(
  \(call) {
    identical(call[[1]], as.name("mcmc")) &&
      length(call) > 1 &&
      identical(call[[2]], as.name("m"))
  }
) |>
  deparse(width.cutoff = 50)
```

## What was compared

```{r}
#| label: versions
provenance |>
  distinct(version, greta_version, greta_sha, tensorflow_version, tfp_version, r_version) |>
  arrange(version) |>
  mutate(
    installed_from = versions[as.character(version)],
    # pak records a CRAN install's version in place of a commit, so only a
    # GitHub install gets a link to its commit
    installed_from = ifelse(
      grepl("^[0-9a-f]{40}$", greta_sha),
      paste0(
        "[",
        sub("@.*", "", installed_from),
        "@",
        substr(greta_sha, 1, 8),
        "](https://github.com/greta-dev/greta/commit/",
        greta_sha,
        ")"
      ),
      installed_from
    )
  ) |>
  select(
    version,
    `installed from` = installed_from,
    greta = greta_version,
    TensorFlow = tensorflow_version,
    TFP = tfp_version,
    R = r_version
  ) |>
  knitr::kable()

python_stacks <- provenance |>
  distinct(version, tensorflow_version, tfp_version) |>
  group_by(tensorflow_version, tfp_version) |>
  summarise(
    versions = paste_and(as.character(sort(version))),
    .groups = "drop"
  )
python_stack_text <- paste0(
  python_stacks$versions,
  " ran on TensorFlow ",
  python_stacks$tensorflow_version,
  " and TensorFlow Probability ",
  python_stacks$tfp_version,
  collapse = "; "
)
```

`{r} python_stack_text`.

Each version ran in `{r} n_sessions` sessions, each a fresh R session per version, on `{r} unique(provenance$machine)` with `{r} unique(provenance$cores)` cores, with the versions in a new random order each session. Within a session, each version's process ran every measurement below in the order of [The code](#the-code), the models in the order listed.

<details><summary>Table: the order of each session</summary>

```{r}
#| label: table-order
sessions |>
  lapply(\(session) {
    data.frame(
      session = session$session,
      position = seq_along(session$order),
      version = session$order
    )
  }) |>
  bind_rows() |>
  dt_table()
```

</details>

Every `mcmc()` call, on every version, was this call from `run.R`, with `warmup = {r} settings$warmup` and `n_samples = {r} settings$n_samples` draws, `chains = {r} settings$chains` and `n_cores = {r} settings$n_cores`:

```{r}
#| label: mcmc-call
#| code: !expr mcmc_call
#| eval: false
#| code-fold: false
```

<details><summary>Table: every setting in `run.R`</summary>

```{r}
#| label: table-settings
tibble::enframe(unlist(settings), name = "setting", value = "value") |>
  dt_table()
```

</details>

## The models

Each model is from greta's `inst/examples/`, and is defined in maths and in the greta code `run.R` builds it with; the code below is read out of `run.R`. Normal and log-normal distributions are written with their mean and standard deviation, as greta's `normal()` and `lognormal()` take them; $\text{Cauchy}^{+}$ is a Cauchy truncated to positive values.

```{r}
#| label: model-code
model_code <- lapply(as.list(assigned_to("models"))[-1], function_body_source)
hessian_code <- function_body_source(assigned_to("hessian_model"))

parameter_count <- function(name) {
  posterior |>
    filter(model == name) |>
    distinct(variable) |>
    nrow()
}
```

::: panel-tabset

### linear

A regression of each department's rating in `attitude` on its number of complaints. `{r} parameter_count("linear")` parameters.

$$
\begin{aligned}
\text{rating}_i &\sim \text{Normal}(\mu_i, \sigma), \quad i = 1, \dots, 30 \\
\mu_i &= \alpha + \beta \, \text{complaints}_i \\
\alpha, \beta &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Cauchy}^{+}(0, 3)
\end{aligned}
$$

```{r}
#| label: code-linear
#| code: !expr model_code$linear
#| eval: false
#| code-fold: false
```

### multiple_linear

The same rating on all six of `attitude`'s other columns, $X$, through a matrix multiply. `{r} parameter_count("multiple_linear")` parameters.

$$
\begin{aligned}
\text{rating}_i &\sim \text{Normal}(\mu_i, \sigma), \quad i = 1, \dots, 30 \\
\mu_i &= \alpha + \sum_{j = 1}^{6} X_{ij} \, \beta_j \\
\alpha, \beta_j &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Cauchy}^{+}(0, 3)
\end{aligned}
$$

```{r}
#| label: code-multiple-linear
#| code: !expr model_code$multiple_linear
#| eval: false
#| code-fold: false
```

### hierarchical_linear

Sepal length on sepal width in `iris`, with an offset for each species, $s[i]$ being flower $i$'s. Setosa's offset is zero and the other two share a scale, $\tau$. `{r} parameter_count("hierarchical_linear")` parameters.

$$
\begin{aligned}
\text{Sepal.Length}_i &\sim \text{Normal}(\mu_i, \sigma), \quad i = 1, \dots, 150 \\
\mu_i &= \alpha + \beta \, \text{Sepal.Width}_i + \gamma_{s[i]} \\
\gamma_1 &= 0, \qquad \gamma_2, \gamma_3 \sim \text{Normal}(0, \tau) \\
\tau &\sim \text{LogNormal}(0, 1) \\
\alpha, \beta &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Cauchy}^{+}(0, 3)
\end{aligned}
$$

```{r}
#| label: code-hierarchical-linear
#| code: !expr model_code$hierarchical_linear
#| eval: false
#| code-fold: false
```

### eight_schools

The effect of coaching in eight schools, $y_j$, each measured with a known standard error, $\sigma_j$. The scale of the school effects, $\sigma_\eta$, is itself a parameter, which makes the posterior a funnel. `{r} parameter_count("eight_schools")` parameters.

$$
\begin{aligned}
y_j &\sim \text{Normal}(\theta_j, \sigma_j), \quad j = 1, \dots, 8 \\
\theta_j &= \mu + \xi \, \eta_j \\
\eta_j &\sim \text{Normal}(0, \sigma_\eta) \\
\sigma_\eta &\sim \text{InverseGamma}(1, 1) \\
\mu &\sim \text{Normal}(0, 100) \\
\xi &\sim \text{Normal}(0, 5)
\end{aligned}
$$

```{r}
#| label: code-eight-schools
#| code: !expr model_code$eight_schools
#| eval: false
#| code-fold: false
```

### cjs

A Cormack-Jolly-Seber capture-recapture model: survival, $\phi_t$, and recapture, $p_t$, on each of 20 occasions, from the simulated capture histories $y_{it}$ of 100 animals. Animal $i$ is first caught on occasion $f_i$ and last on $l_i$, so it is known to be alive on every occasion between. $\chi_t$ is the probability that an animal alive on occasion $t$ is never seen again, built by a recursion over the 19 earlier occasions. `{r} parameter_count("cjs")` parameters.

$$
\begin{aligned}
\phi_t, \, p_t &\sim \text{Beta}(1, 1), \quad t = 1, \dots, 20 \\
\chi_{20} &= 1, \qquad \chi_t = (1 - \phi_t) + \phi_t \, (1 - p_{t + 1}) \, \chi_{t + 1} \\
1 &\sim \text{Bernoulli}(\phi_{t - 1}), \quad f_i < t \le l_i \\
y_{it} &\sim \text{Bernoulli}(p_t), \quad f_i < t \le l_i \\
1 &\sim \text{Bernoulli}(\chi_{l_i}), \quad \text{if } l_i < 20
\end{aligned}
$$

```{r}
#| label: code-cjs
#| code: !expr model_code$cjs
#| eval: false
#| code-fold: false
```

### hessian model

The model greta#546 reported, used only for `opt(hessian = TRUE)`: twenty observations $y_k$, simulated, each with its own mean $b_k$, so `opt(hessian = TRUE)` takes twenty hessians. greta's `variable()` gives $b_k$ a flat prior, so it has no distribution line.

$$
y_k \sim \text{Normal}(b_k, 1), \quad k = 1, \dots, 20
$$

```{r}
#| label: code-hessian-model
#| code: !expr hessian_code
#| eval: false
#| code-fold: false
```

:::

## Sampler iterations per kept draw

The same `mcmc()` arguments are the same work only if each version's sampler runs the same number of iterations for each draw it keeps. `run.R` measures this in every session, from a random walk that accepts every step: the variance between kept draws, over one step's variance, counts the sampler iterations between them.

```{r}
#| label: sampler-iterations
sampler_iterations_values <- sampler_iterations_per_draw |>
  group_by(version) |>
  summarise(values = paste(sort(unique(value)), collapse = " or "))
```

Sampler iterations per kept draw, in every session: `{r} paste(sampler_iterations_values$version, sampler_iterations_values$values, collapse = ", ")`.

<details><summary>Table: every session</summary>

```{r}
#| label: table-sampler-iterations
sampler_iterations_per_draw |>
  select(session, version, sampler_iterations_per_draw = value) |>
  dt_table()
```

</details>

## Speed

Every call below is timed with {bench}. The first `opt()`, `opt(hessian = TRUE)` or `mcmc()` call on a new model traces the model's TensorFlow functions; later calls on the same model reuse them. So the plots split the first call from the later ones.

::: panel-tabset

### Building the model

Calling the model's function in [The models](#the-models): creating its greta arrays and calling `model()`, `{r} settings$build_calls` times per session.

```{r}
#| label: build-summaries
build_means <- build |>
  group_by(model, version) |>
  summarise(seconds = mean(seconds), .groups = "drop")

build_medians <- build |>
  group_by(version) |>
  summarise(seconds = median(seconds))
```

Median seconds per call, over all models and sessions: `{r} versions_text(build_medians$version, build_medians$seconds)`.

```{r}
#| label: plot-build
#| fig-cap: !expr call_caption
#| fig-height: 7
ggplot(build, aes(x = seconds, y = version, colour = version)) +
  geom_point(data = build_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(~model, ncol = 1, scales = "free_x") +
  labs(x = "Seconds", y = NULL)
```

<details><summary>Table: mean seconds</summary>

```{r}
#| label: table-build-means
dt_table(build_means)
```

</details>

<details><summary>Table: every call</summary>

```{r}
#| label: table-build
build |>
  select(session, version, model, call, seconds) |>
  dt_table()
```

</details>

### opt()

`opt(m)`, `{r} settings$opt_calls` times per session on a model built for it.

```{r}
#| label: opt-summaries
opt_means <- opt_timings |>
  group_by(model, calls, version) |>
  summarise(seconds = mean(seconds), .groups = "drop")

opt_medians <- opt_timings |>
  group_by(version) |>
  summarise(seconds = median(seconds))
```

Median seconds per call, over all models, calls and sessions: `{r} versions_text(opt_medians$version, opt_medians$seconds)`.

```{r}
#| label: plot-opt
#| fig-cap: !expr call_caption
#| fig-height: 9
ggplot(opt_timings, aes(x = seconds, y = version, colour = version)) +
  geom_point(data = opt_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(vars(model, calls), ncol = 2, scales = "free_x") +
  labs(x = "Seconds", y = NULL)
```

<details><summary>Table: mean seconds</summary>

```{r}
#| label: table-opt-means
dt_table(opt_means)
```

</details>

<details><summary>Table: every call</summary>

```{r}
#| label: table-opt
opt_timings |>
  select(session, version, model, call, seconds) |>
  dt_table()
```

</details>

### opt(hessian = TRUE)

`opt(m, hessian = TRUE)` on the hessian model, `{r} settings$hessian_calls` times per session on the same model.

```{r}
#| label: hessian-summaries
hessian_means <- hessian |>
  group_by(calls, version) |>
  summarise(seconds = mean(seconds), .groups = "drop")

hessian_medians <- hessian |>
  group_by(version) |>
  summarise(seconds = median(seconds))
```

Median seconds per call, over all calls and sessions: `{r} versions_text(hessian_medians$version, hessian_medians$seconds)`.

```{r}
#| label: plot-hessian
#| fig-cap: !expr call_caption
#| fig-height: 4
ggplot(hessian, aes(x = seconds, y = version, colour = version)) +
  geom_point(data = hessian_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(~calls, ncol = 1) +
  labs(x = "Seconds", y = NULL)
```

<details><summary>Table: mean seconds</summary>

```{r}
#| label: table-hessian-means
dt_table(hessian_means)
```

</details>

<details><summary>Table: every call</summary>

```{r}
#| label: table-hessian
hessian |>
  select(session, version, call, seconds) |>
  dt_table()
```

</details>

### mcmc()

Two `mcmc()` calls per session on each model: the first on a new model, and the second on the same model, which reuses the functions the first call traced.

```{r}
#| label: mcmc-summaries
mcmc_means <- mcmc_calls |>
  group_by(model, calls, version) |>
  summarise(seconds = mean(seconds), .groups = "drop")

mcmc_medians <- mcmc_calls |>
  group_by(version) |>
  summarise(seconds = median(seconds))
```

Median seconds per call, over all models, calls and sessions: `{r} versions_text(mcmc_medians$version, mcmc_medians$seconds)`.

```{r}
#| label: plot-mcmc
#| fig-cap: !expr call_caption
#| fig-height: 9
ggplot(mcmc_calls, aes(x = seconds, y = version, colour = version)) +
  geom_point(data = mcmc_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(vars(model, calls), ncol = 2, scales = "free_x") +
  labs(x = "Seconds", y = NULL)
```

<details><summary>Table: mean seconds</summary>

```{r}
#| label: table-mcmc-means
dt_table(mcmc_means)
```

</details>

<details><summary>Table: every call</summary>

```{r}
#| label: table-mcmc
mcmc_calls |>
  select(session, version, model, call, seconds) |>
  dt_table()
```

</details>

:::

## Tracing

After each `mcmc()` call, `run.R` asks each of the model's TensorFlow functions how many times it has been traced, so the count after the second call includes the first. It also counts the retracing warnings TensorFlow logged during the call.

```{r}
#| label: tracing-counts
tracing <- mcmc_calls |>
  select(session, version, model, calls, retracing_warnings, ends_with("_traces")) |>
  pivot_longer(
    c(retracing_warnings, ends_with("_traces")),
    names_to = "measure",
    values_to = "count"
  ) |>
  mutate(
    measure = factor(
      measure,
      levels = c("log_prob_traces", "trace_values_traces", "sampler_traces", "retracing_warnings"),
      labels = c("log density traces", "trace values traces", "sampler traces", "retracing warnings")
    )
  )

# whether every session gave the same count, so one bar can stand for them all
counts_per_session <- tracing |>
  group_by(version, model, calls, measure) |>
  summarise(distinct_counts = n_distinct(count), .groups = "drop")
same_in_every_session <- all(counts_per_session$distinct_counts == 1)
```

`{r} if (same_in_every_session) paste0("Each count was the same in all ", n_sessions, " sessions, so each bar is that count.") else "The counts differ between sessions, so each dot is one session."`

```{r}
#| label: plot-tracing
#| fig-height: 8
#| warning: false
tracing_plot <- ggplot(tracing, aes(x = count, y = model, fill = version, colour = version)) +
  scale_fill_manual(values = version_colours) +
  scale_colour_manual(values = version_colours) +
  scale_x_continuous(breaks = \(limits) unique(floor(pretty(limits)))) +
  facet_grid(measure ~ calls, scales = "free_x") +
  labs(x = "Count", y = NULL, fill = NULL, colour = NULL)

if (same_in_every_session) {
  tracing_plot +
    geom_col(
      data = distinct(tracing, version, model, calls, measure, count),
      position = position_dodge(width = 0.8),
      width = 0.75,
      colour = NA
    )
} else {
  tracing_plot +
    geom_point(position = position_dodge(width = 0.8))
}
```

<details><summary>Table: every count</summary>

```{r}
#| label: table-tracing
tracing |>
  select(session, version, model, calls, measure, count) |>
  dt_table()
```

</details>

```{r}
#| label: hessian-warnings
hessian_warning_counts <- hessian |>
  distinct(session, version, retracing_warnings_in_all_calls) |>
  group_by(version) |>
  summarise(counts = paste(sort(unique(retracing_warnings_in_all_calls)), collapse = " or "))
```

Retracing warnings TensorFlow logged during each session's `{r} settings$hessian_calls` calls of `opt(m, hessian = TRUE)`: `{r} paste(hessian_warning_counts$version, hessian_warning_counts$counts, collapse = ", ")`.

## Efficiency

Bulk effective sample size (ESS) per second of `mcmc()`, for the parameter with the smallest bulk ESS in each call, and for the median parameter. The seconds are the whole `mcmc()` call, so a first call's include its tracing.

```{r}
#| label: efficiency
efficiency <- posterior |>
  group_by(session, version, model, calls) |>
  summarise(
    smallest_ess_bulk = min(ess_bulk),
    median_ess_bulk = median(ess_bulk),
    .groups = "drop"
  ) |>
  left_join(
    select(mcmc_calls, session, version, model, calls, seconds),
    by = c("session", "version", "model", "calls")
  ) |>
  mutate(
    smallest_ess_per_second = smallest_ess_bulk / seconds,
    median_ess_per_second = median_ess_bulk / seconds
  )

efficiency_means <- efficiency |>
  group_by(model, calls, version) |>
  summarise(
    smallest_ess_per_second = mean(smallest_ess_per_second),
    median_ess_per_second = mean(median_ess_per_second),
    .groups = "drop"
  )
```

::: panel-tabset

### Smallest bulk ESS per second

```{r}
#| label: plot-efficiency-smallest
#| fig-cap: !expr call_caption
#| fig-height: 9
ggplot(efficiency, aes(x = smallest_ess_per_second, y = version, colour = version)) +
  geom_point(data = efficiency_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(vars(model, calls), ncol = 2, scales = "free_x") +
  labs(x = "Smallest bulk ESS per second", y = NULL)
```

### Median bulk ESS per second

```{r}
#| label: plot-efficiency-median
#| fig-cap: !expr call_caption
#| fig-height: 9
ggplot(efficiency, aes(x = median_ess_per_second, y = version, colour = version)) +
  geom_point(data = efficiency_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(vars(model, calls), ncol = 2, scales = "free_x") +
  labs(x = "Median bulk ESS per second", y = NULL)
```

:::

<details><summary>Table: mean ESS per second</summary>

```{r}
#| label: table-efficiency-means
dt_table(efficiency_means, digits = 1)
```

</details>

<details><summary>Table: every call</summary>

```{r}
#| label: table-efficiency
efficiency |>
  select(session, version, model, calls, seconds, smallest_ess_bulk, median_ess_bulk, smallest_ess_per_second, median_ess_per_second) |>
  dt_table(digits = 1)
```

</details>

## Convergence

For each `mcmc()` call, the largest rank-normalised split-$\hat{R}$ and the smallest bulk ESS over the model's parameters [@vehtari2021].

```{r}
#| label: convergence
convergence <- posterior |>
  group_by(session, version, model, calls) |>
  summarise(
    largest_rhat = max(rhat),
    smallest_ess_bulk = min(ess_bulk),
    .groups = "drop"
  )

convergence_means <- convergence |>
  group_by(model, calls, version) |>
  summarise(
    largest_rhat = mean(largest_rhat),
    smallest_ess_bulk = mean(smallest_ess_bulk),
    .groups = "drop"
  )
```

::: panel-tabset

### Largest R-hat

```{r}
#| label: plot-rhat
#| fig-cap: !expr call_caption
#| fig-height: 9
ggplot(convergence, aes(x = largest_rhat, y = version, colour = version)) +
  geom_point(data = convergence_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(vars(model, calls), ncol = 2, scales = "free_x") +
  labs(x = "Largest R-hat", y = NULL)
```

### Smallest bulk ESS

```{r}
#| label: plot-ess
#| fig-cap: !expr call_caption
#| fig-height: 9
ggplot(convergence, aes(x = smallest_ess_bulk, y = version, colour = version)) +
  geom_point(data = convergence_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(vars(model, calls), ncol = 2, scales = "free_x") +
  labs(x = "Smallest bulk ESS", y = NULL)
```

:::

<details><summary>Table: every call</summary>

```{r}
#| label: table-convergence
dt_table(convergence)
```

</details>

## Agreement

Whether two versions estimate the same posterior means. The difference between two versions' means is divided by the Monte Carlo standard error (MCSE) of that difference, so a difference can be told apart from Monte Carlo noise [@vehtari2021; @magnusson2025]. For each parameter, each version's means and MCSEs are pooled over all of its `mcmc()` calls, and

$$
z = \frac{\bar\theta_{\text{b}} - \bar\theta_{\text{a}}}{\sqrt{\text{se}^2_{\text{a}} + \text{se}^2_{\text{b}}}}.
$$

The calls are independent, so a version's pooled standard error is $\sqrt{\sum \text{mcse}^2} / n$ over its $n$ calls. If both versions sample the same posterior, each $z$ is close to a standard normal draw. Seeded calls of two versions that both respect `set.seed()` return the same draws, and so a $z$ of zero.

```{r}
#| label: agreement
pooled <- posterior |>
  group_by(version, model, variable) |>
  summarise(
    calls = n(),
    mean = mean(mean),
    se = sqrt(sum(mcse_mean^2)) / n(),
    .groups = "drop"
  )

pairs <- utils::combn(version_levels, 2)
agreement <- lapply(seq_len(ncol(pairs)), function(i) {
  reference <- filter(pooled, version == pairs[1, i])
  compared <- filter(pooled, version == pairs[2, i])
  inner_join(reference, compared, by = c("model", "variable"), suffix = c("_a", "_b")) |>
    mutate(
      comparison = paste(pairs[2, i], "vs", pairs[1, i]),
      z = (mean_b - mean_a) / sqrt(se_a^2 + se_b^2)
    )
}) |>
  bind_rows() |>
  mutate(comparison = factor(comparison, levels = unique(comparison)))

# the band all of a comparison's z fall inside with probability 0.95 when both
# versions sample the same posterior: a Bonferroni bound over its parameters
parameters_compared <- n_distinct(paste(agreement$model, agreement$variable))
band <- qnorm(1 - 0.05 / (2 * parameters_compared))

outside_band <- agreement |>
  group_by(comparison) |>
  summarise(outside = sum(abs(z) > band), largest_abs_z = max(abs(z)))
```

The grey band is $\pm$ `{r} format_number(band)`: if two versions sample the same posterior, all `{r} parameters_compared` parameters' $z$ fall inside it with probability at least 0.95, by a Bonferroni bound. Parameters outside it, per comparison: `{r} paste(outside_band$comparison, outside_band$outside, collapse = ", ")`.

```{r}
#| label: plot-agreement
#| fig-cap: "Each dot is one parameter. The grey band is where every parameter's z falls, with probability at least 0.95, when both versions sample the same posterior."
#| fig-height: 7
ggplot(agreement, aes(x = z, y = model)) +
  annotate(
    "rect",
    xmin = -band,
    xmax = band,
    ymin = -Inf,
    ymax = Inf,
    fill = "grey80",
    alpha = 0.5
  ) +
  geom_point(position = jitter, alpha = 0.7) +
  facet_wrap(~comparison, ncol = 1) +
  labs(x = "z", y = NULL)
```

<details><summary>Table: largest |z| per comparison and model</summary>

```{r}
#| label: table-agreement-summary
agreement |>
  group_by(comparison, model) |>
  summarise(
    parameters = n(),
    largest_abs_z = max(abs(z)),
    parameter = variable[which.max(abs(z))],
    .groups = "drop"
  ) |>
  dt_table(digits = 2)
```

</details>

<details><summary>Table: every parameter</summary>

```{r}
#| label: table-agreement
agreement |>
  select(comparison, model, variable, mean_a, se_a, mean_b, se_b, z) |>
  dt_table()
```

</details>

## Traceplots

Every `{r} settings$keep_every_nth_draw`th draw of each chain from session 1's first `mcmc()` call on each model, numbered from the end of warmup.

```{r}
#| label: traceplots
gg_trace <- function(name, version_label) {
  draws |>
    filter(session == 1, model == name, version == version_label) |>
    ggplot(aes(x = draw, y = value, colour = factor(chain))) +
    geom_line(linewidth = 0.3) +
    facet_wrap(~variable, scales = "free_y", ncol = 4) +
    labs(x = "Draw", y = NULL, colour = "Chain")
}

trace_height <- function(name) {
  1 + 1.6 * ceiling(parameter_count(name) / 4)
}
```

::: panel-tabset

### linear

::: panel-tabset

#### CRAN

```{r}
#| label: plot-trace-linear-cran
#| fig-height: !expr trace_height("linear")
gg_trace("linear", "CRAN")
```

#### main

```{r}
#| label: plot-trace-linear-main
#| fig-height: !expr trace_height("linear")
gg_trace("linear", "main")
```

#### #843

```{r}
#| label: plot-trace-linear-843
#| fig-height: !expr trace_height("linear")
gg_trace("linear", "#843")
```

:::

### multiple_linear

::: panel-tabset

#### CRAN

```{r}
#| label: plot-trace-multiple-linear-cran
#| fig-height: !expr trace_height("multiple_linear")
gg_trace("multiple_linear", "CRAN")
```

#### main

```{r}
#| label: plot-trace-multiple-linear-main
#| fig-height: !expr trace_height("multiple_linear")
gg_trace("multiple_linear", "main")
```

#### #843

```{r}
#| label: plot-trace-multiple-linear-843
#| fig-height: !expr trace_height("multiple_linear")
gg_trace("multiple_linear", "#843")
```

:::

### hierarchical_linear

::: panel-tabset

#### CRAN

```{r}
#| label: plot-trace-hierarchical-linear-cran
#| fig-height: !expr trace_height("hierarchical_linear")
gg_trace("hierarchical_linear", "CRAN")
```

#### main

```{r}
#| label: plot-trace-hierarchical-linear-main
#| fig-height: !expr trace_height("hierarchical_linear")
gg_trace("hierarchical_linear", "main")
```

#### #843

```{r}
#| label: plot-trace-hierarchical-linear-843
#| fig-height: !expr trace_height("hierarchical_linear")
gg_trace("hierarchical_linear", "#843")
```

:::

### eight_schools

::: panel-tabset

#### CRAN

```{r}
#| label: plot-trace-eight-schools-cran
#| fig-height: !expr trace_height("eight_schools")
gg_trace("eight_schools", "CRAN")
```

#### main

```{r}
#| label: plot-trace-eight-schools-main
#| fig-height: !expr trace_height("eight_schools")
gg_trace("eight_schools", "main")
```

#### #843

```{r}
#| label: plot-trace-eight-schools-843
#| fig-height: !expr trace_height("eight_schools")
gg_trace("eight_schools", "#843")
```

:::

### cjs

::: panel-tabset

#### CRAN

```{r}
#| label: plot-trace-cjs-cran
#| fig-height: !expr trace_height("cjs")
gg_trace("cjs", "CRAN")
```

#### main

```{r}
#| label: plot-trace-cjs-main
#| fig-height: !expr trace_height("cjs")
gg_trace("cjs", "main")
```

#### #843

```{r}
#| label: plot-trace-cjs-843
#| fig-height: !expr trace_height("cjs")
gg_trace("cjs", "#843")
```

:::

:::

## Posterior densities

The same kept draws, from every session's first `mcmc()` call, pooled over sessions.

```{r}
#| label: densities
gg_densities <- function(name) {
  draws |>
    filter(model == name) |>
    ggplot(aes(x = value, colour = version)) +
    geom_density() +
    scale_colour_manual(values = version_colours) +
    facet_wrap(~variable, scales = "free", ncol = 4) +
    labs(x = NULL, y = NULL, colour = NULL)
}

# each version's draws keep the order run.R saved them in within a session
draws_identical <- function(pair) {
  first <- filter(draws, version == pair[1]) |> arrange(session)
  second <- filter(draws, version == pair[2]) |> arrange(session)
  identical(first$value, second$value)
}
identical_pairs <- Filter(
  draws_identical,
  utils::combn(version_levels, 2, simplify = FALSE)
)
identical_text <- if (length(identical_pairs) == 0) {
  ""
} else {
  paste0(
    vapply(identical_pairs, \(pair) paste(pair, collapse = "'s and "), character(1)),
    "'s draws are identical in every session, so their lines lie on top of each other.",
    collapse = " "
  )
}
```

`{r} identical_text`

::: panel-tabset

### linear

```{r}
#| label: plot-densities-linear
#| fig-height: !expr trace_height("linear")
gg_densities("linear")
```

### multiple_linear

```{r}
#| label: plot-densities-multiple-linear
#| fig-height: !expr trace_height("multiple_linear")
gg_densities("multiple_linear")
```

### hierarchical_linear

```{r}
#| label: plot-densities-hierarchical-linear
#| fig-height: !expr trace_height("hierarchical_linear")
gg_densities("hierarchical_linear")
```

### eight_schools

```{r}
#| label: plot-densities-eight-schools
#| fig-height: !expr trace_height("eight_schools")
gg_densities("eight_schools")
```

### cjs

```{r}
#| label: plot-densities-cjs
#| fig-height: !expr trace_height("cjs")
gg_densities("cjs")
```

:::

## The code

`run.R`, which made every result on this page. [Download it](run.R), or find it with its results in [greta-dev/greta.benchmarks](https://github.com/greta-dev/greta.benchmarks).

```{r}
#| label: run-script
#| file: run.R
#| eval: false
#| code-fold: false
```

## References

::: {#refs}
:::
