---
title: "greta#843 (faster-hessians-i546): retracing and hessians"
description: "Building models, opt(), opt(hessian = TRUE) and mcmc() on greta 0.6.0 from CRAN, on main and on greta#843. greta#843 traces each model's TensorFlow functions fewer times, and opt(hessian = TRUE) is much faster on it."
date: 2026-10-05
bibliography: ../../references.bib
---

```{r}
#| label: setup
#| code-summary: "Setup"
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(bench)
  library(here)
})

post_path <- "posts/2026-10-05-retracing-i546"

# each `# ---- label ----` section of the scripts, shown below by its label
knitr::read_chunk(here(post_path, "benchmark.R"))
knitr::read_chunk(here(post_path, "run.R"))

# links to the scripts on GitHub, at the commit being rendered when GitHub
# Actions renders the site
github_ref <- Sys.getenv("GITHUB_SHA", "main")
github_file_url <- function(file) {
  paste0(
    "https://github.com/greta-dev/greta.benchmarks/blob/",
    github_ref,
    "/",
    post_path,
    "/",
    file
  )
}
# Quarto escapes the markdown in an inline `{r}` result unless it is wrapped
# in I(), so the link helpers return I()
file_link <- function(file) {
  I(paste0("[`", file, "`](", github_file_url(file), ")"))
}

# a link to the lines of a script's `# ---- label ----` section
section_link <- function(file, label) {
  lines <- readLines(here(post_path, file))
  headings <- grep("^# ---- .+ ----$", lines)
  start <- headings[lines[headings] == paste("# ----", label, "----")]
  end <- c(headings, length(lines) + 1)[match(start, headings) + 1] - 1
  while (end > start && !nzchar(trimws(lines[end]))) {
    end <- end - 1
  }
  I(paste0(
    "[`", label, "`](", github_file_url(file), "#L", start, "-L", end, ")"
  ))
}

# "CRAN, main and #843"
paste_and <- function(x) {
  if (length(x) <= 1) {
    return(x)
  }
  paste(paste(x[-length(x)], collapse = ", "), "and", x[length(x)])
}

# links to several sections of a script, as "a, b and c"
section_links <- function(file, labels) {
  links <- vapply(
    labels,
    \(label) as.character(section_link(file, label)),
    character(1)
  )
  I(paste_and(links))
}

# white panels, so the grey band in the agreement plot shows
theme_set(theme_bw())

sessions <- list.files(
  here(post_path, "results"),
  pattern = "^session-[0-9]+[.]rds$",
  full.names = TRUE
) |>
  lapply(readRDS)

versions <- sessions[[1]]$versions
version_levels <- names(versions)
version_colours <- setNames(
  c("#E69F00", "#56B4E9", "#009E73")[seq_along(version_levels)],
  version_levels
)
# the version this post is about: every ratio is against it
pr_version <- "#843"
bench_settings <- sessions[[1]]$results[[1]]$bench_settings
n_sessions <- length(sessions)

# one row per session and version, holding that version's results
runs <- lapply(sessions, function(session_file) {
  tibble(
    session = session_file$session,
    version = factor(names(session_file$results), levels = version_levels),
    position = match(names(session_file$results), session_file$order),
    results = unname(session_file$results)
  )
}) |>
  bind_rows()

model_levels <- names(runs$results[[1]]$build)
# benchmark.R's section labels spell each model with hyphens
model_labels <- gsub("_", "-", model_levels)

# a part of the results that is a table or a value per run, stacked, with the
# run it came from
stack_tables <- function(part) {
  pieces <- lapply(seq_len(nrow(runs)), function(i) {
    values <- runs$results[[i]][[part]]
    if (!is.data.frame(values)) {
      values <- tibble(value = values)
    }
    tibble(session = runs$session[i], version = runs$version[i], values)
  })
  bind_rows(pieces)
}

# a part of the results that is a bench_mark per model, or one per run,
# stacked into one bench_mark, with the run and model each came from
stack_marks <- function(part) {
  pieces <- list()
  for (i in seq_len(nrow(runs))) {
    marks <- runs$results[[i]][[part]]
    if (inherits(marks, "bench_mark")) {
      marks <- list(marks)
    }
    for (j in seq_along(marks)) {
      mark <- marks[[j]]
      mark$session <- runs$session[i]
      mark$version <- runs$version[i]
      mark$model <- if (is.null(names(marks))) NA_character_ else names(marks)[j]
      pieces[[length(pieces) + 1]] <- mark
    }
  }
  vctrs::vec_rbind(!!!pieces)
}

provenance <- stack_tables("provenance")
sampler_iterations_per_draw <- stack_tables("sampler_iterations_per_draw")
build_marks <- stack_marks("build")
opt_marks <- stack_marks("opt")
hessian_marks <- stack_marks("hessian")
hessian_retracing_warnings <- stack_tables("hessian_retracing_warnings")
mcmc_marks <- stack_marks("mcmc")
mcmc_tracing <- stack_tables("mcmc_tracing")

# one row per timed call: bench keeps every call's time in the `time` column
every_call <- function(marks) {
  marks |>
    as_tibble() |>
    select(session, version, model, time) |>
    mutate(call = lapply(time, seq_along)) |>
    unnest(c(time, call)) |>
    mutate(
      seconds = as.numeric(time),
      model = factor(model, levels = model_levels),
      .keep = "unused"
    )
}

# bench's own summary() of each mark, with the mean and longest of its calls
# added, all in seconds
bench_summary <- function(marks) {
  summaries <- summary(marks, filter_gc = FALSE)
  tibble(
    session = summaries$session,
    version = summaries$version,
    model = summaries$model,
    min = as.numeric(summaries$min),
    median = as.numeric(summaries$median),
    mean = vapply(marks$time, \(time) mean(as.numeric(time)), numeric(1)),
    max = vapply(marks$time, \(time) max(as.numeric(time)), numeric(1)),
    `itr/sec` = summaries$`itr/sec`,
    n_itr = summaries$n_itr,
    total_time = as.numeric(summaries$total_time)
  )
}

# each version's median call, from a table of every call
version_medians <- function(calls) {
  calls |>
    group_by(version) |>
    summarise(seconds = median(seconds))
}

# The first call on a model traces its TensorFlow functions; later calls on the
# same model reuse them.
label_calls <- function(call) {
  last <- max(call)
  later <- if (last == 2) {
    "second call on the same model"
  } else if (last == 3) {
    "calls 2 and 3 on the same model"
  } else {
    paste0("calls 2 to ", last, " on the same model")
  }
  factor(
    ifelse(call == 1, "first call on a new model", later),
    levels = c("first call on a new model", later)
  )
}

# three significant figures each, where format() would pad a vector to the
# decimal places of its smallest value; formatC() pads each to a common width,
# so the spaces are trimmed
format_number <- function(x) {
  trimws(formatC(signif(x, 3), format = "fg", digits = 3, big.mark = ","))
}

# "CRAN 10.6, main 11.2, #843 0.643"
versions_text <- function(version, value) {
  paste(paste(version, format_number(value)), collapse = ", ")
}

# a version's value, from a table with one row per version
value_of <- function(table, version_name, column = "seconds") {
  table[[column]][table$version == version_name]
}

# how many times as long as #843 a version takes, from a table of medians
times_as_long <- function(table, version_name) {
  format_number(value_of(table, version_name) / value_of(table, pr_version))
}

# "once", "twice" or "3 times"
times_word <- function(n) {
  ifelse(n == 1, "once", ifelse(n == 2, "twice", paste(n, "times")))
}

version_scale <- scale_colour_manual(values = version_colours, guide = "none")
version_fill <- scale_fill_manual(values = version_colours, guide = "none")
jitter <- position_jitter(width = 0, height = 0.2, seed = 2026)

# A raincloud for each version: the density of its calls above, their box plot
# in the middle, and every call as a dot below. The summaries are drawn first
# and faint, so the dots sit on top.
raincloud <- list(
  ggdist::stat_halfeye(
    aes(fill = version),
    width = 0.4,
    justification = -0.3,
    .width = 0,
    point_colour = NA,
    alpha = 0.3
  ),
  geom_boxplot(
    aes(fill = version),
    width = 0.25,
    outlier.shape = NA,
    alpha = 0.3
  ),
  # a fixed dot size, so a call looks the same size in every panel
  ggdist::stat_dots(
    aes(fill = version),
    side = "bottom",
    layout = "swarm",
    justification = 1.3,
    scale = 0.75,
    binwidth = unit(2.6, "mm"),
    overflow = "compress",
    colour = NA,
    alpha = 0.6
  ),
  version_scale,
  version_fill,
  theme_bw(base_size = 13)
)
raincloud_caption <- "Each dot is one call, from every session. Above the dots, the curve is the density of the calls and the box marks their median and quartiles."
session_caption <- "Each small dot is one session's second call. The large faint dot behind them is their mean."

# a searchable table, its numbers rounded for reading
dt_table <- function(data, digits = 3) {
  table <- DT::datatable(
    data,
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 10, autoWidth = TRUE)
  )
  decimal_columns <- names(data)[vapply(data, is.double, logical(1))]
  if (length(decimal_columns) == 0) {
    return(table)
  }
  DT::formatRound(table, decimal_columns, digits)
}

# a model's draws array is draws by chains by parameters
parameter_count <- function(name) {
  dim(runs$results[[1]]$mcmc_draws[[name]])[3]
}

# wide enough for a column of panels per parameter
draws_width <- function(name) {
  max(8, 1.6 * parameter_count(name))
}

# the opening's numbers: each version's median opt(hessian = TRUE) call, and
# how many times mcmc() traced each model's log density
opening_hessian <- version_medians(every_call(hessian_marks))
opening_traces <- mcmc_tracing |>
  group_by(version) |>
  summarise(
    traces = paste(times_word(sort(unique(log_prob_traces))), collapse = " or ")
  )
```

[greta#843](https://github.com/greta-dev/greta/pull/843) traces each model's log-density and trace-values functions once, whatever the batch size, and takes all of `opt(hessian = TRUE)`'s hessians in one graph build. This post compares it with greta 0.6.0 on CRAN and with main. On a model with twenty hessians, an `opt(hessian = TRUE)` call takes a median of `{r} format_number(value_of(opening_hessian, pr_version))` seconds on #843; CRAN takes `{r} times_as_long(opening_hessian, "CRAN")` times as long, and main `{r} times_as_long(opening_hessian, "main")` times as long. In `mcmc()`, #843 traces each model's log density `{r} value_of(opening_traces, pr_version, "traces")`, where CRAN traces it `{r} value_of(opening_traces, "CRAN", "traces")` and main `{r} value_of(opening_traces, "main", "traces")`.

## How long does `opt(hessian = TRUE)` take? {#speed}

Each session called `opt(hessian = TRUE)` `{r} bench_settings$hessian_calls` times on the [hessian model](#the-models), whose twenty scalar parameters each need a hessian. The first call traces the model's TensorFlow functions and later calls reuse them, so they are plotted apart.

```{r}
#| label: hessian-summaries
hessian <- every_call(hessian_marks) |>
  select(-model) |>
  mutate(calls = label_calls(call))
hessian_summary <- bench_summary(hessian_marks) |>
  select(-model)

# each version's median first call, and its median later call
hessian_medians <- hessian |>
  group_by(version, first_call = call == 1) |>
  summarise(seconds = median(seconds), .groups = "drop")
hessian_first <- filter(hessian_medians, first_call)
hessian_later <- filter(hessian_medians, !first_call)

hessian_warning_counts <- hessian_retracing_warnings |>
  group_by(version) |>
  summarise(counts = paste(sort(unique(value)), collapse = " or "))
```

```{r}
#| label: plot-hessian
#| fig-cap: !expr raincloud_caption
#| fig-height: 11
ggplot(hessian, aes(x = seconds, y = version, colour = version)) +
  raincloud +
  facet_wrap(~calls, ncol = 1) +
  labs(x = "Seconds", y = NULL)
```

Timed by `benchmark.R`'s `{r} section_link("benchmark.R", "bench-hessian")` section.

The median first call takes `{r} versions_text(hessian_first$version, hessian_first$seconds)` seconds, and the median later call `{r} versions_text(hessian_later$version, hessian_later$seconds)`. So on later calls, CRAN takes `{r} times_as_long(hessian_later, "CRAN")` times as long as #843, and main `{r} times_as_long(hessian_later, "main")` times as long. TensorFlow's retracing warnings during each session's calls, counted in the same section: `{r} paste(hessian_warning_counts$version, hessian_warning_counts$counts, collapse = ", ")`.

<details><summary>Table: bench's summary() of every mark</summary>

```{r}
#| label: table-hessian-summary
dt_table(hessian_summary, digits = 4)
```

</details>

<details><summary>Table: every call</summary>

```{r}
#| label: table-hessian
dt_table(hessian, digits = 4)
```

</details>

## How many times does `mcmc()` trace a model?

After both of a model's `mcmc()` calls, each session asked each of the model's TensorFlow functions how many times it had been traced, and counted the retracing warnings TensorFlow logged during the two calls.

```{r}
#| label: tracing-counts
tracing <- mcmc_tracing |>
  pivot_longer(
    c(retracing_warnings, ends_with("_traces")),
    names_to = "measure",
    values_to = "count"
  ) |>
  mutate(
    model = factor(model, levels = model_levels),
    measure = factor(
      measure,
      levels = c("log_prob_traces", "trace_values_traces", "sampler_traces", "retracing_warnings"),
      labels = c("log density traces", "trace values traces", "sampler traces", "retracing warnings")
    )
  )

# whether every session gave the same count, so one bar can stand for them all
counts_per_session <- tracing |>
  group_by(version, model, measure) |>
  summarise(distinct_counts = n_distinct(count), .groups = "drop")
same_in_every_session <- all(counts_per_session$distinct_counts == 1)

# how many times a version traced a function, over every model and session
traced <- function(measure_name, version_name) {
  counts <- tracing$count[
    tracing$measure == measure_name & tracing$version == version_name
  ]
  paste(times_word(sort(unique(counts))), collapse = " or ")
}
```

`{r} if (same_in_every_session) paste0("Each count was the same in all ", n_sessions, " sessions, so each bar is that count.") else "The counts differ between sessions, so each dot is one session."`

```{r}
#| label: plot-tracing
#| fig-height: 5
tracing_plot <- ggplot(tracing, aes(x = count, y = model)) +
  scale_x_continuous(breaks = \(limits) unique(floor(pretty(limits)))) +
  facet_wrap(~measure, ncol = 2, scales = "free_x") +
  labs(x = "Count", y = NULL, fill = NULL, colour = NULL)

# bars are filled and dots coloured, so each draws only the scale it uses
if (same_in_every_session) {
  tracing_plot +
    geom_col(
      aes(fill = version),
      data = distinct(tracing, version, model, measure, count),
      position = position_dodge(width = 0.8),
      width = 0.75
    ) +
    scale_fill_manual(values = version_colours)
} else {
  tracing_plot +
    geom_point(aes(colour = version), position = position_dodge(width = 0.8)) +
    scale_colour_manual(values = version_colours)
}
```

Counted by `benchmark.R`'s `{r} section_link("benchmark.R", "mcmc-tracing")` section, with the helpers in its `{r} section_link("benchmark.R", "benchmark-helpers")` section.

On #843, each model's log density is traced `{r} traced("log density traces", pr_version)` and its trace values `{r} traced("trace values traces", pr_version)`; on CRAN, `{r} traced("log density traces", "CRAN")` and `{r} traced("trace values traces", "CRAN")`; on main, `{r} traced("log density traces", "main")` and `{r} traced("trace values traces", "main")`. TensorFlow logged `{r} sum(mcmc_tracing$retracing_warnings)` retracing warnings during `mcmc()`, over every version, model and session.

<details><summary>Table: every count</summary>

```{r}
#| label: table-tracing
dt_table(tracing)
```

</details>

## How long does `mcmc()` take?

Each session called `mcmc()` twice on each model, with `{r} bench_settings$warmup` warmup and `{r} bench_settings$n_samples` kept draws on each of `{r} bench_settings$chains` chains: first on a new model, which traces its TensorFlow functions, then on the same model again.

```{r}
#| label: mcmc-summaries
mcmc_timings <- every_call(mcmc_marks) |>
  mutate(calls = label_calls(call))
mcmc_summary <- bench_summary(mcmc_marks)
mcmc_medians <- version_medians(mcmc_timings)
```

```{r}
#| label: plot-mcmc
#| fig-cap: !expr raincloud_caption
#| fig-height: 16
ggplot(mcmc_timings, aes(x = seconds, y = version, colour = version)) +
  raincloud +
  facet_wrap(vars(model, calls), ncol = 2, scales = "free_x") +
  labs(x = "Seconds", y = NULL)
```

Timed by `benchmark.R`'s `{r} section_links("benchmark.R", paste0("bench-mcmc-", model_labels))` sections.

Over all models and both calls, the median call takes `{r} format_number(value_of(mcmc_medians, pr_version))` seconds on #843; CRAN takes `{r} times_as_long(mcmc_medians, "CRAN")` times as long, and main `{r} times_as_long(mcmc_medians, "main")` times as long.

<details><summary>Table: bench's summary() of every mark</summary>

```{r}
#| label: table-mcmc-summary
dt_table(mcmc_summary, digits = 4)
```

</details>

<details><summary>Table: every call</summary>

```{r}
#| label: table-mcmc
dt_table(mcmc_timings, digits = 4)
```

</details>

## How many effective draws per second does `mcmc()` give?

Efficiency is the bulk effective sample size (ESS) per second, the time-normalised ESS of @girolami2011, for the parameter with the smallest bulk ESS, as @hoffman2014 report, and for the median parameter [@vehtari2021]. It uses every draw of each model's second `mcmc()` call, and that call's seconds, warmup included.

```{r}
#| label: efficiency
posterior <- lapply(seq_len(nrow(runs)), function(i) {
  draws <- runs$results[[i]]$mcmc_draws
  lapply(names(draws), function(model_name) {
    tibble(
      session = runs$session[i],
      version = runs$version[i],
      model = model_name,
      posterior::summarise_draws(
        draws[[model_name]],
        "mean",
        "sd",
        "mcse_mean",
        "rhat",
        "ess_bulk",
        "ess_tail"
      )
    )
  }) |>
    bind_rows()
}) |>
  bind_rows() |>
  mutate(model = factor(model, levels = model_levels))

second_calls <- mcmc_timings |>
  filter(call == 2) |>
  select(session, version, model, seconds)

efficiency <- posterior |>
  group_by(session, version, model) |>
  summarise(
    smallest_ess_bulk = min(ess_bulk),
    median_ess_bulk = median(ess_bulk),
    .groups = "drop"
  ) |>
  left_join(second_calls, by = c("session", "version", "model")) |>
  mutate(
    smallest_ess_per_second = smallest_ess_bulk / seconds,
    median_ess_per_second = median_ess_bulk / seconds
  )

efficiency_means <- efficiency |>
  group_by(model, version) |>
  summarise(
    smallest_ess_per_second = mean(smallest_ess_per_second),
    median_ess_per_second = mean(median_ess_per_second),
    .groups = "drop"
  )

# one row per session, model, version and measure, for one panel per measure
efficiency_measures <- c(
  smallest_ess_per_second = "Smallest bulk ESS per second",
  median_ess_per_second = "Median bulk ESS per second"
)
longer_efficiency <- function(data) {
  data |>
    pivot_longer(
      all_of(names(efficiency_measures)),
      names_to = "measure",
      values_to = "ess_per_second"
    ) |>
    mutate(
      measure = factor(
        efficiency_measures[measure],
        levels = efficiency_measures
      )
    )
}

# #843's mean smallest bulk ESS per second over another version's, per model
efficiency_ratios <- efficiency_means |>
  group_by(model) |>
  summarise(
    of_cran = smallest_ess_per_second[version == pr_version] /
      smallest_ess_per_second[version == "CRAN"],
    of_main = smallest_ess_per_second[version == pr_version] /
      smallest_ess_per_second[version == "main"]
  )
```

```{r}
#| label: plot-efficiency
#| fig-cap: !expr session_caption
#| fig-height: 10
ggplot(
  longer_efficiency(efficiency),
  aes(x = ess_per_second, y = version, colour = version)
) +
  geom_point(
    data = longer_efficiency(efficiency_means),
    size = 5,
    alpha = 0.25
  ) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(vars(model, measure), ncol = 2, scales = "free_x") +
  labs(x = "Bulk ESS per second", y = NULL)
```

The draws were kept by `benchmark.R`'s `{r} section_link("benchmark.R", "mcmc-draws")` section, and the calls timed by its `{r} section_links("benchmark.R", paste0("bench-mcmc-", model_labels))` sections.

Depending on the model, #843 gives between `{r} format_number(min(efficiency_ratios$of_cran))` and `{r} format_number(max(efficiency_ratios$of_cran))` times CRAN's mean smallest bulk ESS per second, and between `{r} format_number(min(efficiency_ratios$of_main))` and `{r} format_number(max(efficiency_ratios$of_main))` times main's.

<details><summary>Table: mean ESS per second</summary>

```{r}
#| label: table-efficiency-means
dt_table(efficiency_means, digits = 1)
```

</details>

<details><summary>Table: every session</summary>

```{r}
#| label: table-efficiency
dt_table(efficiency, digits = 1)
```

</details>

## How long does `opt()` take?

Each session called `opt()` `{r} bench_settings$opt_calls` times on each model, built once for it. The first call traces the model's TensorFlow functions and later calls reuse them.

```{r}
#| label: opt-summaries
opt_timings <- every_call(opt_marks) |>
  mutate(calls = label_calls(call))
opt_summary <- bench_summary(opt_marks)
opt_medians <- version_medians(opt_timings)
```

```{r}
#| label: plot-opt
#| fig-cap: !expr raincloud_caption
#| fig-height: 16
ggplot(opt_timings, aes(x = seconds, y = version, colour = version)) +
  raincloud +
  facet_wrap(vars(model, calls), ncol = 2, scales = "free_x") +
  labs(x = "Seconds", y = NULL)
```

Timed by `benchmark.R`'s `{r} section_links("benchmark.R", paste0("bench-opt-", model_labels))` sections.

Over all models and calls, the median call takes `{r} format_number(value_of(opt_medians, pr_version))` seconds on #843; CRAN takes `{r} times_as_long(opt_medians, "CRAN")` times as long, and main `{r} times_as_long(opt_medians, "main")` times as long.

<details><summary>Table: bench's summary() of every mark</summary>

```{r}
#| label: table-opt-summary
dt_table(opt_summary, digits = 4)
```

</details>

<details><summary>Table: every call</summary>

```{r}
#| label: table-opt
dt_table(opt_timings, digits = 4)
```

</details>

## How long does building a model take?

Each session built each model `{r} bench_settings$build_calls` times: creating its greta arrays and calling `model()`.

```{r}
#| label: build-summaries
build <- every_call(build_marks)
build_summary <- bench_summary(build_marks)
build_medians <- version_medians(build)
```

```{r}
#| label: plot-build
#| fig-cap: !expr raincloud_caption
#| fig-height: 13
ggplot(build, aes(x = seconds, y = version, colour = version)) +
  raincloud +
  facet_wrap(~model, ncol = 1, scales = "free_x") +
  labs(x = "Seconds", y = NULL)
```

Timed by `benchmark.R`'s `{r} section_links("benchmark.R", paste0("bench-build-", model_labels))` sections.

Over all models, the median build takes `{r} format_number(value_of(build_medians, pr_version))` seconds on #843; CRAN takes `{r} times_as_long(build_medians, "CRAN")` times as long, and main `{r} times_as_long(build_medians, "main")` times as long.

<details><summary>Table: bench's summary() of every mark</summary>

```{r}
#| label: table-build-summary
dt_table(build_summary, digits = 4)
```

</details>

<details><summary>Table: every call</summary>

```{r}
#| label: table-build
dt_table(build, digits = 4)
```

</details>

## Do the versions sample the same posterior?

For each parameter, each version's posterior means and their Monte Carlo standard errors (MCSE) are pooled over its sessions, and two versions' pooled means are compared by

$$
z = \frac{\bar\theta_{\text{b}} - \bar\theta_{\text{a}}}{\sqrt{\text{se}^2_{\text{a}} + \text{se}^2_{\text{b}}}},
$$

so a difference can be told apart from Monte Carlo noise [@vehtari2021; @magnusson2025]. Each session has a seed of its own, so the sessions are independent, and a version's pooled standard error is $\sqrt{\sum \text{mcse}^2} / n$ over its $n$ sessions. If both versions sample the same posterior, each $z$ is close to a standard normal draw; two versions that return the same draws from the same seed give a $z$ of zero.

```{r}
#| label: agreement
pooled <- posterior |>
  group_by(version, model, variable) |>
  summarise(
    sessions = n(),
    mean = mean(mean),
    se = sqrt(sum(mcse_mean^2)) / n(),
    .groups = "drop"
  )

pairs <- utils::combn(version_levels, 2)
agreement <- lapply(seq_len(ncol(pairs)), function(i) {
  reference <- filter(pooled, version == pairs[1, i])
  compared <- filter(pooled, version == pairs[2, i])
  inner_join(reference, compared, by = c("model", "variable"), suffix = c("_a", "_b")) |>
    mutate(
      comparison = paste(pairs[2, i], "vs", pairs[1, i]),
      z = (mean_b - mean_a) / sqrt(se_a^2 + se_b^2)
    )
}) |>
  bind_rows() |>
  mutate(comparison = factor(comparison, levels = unique(comparison)))

# the band all of a comparison's z fall inside with probability 0.95 when both
# versions sample the same posterior: a Bonferroni bound over its parameters
parameters_compared <- n_distinct(paste(agreement$model, agreement$variable))
band <- qnorm(1 - 0.05 / (2 * parameters_compared))

outside_band <- agreement |>
  group_by(comparison) |>
  summarise(outside = sum(abs(z) > band), largest_abs_z = max(abs(z)))

# whether two versions returned the same draws in every session
draws_identical <- function(pair) {
  all(vapply(
    unique(runs$session),
    function(session_number) {
      first <- runs$results[[which(runs$session == session_number & runs$version == pair[1])]]
      second <- runs$results[[which(runs$session == session_number & runs$version == pair[2])]]
      identical(first$mcmc_draws, second$mcmc_draws)
    },
    logical(1)
  ))
}
identical_pairs <- Filter(
  draws_identical,
  utils::combn(version_levels, 2, simplify = FALSE)
)
identical_text <- if (length(identical_pairs) == 0) {
  "No two versions returned the same draws."
} else {
  paste0(
    vapply(identical_pairs, \(pair) paste(pair, collapse = "'s and "), character(1)),
    "'s draws are identical in every session.",
    collapse = " "
  )
}
```

```{r}
#| label: plot-agreement
#| fig-cap: "Each dot is one parameter. The grey band is where every parameter's z falls, with probability at least 0.95, when both versions sample the same posterior."
#| fig-height: 6
ggplot(agreement, aes(x = z, y = model)) +
  annotate(
    "rect",
    xmin = -band,
    xmax = band,
    ymin = -Inf,
    ymax = Inf,
    fill = "grey80",
    alpha = 0.5
  ) +
  geom_point(position = jitter, alpha = 0.7) +
  facet_wrap(~comparison, ncol = 1) +
  labs(x = "z", y = NULL)
```

From the draws of each model's second `mcmc()` call, kept by `benchmark.R`'s `{r} section_link("benchmark.R", "mcmc-draws")` section.

The grey band is $\pm$ `{r} format_number(band)`, a Bonferroni bound over the `{r} parameters_compared` parameters. Parameters outside it, per comparison: `{r} paste(outside_band$comparison, outside_band$outside, collapse = ", ")`. `{r} identical_text`

<details><summary>Table: largest |z| per comparison and model</summary>

```{r}
#| label: table-agreement-summary
agreement |>
  group_by(comparison, model) |>
  summarise(
    parameters = n(),
    largest_abs_z = max(abs(z)),
    parameter = variable[which.max(abs(z))],
    .groups = "drop"
  ) |>
  dt_table(digits = 2)
```

</details>

<details><summary>Table: every parameter</summary>

```{r}
#| label: table-agreement
agreement |>
  select(comparison, model, variable, mean_a, se_a, mean_b, se_b, z) |>
  dt_table()
```

</details>

<details><summary>Traceplots</summary>

Every draw of each chain from session 1's second `mcmc()` call on each model, numbered from the end of warmup, with a row of panels for each version.

```{r}
#| label: traceplots
# every draw of every run, one row per draw and parameter
draws <- lapply(seq_len(nrow(runs)), function(i) {
  model_draws <- runs$results[[i]]$mcmc_draws
  lapply(names(model_draws), function(model_name) {
    posterior::as_draws_df(model_draws[[model_name]]) |>
      as.data.frame() |>
      pivot_longer(
        -c(.chain, .iteration, .draw),
        names_to = "variable",
        values_to = "value"
      ) |>
      mutate(
        session = runs$session[i],
        version = runs$version[i],
        model = model_name
      )
  }) |>
    bind_rows()
}) |>
  bind_rows()

gg_trace <- function(name) {
  model_draws <- draws |>
    filter(session == 1, model == name) |>
    mutate(variable = factor(variable, levels = unique(variable)))
  # each parameter's panels share its range across the versions, so they can
  # be compared down a column
  parameter_ranges <- model_draws |>
    group_by(variable) |>
    reframe(value = range(value)) |>
    crossing(version = unique(model_draws$version)) |>
    mutate(.iteration = 1L)
  ggplot(model_draws, aes(x = .iteration, y = value, colour = factor(.chain))) +
    geom_line(linewidth = 0.2) +
    geom_blank(
      data = parameter_ranges,
      aes(x = .iteration, y = value),
      inherit.aes = FALSE
    ) +
    facet_wrap(
      vars(version, variable),
      ncol = n_distinct(model_draws$variable),
      scales = "free_y"
    ) +
    labs(x = "Draw", y = NULL, colour = "Chain")
}
```

::: panel-tabset

### linear

```{r}
#| label: plot-trace-linear
#| fig-height: 6
#| fig-width: !expr draws_width("linear")
gg_trace("linear")
```

### multiple_linear

```{r}
#| label: plot-trace-multiple-linear
#| fig-height: 6
#| fig-width: !expr draws_width("multiple_linear")
gg_trace("multiple_linear")
```

### hierarchical_linear

```{r}
#| label: plot-trace-hierarchical-linear
#| fig-height: 6
#| fig-width: !expr draws_width("hierarchical_linear")
gg_trace("hierarchical_linear")
```

### eight_schools

```{r}
#| label: plot-trace-eight-schools
#| fig-height: 6
#| fig-width: !expr draws_width("eight_schools")
gg_trace("eight_schools")
```

:::

</details>

<details><summary>Posterior densities</summary>

Every draw of every chain, from each session's second `mcmc()` call, pooled over sessions, with every version's density overlaid in one panel per parameter. Where two versions' draws are identical, their lines lie on top of each other.

```{r}
#| label: densities
# every version's density in one panel per parameter, each with its own line
# type as well as colour, so a line drawn on top of an identical one still shows
gg_densities <- function(name) {
  draws |>
    filter(model == name) |>
    mutate(variable = factor(variable, levels = unique(variable))) |>
    ggplot(aes(x = value, colour = version, linetype = version)) +
    geom_density(linewidth = 0.7) +
    scale_colour_manual(values = version_colours) +
    facet_wrap(~variable, scales = "free", ncol = 4) +
    labs(x = NULL, y = NULL, colour = NULL, linetype = NULL) +
    theme(legend.position = "top")
}

densities_height <- function(name) {
  1.5 + 2 * ceiling(parameter_count(name) / 4)
}
```

::: panel-tabset

### linear

```{r}
#| label: plot-densities-linear
#| fig-height: !expr densities_height("linear")
gg_densities("linear")
```

### multiple_linear

```{r}
#| label: plot-densities-multiple-linear
#| fig-height: !expr densities_height("multiple_linear")
gg_densities("multiple_linear")
```

### hierarchical_linear

```{r}
#| label: plot-densities-hierarchical-linear
#| fig-height: !expr densities_height("hierarchical_linear")
gg_densities("hierarchical_linear")
```

### eight_schools

```{r}
#| label: plot-densities-eight-schools
#| fig-height: !expr densities_height("eight_schools")
gg_densities("eight_schools")
```

:::

</details>

## Do the chains converge?

For each session's second `mcmc()` call, the largest rank-normalised split-$\hat{R}$ and the smallest bulk ESS over each model's parameters [@vehtari2021].

```{r}
#| label: convergence
convergence <- posterior |>
  group_by(session, version, model) |>
  summarise(
    largest_rhat = max(rhat),
    smallest_ess_bulk = min(ess_bulk),
    .groups = "drop"
  )

convergence_means <- convergence |>
  group_by(model, version) |>
  summarise(
    largest_rhat = mean(largest_rhat),
    smallest_ess_bulk = mean(smallest_ess_bulk),
    .groups = "drop"
  )

# one row per session, model, version and measure, for one panel per measure
convergence_measures <- c(
  largest_rhat = "Largest R-hat",
  smallest_ess_bulk = "Smallest bulk ESS"
)
longer_convergence <- function(data) {
  data |>
    pivot_longer(
      all_of(names(convergence_measures)),
      names_to = "measure",
      values_to = "value"
    ) |>
    mutate(
      measure = factor(
        convergence_measures[measure],
        levels = convergence_measures
      )
    )
}

# the model whose mean largest R-hat is largest, on any version
rhat_by_model <- convergence_means |>
  group_by(model) |>
  summarise(largest_rhat = max(largest_rhat))
worst_model <- as.character(rhat_by_model$model[which.max(rhat_by_model$largest_rhat)])
worst_rhat <- filter(convergence_means, model == worst_model)
other_models_rhat <- max(rhat_by_model$largest_rhat[rhat_by_model$model != worst_model])
```

```{r}
#| label: plot-convergence
#| fig-cap: !expr session_caption
#| fig-height: 10
ggplot(
  longer_convergence(convergence),
  aes(x = value, y = version, colour = version)
) +
  geom_point(
    data = longer_convergence(convergence_means),
    size = 5,
    alpha = 0.25
  ) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(vars(model, measure), ncol = 2, scales = "free_x") +
  labs(x = NULL, y = NULL)
```

From the draws kept by `benchmark.R`'s `{r} section_link("benchmark.R", "mcmc-draws")` section.

Averaged over sessions, the largest R-hat is highest on `{r} worst_model`: `{r} versions_text(worst_rhat$version, worst_rhat$largest_rhat)`. On every other model it is at most `{r} format_number(other_models_rhat)`, on every version.

<details><summary>Table: every session</summary>

```{r}
#| label: table-convergence
dt_table(convergence)
```

</details>

## How it was measured

```{r}
#| label: versions
provenance |>
  distinct(version, greta_version, greta_sha, tensorflow_version, tfp_version, r_version) |>
  arrange(version) |>
  mutate(
    installed_from = versions[as.character(version)],
    # pak records a CRAN install's version in place of a commit, so only a
    # GitHub install gets a link to its commit
    installed_from = ifelse(
      grepl("^[0-9a-f]{40}$", greta_sha),
      paste0(
        "[",
        sub("@.*", "", installed_from),
        "@",
        substr(greta_sha, 1, 8),
        "](https://github.com/greta-dev/greta/commit/",
        greta_sha,
        ")"
      ),
      installed_from
    )
  ) |>
  select(
    version,
    `installed from` = installed_from,
    greta = greta_version,
    TensorFlow = tensorflow_version,
    TFP = tfp_version,
    R = r_version
  ) |>
  knitr::kable()

python_stacks <- provenance |>
  distinct(version, tensorflow_version, tfp_version) |>
  group_by(tensorflow_version, tfp_version) |>
  summarise(
    versions = paste_and(as.character(sort(version))),
    .groups = "drop"
  )
python_stack_text <- paste0(
  python_stacks$versions,
  " ran on TensorFlow ",
  python_stacks$tensorflow_version,
  " and TensorFlow Probability ",
  python_stacks$tfp_version,
  collapse = "; "
)

sampler_iterations_values <- sampler_iterations_per_draw |>
  group_by(version) |>
  summarise(values = paste(sort(unique(value)), collapse = " or "))
```

The versions are pinned to these commits in `run.R`'s `{r} section_link("run.R", "run-versions")` section. `{r} python_stack_text`.

Each version ran in `{r} n_sessions` sessions, on `{r} unique(provenance$machine)` with `{r} unique(provenance$cores)` cores. In each session, `run.R`'s `{r} section_link("run.R", "run-sessions")` section ran `benchmark.R` once per version, each in a fresh R session, with the versions in a new random order. Each session has a seed of its own, set in `benchmark.R`'s `{r} section_link("benchmark.R", "benchmark-setup")` section, and the same settings, in its `{r} section_link("benchmark.R", "benchmark-settings")` section.

<details><summary>Table: the order of each session</summary>

```{r}
#| label: table-order
runs |>
  select(session, position, version) |>
  arrange(session, position) |>
  dt_table()
```

</details>

Every call is timed with {bench}'s `mark()` and saved as `mark()` made it, so `summary()` and `autoplot()` work on it. Under each timing, one table is bench's `summary()` of every mark, with the mean and longest of its calls added, and the other has every call.

The same `mcmc()` arguments are the same work only if each version's sampler runs the same number of iterations for each draw it keeps. `benchmark.R`'s `{r} section_link("benchmark.R", "benchmark-sampler-iterations")` section counts them, and in every session they were `{r} paste(sampler_iterations_values$version, sampler_iterations_values$values, collapse = ", ")`.

<details><summary>Table: sampler iterations per kept draw, in every session</summary>

```{r}
#| label: table-sampler-iterations
sampler_iterations_per_draw |>
  select(session, version, sampler_iterations_per_draw = value) |>
  dt_table()
```

</details>

Two scripts made every result here: `{r} file_link("benchmark.R")` is what each version runs, and `{r} file_link("run.R")` runs it for each version in each session. Both in full:

```{r}
#| label: run-script
#| file: run.R
#| eval: false
#| code-summary: "run.R"
```

```{r}
#| label: benchmark-script
#| file: benchmark.R
#| eval: false
#| code-summary: "benchmark.R"
```

### The models

Four models are from greta's `inst/examples/`; the hessian model is the one greta#546 reported. Normal and log-normal distributions are written with their mean and standard deviation, as greta's `normal()` and `lognormal()` take them; $\text{Cauchy}^{+}$ is a Cauchy truncated to positive values.

::: panel-tabset

#### linear

A regression of each department's rating in `attitude` on its number of complaints. `{r} parameter_count("linear")` parameters.

$$
\begin{aligned}
\text{rating}_i &\sim \text{Normal}(\mu_i, \sigma), \quad i = 1, \dots, 30 \\
\mu_i &= \alpha + \beta \, \text{complaints}_i \\
\alpha, \beta &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Cauchy}^{+}(0, 3)
\end{aligned}
$$

```{r}
#| label: model-linear
#| eval: false
#| code-summary: "benchmark.R: model-linear"
```

#### multiple_linear

The same rating on all six of `attitude`'s other columns, $X$, through a matrix multiply. `{r} parameter_count("multiple_linear")` parameters.

$$
\begin{aligned}
\text{rating}_i &\sim \text{Normal}(\mu_i, \sigma), \quad i = 1, \dots, 30 \\
\mu_i &= \alpha + \sum_{j = 1}^{6} X_{ij} \, \beta_j \\
\alpha, \beta_j &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Cauchy}^{+}(0, 3)
\end{aligned}
$$

```{r}
#| label: model-multiple-linear
#| eval: false
#| code-summary: "benchmark.R: model-multiple-linear"
```

#### hierarchical_linear

Sepal length on sepal width in `iris`, with an offset for each species, $s[i]$ being flower $i$'s. Setosa's offset is zero and the other two share a scale, $\tau$. `{r} parameter_count("hierarchical_linear")` parameters.

$$
\begin{aligned}
\text{Sepal.Length}_i &\sim \text{Normal}(\mu_i, \sigma), \quad i = 1, \dots, 150 \\
\mu_i &= \alpha + \beta \, \text{Sepal.Width}_i + \gamma_{s[i]} \\
\gamma_1 &= 0, \qquad \gamma_2, \gamma_3 \sim \text{Normal}(0, \tau) \\
\tau &\sim \text{LogNormal}(0, 1) \\
\alpha, \beta &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Cauchy}^{+}(0, 3)
\end{aligned}
$$

```{r}
#| label: model-hierarchical-linear
#| eval: false
#| code-summary: "benchmark.R: model-hierarchical-linear"
```

#### eight_schools

The effect of coaching in eight schools, $y_j$, each measured with a known standard error, $\sigma_j$. The scale of the school effects, $\sigma_\eta$, is itself a parameter, which makes the posterior a funnel. `{r} parameter_count("eight_schools")` parameters.

$$
\begin{aligned}
y_j &\sim \text{Normal}(\theta_j, \sigma_j), \quad j = 1, \dots, 8 \\
\theta_j &= \mu + \xi \, \eta_j \\
\eta_j &\sim \text{Normal}(0, \sigma_\eta) \\
\sigma_\eta &\sim \text{InverseGamma}(1, 1) \\
\mu &\sim \text{Normal}(0, 100) \\
\xi &\sim \text{Normal}(0, 5)
\end{aligned}
$$

```{r}
#| label: model-eight-schools
#| eval: false
#| code-summary: "benchmark.R: model-eight-schools"
```

#### hessian model

Used only for `opt(hessian = TRUE)`: twenty simulated observations $y_k$, each with its own mean $b_k$, so each call takes twenty hessians. greta's `variable()` gives $b_k$ a flat prior, so it has no distribution line.

$$
y_k \sim \text{Normal}(b_k, 1), \quad k = 1, \dots, 20
$$

```{r}
#| label: model-hessian
#| eval: false
#| code-summary: "benchmark.R: model-hessian"
```

:::

## Session details

What each version reported about itself in session 1, from `benchmark.R`'s `{r} section_link("benchmark.R", "benchmark-provenance")` section.

```{r}
#| label: session-one
session_one <- runs |>
  filter(session == 1) |>
  arrange(version)

# a saved session_info() prints with sessioninfo's print method, which is
# registered only once its namespace is loaded
invisible(loadNamespace("sessioninfo"))
```

::: panel-tabset

### CRAN

<details><summary>`greta_sitrep()`</summary>

```{r}
#| label: sitrep-cran
writeLines(session_one$results[[which(session_one$version == "CRAN")]]$greta_sitrep_report)
```

</details>

<details><summary>`sessioninfo::session_info()`</summary>

```{r}
#| label: session-info-cran
session_one$results[[which(session_one$version == "CRAN")]]$session_info
```

</details>

### main

<details><summary>`greta_sitrep()`</summary>

```{r}
#| label: sitrep-main
writeLines(session_one$results[[which(session_one$version == "main")]]$greta_sitrep_report)
```

</details>

<details><summary>`sessioninfo::session_info()`</summary>

```{r}
#| label: session-info-main
session_one$results[[which(session_one$version == "main")]]$session_info
```

</details>

### #843

<details><summary>`greta_sitrep()`</summary>

```{r}
#| label: sitrep-843
writeLines(session_one$results[[which(session_one$version == "#843")]]$greta_sitrep_report)
```

</details>

<details><summary>`sessioninfo::session_info()`</summary>

```{r}
#| label: session-info-843
session_one$results[[which(session_one$version == "#843")]]$session_info
```

</details>

:::

## References

::: {#refs}
:::
