---
title: "greta#855 (tf-warmup-i547): TFP's loops, tried and not adopted"
description: "TensorFlow Probability's sample_chain() and minimize() timed against greta's own sampling loop and opt(), on the PR's branch and two models. Once traced, sample_chain() takes as long as greta's loop, and minimize() takes longer than opt(), so the PR keeps greta's own loops."
date: 2026-10-07
bibliography: ../../references.bib
---

```{r}
#| label: setup
#| code-summary: "Setup"
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(bench)
  library(here)
})

post_path <- "posts/2026-10-07-tfp-alternatives-i547"

# each `# ---- label ----` section of the scripts, shown below by its label
knitr::read_chunk(here(post_path, "benchmark.R"))
knitr::read_chunk(here(post_path, "run.R"))

# links to the scripts on GitHub, at the commit being rendered when GitHub
# Actions renders the site
github_ref <- Sys.getenv("GITHUB_SHA", "main")
github_file_url <- function(file) {
  paste0(
    "https://github.com/greta-dev/greta.benchmarks/blob/",
    github_ref,
    "/",
    post_path,
    "/",
    file
  )
}
# Quarto escapes the markdown in an inline `{r}` result unless it is wrapped
# in I(), so both link helpers return I()
file_link <- function(file) {
  I(paste0("[`", file, "`](", github_file_url(file), ")"))
}

# a link to the lines of a script's `# ---- label ----` section
section_link <- function(file, label) {
  lines <- readLines(here(post_path, file))
  headings <- grep("^# ---- .+ ----$", lines)
  start <- headings[lines[headings] == paste("# ----", label, "----")]
  end <- c(headings, length(lines) + 1)[match(start, headings) + 1] - 1
  while (end > start && !nzchar(trimws(lines[end]))) {
    end <- end - 1
  }
  I(paste0(
    "[`", label, "`](", github_file_url(file), "#L", start, "-L", end, ")"
  ))
}

theme_set(theme_bw(base_size = 13))

sessions <- list.files(
  here(post_path, "results"),
  pattern = "^session-[0-9]+[.]rds$",
  full.names = TRUE
) |>
  lapply(readRDS)
n_sessions <- length(sessions)
results <- lapply(sessions, \(session_file) session_file$results$branch)
bench_settings <- results[[1]]$bench_settings
provenance <- results[[1]]$provenance
python_version <- sub(
  " .*",
  "",
  results[[1]]$session_info$python$version_string
)
model_levels <- names(results[[1]]$sampling)

loop_levels <- c("greta", "tfp")
loop_colours <- c(greta = "#009E73", tfp = "#E69F00")

# a loop's first timed call on a model, apart from the calls after it
label_calls <- function(call) {
  factor(
    ifelse(call == 1, "first call", "later calls"),
    levels = c("first call", "later calls")
  )
}

# one row per timed call: bench keeps every call's time in the `time` column
every_call <- function(marks, ...) {
  marks |>
    as_tibble() |>
    mutate(loop = factor(as.character(expression), levels = loop_levels)) |>
    select(loop, ..., time) |>
    mutate(call = lapply(time, seq_along)) |>
    unnest(c(time, call)) |>
    mutate(seconds = as.numeric(time), calls = label_calls(call), .keep = "unused")
}

# every timed call, from every session and model
sampling_calls <- lapply(seq_along(results), function(i) {
  lapply(model_levels, function(model_name) {
    every_call(results[[i]]$sampling[[model_name]]$marks) |>
      mutate(session = sessions[[i]]$session, model = model_name)
  }) |>
    bind_rows()
}) |>
  bind_rows() |>
  mutate(model = factor(model, levels = model_levels))

opt_calls <- lapply(seq_along(results), function(i) {
  lapply(model_levels, function(model_name) {
    every_call(results[[i]]$opt[[model_name]], max_iterations) |>
      mutate(session = sessions[[i]]$session, model = model_name)
  }) |>
    bind_rows()
}) |>
  bind_rows() |>
  mutate(
    model = factor(model, levels = model_levels),
    iterations = factor(paste(max_iterations, "iterations"))
  )

# each loop's median call, and how many times as long TFP's loop takes as
# greta's
loop_medians <- function(calls, ...) {
  calls |>
    group_by(..., loop) |>
    summarise(seconds = median(seconds), .groups = "drop") |>
    pivot_wider(names_from = loop, values_from = seconds) |>
    mutate(tfp_over_greta = tfp / greta)
}
sampling_medians <- loop_medians(sampling_calls, model, calls)
sampling_ratios <- split(
  sampling_medians$tfp_over_greta,
  sampling_medians$calls
)
opt_medians <- loop_medians(opt_calls, model, max_iterations)

# three significant figures each, where format() would pad a vector to the
# decimal places of its smallest value
format_number <- function(x) {
  formatC(signif(x, 3), format = "fg", digits = 3, big.mark = ",")
}

# "a to b", or one number when the two round the same
format_range <- function(x) {
  ends <- format_number(range(x))
  if (ends[1] == ends[2]) {
    return(ends[1])
  }
  paste(ends, collapse = " to ")
}

# A raincloud for each loop: the density of its calls above, their box plot
# in the middle, and every call as a dot below. The summaries are drawn first
# and faint, so the dots sit on top.
raincloud <- list(
  ggdist::stat_halfeye(
    aes(fill = loop),
    width = 0.4,
    justification = -0.3,
    .width = 0,
    point_colour = NA,
    alpha = 0.3
  ),
  geom_boxplot(aes(fill = loop), width = 0.25, outlier.shape = NA, alpha = 0.3),
  # a fixed dot size, so a call looks the same size in every panel
  ggdist::stat_dots(
    aes(fill = loop),
    side = "bottom",
    layout = "swarm",
    justification = 1.3,
    scale = 0.75,
    binwidth = unit(2.6, "mm"),
    overflow = "compress",
    colour = NA,
    alpha = 0.6
  ),
  scale_colour_manual(values = loop_colours, guide = "none"),
  scale_fill_manual(values = loop_colours, guide = "none")
)
raincloud_caption <- "Each dot is one call, from every session. Above the dots, the curve is the density of the calls and the box marks their median and quartiles."

# a searchable table, its numbers rounded for reading
dt_table <- function(data, digits = 3) {
  table <- DT::datatable(
    data,
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 10, autoWidth = TRUE)
  )
  decimal_columns <- names(data)[vapply(data, is.double, logical(1))]
  if (length(decimal_columns) == 0) {
    return(table)
  }
  DT::formatRound(table, decimal_columns, digits)
}
```

[greta#855](https://github.com/greta-dev/greta/pull/855) runs `mcmc()`'s sampling and `opt()`'s optimiser in loops of greta's own, inside TensorFlow; [the main post](../2026-10-07-tf-warmup-i547/) times them. This post times TensorFlow Probability's loops for the same jobs, `sample_chain()` and `minimize()`, against greta's, on the PR's branch and two models. Once traced, `sample_chain()` takes `{r} format_range(sampling_ratios[["later calls"]])` times as long as greta's loop, and `minimize()` takes `{r} format_range(opt_medians$tfp_over_greta)` times as long as `opt()`. Neither is faster, so the PR keeps greta's own loops.

## How long does `sample_chain()` take?

Each session ran `hmc()` on each model for `{r} bench_settings$sampling_iterations` iterations on `{r} bench_settings$chains` chains, `{r} bench_settings$sampling_calls` times through each loop, from the state one `mcmc()` call had tuned.

```{r}
#| label: plot-sampling
#| fig-cap: !expr raincloud_caption
#| fig-height: 7
ggplot(sampling_calls, aes(x = seconds, y = loop, colour = loop)) +
  raincloud +
  scale_y_discrete(labels = c(greta = "greta's loop", tfp = "sample_chain()")) +
  facet_wrap(vars(model, calls), ncol = 2, scales = "free_x") +
  labs(x = "Seconds", y = NULL)
```

Timed by `benchmark.R`'s `{r} section_link("benchmark.R", "benchmark-sampling")` section, with the kernel from its `{r} section_link("benchmark.R", "tfp-leapfrog-kernel")` section.

Once traced, `sample_chain()` takes `{r} format_range(sampling_ratios[["later calls"]])` times as long as greta's loop. Its first call, which traces it, takes `{r} format_range(sampling_ratios[["first call"]])` times as long as greta's first call, which reuses the trace of the `mcmc()` call that tuned the sampler.

<details><summary>Table: median seconds of each loop's calls</summary>

```{r}
#| label: table-sampling
dt_table(sampling_medians, digits = 3)
```

</details>

## How long does `minimize()` take?

Each session ran `opt()` with `adam()`, and `minimize()` with the same Keras optimiser, `{r} bench_settings$opt_calls` times each on each model at `{r} paste(bench_settings$opt_max_iterations, collapse = " and ")` iterations, building and tracing the loop afresh on every call, as `opt()` does.

```{r}
#| label: plot-opt
#| fig-cap: !expr raincloud_caption
#| fig-height: 7
# what the extra iterations cost opt(): the difference between its median call
# at the most and at the fewest iterations
iteration_cost <- opt_medians |>
  group_by(model) |>
  summarise(
    fewest = greta[which.min(max_iterations)],
    extra = greta[which.max(max_iterations)] - fewest,
    .groups = "drop"
  )

ggplot(opt_calls, aes(x = seconds, y = loop, colour = loop)) +
  raincloud +
  scale_y_discrete(labels = c(greta = "opt()", tfp = "minimize()")) +
  facet_wrap(vars(model, iterations), ncol = 2, scales = "free_x") +
  labs(x = "Seconds", y = NULL)
```

Timed by `benchmark.R`'s `{r} section_link("benchmark.R", "benchmark-opt")` section.

`minimize()` takes `{r} format_range(opt_medians$tfp_over_greta)` times as long as `opt()`. Most of an `opt()` call is building and tracing its loop: on the two models, its median call takes `{r} format_range(iteration_cost$fewest)` seconds at `{r} min(bench_settings$opt_max_iterations)` iterations, and `{r} diff(range(bench_settings$opt_max_iterations))` more iterations add `{r} format_range(iteration_cost$extra)` seconds.

<details><summary>Table: median seconds of each loop's calls</summary>

```{r}
#| label: table-opt
dt_table(opt_medians, digits = 3)
```

</details>

## Do both loops sample the same posterior?

For each free parameter in each session, the difference between the two loops' posterior means, in units of its Monte Carlo standard error [@vehtari2021]: a z beyond about ±3 would say they sample different posteriors.

```{r}
#| label: plot-agreement
#| fig-cap: "Each dot is one free parameter in one session. The grey band marks z between -3 and 3."
#| fig-height: 4
loop_means <- lapply(seq_along(results), function(i) {
  lapply(model_levels, function(model_name) {
    draws <- results[[i]]$sampling[[model_name]]$draws
    lapply(loop_levels, function(loop) {
      tibble(
        session = sessions[[i]]$session,
        model = model_name,
        loop = loop,
        posterior::summarise_draws(draws[[loop]], "mean", "mcse_mean")
      )
    }) |>
      bind_rows()
  }) |>
    bind_rows()
}) |>
  bind_rows()

agreement <- loop_means |>
  pivot_wider(
    id_cols = c(session, model, variable),
    names_from = loop,
    values_from = c(mean, mcse_mean)
  ) |>
  mutate(
    model = factor(model, levels = model_levels),
    # the draws are of the free state, whose columns have no names, so
    # posterior names them ...1, ...2 and so on
    variable = sub("^[.]+", "free parameter ", variable),
    z = (mean_greta - mean_tfp) / sqrt(mcse_mean_greta^2 + mcse_mean_tfp^2)
  )

ggplot(agreement, aes(x = z, y = variable)) +
  annotate("rect", xmin = -3, xmax = 3, ymin = -Inf, ymax = Inf, alpha = 0.15) +
  geom_point(alpha = 0.6, size = 2) +
  facet_wrap(vars(model), scales = "free_y") +
  expand_limits(x = c(-4, 4)) +
  labs(x = "z", y = NULL)
```

Drawn by `benchmark.R`'s `{r} section_link("benchmark.R", "benchmark-sampling")` section, in each loop's last call.

The largest |z| of the `{r} nrow(agreement)` comparisons is `{r} format_number(max(abs(agreement$z)))`, so none says the loops sample different posteriors.

<details><summary>Table: every comparison</summary>

```{r}
#| label: table-agreement
dt_table(agreement, digits = 3)
```

</details>

## How it was measured

`{r} file_link("run.R")` ran `{r} file_link("benchmark.R")` top to bottom in `{r} n_sessions` fresh R sessions, with the settings in its `{r} section_link("benchmark.R", "benchmark-settings")` section. Each session ran one version of greta, the PR's branch at commit `5bdab31e`, which is not on GitHub, so pak installed it from a local worktree, as `{r} section_link("run.R", "run-versions")` says. Both loops of each pair ran in the same session, greta's calls before TFP's. The sessions ran on Python `{r} python_version`, TensorFlow `{r} provenance$tensorflow_version`, TensorFlow Probability `{r} provenance$tfp_version` and reticulate `{r} provenance$reticulate_version`, on `{r} provenance$machine` with `{r} provenance$cores` cores.

```{r}
#| label: run-script
#| file: run.R
#| eval: false
#| code-summary: "run.R"
```

```{r}
#| label: benchmark-script
#| file: benchmark.R
#| eval: false
#| code-summary: "benchmark.R"
```

### The TFP loops

Neither TFP loop fits greta as it stands, so each needed a subclass of a TFP class, written in R with `reticulate::PyClass()`:

- `sample_chain()` steps one kernel, built before its loop, so `hmc()`'s number of leapfrog steps would be fixed for the whole call, where greta draws it afresh every iteration. TFP's `HamiltonianMonteCarlo` reads the number from its kernel results when built with `store_parameters_in_results = TRUE`, so a wrapper kernel, `RandomLeapfrog`, writes a fresh number there every step. Both loops start from the same state, with the same step sizes and range of leapfrog steps.
- `minimize()` stops when a convergence criterion says so, and none of TFP's is greta's: stop once two successive objectives differ by at most `tolerance`. The `AbsoluteChange` criterion does that. The tolerance here is 0, so both loops run every iteration. `opt()` starts from small random values and `minimize()` from zero.

```{r}
#| label: tfp-leapfrog-kernel
#| eval: false
#| code-summary: "benchmark.R: tfp-leapfrog-kernel"
```

```{r}
#| label: benchmark-opt
#| eval: false
#| code-summary: "benchmark.R: benchmark-opt"
```

### The models

Two of greta's example models: `{r} section_link("benchmark.R", "model-linear")`, a linear regression of `attitude`'s ratings on complaints, and `{r} section_link("benchmark.R", "model-hierarchical-linear")`, which regresses `iris`'s sepal length on sepal width with an offset for each species.

```{r}
#| label: model-linear
#| eval: false
#| code-summary: "benchmark.R: model-linear"
```

```{r}
#| label: model-hierarchical-linear
#| eval: false
#| code-summary: "benchmark.R: model-hierarchical-linear"
```

## Session details

What the branch reported about itself in session 1, from `benchmark.R`'s `{r} section_link("benchmark.R", "benchmark-provenance")` section.

<details><summary>`greta_sitrep()`</summary>

```{r}
#| label: sitrep
writeLines(results[[1]]$greta_sitrep_report)
```

</details>

<details><summary>`sessioninfo::session_info()`</summary>

```{r}
#| label: session-info
# a saved session_info() prints as a plain list unless sessioninfo's print
# method is registered, which loading its namespace does
invisible(loadNamespace("sessioninfo"))
results[[1]]$session_info
```

</details>

## References
