---
title: "TITLE on CRAN, main and greta#NNN"
description: "Building models, opt() and mcmc() on greta 0.6.0, main and greta#NNN: how long each takes, and the posteriors mcmc() returns."
date: YYYY-MM-DD
bibliography: ../../references.bib
---

[greta#NNN](https://github.com/greta-dev/greta/pull/NNN) WHAT THE PULL REQUEST CHANGES, IN ONE SENTENCE. This post compares it with the version on CRAN and with main.

```{r}
#| label: setup
#| message: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(bench)
library(here)

post_path <- "posts/YYYY-MM-DD-short-name-iNNN"

# each `# ---- label ----` section of the scripts, shown below by its label
knitr::read_chunk(here(post_path, "benchmark.R"))
knitr::read_chunk(here(post_path, "run.R"))

# links to the scripts on GitHub, at the commit being rendered when GitHub
# Actions renders the site
github_ref <- Sys.getenv("GITHUB_SHA", "main")
github_file_url <- function(file) {
  paste0(
    "https://github.com/greta-dev/greta.benchmarks/blob/",
    github_ref,
    "/",
    post_path,
    "/",
    file
  )
}
# Quarto escapes the markdown in an inline `{r}` result unless it is wrapped
# in I(), so both link helpers return I()
file_link <- function(file) {
  I(paste0("[`", file, "`](", github_file_url(file), ")"))
}

# a link to the lines of a script's `# ---- label ----` section
section_link <- function(file, label) {
  lines <- readLines(here(post_path, file))
  headings <- grep("^# ---- .+ ----$", lines)
  start <- headings[lines[headings] == paste("# ----", label, "----")]
  end <- c(headings, length(lines) + 1)[match(start, headings) + 1] - 1
  while (end > start && !nzchar(trimws(lines[end]))) {
    end <- end - 1
  }
  I(paste0(
    "[`", label, "`](", github_file_url(file), "#L", start, "-L", end, ")"
  ))
}

# white panels, so the grey band in Agreement shows
theme_set(theme_bw())

sessions <- list.files(
  here(post_path, "results"),
  pattern = "^session-[0-9]+[.]rds$",
  full.names = TRUE
) |>
  lapply(readRDS)

versions <- sessions[[1]]$versions
version_levels <- names(versions)
version_colours <- setNames(
  c("#E69F00", "#56B4E9", "#009E73")[seq_along(version_levels)],
  version_levels
)
bench_settings <- sessions[[1]]$results[[1]]$bench_settings
n_sessions <- length(sessions)

# one row per session and version, holding that version's results
runs <- lapply(sessions, function(session_file) {
  tibble(
    session = session_file$session,
    version = factor(names(session_file$results), levels = version_levels),
    position = match(names(session_file$results), session_file$order),
    results = unname(session_file$results)
  )
}) |>
  bind_rows()

model_levels <- names(runs$results[[1]]$build)

# a part of the results that is a table or a value per run, stacked, with the
# run it came from
stack_tables <- function(part) {
  pieces <- lapply(seq_len(nrow(runs)), function(i) {
    values <- runs$results[[i]][[part]]
    if (!is.data.frame(values)) {
      values <- tibble(value = values)
    }
    tibble(session = runs$session[i], version = runs$version[i], values)
  })
  bind_rows(pieces)
}

# a part of the results that is a bench_mark per model, or one per run,
# stacked into one bench_mark, with the run and model each came from
stack_marks <- function(part) {
  pieces <- list()
  for (i in seq_len(nrow(runs))) {
    marks <- runs$results[[i]][[part]]
    if (inherits(marks, "bench_mark")) {
      marks <- list(marks)
    }
    for (j in seq_along(marks)) {
      mark <- marks[[j]]
      mark$session <- runs$session[i]
      mark$version <- runs$version[i]
      mark$model <- if (is.null(names(marks))) NA_character_ else names(marks)[j]
      pieces[[length(pieces) + 1]] <- mark
    }
  }
  vctrs::vec_rbind(!!!pieces)
}

provenance <- stack_tables("provenance")
sampler_iterations_per_draw <- stack_tables("sampler_iterations_per_draw")
build_marks <- stack_marks("build")
opt_marks <- stack_marks("opt")
mcmc_marks <- stack_marks("mcmc")

# one row per timed call: bench keeps every call's time in the `time` column
every_call <- function(marks) {
  marks |>
    as_tibble() |>
    select(session, version, model, time) |>
    mutate(call = lapply(time, seq_along)) |>
    unnest(c(time, call)) |>
    mutate(
      seconds = as.numeric(time),
      model = factor(model, levels = model_levels),
      .keep = "unused"
    )
}

# bench's own summary() of each mark, with the mean and longest of its calls
# added, all in seconds
bench_summary <- function(marks) {
  summaries <- summary(marks, filter_gc = FALSE)
  tibble(
    session = summaries$session,
    version = summaries$version,
    model = summaries$model,
    min = as.numeric(summaries$min),
    median = as.numeric(summaries$median),
    mean = vapply(marks$time, \(time) mean(as.numeric(time)), numeric(1)),
    max = vapply(marks$time, \(time) max(as.numeric(time)), numeric(1)),
    `itr/sec` = summaries$`itr/sec`,
    n_itr = summaries$n_itr,
    total_time = as.numeric(summaries$total_time)
  )
}

# The first call on a model traces its TensorFlow functions; later calls on the
# same model reuse them.
label_calls <- function(call) {
  last <- max(call)
  later <- if (last == 2) {
    "second call on the same model"
  } else if (last == 3) {
    "calls 2 and 3 on the same model"
  } else {
    paste0("calls 2 to ", last, " on the same model")
  }
  factor(
    ifelse(call == 1, "first call on a new model", later),
    levels = c("first call on a new model", later)
  )
}

# three significant figures each, where format() would pad a vector to the
# decimal places of its smallest value
format_number <- function(x) {
  formatC(signif(x, 3), format = "fg", digits = 3, big.mark = ",")
}

# "CRAN 10.6, main 11.2, #843 0.643"
versions_text <- function(version, value) {
  paste(paste(version, format_number(value)), collapse = ", ")
}

# "CRAN, main and #843"
paste_and <- function(x) {
  if (length(x) <= 1) {
    return(x)
  }
  paste(paste(x[-length(x)], collapse = ", "), "and", x[length(x)])
}

version_scale <- scale_colour_manual(values = version_colours, guide = "none")
version_fill <- scale_fill_manual(values = version_colours, guide = "none")
jitter <- position_jitter(width = 0, height = 0.2, seed = 2026)

# A raincloud for each version: the density of its calls above, their box plot
# in the middle, and every call as a dot below. The summaries are drawn first
# and faint, so the dots sit on top.
raincloud <- list(
  ggdist::stat_halfeye(
    aes(fill = version),
    width = 0.4,
    justification = -0.3,
    .width = 0,
    point_colour = NA,
    alpha = 0.3
  ),
  geom_boxplot(
    aes(fill = version),
    width = 0.25,
    outlier.shape = NA,
    alpha = 0.3
  ),
  # a fixed dot size, so a call looks the same size in every panel
  ggdist::stat_dots(
    aes(fill = version),
    side = "bottom",
    layout = "swarm",
    justification = 1.3,
    scale = 0.75,
    binwidth = unit(2.6, "mm"),
    overflow = "compress",
    colour = NA,
    alpha = 0.6
  ),
  version_scale,
  version_fill,
  theme_bw(base_size = 13)
)
raincloud_caption <- "Each dot is one call, from every session. Above the dots, the curve is the density of the calls and the box marks their median and quartiles."

# a searchable table, its numbers rounded for reading
dt_table <- function(data, digits = 3) {
  table <- DT::datatable(
    data,
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 10, autoWidth = TRUE)
  )
  decimal_columns <- names(data)[vapply(data, is.double, logical(1))]
  if (length(decimal_columns) == 0) {
    return(table)
  }
  DT::formatRound(table, decimal_columns, digits)
}

# a model's draws array is draws by chains by parameters
parameter_count <- function(name) {
  dim(runs$results[[1]]$mcmc_draws[[name]])[3]
}

# wide enough for a column of panels per parameter
draws_width <- function(name) {
  max(8, 1.6 * parameter_count(name))
}
```

Two scripts made every result here. `{r} file_link("benchmark.R")` is what each version runs, in a fresh R session of its own, and `{r} file_link("run.R")` runs it for each version in each session. Each section below shows the part of a script it is about, read out of the script by its `# ---- label ----` heading and linked to those lines on GitHub, so the code shown is the code that ran. All the code on this page is folded: click a fold to open it. Both scripts in full:

```{r}
#| label: run-script
#| file: run.R
#| eval: false
#| code-summary: "run.R"
```

```{r}
#| label: benchmark-script
#| file: benchmark.R
#| eval: false
#| code-summary: "benchmark.R"
```

## What was compared

```{r}
#| label: versions
provenance |>
  distinct(version, greta_version, greta_sha, tensorflow_version, tfp_version, r_version) |>
  arrange(version) |>
  mutate(
    installed_from = versions[as.character(version)],
    # pak records a CRAN install's version in place of a commit, so only a
    # GitHub install gets a link to its commit
    installed_from = ifelse(
      grepl("^[0-9a-f]{40}$", greta_sha),
      paste0(
        "[",
        sub("@.*", "", installed_from),
        "@",
        substr(greta_sha, 1, 8),
        "](https://github.com/greta-dev/greta/commit/",
        greta_sha,
        ")"
      ),
      installed_from
    )
  ) |>
  select(
    version,
    `installed from` = installed_from,
    greta = greta_version,
    TensorFlow = tensorflow_version,
    TFP = tfp_version,
    R = r_version
  ) |>
  knitr::kable()

python_stacks <- provenance |>
  distinct(version, tensorflow_version, tfp_version) |>
  group_by(tensorflow_version, tfp_version) |>
  summarise(
    versions = paste_and(as.character(sort(version))),
    .groups = "drop"
  )
python_stack_text <- paste0(
  python_stacks$versions,
  " ran on TensorFlow ",
  python_stacks$tensorflow_version,
  " and TensorFlow Probability ",
  python_stacks$tfp_version,
  collapse = "; "
)
```

`{r} python_stack_text`. Every version's `greta_sitrep()` report and `sessioninfo::session_info()` are in [Session details](#session-details).

The versions, pinned to commits, are in `run.R`'s `{r} section_link("run.R", "run-versions")` section:

```{r}
#| label: run-versions
#| eval: false
#| code-summary: "run.R: run-versions"
```

Each version ran in `{r} n_sessions` sessions, on `{r} unique(provenance$machine)` with `{r} unique(provenance$cores)` cores. In each session, `run.R`'s `{r} section_link("run.R", "run-sessions")` section ran `benchmark.R` once per version, each in a fresh R session, with the versions in a new random order:

```{r}
#| label: run-sessions
#| eval: false
#| code-summary: "run.R: run-sessions"
```

<details><summary>Table: the order of each session</summary>

```{r}
#| label: table-order
runs |>
  select(session, position, version) |>
  arrange(session, position) |>
  dt_table()
```

</details>

Each version's session ran `benchmark.R` from top to bottom, with the seed and settings in its `{r} section_link("benchmark.R", "benchmark-setup")` and `{r} section_link("benchmark.R", "benchmark-settings")` sections:

```{r}
#| label: benchmark-setup
#| eval: false
#| code-summary: "benchmark.R: benchmark-setup"
```

```{r}
#| label: benchmark-settings
#| eval: false
#| code-summary: "benchmark.R: benchmark-settings"
```

## The models

Each model is from greta's `inst/examples/`, and is defined in maths and by its section of `benchmark.R`. Normal and log-normal distributions are written with their mean and standard deviation, as greta's `normal()` and `lognormal()` take them; $\text{Cauchy}^{+}$ is a Cauchy truncated to positive values. Defining a model's function needs no greta, so these sections are run here, which lets the timing sections below print each model.

::: panel-tabset

### linear

A regression of each department's rating in `attitude` on its number of complaints. `{r} parameter_count("linear")` parameters. Its section is `{r} section_link("benchmark.R", "model-linear")`.

$$
\begin{aligned}
\text{rating}_i &\sim \text{Normal}(\mu_i, \sigma), \quad i = 1, \dots, 30 \\
\mu_i &= \alpha + \beta \, \text{complaints}_i \\
\alpha, \beta &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Cauchy}^{+}(0, 3)
\end{aligned}
$$

```{r}
#| label: model-linear
#| code-summary: "benchmark.R: model-linear"
```

### multiple_linear

The same rating on all six of `attitude`'s other columns, $X$, through a matrix multiply. `{r} parameter_count("multiple_linear")` parameters. Its section is `{r} section_link("benchmark.R", "model-multiple-linear")`.

$$
\begin{aligned}
\text{rating}_i &\sim \text{Normal}(\mu_i, \sigma), \quad i = 1, \dots, 30 \\
\mu_i &= \alpha + \sum_{j = 1}^{6} X_{ij} \, \beta_j \\
\alpha, \beta_j &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Cauchy}^{+}(0, 3)
\end{aligned}
$$

```{r}
#| label: model-multiple-linear
#| code-summary: "benchmark.R: model-multiple-linear"
```

### hierarchical_linear

Sepal length on sepal width in `iris`, with an offset for each species, $s[i]$ being flower $i$'s. Setosa's offset is zero and the other two share a scale, $\tau$. `{r} parameter_count("hierarchical_linear")` parameters. Its section is `{r} section_link("benchmark.R", "model-hierarchical-linear")`.

$$
\begin{aligned}
\text{Sepal.Length}_i &\sim \text{Normal}(\mu_i, \sigma), \quad i = 1, \dots, 150 \\
\mu_i &= \alpha + \beta \, \text{Sepal.Width}_i + \gamma_{s[i]} \\
\gamma_1 &= 0, \qquad \gamma_2, \gamma_3 \sim \text{Normal}(0, \tau) \\
\tau &\sim \text{LogNormal}(0, 1) \\
\alpha, \beta &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Cauchy}^{+}(0, 3)
\end{aligned}
$$

```{r}
#| label: model-hierarchical-linear
#| code-summary: "benchmark.R: model-hierarchical-linear"
```

### eight_schools

The effect of coaching in eight schools, $y_j$, each measured with a known standard error, $\sigma_j$. The scale of the school effects, $\sigma_\eta$, is itself a parameter, which makes the posterior a funnel. `{r} parameter_count("eight_schools")` parameters. Its section is `{r} section_link("benchmark.R", "model-eight-schools")`.

$$
\begin{aligned}
y_j &\sim \text{Normal}(\theta_j, \sigma_j), \quad j = 1, \dots, 8 \\
\theta_j &= \mu + \xi \, \eta_j \\
\eta_j &\sim \text{Normal}(0, \sigma_\eta) \\
\sigma_\eta &\sim \text{InverseGamma}(1, 1) \\
\mu &\sim \text{Normal}(0, 100) \\
\xi &\sim \text{Normal}(0, 5)
\end{aligned}
$$

```{r}
#| label: model-eight-schools
#| code-summary: "benchmark.R: model-eight-schools"
```

:::

## Sampler iterations per kept draw

The same `mcmc()` arguments are the same work only if each version's sampler runs the same number of iterations for each draw it keeps. `benchmark.R` measures this in its `{r} section_link("benchmark.R", "benchmark-sampler-iterations")` section:

```{r}
#| label: benchmark-sampler-iterations
#| eval: false
#| code-summary: "benchmark.R: benchmark-sampler-iterations"
```

```{r}
#| label: sampler-iterations
sampler_iterations_values <- sampler_iterations_per_draw |>
  group_by(version) |>
  summarise(values = paste(sort(unique(value)), collapse = " or "))
```

Sampler iterations per kept draw, in every session: `{r} paste(sampler_iterations_values$version, sampler_iterations_values$values, collapse = ", ")`.

<details><summary>Table: every session</summary>

```{r}
#| label: table-sampler-iterations
sampler_iterations_per_draw |>
  select(session, version, sampler_iterations_per_draw = value) |>
  dt_table()
```

</details>

## Speed

Every call below is timed with {bench}'s `mark()`, which keeps each call's time in its result's `time` column. `benchmark.R` saves each result as `mark()` made it, so `summary()` and `autoplot()` work on them, and each table below is bench's `summary()` of every mark, with the mean and longest of its calls added. The first `opt()` or `mcmc()` call on a new model traces the model's TensorFlow functions; later calls on the same model reuse them. So the plots split the first call from the later ones.

Each timing section starts by printing the model it times. This page runs only that line, which prints the model's function below the folded code; knitr shows the lines it does not run here with `##` in front of them.

::: panel-tabset

### Building the model

Calling each model's function: creating its greta arrays and calling `model()`. The timed code for each model is its `bench-build-` section of `benchmark.R`: `{r} section_link("benchmark.R", "bench-build-linear")`, `{r} section_link("benchmark.R", "bench-build-multiple-linear")`, `{r} section_link("benchmark.R", "bench-build-hierarchical-linear")` and `{r} section_link("benchmark.R", "bench-build-eight-schools")`.

::: panel-tabset

#### linear

```{r}
#| label: bench-build-linear
#| eval: !expr 1
#| code-summary: "benchmark.R: bench-build-linear"
```

#### multiple_linear

```{r}
#| label: bench-build-multiple-linear
#| eval: !expr 1
#| code-summary: "benchmark.R: bench-build-multiple-linear"
```

#### hierarchical_linear

```{r}
#| label: bench-build-hierarchical-linear
#| eval: !expr 1
#| code-summary: "benchmark.R: bench-build-hierarchical-linear"
```

#### eight_schools

```{r}
#| label: bench-build-eight-schools
#| eval: !expr 1
#| code-summary: "benchmark.R: bench-build-eight-schools"
```

:::

```{r}
#| label: build-summaries
build <- every_call(build_marks)
build_summary <- bench_summary(build_marks)

build_medians <- build |>
  group_by(version) |>
  summarise(seconds = median(seconds))
```

Median seconds per call, over all models and sessions: `{r} versions_text(build_medians$version, build_medians$seconds)`.

```{r}
#| label: plot-build
#| fig-cap: !expr raincloud_caption
#| fig-height: 13
ggplot(build, aes(x = seconds, y = version, colour = version)) +
  raincloud +
  facet_wrap(~model, ncol = 1, scales = "free_x") +
  labs(x = "Seconds", y = NULL)
```

<details><summary>Table: bench's summary() of every mark</summary>

```{r}
#| label: table-build-summary
dt_table(build_summary, digits = 4)
```

</details>

<details><summary>Table: every call</summary>

```{r}
#| label: table-build
dt_table(build, digits = 4)
```

</details>

### opt()

`opt()`, `{r} bench_settings$opt_calls` times on a model built for it. The timed code for each model is its `bench-opt-` section of `benchmark.R`, after the note in `{r} section_link("benchmark.R", "bench-opt")`: `{r} section_link("benchmark.R", "bench-opt-linear")`, `{r} section_link("benchmark.R", "bench-opt-multiple-linear")`, `{r} section_link("benchmark.R", "bench-opt-hierarchical-linear")` and `{r} section_link("benchmark.R", "bench-opt-eight-schools")`.

```{r}
#| label: bench-opt
#| eval: false
#| code-summary: "benchmark.R: bench-opt"
```

::: panel-tabset

#### linear

```{r}
#| label: bench-opt-linear
#| eval: !expr 1
#| code-summary: "benchmark.R: bench-opt-linear"
```

#### multiple_linear

```{r}
#| label: bench-opt-multiple-linear
#| eval: !expr 1
#| code-summary: "benchmark.R: bench-opt-multiple-linear"
```

#### hierarchical_linear

```{r}
#| label: bench-opt-hierarchical-linear
#| eval: !expr 1
#| code-summary: "benchmark.R: bench-opt-hierarchical-linear"
```

#### eight_schools

```{r}
#| label: bench-opt-eight-schools
#| eval: !expr 1
#| code-summary: "benchmark.R: bench-opt-eight-schools"
```

:::

```{r}
#| label: opt-summaries
opt_timings <- every_call(opt_marks) |>
  mutate(calls = label_calls(call))
opt_summary <- bench_summary(opt_marks)

opt_medians <- opt_timings |>
  group_by(version) |>
  summarise(seconds = median(seconds))
```

Median seconds per call, over all models, calls and sessions: `{r} versions_text(opt_medians$version, opt_medians$seconds)`.

```{r}
#| label: plot-opt
#| fig-cap: !expr raincloud_caption
#| fig-height: 16
ggplot(opt_timings, aes(x = seconds, y = version, colour = version)) +
  raincloud +
  facet_wrap(vars(model, calls), ncol = 2, scales = "free_x") +
  labs(x = "Seconds", y = NULL)
```

<details><summary>Table: bench's summary() of every mark</summary>

```{r}
#| label: table-opt-summary
dt_table(opt_summary, digits = 4)
```

</details>

<details><summary>Table: every call</summary>

```{r}
#| label: table-opt
dt_table(opt_timings, digits = 4)
```

</details>

### mcmc()

`mcmc()` twice on each model: first on a new model, then on the same model again, which reuses the functions the first call traced. The timed code for each model is its `bench-mcmc-` section of `benchmark.R`, after the note in `{r} section_link("benchmark.R", "bench-mcmc")`: `{r} section_link("benchmark.R", "bench-mcmc-linear")`, `{r} section_link("benchmark.R", "bench-mcmc-multiple-linear")`, `{r} section_link("benchmark.R", "bench-mcmc-hierarchical-linear")` and `{r} section_link("benchmark.R", "bench-mcmc-eight-schools")`.

```{r}
#| label: bench-mcmc
#| eval: false
#| code-summary: "benchmark.R: bench-mcmc"
```

::: panel-tabset

#### linear

```{r}
#| label: bench-mcmc-linear
#| eval: !expr 1
#| code-summary: "benchmark.R: bench-mcmc-linear"
```

#### multiple_linear

```{r}
#| label: bench-mcmc-multiple-linear
#| eval: !expr 1
#| code-summary: "benchmark.R: bench-mcmc-multiple-linear"
```

#### hierarchical_linear

```{r}
#| label: bench-mcmc-hierarchical-linear
#| eval: !expr 1
#| code-summary: "benchmark.R: bench-mcmc-hierarchical-linear"
```

#### eight_schools

```{r}
#| label: bench-mcmc-eight-schools
#| eval: !expr 1
#| code-summary: "benchmark.R: bench-mcmc-eight-schools"
```

:::

```{r}
#| label: mcmc-summaries
mcmc_timings <- every_call(mcmc_marks) |>
  mutate(calls = label_calls(call))
mcmc_summary <- bench_summary(mcmc_marks)

mcmc_medians <- mcmc_timings |>
  group_by(version) |>
  summarise(seconds = median(seconds))
```

Median seconds per call, over all models, calls and sessions: `{r} versions_text(mcmc_medians$version, mcmc_medians$seconds)`.

```{r}
#| label: plot-mcmc
#| fig-cap: !expr raincloud_caption
#| fig-height: 16
ggplot(mcmc_timings, aes(x = seconds, y = version, colour = version)) +
  raincloud +
  facet_wrap(vars(model, calls), ncol = 2, scales = "free_x") +
  labs(x = "Seconds", y = NULL)
```

<details><summary>Table: bench's summary() of every mark</summary>

```{r}
#| label: table-mcmc-summary
dt_table(mcmc_summary, digits = 4)
```

</details>

<details><summary>Table: every call</summary>

```{r}
#| label: table-mcmc
dt_table(mcmc_timings, digits = 4)
```

</details>

:::

## Efficiency

The posterior checks from here on use every draw of every chain from each model's second `mcmc()` call, which `benchmark.R`'s `{r} section_link("benchmark.R", "mcmc-draws")` section keeps:

```{r}
#| label: mcmc-draws
#| eval: false
#| code-summary: "benchmark.R: mcmc-draws"
```

Each model's posterior summary, in each session, from those draws [@vehtari2021]:

```{r}
#| label: posterior-summaries
posterior <- lapply(seq_len(nrow(runs)), function(i) {
  draws <- runs$results[[i]]$mcmc_draws
  lapply(names(draws), function(model_name) {
    tibble(
      session = runs$session[i],
      version = runs$version[i],
      model = model_name,
      posterior::summarise_draws(
        draws[[model_name]],
        "mean",
        "sd",
        "mcse_mean",
        "rhat",
        "ess_bulk",
        "ess_tail"
      )
    )
  }) |>
    bind_rows()
}) |>
  bind_rows() |>
  mutate(model = factor(model, levels = model_levels))
```

Efficiency is the bulk effective sample size (ESS) per second of sampling, the time-normalised ESS of @girolami2011, for the parameter with the smallest bulk ESS, as @hoffman2014 report, and for the median parameter. The draws are the second call's, so the seconds are the second call's, warmup included.

```{r}
#| label: efficiency
second_calls <- mcmc_timings |>
  filter(call == 2) |>
  select(session, version, model, seconds)

efficiency <- posterior |>
  group_by(session, version, model) |>
  summarise(
    smallest_ess_bulk = min(ess_bulk),
    median_ess_bulk = median(ess_bulk),
    .groups = "drop"
  ) |>
  left_join(second_calls, by = c("session", "version", "model")) |>
  mutate(
    smallest_ess_per_second = smallest_ess_bulk / seconds,
    median_ess_per_second = median_ess_bulk / seconds
  )

efficiency_means <- efficiency |>
  group_by(model, version) |>
  summarise(
    smallest_ess_per_second = mean(smallest_ess_per_second),
    median_ess_per_second = mean(median_ess_per_second),
    .groups = "drop"
  )

session_caption <- "Each small dot is one session's second call. The large faint dot behind them is their mean."
```

::: panel-tabset

### Smallest bulk ESS per second

```{r}
#| label: plot-efficiency-smallest
#| fig-cap: !expr session_caption
#| fig-height: 6
ggplot(efficiency, aes(x = smallest_ess_per_second, y = version, colour = version)) +
  geom_point(data = efficiency_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(~model, ncol = 1, scales = "free_x") +
  labs(x = "Smallest bulk ESS per second", y = NULL)
```

### Median bulk ESS per second

```{r}
#| label: plot-efficiency-median
#| fig-cap: !expr session_caption
#| fig-height: 6
ggplot(efficiency, aes(x = median_ess_per_second, y = version, colour = version)) +
  geom_point(data = efficiency_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(~model, ncol = 1, scales = "free_x") +
  labs(x = "Median bulk ESS per second", y = NULL)
```

:::

<details><summary>Table: mean ESS per second</summary>

```{r}
#| label: table-efficiency-means
dt_table(efficiency_means, digits = 1)
```

</details>

<details><summary>Table: every session</summary>

```{r}
#| label: table-efficiency
dt_table(efficiency, digits = 1)
```

</details>

## Convergence

For each session's second `mcmc()` call, the largest rank-normalised split-$\hat{R}$ and the smallest bulk ESS over the model's parameters [@vehtari2021].

```{r}
#| label: convergence
convergence <- posterior |>
  group_by(session, version, model) |>
  summarise(
    largest_rhat = max(rhat),
    smallest_ess_bulk = min(ess_bulk),
    .groups = "drop"
  )

convergence_means <- convergence |>
  group_by(model, version) |>
  summarise(
    largest_rhat = mean(largest_rhat),
    smallest_ess_bulk = mean(smallest_ess_bulk),
    .groups = "drop"
  )
```

::: panel-tabset

### Largest R-hat

```{r}
#| label: plot-rhat
#| fig-cap: !expr session_caption
#| fig-height: 6
ggplot(convergence, aes(x = largest_rhat, y = version, colour = version)) +
  geom_point(data = convergence_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(~model, ncol = 1, scales = "free_x") +
  labs(x = "Largest R-hat", y = NULL)
```

### Smallest bulk ESS

```{r}
#| label: plot-ess
#| fig-cap: !expr session_caption
#| fig-height: 6
ggplot(convergence, aes(x = smallest_ess_bulk, y = version, colour = version)) +
  geom_point(data = convergence_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(~model, ncol = 1, scales = "free_x") +
  labs(x = "Smallest bulk ESS", y = NULL)
```

:::

<details><summary>Table: every session</summary>

```{r}
#| label: table-convergence
dt_table(convergence)
```

</details>

## Agreement

Whether two versions estimate the same posterior means. The difference between two versions' means is divided by the Monte Carlo standard error (MCSE) of that difference, so a difference can be told apart from Monte Carlo noise [@vehtari2021; @magnusson2025]. For each parameter, each version's means and MCSEs are pooled over its sessions, and

$$
z = \frac{\bar\theta_{\text{b}} - \bar\theta_{\text{a}}}{\sqrt{\text{se}^2_{\text{a}} + \text{se}^2_{\text{b}}}}.
$$

Each session has a seed of its own, so the sessions are independent, and a version's pooled standard error is $\sqrt{\sum \text{mcse}^2} / n$ over its $n$ sessions. If both versions sample the same posterior, each $z$ is close to a standard normal draw. Two versions that both respect `set.seed()` return the same draws from the same seed, and so a $z$ of zero.

```{r}
#| label: agreement
pooled <- posterior |>
  group_by(version, model, variable) |>
  summarise(
    sessions = n(),
    mean = mean(mean),
    se = sqrt(sum(mcse_mean^2)) / n(),
    .groups = "drop"
  )

pairs <- utils::combn(version_levels, 2)
agreement <- lapply(seq_len(ncol(pairs)), function(i) {
  reference <- filter(pooled, version == pairs[1, i])
  compared <- filter(pooled, version == pairs[2, i])
  inner_join(reference, compared, by = c("model", "variable"), suffix = c("_a", "_b")) |>
    mutate(
      comparison = paste(pairs[2, i], "vs", pairs[1, i]),
      z = (mean_b - mean_a) / sqrt(se_a^2 + se_b^2)
    )
}) |>
  bind_rows() |>
  mutate(comparison = factor(comparison, levels = unique(comparison)))

# the band all of a comparison's z fall inside with probability 0.95 when both
# versions sample the same posterior: a Bonferroni bound over its parameters
parameters_compared <- n_distinct(paste(agreement$model, agreement$variable))
band <- qnorm(1 - 0.05 / (2 * parameters_compared))

outside_band <- agreement |>
  group_by(comparison) |>
  summarise(outside = sum(abs(z) > band), largest_abs_z = max(abs(z)))
```

The grey band is $\pm$ `{r} format_number(band)`: if two versions sample the same posterior, all `{r} parameters_compared` parameters' $z$ fall inside it with probability at least 0.95, by a Bonferroni bound. Parameters outside it, per comparison: `{r} paste(outside_band$comparison, outside_band$outside, collapse = ", ")`.

```{r}
#| label: plot-agreement
#| fig-cap: "Each dot is one parameter. The grey band is where every parameter's z falls, with probability at least 0.95, when both versions sample the same posterior."
#| fig-height: 6
ggplot(agreement, aes(x = z, y = model)) +
  annotate(
    "rect",
    xmin = -band,
    xmax = band,
    ymin = -Inf,
    ymax = Inf,
    fill = "grey80",
    alpha = 0.5
  ) +
  geom_point(position = jitter, alpha = 0.7) +
  facet_wrap(~comparison, ncol = 1) +
  labs(x = "z", y = NULL)
```

<details><summary>Table: largest |z| per comparison and model</summary>

```{r}
#| label: table-agreement-summary
agreement |>
  group_by(comparison, model) |>
  summarise(
    parameters = n(),
    largest_abs_z = max(abs(z)),
    parameter = variable[which.max(abs(z))],
    .groups = "drop"
  ) |>
  dt_table(digits = 2)
```

</details>

<details><summary>Table: every parameter</summary>

```{r}
#| label: table-agreement
agreement |>
  select(comparison, model, variable, mean_a, se_a, mean_b, se_b, z) |>
  dt_table()
```

</details>

## Traceplots

Every draw of each chain from session 1's second `mcmc()` call on each model, numbered from the end of warmup, with a row of panels for each version.

```{r}
#| label: traceplots
# every draw of every run, one row per draw and parameter
draws <- lapply(seq_len(nrow(runs)), function(i) {
  model_draws <- runs$results[[i]]$mcmc_draws
  lapply(names(model_draws), function(model_name) {
    posterior::as_draws_df(model_draws[[model_name]]) |>
      as.data.frame() |>
      pivot_longer(
        -c(.chain, .iteration, .draw),
        names_to = "variable",
        values_to = "value"
      ) |>
      mutate(
        session = runs$session[i],
        version = runs$version[i],
        model = model_name
      )
  }) |>
    bind_rows()
}) |>
  bind_rows()

gg_trace <- function(name) {
  model_draws <- draws |>
    filter(session == 1, model == name) |>
    mutate(variable = factor(variable, levels = unique(variable)))
  # each parameter's panels share its range across the versions, so they can
  # be compared down a column
  parameter_ranges <- model_draws |>
    group_by(variable) |>
    reframe(value = range(value)) |>
    crossing(version = unique(model_draws$version)) |>
    mutate(.iteration = 1L)
  ggplot(model_draws, aes(x = .iteration, y = value, colour = factor(.chain))) +
    geom_line(linewidth = 0.2) +
    geom_blank(
      data = parameter_ranges,
      aes(x = .iteration, y = value),
      inherit.aes = FALSE
    ) +
    facet_wrap(
      vars(version, variable),
      ncol = n_distinct(model_draws$variable),
      scales = "free_y"
    ) +
    labs(x = "Draw", y = NULL, colour = "Chain")
}
```

::: panel-tabset

### linear

```{r}
#| label: plot-trace-linear
#| fig-height: 6
#| fig-width: !expr draws_width("linear")
#| column: page
gg_trace("linear")
```

### multiple_linear

```{r}
#| label: plot-trace-multiple-linear
#| fig-height: 6
#| fig-width: !expr draws_width("multiple_linear")
#| column: page
gg_trace("multiple_linear")
```

### hierarchical_linear

```{r}
#| label: plot-trace-hierarchical-linear
#| fig-height: 6
#| fig-width: !expr draws_width("hierarchical_linear")
#| column: page
gg_trace("hierarchical_linear")
```

### eight_schools

```{r}
#| label: plot-trace-eight-schools
#| fig-height: 6
#| fig-width: !expr draws_width("eight_schools")
#| column: page
gg_trace("eight_schools")
```

:::

## Posterior densities

Every draw of every chain, from each session's second `mcmc()` call, pooled over sessions, with every version's density overlaid in one panel per parameter.

```{r}
#| label: densities
# every version's density in one panel per parameter, each with its own line
# type as well as colour, so a line drawn on top of an identical one still shows
gg_densities <- function(name) {
  draws |>
    filter(model == name) |>
    mutate(variable = factor(variable, levels = unique(variable))) |>
    ggplot(aes(x = value, colour = version, linetype = version)) +
    geom_density(linewidth = 0.7) +
    scale_colour_manual(values = version_colours) +
    facet_wrap(~variable, scales = "free", ncol = 4) +
    labs(x = NULL, y = NULL, colour = NULL, linetype = NULL) +
    theme(legend.position = "top")
}

densities_height <- function(name) {
  1.5 + 2 * ceiling(parameter_count(name) / 4)
}

# whether two versions returned the same draws in every session
draws_identical <- function(pair) {
  all(vapply(
    unique(runs$session),
    function(session_number) {
      first <- runs$results[[which(runs$session == session_number & runs$version == pair[1])]]
      second <- runs$results[[which(runs$session == session_number & runs$version == pair[2])]]
      identical(first$mcmc_draws, second$mcmc_draws)
    },
    logical(1)
  ))
}
identical_pairs <- Filter(
  draws_identical,
  utils::combn(version_levels, 2, simplify = FALSE)
)
identical_text <- if (length(identical_pairs) == 0) {
  ""
} else {
  paste0(
    vapply(identical_pairs, \(pair) paste(pair, collapse = "'s and "), character(1)),
    "'s draws are identical in every session, so their lines lie on top of each other.",
    collapse = " "
  )
}
```

`{r} identical_text`

::: panel-tabset

### linear

```{r}
#| label: plot-densities-linear
#| fig-height: !expr densities_height("linear")
gg_densities("linear")
```

### multiple_linear

```{r}
#| label: plot-densities-multiple-linear
#| fig-height: !expr densities_height("multiple_linear")
gg_densities("multiple_linear")
```

### hierarchical_linear

```{r}
#| label: plot-densities-hierarchical-linear
#| fig-height: !expr densities_height("hierarchical_linear")
gg_densities("hierarchical_linear")
```

### eight_schools

```{r}
#| label: plot-densities-eight-schools
#| fig-height: !expr densities_height("eight_schools")
gg_densities("eight_schools")
```

:::

## Session details

What each version reported about itself in session 1, from `benchmark.R`'s `{r} section_link("benchmark.R", "benchmark-provenance")` section:

```{r}
#| label: benchmark-provenance
#| eval: false
#| code-summary: "benchmark.R: benchmark-provenance"
```

```{r}
#| label: session-one
session_one <- runs |>
  filter(session == 1) |>
  arrange(version)
```

::: panel-tabset

### CRAN

<details><summary>`greta_sitrep()`</summary>

```{r}
#| label: sitrep-cran
writeLines(session_one$results[[which(session_one$version == "CRAN")]]$greta_sitrep_report)
```

</details>

<details><summary>`sessioninfo::session_info()`</summary>

```{r}
#| label: session-info-cran
session_one$results[[which(session_one$version == "CRAN")]]$session_info
```

</details>

### main

<details><summary>`greta_sitrep()`</summary>

```{r}
#| label: sitrep-main
writeLines(session_one$results[[which(session_one$version == "main")]]$greta_sitrep_report)
```

</details>

<details><summary>`sessioninfo::session_info()`</summary>

```{r}
#| label: session-info-main
session_one$results[[which(session_one$version == "main")]]$session_info
```

</details>

### #NNN

<details><summary>`greta_sitrep()`</summary>

```{r}
#| label: sitrep-nnn
writeLines(session_one$results[[which(session_one$version == "#NNN")]]$greta_sitrep_report)
```

</details>

<details><summary>`sessioninfo::session_info()`</summary>

```{r}
#| label: session-info-nnn
session_one$results[[which(session_one$version == "#NNN")]]$session_info
```

</details>

:::

## References

::: {#refs}
:::
