---
title: "TITLE on CRAN, main and greta#NNN"
description: "Building models, opt() and mcmc() on greta 0.6.0, main and greta#NNN: how long each takes, and the posteriors mcmc() returns."
date: YYYY-MM-DD
bibliography: ../../references.bib
resources:
  - run.R
  - benchmark.R
---

[greta#NNN](https://github.com/greta-dev/greta/pull/NNN) WHAT THE PULL REQUEST CHANGES, IN ONE SENTENCE. This post compares it with the version on CRAN and with main.

Two scripts made every result here. [`benchmark.R`](benchmark.R) is what each version runs, in a fresh R session of its own, and [`run.R`](run.R) runs it for each version in each session. Each section below shows the part of a script it is about, read out of the script by its `# ---- label ----` heading, so the code shown is the code that ran. Both scripts are in full at the end.

```{r}
#| label: setup
#| message: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(bench)

# each `# ---- label ----` section of the scripts, shown below by its label
knitr::read_chunk("benchmark.R")
knitr::read_chunk("run.R")

# white panels, so the grey band in Agreement shows
theme_set(theme_bw())

sessions <- list.files(
  "results",
  pattern = "^session-[0-9]+[.]rds$",
  full.names = TRUE
) |>
  lapply(readRDS)

versions <- sessions[[1]]$versions
version_levels <- names(versions)
version_colours <- setNames(
  c("#E69F00", "#56B4E9", "#009E73")[seq_along(version_levels)],
  version_levels
)
settings <- sessions[[1]]$results[[1]]$settings
n_sessions <- length(sessions)

# `part` from every version in every session, stacked, with the session and
# version it came from; bench_mark objects stay bench_mark objects
collect <- function(part) {
  pieces <- list()
  for (session in sessions) {
    for (version in names(session$results)) {
      values <- session$results[[version]][[part]]
      if (!is.data.frame(values)) {
        values <- data.frame(value = values)
      }
      values$session <- session$session
      values$version <- factor(version, levels = version_levels)
      values$position <- match(version, session$order)
      pieces[[length(pieces) + 1]] <- values
    }
  }
  vctrs::vec_rbind(!!!pieces)
}

provenance <- collect("provenance")
sampler_iterations_per_draw <- collect("sampler_iterations_per_draw")
build_marks <- collect("build")
opt_marks <- collect("opt")
mcmc_marks <- collect("mcmc")
model_levels <- unique(build_marks$model)

# one row per timed call: bench keeps every call's time in the `time` column
every_call <- function(marks) {
  marks |>
    as_tibble() |>
    select(session, version, any_of("model"), time) |>
    mutate(call = lapply(time, seq_along)) |>
    unnest(c(time, call)) |>
    mutate(seconds = as.numeric(time), .keep = "unused") |>
    mutate(across(any_of("model"), \(model) factor(model, levels = model_levels)))
}

# The first call on a model traces its TensorFlow functions; later calls on the
# same model reuse them.
label_calls <- function(call) {
  last <- max(call)
  later <- if (last == 2) {
    "second call on the same model"
  } else if (last == 3) {
    "calls 2 and 3 on the same model"
  } else {
    paste0("calls 2 to ", last, " on the same model")
  }
  factor(
    ifelse(call == 1, "first call on a new model", later),
    levels = c("first call on a new model", later)
  )
}

# three significant figures each, where format() would pad a vector to the
# decimal places of its smallest value
format_number <- function(x) {
  formatC(signif(x, 3), format = "fg", digits = 3, big.mark = ",")
}

# "CRAN 10.6, main 11.2, #843 0.643"
versions_text <- function(version, value) {
  paste(paste(version, format_number(value)), collapse = ", ")
}

# "CRAN, main and #843"
paste_and <- function(x) {
  if (length(x) <= 1) {
    return(x)
  }
  paste(paste(x[-length(x)], collapse = ", "), "and", x[length(x)])
}

version_scale <- scale_colour_manual(values = version_colours, guide = "none")
jitter <- position_jitter(width = 0, height = 0.2, seed = 2026)
call_caption <- "Each small dot is one call, from every session. The large faint dot behind them is their mean."

# a searchable table, its numbers rounded for reading
dt_table <- function(data, digits = 3) {
  table <- DT::datatable(
    data,
    rownames = FALSE,
    filter = "top",
    options = list(pageLength = 10, autoWidth = TRUE)
  )
  decimal_columns <- names(data)[vapply(data, is.double, logical(1))]
  if (length(decimal_columns) == 0) {
    return(table)
  }
  DT::formatRound(table, decimal_columns, digits)
}

# every posterior summary of a model has one row per parameter
parameter_count <- function(name) {
  nrow(mcmc_marks$posterior[[match(name, mcmc_marks$model)]])
}
```

## What was compared

```{r}
#| label: versions
provenance |>
  distinct(version, greta_version, greta_sha, tensorflow_version, tfp_version, r_version) |>
  arrange(version) |>
  mutate(
    installed_from = versions[as.character(version)],
    # pak records a CRAN install's version in place of a commit, so only a
    # GitHub install gets a link to its commit
    installed_from = ifelse(
      grepl("^[0-9a-f]{40}$", greta_sha),
      paste0(
        "[",
        sub("@.*", "", installed_from),
        "@",
        substr(greta_sha, 1, 8),
        "](https://github.com/greta-dev/greta/commit/",
        greta_sha,
        ")"
      ),
      installed_from
    )
  ) |>
  select(
    version,
    `installed from` = installed_from,
    greta = greta_version,
    TensorFlow = tensorflow_version,
    TFP = tfp_version,
    R = r_version
  ) |>
  knitr::kable()

python_stacks <- provenance |>
  distinct(version, tensorflow_version, tfp_version) |>
  group_by(tensorflow_version, tfp_version) |>
  summarise(
    versions = paste_and(as.character(sort(version))),
    .groups = "drop"
  )
python_stack_text <- paste0(
  python_stacks$versions,
  " ran on TensorFlow ",
  python_stacks$tensorflow_version,
  " and TensorFlow Probability ",
  python_stacks$tfp_version,
  collapse = "; "
)
```

`{r} python_stack_text`.

The versions, pinned to commits, from `run.R`'s `run-versions` section:

```{r}
#| label: run-versions
#| eval: false
#| code-fold: false
```

Each version ran in `{r} n_sessions` sessions, on `{r} unique(provenance$machine)` with `{r} unique(provenance$cores)` cores. In each session, `run.R`'s `run-sessions` section ran `benchmark.R` once per version, each in a fresh R session, with the versions in a new random order:

```{r}
#| label: run-sessions
#| eval: false
#| code-fold: false
```

<details><summary>Table: the order of each session</summary>

```{r}
#| label: table-order
sessions |>
  lapply(\(session) {
    data.frame(
      session = session$session,
      position = seq_along(session$order),
      version = session$order
    )
  }) |>
  bind_rows() |>
  dt_table()
```

</details>

Each version's session ran `benchmark.R` from top to bottom, with these settings, from its `benchmark-settings` section:

```{r}
#| label: benchmark-settings
#| eval: false
#| code-fold: false
```

## The models

Each model is from greta's `inst/examples/`, and is defined in maths and by its section of `benchmark.R`. Normal and log-normal distributions are written with their mean and standard deviation, as greta's `normal()` and `lognormal()` take them; $\text{Cauchy}^{+}$ is a Cauchy truncated to positive values.

::: panel-tabset

### linear

A regression of each department's rating in `attitude` on its number of complaints. `{r} parameter_count("linear")` parameters.

$$
\begin{aligned}
\text{rating}_i &\sim \text{Normal}(\mu_i, \sigma), \quad i = 1, \dots, 30 \\
\mu_i &= \alpha + \beta \, \text{complaints}_i \\
\alpha, \beta &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Cauchy}^{+}(0, 3)
\end{aligned}
$$

```{r}
#| label: model-linear
#| eval: false
#| code-fold: false
```

### multiple_linear

The same rating on all six of `attitude`'s other columns, $X$, through a matrix multiply. `{r} parameter_count("multiple_linear")` parameters.

$$
\begin{aligned}
\text{rating}_i &\sim \text{Normal}(\mu_i, \sigma), \quad i = 1, \dots, 30 \\
\mu_i &= \alpha + \sum_{j = 1}^{6} X_{ij} \, \beta_j \\
\alpha, \beta_j &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Cauchy}^{+}(0, 3)
\end{aligned}
$$

```{r}
#| label: model-multiple-linear
#| eval: false
#| code-fold: false
```

### hierarchical_linear

Sepal length on sepal width in `iris`, with an offset for each species, $s[i]$ being flower $i$'s. Setosa's offset is zero and the other two share a scale, $\tau$. `{r} parameter_count("hierarchical_linear")` parameters.

$$
\begin{aligned}
\text{Sepal.Length}_i &\sim \text{Normal}(\mu_i, \sigma), \quad i = 1, \dots, 150 \\
\mu_i &= \alpha + \beta \, \text{Sepal.Width}_i + \gamma_{s[i]} \\
\gamma_1 &= 0, \qquad \gamma_2, \gamma_3 \sim \text{Normal}(0, \tau) \\
\tau &\sim \text{LogNormal}(0, 1) \\
\alpha, \beta &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Cauchy}^{+}(0, 3)
\end{aligned}
$$

```{r}
#| label: model-hierarchical-linear
#| eval: false
#| code-fold: false
```

### eight_schools

The effect of coaching in eight schools, $y_j$, each measured with a known standard error, $\sigma_j$. The scale of the school effects, $\sigma_\eta$, is itself a parameter, which makes the posterior a funnel. `{r} parameter_count("eight_schools")` parameters.

$$
\begin{aligned}
y_j &\sim \text{Normal}(\theta_j, \sigma_j), \quad j = 1, \dots, 8 \\
\theta_j &= \mu + \xi \, \eta_j \\
\eta_j &\sim \text{Normal}(0, \sigma_\eta) \\
\sigma_\eta &\sim \text{InverseGamma}(1, 1) \\
\mu &\sim \text{Normal}(0, 100) \\
\xi &\sim \text{Normal}(0, 5)
\end{aligned}
$$

```{r}
#| label: model-eight-schools
#| eval: false
#| code-fold: false
```

### cjs

A Cormack-Jolly-Seber capture-recapture model: survival, $\phi_t$, and recapture, $p_t$, on each of 20 occasions, from the simulated capture histories $y_{it}$ of 100 animals. Animal $i$ is first caught on occasion $f_i$ and last on $l_i$, so it is known to be alive on every occasion between. $\chi_t$ is the probability that an animal alive on occasion $t$ is never seen again, built by a recursion over the 19 earlier occasions. `{r} parameter_count("cjs")` parameters.

$$
\begin{aligned}
\phi_t, \, p_t &\sim \text{Beta}(1, 1), \quad t = 1, \dots, 20 \\
\chi_{20} &= 1, \qquad \chi_t = (1 - \phi_t) + \phi_t \, (1 - p_{t + 1}) \, \chi_{t + 1} \\
1 &\sim \text{Bernoulli}(\phi_{t - 1}), \quad f_i < t \le l_i \\
y_{it} &\sim \text{Bernoulli}(p_t), \quad f_i < t \le l_i \\
1 &\sim \text{Bernoulli}(\chi_{l_i}), \quad \text{if } l_i < 20
\end{aligned}
$$

```{r}
#| label: model-cjs
#| eval: false
#| code-fold: false
```

:::

## Sampler iterations per kept draw

The same `mcmc()` arguments are the same work only if each version's sampler runs the same number of iterations for each draw it keeps. `benchmark.R` measures this in its `benchmark-sampler-iterations` section:

```{r}
#| label: benchmark-sampler-iterations
#| eval: false
#| code-fold: false
```

```{r}
#| label: sampler-iterations
sampler_iterations_values <- sampler_iterations_per_draw |>
  group_by(version) |>
  summarise(values = paste(sort(unique(value)), collapse = " or "))
```

Sampler iterations per kept draw, in every session: `{r} paste(sampler_iterations_values$version, sampler_iterations_values$values, collapse = ", ")`.

<details><summary>Table: every session</summary>

```{r}
#| label: table-sampler-iterations
sampler_iterations_per_draw |>
  select(session, version, sampler_iterations_per_draw = value) |>
  dt_table()
```

</details>

## Speed

Every call below is timed with {bench}, which keeps each call's time in its result's `time` column. `benchmark.R` saves those results as {bench} made them, so `summary()` and `autoplot()` work on them. The first `opt()` or `mcmc()` call on a new model traces the model's TensorFlow functions; later calls on the same model reuse them. So the plots split the first call from the later ones.

::: panel-tabset

### Building the model

Calling each model's function in [The models](#the-models): creating its greta arrays and calling `model()`. The timed code, `benchmark.R`'s `benchmark-build` section:

```{r}
#| label: benchmark-build
#| eval: false
#| code-fold: false
```

```{r}
#| label: build-summaries
build <- every_call(build_marks)

build_means <- build |>
  group_by(model, version) |>
  summarise(seconds = mean(seconds), .groups = "drop")

build_medians <- build |>
  group_by(version) |>
  summarise(seconds = median(seconds))
```

Median seconds per call, over all models and sessions: `{r} versions_text(build_medians$version, build_medians$seconds)`.

```{r}
#| label: plot-build
#| fig-cap: !expr call_caption
#| fig-height: 7
ggplot(build, aes(x = seconds, y = version, colour = version)) +
  geom_point(data = build_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(~model, ncol = 1, scales = "free_x") +
  labs(x = "Seconds", y = NULL)
```

<details><summary>Table: mean seconds</summary>

```{r}
#| label: table-build-means
dt_table(build_means)
```

</details>

<details><summary>Table: every call</summary>

```{r}
#| label: table-build
dt_table(build)
```

</details>

### opt()

`opt(m)`, `{r} settings$opt_calls` times on a model built for it. The timed code, `benchmark.R`'s `benchmark-opt` section:

```{r}
#| label: benchmark-opt
#| eval: false
#| code-fold: false
```

```{r}
#| label: opt-summaries
opt_timings <- every_call(opt_marks) |>
  mutate(calls = label_calls(call))

opt_means <- opt_timings |>
  group_by(model, calls, version) |>
  summarise(seconds = mean(seconds), .groups = "drop")

opt_medians <- opt_timings |>
  group_by(version) |>
  summarise(seconds = median(seconds))
```

Median seconds per call, over all models, calls and sessions: `{r} versions_text(opt_medians$version, opt_medians$seconds)`.

```{r}
#| label: plot-opt
#| fig-cap: !expr call_caption
#| fig-height: 9
ggplot(opt_timings, aes(x = seconds, y = version, colour = version)) +
  geom_point(data = opt_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(vars(model, calls), ncol = 2, scales = "free_x") +
  labs(x = "Seconds", y = NULL)
```

<details><summary>Table: mean seconds</summary>

```{r}
#| label: table-opt-means
dt_table(opt_means)
```

</details>

<details><summary>Table: every call</summary>

```{r}
#| label: table-opt
dt_table(opt_timings)
```

</details>

### mcmc()

`mcmc()` twice on each model: first on a new model, then on the same model again, which reuses the functions the first call traced. The timed code, `benchmark.R`'s `benchmark-mcmc` section, which also keeps the second call's draws for the posterior checks below:

```{r}
#| label: benchmark-mcmc
#| eval: false
#| code-fold: false
```

```{r}
#| label: mcmc-summaries
mcmc_timings <- every_call(mcmc_marks) |>
  mutate(calls = label_calls(call))

mcmc_means <- mcmc_timings |>
  group_by(model, calls, version) |>
  summarise(seconds = mean(seconds), .groups = "drop")

mcmc_medians <- mcmc_timings |>
  group_by(version) |>
  summarise(seconds = median(seconds))
```

Median seconds per call, over all models, calls and sessions: `{r} versions_text(mcmc_medians$version, mcmc_medians$seconds)`.

```{r}
#| label: plot-mcmc
#| fig-cap: !expr call_caption
#| fig-height: 9
ggplot(mcmc_timings, aes(x = seconds, y = version, colour = version)) +
  geom_point(data = mcmc_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(vars(model, calls), ncol = 2, scales = "free_x") +
  labs(x = "Seconds", y = NULL)
```

<details><summary>Table: mean seconds</summary>

```{r}
#| label: table-mcmc-means
dt_table(mcmc_means)
```

</details>

<details><summary>Table: every call</summary>

```{r}
#| label: table-mcmc
dt_table(mcmc_timings)
```

</details>

:::

## Efficiency

Bulk effective sample size (ESS) per second, for the parameter with the smallest bulk ESS and for the median parameter. The ESS is from the second `mcmc()` call's draws, the ones `benchmark-mcmc` keeps, so the seconds are the second call's.

```{r}
#| label: efficiency
posterior <- mcmc_marks |>
  as_tibble() |>
  select(session, version, model, posterior) |>
  unnest(posterior) |>
  mutate(model = factor(model, levels = model_levels))

second_calls <- mcmc_timings |>
  filter(call == 2) |>
  select(session, version, model, seconds)

efficiency <- posterior |>
  group_by(session, version, model) |>
  summarise(
    smallest_ess_bulk = min(ess_bulk),
    median_ess_bulk = median(ess_bulk),
    .groups = "drop"
  ) |>
  left_join(second_calls, by = c("session", "version", "model")) |>
  mutate(
    smallest_ess_per_second = smallest_ess_bulk / seconds,
    median_ess_per_second = median_ess_bulk / seconds
  )

efficiency_means <- efficiency |>
  group_by(model, version) |>
  summarise(
    smallest_ess_per_second = mean(smallest_ess_per_second),
    median_ess_per_second = mean(median_ess_per_second),
    .groups = "drop"
  )

session_caption <- "Each small dot is one session's second call. The large faint dot behind them is their mean."
```

::: panel-tabset

### Smallest bulk ESS per second

```{r}
#| label: plot-efficiency-smallest
#| fig-cap: !expr session_caption
#| fig-height: 7
ggplot(efficiency, aes(x = smallest_ess_per_second, y = version, colour = version)) +
  geom_point(data = efficiency_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(~model, ncol = 1, scales = "free_x") +
  labs(x = "Smallest bulk ESS per second", y = NULL)
```

### Median bulk ESS per second

```{r}
#| label: plot-efficiency-median
#| fig-cap: !expr session_caption
#| fig-height: 7
ggplot(efficiency, aes(x = median_ess_per_second, y = version, colour = version)) +
  geom_point(data = efficiency_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(~model, ncol = 1, scales = "free_x") +
  labs(x = "Median bulk ESS per second", y = NULL)
```

:::

<details><summary>Table: mean ESS per second</summary>

```{r}
#| label: table-efficiency-means
dt_table(efficiency_means, digits = 1)
```

</details>

<details><summary>Table: every session</summary>

```{r}
#| label: table-efficiency
dt_table(efficiency, digits = 1)
```

</details>

## Convergence

For each session's second `mcmc()` call, the largest rank-normalised split-$\hat{R}$ and the smallest bulk ESS over the model's parameters [@vehtari2021].

```{r}
#| label: convergence
convergence <- posterior |>
  group_by(session, version, model) |>
  summarise(
    largest_rhat = max(rhat),
    smallest_ess_bulk = min(ess_bulk),
    .groups = "drop"
  )

convergence_means <- convergence |>
  group_by(model, version) |>
  summarise(
    largest_rhat = mean(largest_rhat),
    smallest_ess_bulk = mean(smallest_ess_bulk),
    .groups = "drop"
  )
```

::: panel-tabset

### Largest R-hat

```{r}
#| label: plot-rhat
#| fig-cap: !expr session_caption
#| fig-height: 7
ggplot(convergence, aes(x = largest_rhat, y = version, colour = version)) +
  geom_point(data = convergence_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(~model, ncol = 1, scales = "free_x") +
  labs(x = "Largest R-hat", y = NULL)
```

### Smallest bulk ESS

```{r}
#| label: plot-ess
#| fig-cap: !expr session_caption
#| fig-height: 7
ggplot(convergence, aes(x = smallest_ess_bulk, y = version, colour = version)) +
  geom_point(data = convergence_means, size = 5, alpha = 0.25) +
  geom_point(position = jitter, size = 1.2) +
  version_scale +
  facet_wrap(~model, ncol = 1, scales = "free_x") +
  labs(x = "Smallest bulk ESS", y = NULL)
```

:::

<details><summary>Table: every session</summary>

```{r}
#| label: table-convergence
dt_table(convergence)
```

</details>

## Agreement

Whether two versions estimate the same posterior means. The difference between two versions' means is divided by the Monte Carlo standard error (MCSE) of that difference, so a difference can be told apart from Monte Carlo noise [@vehtari2021; @magnusson2025]. For each parameter, each version's means and MCSEs are pooled over its sessions, and

$$
z = \frac{\bar\theta_{\text{b}} - \bar\theta_{\text{a}}}{\sqrt{\text{se}^2_{\text{a}} + \text{se}^2_{\text{b}}}}.
$$

The sessions are independent, so a version's pooled standard error is $\sqrt{\sum \text{mcse}^2} / n$ over its $n$ sessions. If both versions sample the same posterior, each $z$ is close to a standard normal draw. Seeded calls of two versions that both respect `set.seed()` return the same draws, and so a $z$ of zero.

```{r}
#| label: agreement
pooled <- posterior |>
  group_by(version, model, variable) |>
  summarise(
    sessions = n(),
    mean = mean(mean),
    se = sqrt(sum(mcse_mean^2)) / n(),
    .groups = "drop"
  )

pairs <- utils::combn(version_levels, 2)
agreement <- lapply(seq_len(ncol(pairs)), function(i) {
  reference <- filter(pooled, version == pairs[1, i])
  compared <- filter(pooled, version == pairs[2, i])
  inner_join(reference, compared, by = c("model", "variable"), suffix = c("_a", "_b")) |>
    mutate(
      comparison = paste(pairs[2, i], "vs", pairs[1, i]),
      z = (mean_b - mean_a) / sqrt(se_a^2 + se_b^2)
    )
}) |>
  bind_rows() |>
  mutate(comparison = factor(comparison, levels = unique(comparison)))

# the band all of a comparison's z fall inside with probability 0.95 when both
# versions sample the same posterior: a Bonferroni bound over its parameters
parameters_compared <- n_distinct(paste(agreement$model, agreement$variable))
band <- qnorm(1 - 0.05 / (2 * parameters_compared))

outside_band <- agreement |>
  group_by(comparison) |>
  summarise(outside = sum(abs(z) > band), largest_abs_z = max(abs(z)))
```

The grey band is $\pm$ `{r} format_number(band)`: if two versions sample the same posterior, all `{r} parameters_compared` parameters' $z$ fall inside it with probability at least 0.95, by a Bonferroni bound. Parameters outside it, per comparison: `{r} paste(outside_band$comparison, outside_band$outside, collapse = ", ")`.

```{r}
#| label: plot-agreement
#| fig-cap: "Each dot is one parameter. The grey band is where every parameter's z falls, with probability at least 0.95, when both versions sample the same posterior."
#| fig-height: 7
ggplot(agreement, aes(x = z, y = model)) +
  annotate(
    "rect",
    xmin = -band,
    xmax = band,
    ymin = -Inf,
    ymax = Inf,
    fill = "grey80",
    alpha = 0.5
  ) +
  geom_point(position = jitter, alpha = 0.7) +
  facet_wrap(~comparison, ncol = 1) +
  labs(x = "z", y = NULL)
```

<details><summary>Table: largest |z| per comparison and model</summary>

```{r}
#| label: table-agreement-summary
agreement |>
  group_by(comparison, model) |>
  summarise(
    parameters = n(),
    largest_abs_z = max(abs(z)),
    parameter = variable[which.max(abs(z))],
    .groups = "drop"
  ) |>
  dt_table(digits = 2)
```

</details>

<details><summary>Table: every parameter</summary>

```{r}
#| label: table-agreement
agreement |>
  select(comparison, model, variable, mean_a, se_a, mean_b, se_b, z) |>
  dt_table()
```

</details>

## Traceplots

Every `{r} settings$keep_every_nth_draw`th draw of each chain from session 1's second `mcmc()` call on each model, numbered from the end of warmup, as `benchmark-mcmc` keeps them.

```{r}
#| label: traceplots
draws <- mcmc_marks |>
  as_tibble() |>
  select(session, version, model, kept_draws) |>
  unnest(kept_draws)

gg_trace <- function(name, version_label) {
  draws |>
    filter(session == 1, model == name, version == version_label) |>
    ggplot(aes(x = draw, y = value, colour = factor(chain))) +
    geom_line(linewidth = 0.3) +
    facet_wrap(~variable, scales = "free_y", ncol = 4) +
    labs(x = "Draw", y = NULL, colour = "Chain")
}

trace_height <- function(name) {
  1 + 1.6 * ceiling(parameter_count(name) / 4)
}
```

::: panel-tabset

### linear

::: panel-tabset

#### CRAN

```{r}
#| label: plot-trace-linear-cran
#| fig-height: !expr trace_height("linear")
gg_trace("linear", "CRAN")
```

#### main

```{r}
#| label: plot-trace-linear-main
#| fig-height: !expr trace_height("linear")
gg_trace("linear", "main")
```

#### #NNN

```{r}
#| label: plot-trace-linear-nnn
#| fig-height: !expr trace_height("linear")
gg_trace("linear", "#NNN")
```

:::

### multiple_linear

::: panel-tabset

#### CRAN

```{r}
#| label: plot-trace-multiple-linear-cran
#| fig-height: !expr trace_height("multiple_linear")
gg_trace("multiple_linear", "CRAN")
```

#### main

```{r}
#| label: plot-trace-multiple-linear-main
#| fig-height: !expr trace_height("multiple_linear")
gg_trace("multiple_linear", "main")
```

#### #NNN

```{r}
#| label: plot-trace-multiple-linear-nnn
#| fig-height: !expr trace_height("multiple_linear")
gg_trace("multiple_linear", "#NNN")
```

:::

### hierarchical_linear

::: panel-tabset

#### CRAN

```{r}
#| label: plot-trace-hierarchical-linear-cran
#| fig-height: !expr trace_height("hierarchical_linear")
gg_trace("hierarchical_linear", "CRAN")
```

#### main

```{r}
#| label: plot-trace-hierarchical-linear-main
#| fig-height: !expr trace_height("hierarchical_linear")
gg_trace("hierarchical_linear", "main")
```

#### #NNN

```{r}
#| label: plot-trace-hierarchical-linear-nnn
#| fig-height: !expr trace_height("hierarchical_linear")
gg_trace("hierarchical_linear", "#NNN")
```

:::

### eight_schools

::: panel-tabset

#### CRAN

```{r}
#| label: plot-trace-eight-schools-cran
#| fig-height: !expr trace_height("eight_schools")
gg_trace("eight_schools", "CRAN")
```

#### main

```{r}
#| label: plot-trace-eight-schools-main
#| fig-height: !expr trace_height("eight_schools")
gg_trace("eight_schools", "main")
```

#### #NNN

```{r}
#| label: plot-trace-eight-schools-nnn
#| fig-height: !expr trace_height("eight_schools")
gg_trace("eight_schools", "#NNN")
```

:::

### cjs

::: panel-tabset

#### CRAN

```{r}
#| label: plot-trace-cjs-cran
#| fig-height: !expr trace_height("cjs")
gg_trace("cjs", "CRAN")
```

#### main

```{r}
#| label: plot-trace-cjs-main
#| fig-height: !expr trace_height("cjs")
gg_trace("cjs", "main")
```

#### #NNN

```{r}
#| label: plot-trace-cjs-nnn
#| fig-height: !expr trace_height("cjs")
gg_trace("cjs", "#NNN")
```

:::

:::

## Posterior densities

The same kept draws, from every session, pooled over sessions.

```{r}
#| label: densities
gg_densities <- function(name) {
  draws |>
    filter(model == name) |>
    ggplot(aes(x = value, colour = version)) +
    geom_density() +
    scale_colour_manual(values = version_colours) +
    facet_wrap(~variable, scales = "free", ncol = 4) +
    labs(x = NULL, y = NULL, colour = NULL)
}

# each version's draws keep the order benchmark.R saved them in within a session
draws_identical <- function(pair) {
  first <- filter(draws, version == pair[1]) |> arrange(session)
  second <- filter(draws, version == pair[2]) |> arrange(session)
  identical(first$value, second$value)
}
identical_pairs <- Filter(
  draws_identical,
  utils::combn(version_levels, 2, simplify = FALSE)
)
identical_text <- if (length(identical_pairs) == 0) {
  ""
} else {
  paste0(
    vapply(identical_pairs, \(pair) paste(pair, collapse = "'s and "), character(1)),
    "'s draws are identical in every session, so their lines lie on top of each other.",
    collapse = " "
  )
}
```

`{r} identical_text`

::: panel-tabset

### linear

```{r}
#| label: plot-densities-linear
#| fig-height: !expr trace_height("linear")
gg_densities("linear")
```

### multiple_linear

```{r}
#| label: plot-densities-multiple-linear
#| fig-height: !expr trace_height("multiple_linear")
gg_densities("multiple_linear")
```

### hierarchical_linear

```{r}
#| label: plot-densities-hierarchical-linear
#| fig-height: !expr trace_height("hierarchical_linear")
gg_densities("hierarchical_linear")
```

### eight_schools

```{r}
#| label: plot-densities-eight-schools
#| fig-height: !expr trace_height("eight_schools")
gg_densities("eight_schools")
```

### cjs

```{r}
#| label: plot-densities-cjs
#| fig-height: !expr trace_height("cjs")
gg_densities("cjs")
```

:::

## The code

### run.R

[Download `run.R`](run.R).

```{r}
#| label: run-script
#| file: run.R
#| eval: false
#| code-fold: false
```

### benchmark.R

[Download `benchmark.R`](benchmark.R).

```{r}
#| label: benchmark-script
#| file: benchmark.R
#| eval: false
#| code-fold: false
```

Both, with every session's results in `results/`, are in [greta-dev/greta.benchmarks](https://github.com/greta-dev/greta.benchmarks).

## References

::: {#refs}
:::
