---
title: "greta#NNN (branch-name): what it is about, in a few plain words"
description: "What was compared, on what, and the result in words. One or two plain sentences, only state numbers for speed comparisons."
date: YYYY-MM-DD
---

<!--
How a post reads:

- Title: the PR number and branch name, then what it is about. Use the issue
  number when there is no PR.
- The opening: at most 3 short sentences, before any heading, saying what the
  PR changes and which versions were compared, then the headline timings as
  dot points, every number computed inline. No history, caveats or method:
  those come later.
- Timings are spread out: a dot point per case, and under it a nested dot
  point per comparison, each one sentence.
- Every PR or version named in the text is a link: a PR to its page, a commit
  to its commit, through version_text and version_ref() in the setup.
- Say how many times each method was called on a model, and that the first
  call traces it, rather than writing "a later call".
- Code is in backticks in the title, description and text. Figures and tables
  show it as plain text.
- Headings name the measure and which calls it covers: "Runtime per call,
  once the model is traced", not "How long does it take?". Axis labels and
  tables give the unit: "Runtime per call (seconds)".
- One section per question: a sentence on what is shown, one figure, the
  "Timed by" line and the "Run it yourself" block under it, then the answer in
  a sentence or two, numbers computed inline. Tables go in <details>, unless 
  they are a timed result that is key.
- Ratios name the slower version, so none is below 1: "CRAN takes 2.5 times
  as long as the PR", or "the PR takes 1.1 times as long as main" when the PR
  is the slower. Where more is better (ESS per second), name the version that
  gives more: "the PR gives 2.3 times CRAN's (2300 ESS vs 1000 ESS)". 
  compare_times() and compare_rates() in posts/_helpers.R write these sentences.
- No warning or message prints on the page. Handle each where it arises, and
  say in a sentence what it means.
- "How it was tested" is a collapsed callout straight after the opening: the
  versions and what each one changes, the models in words and maths, what was
  timed with its settings and a link to each section, the sessions and the
  machine, and the scripts in full.
- Session details last.

The helpers (links, numbers, rainclouds, tables, Run it yourself) are in
posts/_helpers.R. Anything that computes a number the post reports stays in
this page.
-->

```{r}
#| label: setup
#| code-summary: "Setup"
source(here::here("posts", "_helpers.R"))

post_path <- "posts/YYYY-MM-DD-short-name-iNNN"
file_link <- file_linker(post_path)
section_link <- section_linker(post_path)
knitr::read_chunk(here(post_path, "benchmark.R"))

sessions <- read_sessions(post_path)
n_sessions <- length(sessions)

# the names run.R gave the versions, and the names this page shows; a version
# pak installed from a local archive maps to its commit on GitHub
version_labels <- c(CRAN = "CRAN", main = "main", `#NNN` = "#NNN")
versions <- display_versions(
  sessions[[1]]$versions,
  labels = version_labels,
  on_github = c()
)
version_levels <- names(versions)
colours <- version_colours(version_levels)
settings <- sessions[[1]]$results[[1]]$settings

# how the text names each version: a PR links to its page, a commit to its
# commit
version_text <- c(
  CRAN = "CRAN",
  main = paste0("[main](", commit_url(versions[["main"]]), ")"),
  `#NNN` = paste0("[#NNN](", pr_url(NNN), ")")
)
version_ref <- function(version) I(version_text[[version]])

# one row per session and version, holding that version's results
runs <- lapply(sessions, function(session_file) {
  tibble(
    session = session_file$session,
    version = factor(
      unname(version_labels[names(session_file$results)]),
      levels = version_levels
    ),
    results = unname(session_file$results)
  )
}) |>
  bind_rows()

# every timed call: one row per call, in every session and version
mcmc_calls <- lapply(seq_len(nrow(runs)), function(i) {
  every_call(runs$results[[i]]$mcmc$linear) |>
    mutate(session = runs$session[i], version = runs$version[i])
}) |>
  bind_rows()

# each version's median call after the first, and how many times as long CRAN
# and main take as the PR
mcmc_medians <- mcmc_calls |>
  filter(calls == "calls after the first") |>
  group_by(version) |>
  summarise(seconds = median(seconds), .groups = "drop") |>
  pivot_wider(names_from = version, values_from = seconds) |>
  mutate(cran_over_pr = CRAN / `#NNN`, main_over_pr = main / `#NNN`)
```

[greta#NNN](https://github.com/greta-dev/greta/pull/NNN) WHAT THE PR CHANGES, IN ONE SENTENCE. This post times `{r} version_ref("#NNN")` against greta `{r} sessions[[1]]$results$CRAN$provenance$greta_version` from CRAN and against `{r} version_ref("main")`. `mcmc()` was called `{r} settings$mcmc_calls` times on the same model, and for the calls after the first, which traces the model:

- `mcmc()`:
    - `{r} compare_times(mcmc_medians$cran_over_pr, version_text[["CRAN"]], version_text[["#NNN"]])`
    - `{r} compare_times(mcmc_medians$main_over_pr, version_text[["main"]], version_text[["#NNN"]])`

::: {.callout-note collapse="true"}
## How it was tested

**The versions.**

- **CRAN**: greta `{r} sessions[[1]]$results$CRAN$provenance$greta_version`.
- **main**: `{r} I(version_link(versions[["main"]]))`.
- **#NNN**: `{r} I(version_link(versions[["#NNN"]]))`. WHAT IT CHANGES, IN A SENTENCE, AND ANY CHECK THAT IT GIVES THE SAME RESULTS.

**The model**, greta's `linear` example: each department's rating in `attitude` regressed on its number of complaints. The normal is written with its mean and standard deviation, as greta's `normal()` takes them; $\text{Cauchy}^{+}$ is a Cauchy truncated to positive values.

$$
\begin{aligned}
\text{rating}_i &\sim \text{Normal}(\mu_i, \sigma), \quad i = 1, \dots, 30 \\
\mu_i &= \alpha + \beta \, \text{complaints}_i \\
\alpha, \beta &\sim \text{Normal}(0, 10) \\
\sigma &\sim \text{Cauchy}^{+}(0, 3)
\end{aligned}
$$

**What was timed.** In each session, each version timed, on the model built afresh for each so that its first call traces:

- `mcmc()` on `{r} settings$chains` chains, with `warmup` of `{r} settings$warmup` and `n_samples` of `{r} settings$n_samples`, `{r} settings$mcmc_calls` times (`benchmark.R`'s `{r} section_link("benchmark.R", "mcmc-linear")` section).
- `opt()` with `adam()` and `max_iterations` of `{r} settings$opt_iterations`, `{r} settings$opt_calls` times (its `{r} section_link("benchmark.R", "opt-linear")` section).

**The sessions.** `{r} file_link("run.R")` installed each version as its `{r} section_link("run.R", "run-versions")` section says, and ran `{r} file_link("benchmark.R")` for each in a fresh R session of its own, in `{r} n_sessions` sessions, the order rotating so that each version ran first equally often (its `{r} section_link("run.R", "run-sessions")` section). Every version ran on TensorFlow `{r} runs$results[[1]]$provenance$tensorflow_version`, TensorFlow Probability `{r} runs$results[[1]]$provenance$tfp_version` and reticulate `{r} runs$results[[1]]$provenance$reticulate_version`, on `{r} runs$results[[1]]$provenance$machine` with `{r} runs$results[[1]]$provenance$cores` cores.

**The scripts in full.**

```{r}
#| label: run-script
#| file: run.R
#| eval: false
#| code-summary: "run.R"
```

```{r}
#| label: benchmark-script
#| file: benchmark.R
#| eval: false
#| code-summary: "benchmark.R"
```

:::

## Runtime per `mcmc()` call {#mcmc}

Every timed `mcmc()` call on greta's `linear` example, split into the first call on the model, which traces its TensorFlow functions, and the calls after it, which reuse them.

```{r}
#| label: plot-mcmc
#| fig-cap: !expr raincloud_caption
ggplot(mcmc_calls, aes(x = seconds, y = version, colour = version)) +
  raincloud(colours) +
  facet_wrap(vars(calls), scales = "free_x") +
  labs(x = "Runtime per call (seconds)", y = NULL)
```

`{r} timed_by(section_link, "benchmark.R", "mcmc-linear")`

```{r}
#| label: run-mcmc
#| eval: false
#| code-summary: "Run it yourself"
#| code: !expr run_it_yourself(post_path, "benchmark.R", c("setup", "mcmc-linear"), versions, "mcmc_linear")
```

ONE OR TWO SENTENCES ANSWERING THE QUESTION, NUMBERS COMPUTED INLINE.

<details><summary>Table: median seconds of the calls after the first</summary>

```{r}
#| label: table-mcmc
dt_table(mcmc_medians, digits = 3)
```

</details>

<!--
Further questions follow the same shape: opt(), how the number of chains
changes time, effective samples per second, and checks that every version
samples the same posterior and converges.
-->

## Session details

```{r}
#| label: session-info-of
first_session <- runs |> filter(session == sessions[[1]]$session)
session_info_of <- function(version) {
  first_session$results[[match(version, first_session$version)]]$session_info
}
```

<details><summary>`sessioninfo::session_info()` for CRAN</summary>

```{r}
#| label: session-info-cran
session_info_of("CRAN")
```

</details>

<details><summary>`sessioninfo::session_info()` for #NNN</summary>

```{r}
#| label: session-info-pr
session_info_of("#NNN")
```

</details>
