---
title: "The cost of greta's discarded warmup trace"
format: gfm
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(here)
library(fs)
library(bench)
library(dplyr)
library(tidyr)
library(rlang)
library(tibble)
library(knitr)

run_dir <- here("2026-08-20-warmup-trace")
results_rds <- path(run_dir, "results.rds")
if (!file_exists(results_rds)) {
  stop("no results.rds - run 2026-08-20-warmup-trace/01-measure.R first", call. = FALSE)
}
results <- readRDS(results_rds)

# refuse to render numbers that predate an edit to the script that produced
# them. One input, compared by content, so this is not general cache
# invalidation - just the one question worth asking.
knitr::read_chunk(path(run_dir, "01-measure.R"))
```

## The question

greta's warmup loop appends each burst's free state onto `traced_free_state`,
then discards the whole thing when warmup ends. What does that cost, and does
it grow faster than linearly?

Written up as [greta #834](https://github.com/greta-dev/greta/issues/834); the
issue text is in `issue.md` beside this file.

Wall time rather than effective samples per second, on purpose: the change this
measures deletes work whose result is thrown away, so the sampler behaves
identically and the draws are unchanged. ESS is a stochastic estimate and would
only add variance. It is also unavailable as a clean comparison anyway, since
`mcmc()` does not respect `set.seed()` ([greta
#285](https://github.com/greta-dev/greta/issues/285),
[#427](https://github.com/greta-dev/greta/issues/427)).

## Part 1: how runtime scales with warmup

Wall-clock seconds for `mcmc(n_samples = 100, chains = 2)`, single runs.

```{r}
#| label: scaling
results$scaling |>
  mutate(n_free = paste("n_free =", n_free)) |>
  pivot_wider(names_from = n_free, values_from = elapsed) |>
  arrange(warmup) |>
  kable(digits = 2)
```

```{r}
#| label: growth
growth <- results$scaling |>
  arrange(warmup) |>
  summarise(fold = last(elapsed) / first(elapsed), .by = n_free)

fold_at <- function(p) sprintf("%.1f", growth$fold[growth$n_free == p])
```

Warmup 500 to 4000 is an eight-fold increase in iterations. Over that range the
time grows `r fold_at(1)`-fold at one free parameter, `r fold_at(20)`-fold at
twenty, and `r fold_at(200)`-fold at 200. Only the largest model grows faster
than its iteration count; the gap is the accumulation cost, and it appears only
once a model has enough parameters for the copied matrix to be large.

These are single runs, so read the shape across the row rather than any
individual cell. The one- and twenty-parameter columns are the same model cost
to within noise, and swap places between runs.

## Part 2: the accumulation on its own

greta removed, so the growth is attributable to `rbind` rather than to anything
in the sampler. This is the accumulation the warmup loop performs:

```{r accumulate}
#| eval: false
#| echo: true
```

Warmup breaks a burst roughly every three iterations once tuning changepoints
are counted, so 167, 334 and 667 bursts stand in for warmups of about 500, 1000
and 2000 iterations. At 200 free parameters:

```{r}
#| label: accumulation
results$accumulation |>
  select(bursts = n_bursts, median, allocated = mem_alloc) |>
  mutate(across(c(median, allocated), format)) |>
  kable(align = "r")
```

Doubling the burst count quadruples both time and memory; quadrupling it
multiplies them by about sixteen. That is the signature of copying the whole
accumulated matrix on every append. At 667 bursts — a warmup of roughly 2000
iterations — greta allocates over a gigabyte for a matrix it then deletes.

## Part 3: what removing it buys

```{r}
#| label: comparison
#| output: asis
if (is.null(results$comparison)) {
  cat(
    "**Not measured.** This needs a branch with the `self$trace()` call removed",
    "from the warmup loop in `R/sampler_class.R`. Create it, then re-run",
    "`01-measure.R` with `GRETA_FIX_BRANCH` set, then `02-report.R`.\n\n",
    "An in-session check by patching the working tree gave 3.23s to 2.18s at",
    "200 free parameters and warmup 2000, and 7.62s to 3.79s at warmup 4000.",
    "That is recorded as motivation only: patching a working tree is not",
    "reproducible by anyone else, which is the whole reason this directory",
    "exists.\n"
  )
} else {
  print(kable(as_tibble(results$comparison)))
}
```

## Environment

```{r}
#| label: provenance
# sessioninfo reports a GitHub source as "Github (owner/repo@sha)". Turn the
# sha into a link so it can be clicked rather than copied.
link_source <- function(source) {
  sub(
    "([A-Za-z0-9_.-]+/[A-Za-z0-9_.-]+)@([0-9a-f]{7,40})",
    "[\\1@\\2](https://github.com/\\1/commit/\\2)",
    source
  )
}

platform <- results$provenance$platform
packages <- as_tibble(results$provenance$packages)

bind_rows(
  tibble(
    item = c("run at", "OS", "system", "CPU", "cores detected", "R"),
    value = c(
      results$provenance$run_at,
      platform$os,
      platform$system,
      results$provenance$cpu,
      as.character(results$provenance$cores),
      platform$version
    )
  ),
  packages |>
    transmute(
      item = paste("R package:", package),
      value = paste0(ondiskversion, " (", link_source(source), ")")
    )
) |>
  set_names(c("", "")) |>
  kable()
```
