---
title: "Tuning diag_sd on the right n"
format: gfm
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(here)
library(fs)
library(dplyr)
library(tidyr)
library(knitr)
library(ggplot2)

run_dir <- here("2026-09-20-tune-diag-sd")
results_rds <- path(run_dir, "results.rds")
if (!file_exists(results_rds)) {
  stop(
    "no results.rds - run 2026-09-20-tune-diag-sd/01-measure.R first",
    call. = FALSE
  )
}
raw <- readRDS(results_rds)
results <- raw$results
design <- raw$design

fold <- function(x, digits = 2) formatC(x, digits = digits, format = "f")
```

## The question

`tune_diag_sd()` shrinks the sample posterior variance towards `5e-3` before
using it as the HMC proposal scale. The shrinkage is a weighted average of that
variance and 5 pseudo-observations of `1e-3`, so `n` has to be the number of
real observations behind the variance. greta passes
`sum(!self$accept_history)`, a count of *rejected* proposals.

Three branches, one baseline:

- **main** — `n` is the rejection count
- **tune-diag-sd-accepted** — drop the `!`, so `n` is the accepted count
- **tune-diag-sd-samples** — `n` is `welford_state$count`, the number of samples
  the variance was actually estimated from, which is what Stan passes

```{r}
#| label: design
tibble::tibble(
  setting = c("warmups", "reps per cell", "chains", "samples"),
  value = c(
    paste(design$warmups, collapse = ", "),
    design$n_reps,
    design$chains,
    design$n_samples
  )
) |>
  kable()
```

Measured on `r nrow(results)` runs across `r length(unique(results$model))`
models. Commits, resolved at measure time:

```{r}
#| label: shas
tibble::tibble(branch = names(raw$shas), sha = substr(raw$shas, 1, 10)) |>
  kable()
```

## Is the effective sample size usable?

`ess_min` collapses to zero whenever a single parameter never moves, which makes
that cell useless for comparing branches however well everything else mixed. So
this has to be checked before any ESS number is read.

```{r}
#| label: zeros
results |>
  group_by(branch) |>
  summarise(
    cells = n(),
    cells_with_a_zero = sum(n_zero_ess > 0),
    params_at_zero = sum(n_zero_ess),
    params_total = sum(n_par),
    .groups = "drop"
  ) |>
  kable()
```

```{r}
#| label: zeros-by-model
results |>
  group_by(model) |>
  summarise(
    cells_with_a_zero = sum(n_zero_ess > 0),
    cells = n(),
    .groups = "drop"
  ) |>
  kable()
```

## Effective samples per second

Three summaries, because they disagree in informative ways: the minimum is what
limits a run but is the one the zeros destroy, the median survives them, and the
mean is what earlier drafts of the measure script reported on its own.

```{r}
#| label: ess-per-sec
results |>
  group_by(branch, warmup) |>
  summarise(
    across(
      c(ess_min_per_sec, ess_median_per_sec, ess_mean_per_sec),
      \(x) mean(x)
    ),
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric) & !warmup, \(x) round(x, 2))) |>
  kable()
```

```{r}
#| label: ess-per-sec-by-model
results |>
  group_by(model, branch) |>
  summarise(ess_median_per_sec = mean(ess_median_per_sec), .groups = "drop") |>
  pivot_wider(names_from = branch, values_from = ess_median_per_sec) |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  kable()
```

## The tuned proposal scale

The quantity the change acts on directly, and the cheapest evidence the branches
did anything different. Bigger is not better: it has to match the posterior
scale, which is what the ESS tables above judge.

```{r}
#| label: diag-sd
results |>
  group_by(model, branch) |>
  summarise(diag_sd = mean(diag_sd), .groups = "drop") |>
  pivot_wider(names_from = branch, values_from = diag_sd) |>
  mutate(across(where(is.numeric), \(x) round(x, 4))) |>
  kable()
```

## Convergence

A gain in ESS is only a gain if the chains still converged.

```{r}
#| label: rhat
results |>
  group_by(branch) |>
  summarise(
    worst_rhat = ifelse(
      all(is.na(rhat_max)),
      NA_real_,
      round(max(rhat_max, na.rm = TRUE), 3)
    ),
    cells_without_rhat = sum(is.na(rhat_max)),
    cells = n(),
    .groups = "drop"
  ) |>
  kable()
```

## Wall time

Recorded so that a change in ESS per second can be attributed to mixing rather
than to the sampler simply running slower.

```{r}
#| label: elapsed
results |>
  group_by(branch, warmup) |>
  summarise(elapsed = round(mean(elapsed), 3), .groups = "drop") |>
  pivot_wider(names_from = branch, values_from = elapsed) |>
  kable()
```

## Calibration

From `01b-geweke.R`, one file per branch. A Geweke test compares prior draws
against chain draws targeting the same joint distribution; if the sampler is
correct they are the same distribution, so the KS test should not reject.
greta's own test asserts `p >= 0.005`.

`slice()` is the control: it sets `uses_metropolis = FALSE` and never calls
`tune_diag_sd()`, so it must be unaffected. `hmc()` and `rwmh()` both tune
`diag_sd` and are the ones that can move.

This is a stochastic test. One p-value per sampler is weak evidence, and an
occasional low value is expected rather than alarming; it is here to catch a
gross calibration break, not to certify correctness.

```{r}
#| label: geweke
geweke_files <- dir_ls(run_dir, glob = "*geweke-*.rds")
if (length(geweke_files) == 0) {
  cat("no geweke-*.rds in this directory - run 01b-geweke.R first\n")
} else {
  geweke <- lapply(geweke_files, \(f) readRDS(f)$results) |>
    bind_rows()

  geweke |>
    select(branch, sampler, statistic, p_value, passes_greta_threshold) |>
    mutate(
      statistic = round(statistic, 4),
      p_value = round(p_value, 4)
    ) |>
    arrange(sampler, branch) |>
    kable()
}
```

A KS p-value compresses the whole comparison to one number. The QQ plots below
are the same data uncompressed: prior quantiles against chain quantiles, so a
sampler that targets the right distribution sits on the diagonal. Drawn from the
stored draws rather than at measure time, so a different plot needs no re-run.

```{r}
#| label: geweke-qq
#| fig-height: 7
#| fig-width: 8
if (length(geweke_files) > 0) {
  all_draws <- lapply(geweke_files, \(f) readRDS(f)$draws) |>
    bind_rows()

  quants <- (1:99) / 100
  qq <- all_draws |>
    group_by(branch, sampler) |>
    reframe(
      prior = quantile(target_theta, quants),
      chain = quantile(greta_theta, quants)
    )

  ggplot(qq, aes(x = chain, y = prior)) +
    geom_abline(slope = 1, intercept = 0, colour = "grey60") +
    geom_point(size = 0.7, alpha = 0.7) +
    facet_grid(sampler ~ branch) +
    labs(
      x = "chain quantiles",
      y = "prior quantiles",
      title = "Geweke QQ, by sampler and branch"
    ) +
    theme_minimal()
}
```

```{r}
#| label: geweke-model
if (length(geweke_files) > 0) {
  seeds <- lapply(geweke_files, \(f) {
    g <- readRDS(f)
    data.frame(branch = g$label, mu1 = g$mu1, sd1 = g$sd1, sd2 = g$sd2)
  }) |>
    bind_rows()

  cat(
    "all branches tested on the same model:",
    ifelse(nrow(unique(seeds[, -1])) == 1, "yes", "NO - compare with care"),
    "\n"
  )
}
```

## Provenance

```{r}
#| label: provenance
str(raw$provenance)
```
