---
title: "Every function greta traces, on main and greta#843"
format:
  gfm: default
  html: default
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(here)
library(fs)
library(dplyr)
library(tidyr)
library(knitr)

run_dir <- here("2026-09-30-trace-census-i546")
raw <- readRDS(path(run_dir, "results.rds"))
census <- raw$census
timings <- raw$timings

build_order <- c("main", "branch", "fix", "fix_check")
scenario_order <- c(
  "model",
  "opt",
  "opt_adam",
  "opt_twice",
  "opt_hessian",
  "mcmc",
  "extra_samples",
  "calculate_values",
  "calculate_nsim"
)
```

Four builds of greta: `main`; greta#843 as pushed (`branch`); #843 with
`opt()`'s steps on a log-density function traced for one row (`fix`, from
`fix.patch`); and that, with `opt()`'s check of its initial values using the
same function (`fix_check`, from `fix-check.patch`).

## The census

Each call below ran in a fresh process on the linear example, with
`tensorflow::tf_function()` wrapped to record every function greta created.
"Functions" is how many it created, "traces" how many times they were traced
in total, and "most" the most any one function was traced. A function traced
more than once is retracing. `opt` is greta's default optimiser, `bfgs()`;
`opt_adam` uses `adam()`, whose own traced step calls the log-density
function.

```{r}
#| label: census-summary
summary_table <- census |>
  group_by(build, scenario) |>
  summarise(
    functions = n(),
    most = max(traces),
    traces = sum(traces),
    .groups = "drop"
  ) |>
  mutate(
    cell = sprintf("%d / %d / %d", functions, traces, most),
    build = factor(build, build_order),
    scenario = factor(scenario, scenario_order)
  ) |>
  select(scenario, build, cell) |>
  arrange(build) |>
  pivot_wider(names_from = build, values_from = cell) |>
  arrange(scenario)
kable(summary_table, caption = "functions / traces / most traces of one function")
```

Every function traced more than once, and every kind of function created more
than once within one call. `rows` is the free state's number of rows in the
function's signature: the log-density function's open-signature and one-row
versions are made by the same method, and this tells them apart.

```{r}
#| label: census-detail
census |>
  mutate(made_by = sub("^self[$]", "", made_by)) |>
  group_by(build, scenario, made_by, rows) |>
  summarise(
    created = n(),
    most = max(traces),
    traces = sum(traces),
    .groups = "drop"
  ) |>
  filter(created > 1 | most > 1) |>
  mutate(
    build = factor(build, build_order),
    scenario = factor(scenario, scenario_order)
  ) |>
  arrange(scenario, build) |>
  kable()
```

## What tracing costs opt()

Seconds for `opt(m, optimiser = adam(), max_iterations = 100)` on each
example, in a fresh process: the first call, which traces, and a second,
which reuses those traces. Medians of `r max(timings$rep)` replicates, with
their range.

```{r}
#| label: timings
timings |>
  group_by(build, example) |>
  summarise(
    first = sprintf(
      "%.2f (%.2f-%.2f)",
      median(first),
      min(first),
      max(first)
    ),
    second = sprintf(
      "%.3f (%.3f-%.3f)",
      median(second),
      min(second),
      max(second)
    ),
    .groups = "drop"
  ) |>
  mutate(build = factor(build, build_order)) |>
  arrange(example, build) |>
  kable()
```

## Builds

```{r}
#| label: builds
data.frame(
  build = names(raw$builds),
  commit = vapply(raw$builds, `[[`, "", "sha"),
  patch = vapply(
    raw$builds,
    function(b) if (is.null(b$patch)) "" else b$patch,
    ""
  )
) |>
  kable(row.names = FALSE)
```
