```{r}
#| label: further-setup
further_dir <- fs::path_dir(params$further)

tracing_runs <- lapply(
  fs::dir_ls(further_dir, regexp = "tracing-run-.*-run[0-9]+[.]rds$"),
  readRDS
)

as_version <- function(x) factor(x, levels = versions)

sha_link <- function(sha) {
  sprintf(
    "[%s](https://github.com/greta-dev/greta/commit/%s)",
    substr(sha, 1, 8),
    sha
  )
}

tracing_versions <- bind_rows(lapply(tracing_runs, function(x) {
  data.frame(version = x$label, branch = x$branch, sha = x$sha)
})) |>
  distinct() |>
  arrange(as_version(version))
runs_per_version <- length(tracing_runs) / nrow(tracing_versions)
```

greta#546 is about tracing: the first time greta calls one of a model's TensorFlow functions in a session, TensorFlow runs greta's R code to record the function, and records it again for each input shape it has not seen. Section 1 times calls after that has happened. These runs time calls that include it.

## Tracing in `mcmc()`

`tracing-run.qmd`, rendered `r runs_per_version` times against each version by `02-tracing-runs.R`, each time in a fresh R session with only that version installed. Each render runs the five models above with `mcmc()`, `r mcmc_chains` chains of `r warmup_iterations` warmup and `r sample_iterations` sampling iterations on `r mcmc_cores` cores. Each call is the first for its model in that session, so its time includes tracing. It records how many times each of the model's three traced functions was traced, and how many retracing warnings TensorFlow logged. The documents linked are each version's last run:

```{r}
#| label: tracing-run-links
#| output: asis
for (i in seq_len(nrow(tracing_versions))) {
  cat(sprintf(
    "- %s: [html](tracing-run-%s.html), [markdown](tracing-run-%s.md), at %s\n",
    tracing_versions$version[i],
    tracing_versions$branch[i],
    tracing_versions$branch[i],
    sha_link(tracing_versions$sha[i])
  ))
}
```

```{r}
#| label: tracing-mcmc
tracing_mcmc <- bind_rows(lapply(tracing_runs, function(x) {
  cbind(version = x$label, run = x$run, x$mcmc)
})) |>
  mutate(
    version = as_version(version),
    model = factor(model, unique(model))
  )

traces_long <- tracing_mcmc |>
  select(
    model,
    version,
    run,
    log_prob_traces,
    trace_values_traces,
    sampler_traces
  ) |>
  pivot_longer(
    -c(model, version, run),
    names_to = "traced_function",
    values_to = "traces"
  ) |>
  mutate(
    traced_function = factor(
      sub("_traces$", "", traced_function),
      levels = c("log_prob", "trace_values", "sampler")
    )
  )
```

```{r}
#| label: fig-traces
#| fig-cap: "How many times each of the model's traced functions was traced during its first mcmc() call. One dot per run; runs with the same count overlap."
#| fig-height: 6
traces_long |>
  ggplot(aes(x = traces, y = version, colour = version)) +
  geom_point(size = 3, alpha = 0.5) +
  facet_grid(model ~ traced_function) +
  scale_x_continuous(breaks = 0:4, limits = c(0, 4)) +
  labs(x = "traces", y = NULL) +
  theme(legend.position = "none", strip.text.y = element_text(angle = 0))
```

```{r}
#| label: fig-tracing-seconds
#| fig-cap: "Seconds for each model's first mcmc() call, tracing included: the small dots are runs and the large dot is their mean. Each model has its own scale, starting at zero."
#| fig-height: 5
tracing_mcmc |>
  ggplot(aes(x = seconds, y = version, colour = version)) +
  geom_point(size = 1.5, alpha = 0.4) +
  stat_summary(fun = mean, geom = "point", size = 4) +
  facet_wrap(~model, ncol = 1, scales = "free_x") +
  expand_limits(x = 0) +
  labs(x = "seconds", y = NULL) +
  theme(legend.position = "none")
```

<details>
<summary>Table of every run</summary>

```{r}
#| label: tracing-mcmc-table
tracing_mcmc |>
  arrange(model, version, run) |>
  select(
    model,
    version,
    run,
    seconds,
    retracing_warnings,
    log_prob_traces,
    trace_values_traces,
    sampler_traces
  ) |>
  kable(
    row.names = FALSE,
    col.names = c(
      "model",
      "version",
      "run",
      "seconds",
      "retracing warnings",
      "log-prob traces",
      "trace-values traces",
      "sampler traces"
    )
  )
```

</details>

## Tracing in `opt(hessian = TRUE)`, on twenty scalar targets

The model greta#546 reported: twenty observations $y_k$, simulated, each with its own mean $b_k$, so `opt(hessian = TRUE)` takes twenty hessians. greta's `variable()` gives $b_k$ a flat prior, so it has no distribution line.

$$
y_k \sim \text{Normal}(b_k, 1), \quad k = 1, \dots, 20
$$

```{r}
#| label: code-hessian-model
#| echo: true
#| eval: false
set.seed(2026 - 09 - 29)
y <- rnorm(20)

# twenty separate scalar parameters, b1 to b20
target_names <- paste0("b", 1:20)
for (name in target_names) {
  assign(name, variable())
}
distribution(y) <- normal(do.call(c, mget(target_names)), 1)

# model() names its targets from the expressions it is given, so the call is
# built from the names: this is model(b1, b2, ..., b20)
m <- eval(as.call(c(quote(model), lapply(target_names, as.name))))
```

`opt(m, hessian = TRUE)`, timed in the same documents, as the first `opt()` call in each session:

```{r}
#| label: tracing-hessian
tracing_hessian <- bind_rows(lapply(tracing_runs, function(x) {
  cbind(version = x$label, run = x$run, x$hessian)
})) |>
  mutate(version = as_version(version))
```

```{r}
#| label: fig-hessian
#| fig-cap: "Seconds for opt(hessian = TRUE) on twenty scalar targets, as the first opt() call in a session: the small dots are runs and the large dot is their mean."
#| fig-height: 2.5
tracing_hessian |>
  ggplot(aes(x = seconds, y = version, colour = version)) +
  geom_point(size = 1.5, alpha = 0.4) +
  stat_summary(fun = mean, geom = "point", size = 4) +
  expand_limits(x = 0) +
  labs(x = "seconds", y = NULL) +
  theme(legend.position = "none")
```

<details>
<summary>Table of every run</summary>

```{r}
#| label: tracing-hessian-table
tracing_hessian |>
  arrange(version, run) |>
  kable(
    row.names = FALSE,
    col.names = c("version", "run", "seconds", "retracing warnings")
  )
```

</details>
