```{r}
#| label: further-setup
further_dir <- fs::path_dir(params$further)

singles <- lapply(
  fs::dir_ls(further_dir, regexp = "single-run-.*[.]rds$"),
  readRDS
)
names(singles) <- vapply(singles, `[[`, "", "label")

as_version <- function(x) factor(x, levels = versions)

sha_link <- function(sha) {
  sprintf(
    "[%s](https://github.com/greta-dev/greta/commit/%s)",
    substr(sha, 1, 8),
    sha
  )
}
```

## One run of each model, first call in a session

The same document, `single-run.qmd`, rendered against each version by `01-single-runs.R`, each with only that version installed:

```{r}
#| label: single-run-links
#| output: asis
for (version in versions) {
  branch <- singles[[version]]$branch
  cat(sprintf(
    "- %s: [html](single-run-%s.html), [markdown](single-run-%s.md), at %s\n",
    version,
    branch,
    branch,
    sha_link(singles[[version]]$sha)
  ))
}
```

Each runs the five models above with `mcmc()`, 4 chains of 1000 warmup and 1000 samples, as the first call in its R session, so the time includes tracing. It records how many times each of the model's three traced functions was traced, and how many retracing warnings TensorFlow logged. One run each.

```{r}
#| label: single-mcmc
single_mcmc <- bind_rows(lapply(singles, function(x) {
  cbind(version = x$label, x$mcmc)
})) |>
  mutate(
    version = as_version(version),
    model = factor(model, unique(model))
  )

traces_long <- single_mcmc |>
  select(model, version, log_prob_traces, trace_values_traces, sampler_traces) |>
  pivot_longer(
    -c(model, version),
    names_to = "traced_function",
    values_to = "traces"
  ) |>
  mutate(
    traced_function = factor(
      sub("_traces$", "", traced_function),
      levels = c("log_prob", "trace_values", "sampler")
    )
  )
```

```{r}
#| label: fig-traces
#| fig-cap: "How many times each of the model's traced functions was traced during one mcmc() run."
#| fig-height: 6
traces_long |>
  ggplot(aes(x = traces, y = version, colour = version)) +
  geom_point(size = 3) +
  facet_grid(model ~ traced_function) +
  scale_x_continuous(breaks = 0:4, limits = c(0, 4)) +
  labs(x = "traces", y = NULL) +
  theme(legend.position = "none", strip.text.y = element_text(angle = 0))
```

```{r}
#| label: fig-single-seconds
#| fig-cap: "Seconds for one mcmc() run, first call in its session, tracing included. Each model has its own scale, starting at zero."
#| fig-height: 5
single_mcmc |>
  ggplot(aes(x = seconds, y = version, colour = version)) +
  geom_point(size = 3) +
  facet_wrap(~model, ncol = 1, scales = "free_x") +
  expand_limits(x = 0) +
  labs(x = "seconds", y = NULL) +
  theme(legend.position = "none")
```

<details>
<summary>Table of the single runs</summary>

```{r}
#| label: single-mcmc-table
single_mcmc |>
  arrange(model, version) |>
  select(
    model,
    version,
    seconds,
    retracing_warnings,
    log_prob_traces,
    trace_values_traces,
    sampler_traces
  ) |>
  kable(
    row.names = FALSE,
    col.names = c(
      "model",
      "version",
      "seconds",
      "retracing warnings",
      "log-prob traces",
      "trace-values traces",
      "sampler traces"
    )
  )
```

</details>

## opt(hessian = TRUE) on twenty scalar targets

The model greta#546 reported: twenty observations $y_k$, simulated, each with its own mean $b_k$, so `opt(hessian = TRUE)` takes twenty hessians. greta's `variable()` gives $b_k$ a flat prior, so it has no distribution line.

$$
y_k \sim \text{Normal}(b_k, 1), \quad k = 1, \dots, 20
$$

```{r}
#| label: code-hessian-model
#| echo: true
#| eval: false
set.seed(2026 - 09 - 29)
y <- rnorm(20)

# twenty separate scalar parameters, b1 to b20
target_names <- paste0("b", 1:20)
for (name in target_names) {
  assign(name, variable())
}
distribution(y) <- normal(do.call(c, mget(target_names)), 1)

# model() names its targets from the expressions it is given, so the call is
# built from the names: this is model(b1, b2, ..., b20)
m <- eval(as.call(c(quote(model), lapply(target_names, as.name))))
```

`opt(m, hessian = TRUE)`, timed as the first call in its session in the same documents:

```{r}
#| label: single-hessian
bind_rows(lapply(singles, function(x) {
  cbind(version = x$label, x$hessian)
})) |>
  arrange(as_version(version)) |>
  kable(
    row.names = FALSE,
    col.names = c("version", "seconds", "retracing warnings")
  )
```
